### 2018-03-29 N_V/EdgeR/edgeR-btwnLane/makeDataMetrics.Rmd
Completed DEG for all platforms for N_V

---
title: "Honey Bees - edgeR"
author: Lindsay Rutter
output:
  packagedocs::package_docs:
    toc: true
    toc_collapse: true
vignette: >
  %\VignetteEngine{packagedocs::redirect}
---

<meta http-equiv="content-type" content="text/html;charset=utf-8" />

```{r global_options, include=FALSE}
# R output pre blocks are styled by default to indicate output
knitr::opts_chunk$set(
  comment = NA,
  cache = TRUE,
  fig.height = 8,
  fig.width = 10
)
```

```{r}
library(edgeR)
library(ggplot2)
library(GGally)
library(EDASeq)
library(utils)

thisPath <- getwd()

beeCounts <- readRDS("../../data/data.Rds")
beeCounts <- as.matrix(beeCounts)

y <- DGEList(counts=beeCounts)
```

edgeR vignette states that a gene is required to have a count of 5-10 in a library to be considered expressed in that library. Here minLib is equal to `r min(y$samples$lib.size)`. A CPM of 4 corresponds to a count of ~1 in the minimum number of samples in a group (6). So, I keep only rows that have that. This reduces the number of genes from 15,314 to 8,672.

```{r}
minLib <- min(y$samples$lib.size)
keep <- rowSums(cpm(y)>4) >= 6
# Number of genes 15,314--> 8,672
y <- y[keep, , keep.lib.sizes=FALSE]
```

Next, I used edgeR normalization.
```{r}
y[[1]] <- betweenLaneNormalization(y[[1]], which="full", round=FALSE)
```

I can create boxplot and RLE boxplot.

```{r}
ggparcoord(data.frame(log(y[[1]]/colMeans(y[[1]]))), columns=1:24, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
```

I can create MDS plots of interest.

```{r}
allGroups <- c(rep("N",12), rep("V",12))
y$samples$group <- allGroups

plotMDS(y, col = c("red","green","blue","purple")[factor(allGroups)], cex=0.6)
plotMDS(y, col = c("blue","blue","red","red")[factor(allGroups)], cex=0.6)
plotMDS(y, col = c("blue","red","blue","red")[factor(allGroups)], cex=0.6)
```

Copying Section (3.3.1 in edgeR vignette)

```{r}
Group = factor(c(rep("N",12), rep("V",12)))
targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

```{r}
fit <- glmFit(y, design)
saveRDS(y[[1]], file="data.Rds")
```

Below we save the DEGs from all pairwise combinations of treatment groups.

```{r}
allPairs = data.frame(Treatment1 = factor(), Treatment2 = factor(), NumberDEG = numeric(), FirstLarger = numeric(), SecondLarger = numeric())

dataMetrics <- list()

for (i in 1:(ncol(fit)-1)){
  for (j in (i+1):ncol(fit)){
    contrast=rep(0,ncol(fit))
    contrast[i]=1
    contrast[j]=-1
    lrt <- glmLRT(fit, contrast=contrast)
    lrt <- topTags(lrt, n = nrow(y[[1]]))[[1]]
    lrt5 <- lrt[which(lrt$FDR<0.05),]

    df <- y[[1]][rownames(y[[1]]) %in% rownames(lrt5), c(which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[i])), which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[j])))]
    firstLarger <- length(which(apply(df, 1, function(x) mean(x[1:6]) > mean(x[7:12]))))
    secondLarger <- length(which(apply(df, 1, function(x) mean(x[1:6]) < mean(x[7:12]))))

    lrtLength <- nrow(lrt5)
    allPairs = rbind(allPairs, data.frame(Treatment1 = factor(colnames(fit)[i]), Treatment2 = factor(colnames(fit)[j]), NumberDEG = lrtLength, FirstLarger = firstLarger, SecondLarger = secondLarger))

    metrics = cbind(ID = rownames(lrt), lrt)
    metrics$ID = as.character(metrics$ID)
    dataMetrics[[paste0(colnames(fit)[i], "_", colnames(fit)[j])]] <- metrics
  }
}
allPairs <- allPairs[order(allPairs$NumberDEG),]
saveRDS(dataMetrics, file="dataMetrics.Rds")
```

```{r}
allPairs
```

### 2018-03-29 N_V/LimmaVoom/makeDataMetrics.Rmd
Completed DEG for all platforms for N_V

---
title: "Limma Tutorial"
author: Lindsay Rutter
output:
  packagedocs::package_docs:
  toc: true
toc_collapse: true
vignette: >
  %\VignetteEngine{packagedocs::redirect}
---

  <meta http-equiv="content-type" content="text/html;charset=utf-8" />

  ```{r global_options, include=FALSE}
# R output pre blocks are styled by default to indicate output
knitr::opts_chunk$set(
  comment = NA,
  cache = TRUE,
  fig.height = 8,
  fig.width = 10
)
```
Try new paper (https://www.bioconductor.org/help/workflows/RNAseq123/)

```{r}
library(limma)
library(Glimma)
library(GGally)
library(ggplot2)
library(edgeR)

thisPath <- getwd()
beeCounts <- readRDS("../data/data.Rds")
beeCounts <- as.matrix(beeCounts)
x <- DGEList(counts=beeCounts)

group <- factor(c(rep("N",12), rep("V",12)))
x$samples$group <- group
lane <- as.factor(c("L12","L12","L12","L12","L34","L34","L12","L12","L34","L34","L34","L34","L12","L12","L12","L12","L34","L34","L12","L12","L34","L34","L34","L34"))
x$samples$lane <- lane
```

Transform and remove low counts.
```{r}
cpm <- cpm(x)
keep.exprs <- rowSums(cpm>1)>=4 # tried filtering up to 24 and not much difference
x <- x[keep.exprs,, keep.lib.sizes=FALSE] # 15,314 to 10,654
dim(x)
x <- calcNormFactors(x, method = "TMM")
```

Make boxplots
```{r}
ggparcoord(data.frame(log(x[[1]]/colMeans(x[[1]]))), columns=1:24, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
```

Create MDS plots
```{r}
library(RColorBrewer)
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
col.lane <- lane
levels(col.lane) <-  brewer.pal(nlevels(col.lane), "Set2")
col.lane <- as.character(col.lane)
```

```{r}
plotMDS(lcpm, labels=colnames(lcpm), col=col.group)
```

```{r}
plotMDS(lcpm, labels=colnames(lcpm), col=col.lane, dim=c(3,4))
```

```{r}
glMDSPlot(lcpm, labels=paste(group, lane, colnames(lcpm), sep="_"), groups=x$samples[,c(1,4)], launch=FALSE)
```

Create design matrix. There are many ways to setup a design matrix. Here, we removed the intercept from group (the first factor), but kept the intercept from lane. This allows us to do contrasts with group more easily.
```{r}
design <- model.matrix(~0+group+lane)
colnames(design) <- gsub("group", "", colnames(design))

contr.matrix <- makeContrasts(
   N_V = N-V,
   levels = colnames(design))
```

```{r}
par(mfrow=c(1,2))
v <- voom(x, design, plot=TRUE)
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean−variance trend")
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table. Significance is defined using an adjusted p-value cutoff that is set at 5% by default. For the comparison between expression levels in basal and LP, 4,127 genes are found to be down-regulated in basal relative to LP and 4,298 genes are up-regulated in basal relative to LP – a total of 8,425 DE genes.

```{r}
summary(decideTests(efit))
```

```{r}
pairNames <- colnames(contr.matrix)
metrics <- list()
for (i in 1:length(pairNames)) {
  temp <- topTreat(efit, coef=i, n=Inf)
  temp <- cbind(ID=rownames(temp), temp)
  temp$ID = as.character(temp$ID)
  metrics[[ pairNames[i] ]] <- temp
}
```

Save the data

```{r}
saveRDS(metrics, file="dataMetrics.Rds")
saveRDS(x, file="data.Rds")
```

### 2018-03-29 N_V/DESeq2/Method2/Clusters-St/Clusters_R.R

@@ -1,149 +1,0 @@
-library(bigPint)
-library(data.table)
-library(EDASeq)
-library(plyr)
-library(gridExtra)
-
-outDir = getwd()
-data <- readRDS("../../../data/data.Rds")
-data <- data[,c(1:6, 13:18, 7:12, 19:24)]
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-
-metrics <- readRDS("../dataMetrics.Rds")[["C_R"]]
-sigMets = metrics[which(metrics$padj<0.05),]
-sigR <- sigMets[which(sigMets$log2FoldChange<0),]
-
-
-# File output information
-plotName = "R"
-logSoy = data
-logSoy[,-1] <- log(data[,-1]+1)
-
-
-
-
-
-
-
-RowSD = function(x) {
-  sqrt(rowSums((x - rowMeans(x))^2)/(dim(x)[2] - 1))
-}
-
-data_Rownames <- data$ID
-data = data[,-1]
-rownames(data) <- data_Rownames
-data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
-data = as.data.frame(data)
-# Add mean and standard deviation for each row/gene
-data = mutate(data, mean = (NC.1+NC.2+NC.3+NC.4+NC.5+NC.6+NR.1+NR.2+NR.3+NR.4+NR.5+NR.6+VC.1+VC.2+VC.3+VC.4+VC.5+VC.6+VR.1+VR.2+VR.3+VR.4+VR.5+VR.6)/6, stdev = RowSD(cbind(NC.1,NC.2,NC.3,NC.4,NC.5,NC.6,NR.1,NR.2,NR.3,NR.4,NR.5,NR.6,VC.1,VC.2,VC.3,VC.4,VC.5,VC.6,VR.1,VR.2,VR.3,VR.4,VR.5,VR.6)))
-rownames(data)=data_Rownames
-data$ID <- data_Rownames
-
-dataqps <- t(apply(as.matrix(data[,1:24]), 1, scale))
-dataqps <- as.data.frame(dataqps)
-colnames(dataqps) <- colnames(data[,1:24])
-dataqps$ID <- rownames(dataqps)
-
-# Comine the filtered and remaining data
-fulls <- dataqps
-boxDat <- melt(fulls, id.vars="ID")
-colnames(boxDat) <- c("ID", "Sample", "Count")
-
[…]

### 2018-03-29 N_V/LimmaVoom/LimmaVoom-Virus/glimma-plots/js/MDS-Plot.js

@@ -1,0 +1,11 @@
+glimma.layout.setupGrid(d3.select(".container"), "md", [1,2]);
+glimma.storage.chartData.push(glimma.transform.toRowMajor({"dim1":[-0.5785,-0.4699,1.369,1.007,-0.3029,1.813,0.06861,-0.951,-0.9368,0.2962,0.01225,0.7433,-1.76,-0.009189,0.09881,0.587,0.6433,0.2241,0.7046,-1.359,-0.8414,0.1238,0.05039,-0.532],"dim2":[-0.5729,-0.6111,0.704,0.1337,-1.342,0.03975,-0.1947,-1.048,-1.22,-0.2123,-0.3622,0.2002,0.7449,-0.359,-0.08591,0.977,-0.3869,0.007298,-0.5821,1.017,0.9579,1.022,-0.5455,1.72],"dim3":[0.3079,1.011,-0.0679,0.8773,0.9045,-0.4713,-0.6728,-1.985,0.2426,-0.4701,-0.5376,-0.5152,0.2599,-0.2708,0.9866,0.05354,0.1361,0.4192,-0.2873,0.02982,-0.1307,0.1864,0.2697,-0.2758],"dim4":[-0.2408,-0.01183,0.8467,-0.5844,1.101,0.09891,0.02653,-0.1172,-0.8096,0.3605,0.4357,0.04951,0.9654,-0.1841,-0.2233,0.4394,0.227,-0.0241,-0.3712,0.2393,-0.105,-0.7516,-0.6815,-0.6855],"dim5":[0.323,0.3594,0.1285,-0.8252,-0.8739,0.1187,-0.257,-0.3374,0.3263,0.5357,-0.4023,-0.397,-0.6353,0.3851,0.03892,0.3664,0.2578,0.8117,0.3891,0.9593,0.3144,-0.457,-0.4409,-0.6883],"dim6":[-0.565,-0.5259,0.2571,-0.003531,0.8797,0.8769,-0.6119,-0.1603,0.6339,-0.4148,-0.1599,-0.3207,-0.4407,0.5818,-0.1726,-0.7251,0.09827,-0.5899,0.07485,0.7675,0.5199,-0.0699,-0.3149,0.3852],"dim7":[0.4807,0.1534,-0.6146,-0.1251,0.5775,-0.03163,1.34,-0.7133,-0.2981,0.5124,-0.09011,-0.422,-0.6375,0.4227,-0.6088,0.
[…]

### 2018-03-29 VirusHoneyBee/data/makeData.R

@@ -1,0 +1,10 @@
+library("merTools")
+library("readr")
+
+d <- read_delim("GSE65659_AntiviralResponseReadCounts.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
+
+data <- as.data.frame(d)
+data <- data[,1:7]
+colnames(data) <- c("ID", "C.1", "C.2", "C.3", "T.1", "T.2", "T.3")
+
+saveRDS(data, "data.Rds")

### 2018-03-29 VirusHoneyBee/makeData.R

@@ -1,9 +1,0 @@
-library("merTools")
-
-d <- read_delim("~/VirusHoneyBee/GSE65659_AntiviralResponseReadCounts.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
-
-data <- as.data.frame(d)
-data <- data[,1:7]
-colnames(data) <- c("ID", "C.1", "C.2", "C.3", "T.1", "T.2", "T.3")
-
-saveRDS(data, "data/data.Rds")

### 2018-03-29 VirusHoneyBee/DESeq2/Method1/makeDataMetrics.Rmd
Add all DEGs for VirusHoneyBee group

---
title: "Honey Bees - DESeq"
author: Lindsay Rutter
output:
  packagedocs::package_docs:
    toc: true
    toc_collapse: true
vignette: >
  %\VignetteEngine{packagedocs::redirect}
---

<meta http-equiv="content-type" content="text/html;charset=utf-8" />

```{r global_options, include=FALSE}
# R output pre blocks are styled by default to indicate output
knitr::opts_chunk$set(
  comment = NA,
  cache = TRUE,
  fig.height = 8,
  fig.width = 10
)
```

```{r}
library(DESeq2)
library(RColorBrewer)
library(data.table)
library(ggplot2)
library(gplots)

thisPath <- getwd()

data <- readRDS("../../data/data.Rds")
data <- as.matrix(data)

coldata = data.frame(row.names = colnames(data), virus = unlist(lapply(colnames(data), function (x) substring(unlist(strsplit(x, "[.]"))[1],1,1))), diet = unlist(lapply(colnames(data), function (x) substring(unlist(strsplit(x, "[.]"))[1],2,2))), treatment = unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1])))

dds = DESeqDataSetFromMatrix(countData = data, colData = coldata, design = ~ treatment)
dds <- DESeq(dds)
```

We can examine pairwise combinations of all samples

```{r}
uTreat <- unique(unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1])))
allPairs = data.frame(Treatment1 = factor(), Treatment2 = factor(), NumberDEG = numeric())
dataMetrics <- list()

for (i in 1:(length(uTreat)-1)){
  for (j in (i+1):length(uTreat)){
    res <- results(dds, contrast=c("treatment",uTreat[i],uTreat[j]))
    hist(res$pvalue, breaks=20, col="grey", main=paste("pvalue for", uTreat[i], "and", uTreat[j]))
    degLength <- length(which((res@listData)$padj <0.05))
    allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength))

    metrics = as.data.frame(res@listData)
    metrics = cbind(ID = res@rownames, metrics)
    metrics$ID = as.character(metrics$ID)
    dataMetrics[[paste0(uTreat[i], "_", uTreat[j])]] <- metrics
  }
}
allPairs <- allPairs[order(allPairs$NumberDEG),]
allPairs
saveRDS(dataMetrics, file="dataMetrics.Rds")
```

```{r}
plotDispEsts(dds, ylim = c(1e-6, 1e1))
```

Taking rlog transformation for EDA.

```{r}
rld <- rlog(dds)
rldDF <- as.data.frame(assay(rld))
saveRDS(rldDF, file="dataRLD.Rds")
```

```{r}
sampleDists <- dist(t(assay(rld)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(rld$treatment, rld$patient, sep="-")
colnames(sampleDistMatrix) <- NULL
colours = colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
heatmap.2(sampleDistMatrix, trace="none", col=colours)
```

```{r}
print(plotPCA(rld, intgroup="treatment") + geom_label(aes(label = name), label.size=0.001))
```

### 2018-03-29 VirusHoneyBee/EdgeR/edgeR-btwnLane/makeDataMetrics.Rmd
Add all DEGs for VirusHoneyBee group

--- 2018-03-29 N_V/EdgeR/edgeR-btwnLane/makeDataMetrics.Rmd
@@
 minLib <- min(y$samples$lib.size)
-keep <- rowSums(cpm(y)>4) >= 6
+keep <- rowSums(cpm(y)>6) >= 3
 # Number of genes 15,314--> 8,672
@@
 ```{r}
-ggparcoord(data.frame(log(y[[1]]/colMeans(y[[1]]))), columns=1:24, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
+ggparcoord(data.frame(log(y[[1]]/colMeans(y[[1]]))), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
 ```
@@
 ```{r}
-allGroups <- c(rep("N",12), rep("V",12))
+allGroups <- c(rep("C",3), rep("T",3))
 y$samples$group <- allGroups
@@
 ```{r}
-Group = factor(c(rep("N",12), rep("V",12)))
-targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
+Group = factor(c(rep("C",3), rep("T",3)))
+targets <- data.frame(rownames=paste0("Sample",1:6), Group = Group)
 design <- model.matrix(~0+Group, data=y$samples)
@@
     df <- y[[1]][rownames(y[[1]]) %in% rownames(lrt5), c(which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[i])), which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[j])))]
-    firstLarger <- length(which(apply(df, 1, function(x) mean(x[1:6]) > mean(x[7:12]))))
-    secondLarger <- length(which(apply(df, 1, function(x) mean(x[1:6]) < mean(x[7:12]))))
+    firstLarger <- length(which(apply(df, 1, function(x) mean(x[1:3]) > mean(x[4:6]))))
+    secondLarger <- length(which(apply(df, 1, function(x) mean(x[1:3]) < mean(x[4:6]))))
 

### 2018-03-29 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
Add all DEGs for VirusHoneyBee group

--- 2018-03-29 VirusHoneyBee/EdgeR/edgeR-btwnLane/makeDataMetrics.Rmd
@@
 beeCounts <- readRDS("../../data/data.Rds")
+
 beeCounts <- as.matrix(beeCounts)
@@
 ```{r}
-y[[1]] <- betweenLaneNormalization(y[[1]], which="full", round=FALSE)
+y <- calcNormFactors(y)
 ```
@@
 Group = factor(c(rep("C",3), rep("T",3)))
-targets <- data.frame(rownames=paste0("Sample",1:6), Group = Group)
+#targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
 design <- model.matrix(~0+Group, data=y$samples)

### 2018-03-29 VirusHoneyBee/LimmaVoom/LimmaVoom-Virus/makeDataMetrics.Rmd
Add all DEGs for VirusHoneyBee group

--- 2018-03-29 N_V/LimmaVoom/makeDataMetrics.Rmd
@@
 thisPath <- getwd()
-beeCounts <- readRDS("../data/data.Rds")
+beeCounts <- readRDS("../../data/data.Rds")
+colnames(beeCounts) <- c(paste0(rep("C",3),".",1:3),paste0(rep("T",3),".",1:3))
 beeCounts <- as.matrix(beeCounts)
@@
 
-group <- factor(c(rep("N",12), rep("V",12)))
+group <- factor(c(rep("C",3), rep("T",3)))
 x$samples$group <- group
-lane <- as.factor(c("L12","L12","L12","L12","L34","L34","L12","L12","L34","L34","L34","L34","L12","L12","L12","L12","L34","L34","L12","L12","L34","L34","L34","L34"))
-x$samples$lane <- lane
 ```
@@
 ```{r}
-ggparcoord(data.frame(log(x[[1]]/colMeans(x[[1]]))), columns=1:24, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
+ggparcoord(data.frame(log(x[[1]]/colMeans(x[[1]]))), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
 ```
@@
 col.group <- as.character(col.group)
-col.lane <- lane
-levels(col.lane) <-  brewer.pal(nlevels(col.lane), "Set2")
-col.lane <- as.character(col.lane)
 ```
@@
 
+Create design matrix. There are many ways to setup a design matrix. Here, we removed the intercept from group (the first factor). This allows us to do contrasts with group more easily.
 ```{r}
-plotMDS(lcpm, labels=colnames(lcpm), col=col.lane, dim=c(3,4))
-```
-
-```{r}
-glMDSPlot(lcpm, labels=paste(group, lane, colnames(lcpm), sep="_"), groups=x$samples[,c(1,4)], launch=FALSE)
-```
-
-Create design matrix.
-There are many ways to setup a design matrix.
-Here, we removed the intercept from group (the first factor), but kept the intercept from lane.
-This allows us to do contrasts with group more easily.
-```{r}
-design <- model.matrix(~0+group+lane)
+design <- model.matrix(~0+group)
 colnames(design) <- gsub("group", "", colnames(design))
@@
 contr.matrix <- makeContrasts(
-   N_V = N-V,
+   C_T = C-T,
    levels = colnames(design))

### 2018-03-29 VirusHoneyBee/LimmaVoom/makeDataMetrics.Rmd
Add all DEGs for VirusHoneyBee group

--- 2018-03-29 VirusHoneyBee/LimmaVoom/LimmaVoom-Virus/makeDataMetrics.Rmd
@@
 thisPath <- getwd()
-beeCounts <- readRDS("../../data/data.Rds")
-colnames(beeCounts) <- c(paste0(rep("C",3),".",1:3),paste0(rep("T",3),".",1:3))
+beeCounts <- readRDS("../data/data.Rds")
 beeCounts <- as.matrix(beeCounts)

### 2018-03-29 VirusHoneyBee/data/makeData.R

@@ -4,7 +4,8 @@ library("readr")
 d <- read_delim("GSE65659_AntiviralResponseReadCounts.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
 
 data <- as.data.frame(d)
-data <- data[,1:7]
-colnames(data) <- c("ID", "C.1", "C.2", "C.3", "T.1", "T.2", "T.3")
+rownames(data) <- data[,1]
+data <- data[,2:7]
+colnames(data) <- c("C.1", "C.2", "C.3", "T.1", "T.2", "T.3")
 
 saveRDS(data, "data.Rds")

### 2018-03-29 VirusHoneyBee/DESeq2/Method1/ClusterStandard/Clustering_data_FDR_05.R
Add standardized scatMat to C_T DESeq2

@@ -1,0 +1,147 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(matrixStats)
+
+getPCP <- function(nC){
+  
+  set.seed(1)
+  colList = scales::hue_pal()(nC+1)
+  k = cutree(hc, k=nC)
+  
+  yMin = min(sigDatas[,1:nColumns])
+  yMax = max(sigDatas[,1:nColumns])
+  
+  ###########################
+  
+  sbsDF <- data.frame()
+  for (i in 1:nC){
+    x = as.data.frame(sigDatas[which(k==i),])
+    xNames = rownames(x)
+    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+    sbsDF = rbind(sbsDF, data.frame(Cluster = paste("Cluster", i), FDR = xFDR))
+  }
+  
+  plot_clusters = lapply(1:nC, function(i){
+    x = as.data.frame(sigDatas[which(k==i),])
+    nGenes = nrow(x)
+    x$cluster = "color"
+    x$cluster2 = factor(x$cluster)
+    xNames = rownames(x)
+    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+    scatMatMetrics = list()
+    scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
+    scatMatMetrics[[currPair]]$FDR = 10e-10
+    scatMatMetrics[[currPair]]$ID = as.factor(as.character(scatMatMetrics[[currPair]]$ID))
+    
+    plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
+    
+    fileName = paste(getwd(), "/", outDir, "/", currPair, "_SM_", nC, "_", i, ".jpg", sep="")
+    plotDEG(data = plotDatas, dataMetrics = scatMatMetrics, option="scatterPoints", threshVar = "padj", threshVal = 0.05, degPointColor = colList[i+1], fileName=fileName)
+    
+    x$ID = xNames
+    
+    pcpDat <- melt(x[,c(1:(nColumns+1))], id.vars="ID")
+    colnames(pcpDat) <- c("ID", "Sample", "Count")
[…]

### 2018-03-31 Manuscript/chapter4.tex
Manuscript for Chapter 4

\documentclass[11pt,a4paper,oldfontcommands,openany]{memoir}
\usepackage{graphicx}
\usepackage[]{color}
\usepackage{courier}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[breakable, theorems, skins]{tcolorbox}
\usepackage[]{media9}
\usepackage{lipsum}
\tcbset{enhanced}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\DeclareRobustCommand{\mybox}[2][gray!15]{%
\begin{tcolorbox}[   %% Adjust the following parameters at will.
        breakable,
        left=0pt,
        right=0pt,
        top=0pt,
        bottom=0pt,
        colback=#1,
        colframe=#1,
        width=\dimexpr\textwidth\relax, 
        enlarge left by=0mm,
        boxsep=5pt,
        arc=0pt,outer arc=0pt,
        ]
        #2
\end{tcolorbox}
}

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\setlength{\parindent}{0pt} % Remove indent at new paragraphs
\setcounter{secnumdepth}{4}  % Remove section numbering at certain depth # If zero, then no numbering of sections
\setcounter{tocdepth}{4} % Determines number of subsections that will have tabs
\usepackage{tabu}
\usepackage{url}

\usepackage[round]{natbib}   % omit 'round' option if you prefer square brackets
\bibliographystyle{plainnat}

\usepackage{fixltx2e}
%\usepackage{graphicx}	% For external pictures
\usepackage{float}
\usepackage{subfig}	% Add subfigures within figures
\usepackage{verbatim}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{amssymb,amsbsy,amsmath}
\usepackage{epsfig}
\usepackage[left=3cm,top=3cm,bottom=3.5cm,right=3cm]{geometry} % For easy document margins
%\usepackage{fancyhdr} % For customization of header/footer
\usepackage{adjustbox}
\usepackage{framed}
\usepackage{enumitem}
\usepackage{caption}
\numberwithin{equation}{section} % Equation numbers relative to sections

%%%%% NEW ADDED FROM THESIS.TEX %%%%% 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
%\usepackage[dvips]{graphicx}
%\usepackage{times} %clash
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\newcommand{\code}[1]{{\texttt{#1}}}
\newcommand{\pkg}[1]{{\texttt{#1}}}
\newcommand{\class}[1]{{\textit{#1}}}
\newcommand{\R}{{\normalfont\textsf{R }}{}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\begin{document}

\sloppy

%%%%%%%%%%%%%%%%%%%%% TITLE PAGE %%%%%%%%%%%%%%%%%%%%%

% {
% \centering
% ~\vspace{\fill}
% 
% \vspace{2.5cm}
% 
% {\LARGE Thesis Proposal}
% 
% \vspace{1cm}
% 
% {\LARGE\textbf{Visualization methods for genealogical and RNA-sequencing datasets}}
% 
% \vspace{1cm}
% 
% {\LARGE Lindsay Rutter}
% 
% \vspace{4cm}
% 
% {\LARGE Program of Study Committee:}
% 
% \vspace{1cm}
% 
% {\LARGE Dianne Cook (Major Professor)}
% 
% \vspace{.25cm}
% 
% {\LARGE Amy Toth (Major Professor)}
% 
% \vspace{.25cm}
% 
% {\LARGE Heike Hofmann}
% 
% \vspace{.25cm}
% 
% {\LARGE Daniel Nettleton}
% 
% \vspace{.25cm}
% 
% {\LARGE James Reecy}
% 
% \vspace{2.5cm}
% 
% {\centerline{\large May 16, 2016}}
% }
% 
% \clearpage
%\cleardoublepage

%%% CHAPTER'S STYLE
\chapterstyle{bianchi}
\setsecheadstyle{\Large\bfseries\sffamily\raggedright}
\setsubsecheadstyle{\large\bfseries\sffamily\raggedright}
\setsubsubsecheadstyle{\bfseries\sffamily\raggedright}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\tableofcontents

\setlength{\parskip}{10pt} % Inter-paragraph spacing

%%% ADDED FROM THESIS.TEX %%%%%%%%%
\OnehalfSpacing

\chapter{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background to data visualization}

\citealt{ped}
%\ref{sec:ggenealogy}
% \code{ggparcoord}
% \Figure \ref{fig:porcupine}

% \begin{figure}[H]
% \centering
%     \begin{framed}
%     \includegraphics[width=0.8\textwidth]{porcupine}
%     \end{framed}
%     \caption{Three}
%     \label{fig:porcupine}
% \end{figure}

% \section{Literature review}
% \label{sec:litReview}

% \pkg{R} statistical programming language allows for tools to be distributed and modified at ease, encourages cross-platform collaboration, and provides a foundation for effective and aesthetic data visualization from the grammar of graphics. There are several useful \pkg{R} packages that offer tools for analyzing and visualizing genealogical datasets. Here, we introduce these packages, and emphasize the shortcomings for which our package \pkg{ggenealogy} brings to this collection of work.

%\clearpage

%\begin{enumerate}
%\item The data can be viewed at \url{http://shiny.soybase.org/CNV/}.
%\end{enumerate}

% \begin{tabular}{|p{2.6cm}|p{5.65cm}|p{5.65cm}|}
%  \hline
%  & \textbf{Drumming Treatment (D)} & \textbf{No Drumming Treatment (N)} \\ 
%  \hline
%  \textbf{Restricted} & \textbf{DR} & \textbf{NR} \\ 
%  \textbf{Nutrition (R)}& Predict most worker-like & Predict some worker-like \\
%  & gene expression & gene expression \\
%  \hline
%  \textbf{Unrestricted} & \textbf{DU} & \textbf{NU} \\ 
%  \textbf{Nutrition (U)} & Predict some worker-like & Predict least worker-like \\
%  & gene expression & gene expression \\
%  \hline
% \end{tabular}

\bibliography{chapter4}

\end{document}

### 2018-03-31 Overlaps/GO/ReadingGO.R

@@ -1,0 +1,2 @@
+dat <- readRDS("gD.rds")
+cat(dat,sep="\n")

### 2018-04-06 Manuscript/chapter4.tex
Materials and ch4 updates
Intro 2 para

@@
 
-\chapter{Introduction}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background to data visualization}
+\chapter{Gene expression responses to interactive stressors of diet quality and viral infection in \textit{Apis mellifera}}
 
-\citealt{ped}
+\section{Introduction}
+
+Commerically managed honeybees have undergone unusually large declines in the United States and parts of Europe over the past decade (\citealt{ccd1}, \citealt{ccd2}, \citealt{ccd3}), with annual mortality rates exceeding what beekeepers consider sustainable (\citealt{ccd5}, \citealt{ccd6}).
+More than 70 percent of major global food crops (including fruits, vegatables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued wordwide at \$175 billion (\citealt{ccd7}).
+As honeybees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications regarding agricultural sustainability (\citealt{ccd4}).
+
+Honeybee declines have been associated with several factors, including pesticide use, parasites, pathogens, habitat loss, and poor nutrition (\citealt{factors}, \citealt{factors2}).
+Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honeybees in interactive fashions as the environment changes (\citealt{interacting}).
+Two factors that have posed heightened danger to honeybee health related to environmental changes are nutrition and viruses.
+
+\section{Methods}
+
+\section{Results}
+
+\section{Discussion}
+
+% SYNTAX
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %\ref{sec:ggenealogy}

### 2018-04-06 Manuscript/chapter4.tex
Nutrition p

@@
 Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honeybees in interactive fashions as the environment changes (\citealt{interacting}).
-Two factors that have posed heightened danger to honeybee health related to environmental changes are nutrition and viruses.
+Nutrition and viral infection are two broad factors that pose heightened dangers to honeybee health in response to recent environmental changes.
+
+Pollen is the main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honeybees (\citealt{source}, \citealt{source2}).
+At the individual level, pollen supplies most of the nutrients necessary for physiological development (\citealt{brodschneider}) and is believed to have considerable impact on longevity (\citealt{longevity}).
+At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen (\citealt{jelly}, \citealt{jelly2}).
+Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honeybees.
+In particular, honeybees face less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets (\citealt{diverse}, \citealt{diverse2}, \citealt{diverse3}).
+Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas (\citealt{undeveloped}), and beekeepers rank poor nutrition as one of the main reasons for colony losses (\citealt{bkLoss}).
+Understanding how undiversified diets affect honeybee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world (\citealt{ag}, \citealt{ag2}).
 

### 2018-04-06 Manuscript/chapter4.tex
Virus p

@@
 
+Viral infection was a comparatively minor problem in honeybees until the last century when Varroa destructor (an ectoparasitic mite) spread worldwide (\citealt{miteSpread}).
+This mite feeds on honeybee hemolymph (\citealt{hemolymph}), transmits cocktails of viruses, and supports replication of certain viruses (\citealt{miteVirus}, \citealt{miteVirus2}, \citealt{miteVirus3}).
+One of these viruses that has been linked to honeybee decline is Israeli Acute Paralysis Virus (IAPV).
+A positive-sense RNA virus of the Dicistroviridae family (\citealt{fam}), IAPV causes infected honeybees to display shivering wings, decreased locomotion, muscle spams, and paralysis, and 80\% of caged infected adult honeybees die prematurely (\citealt{symptoms}).
+IAPV has demonstrated higher infectious capacities than other honeybee viruses in certain conditions (\citealt{carrillo}) and is more prevalent in colonies that do not survive the winter (\citealt{winter}).
+Its role in the rising phenomenon of ``Colony Collapse Disorder'' (in which the majority of worker bees disappear from a hive) remains unclear: It has been implicated in some studies (\citealt{iapvCCD}, \citealt{iapvCCD2}) but not in other studies (\citealt{ccd1}, \citealt{iapvCCD3}, \citealt{fam}).
+Nonetheless, it seems likely that IAPV reduces colony strength and survival.
+
 \section{Methods}

### 2018-04-07 Manuscript/chapter4.tex
Interacting p

@@
 Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honeybees.
-In particular, honeybees face less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets (\citealt{diverse}, \citealt{diverse2}, \citealt{diverse3}).
+In particular, honeybees are confronted with less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets (\citealt{diverse}, \citealt{diverse2}, \citealt{diverse3}).
 Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas (\citealt{undeveloped}), and beekeepers rank poor nutrition as one of the main reasons for colony losses (\citealt{bkLoss}).
@@
 This mite feeds on honeybee hemolymph (\citealt{hemolymph}), transmits cocktails of viruses, and supports replication of certain viruses (\citealt{miteVirus}, \citealt{miteVirus2}, \citealt{miteVirus3}).
+More than 20 honeybee viruses have been identified (\citealt{numVirus}).
 One of these viruses that has been linked to honeybee decline is Israeli Acute Paralysis Virus (IAPV).
@@
 Nonetheless, it seems likely that IAPV reduces colony strength and survival.
+
+Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}).
+We recently used laboratory cages and nucleus hive experiments to investigate how these two factors interact (\citealt{adamInt}).
+As part of our experiment, we specifically studied the interactive effects of IAPV infection and monofloral diet quality on titers on honeybee mortality rates.
+
+There are several reasons why, in part of our study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
+First, when assessing diet diversity, a sugar diet is often used as a control.
+However, such an experimental design does not reflect real-world conditions for honeybees as they rarely face a total lack of pollen (\citealt{DiPasquale}).
+Second, in studies that compared honeybee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet (\citealt{DiPasquale}).
+Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010).
+Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced wtih less diversified diets (monoculture) (\citealt{landscape1}, \citealt{brodschneider}).
+As a result, there is a need to better understand how monofloral diets affect honeybee health as a step toward mitigating the negative impact of human activity on the honeybee population.
 

### 2018-04-07 Manuscript/chapter4.tex
Images

@@
 We recently used laboratory cages and nucleus hive experiments to investigate how these two factors interact (\citealt{adamInt}).
-As part of our experiment, we specifically studied the interactive effects of IAPV infection and monofloral diet quality on titers on honeybee mortality rates.
+As part of our experiment, we specifically studied the interactive effects of IAPV infection and monofloral diet quality on titers and honeybee mortality rates.
 
-There are several reasons why, in part of our study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
+There are several reasons why, in this part of our previous study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
 First, when assessing diet diversity, a sugar diet is often used as a control.
@@
 
+Consequently, in our prior study, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut).
+Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}).
+For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV.
+This experimental design resulted in four treatment groups that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.
+
+We discovered that the higher-quality Castanea diet had the ability to significantly reduce mortality in the presence of IAPV infection (Figure \ref{fig:mortalityAll}).
+
+Mortality of bees 72 h post-inoculation differed among the treatment groups (mixed model ANOVA across all treatment groups, df=7, 108; F=19.28; p<0.0001).
+There was a significant effect of both virus treatment (mixed model ANOVA, df=1, 108; F=95.26; p<0.0001) and diet treatment (mixed model ANOVA, df=3,108; F=9.93; p<0.001) with a significant interaction between the two factors (mixed model ANOVA, df=3, 108; F=3.311, p=0.023).
+The virus treatment was effective-- in all cases, virus-treated bees showed significantly higher mortality than bees from cages fed the same diet but no virus (Tukey HSD, p<0.05).
+Without virus exposure, none of the treatment groups showed differences in mortality (Tukey HSD, p>0.05).
+However, between the virus-treated groups, both bees fed Castanea or polyfloral pollen showed significantly lower mortality than bees fed no pollen (Tukey HSD, p<0.05); bees fed Cistus pollen exhibited intermediate mortality (Tukey HSD, p<0.05; Figure 1).
+Overall, we find that virus infection increases mortality, but some pollen diets are capable of significantly reducing the severity of this effect.
+
+glht (General Linear Hypotheses)
+
+\begin{figure}[H]
+\centering
+  \begin{framed}
+  \includegraphics[width=0.8\textwidth]{Images/mortalityAll}
+  \end{framed}
+  \caption{Mortality rates between the four treatments. Each treatment contains 15 samples (cages).}
+  \label{fig:mortalityAll}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+  \begin{framed}
+  \includegraphics[width=0.8\textwidth]{Images/mortalityRNASeq}
+  \end{framed}
+  \caption{Mortality rates between the four treatments. Out of the original 15 samples (cages), each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing.}
+  \label{fig:mortalityRNASeq}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+  \begin{framed}
+  \includegraphics[width=0.8\textwidth]{Images/IAPVAll}
+  \end{framed}
+  \caption{IAPV titers between the four treatments. Each treatment contains 15 samples (cages).}
+  \label{fig:IAPVAll}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+  \begin{framed}
+  \includegraphics[width=0.8\textwidth]{Images/IAPVRNASeq}
+  \end{framed}
+  \caption{IAPV titers between the four treatments. Out of the original 15 samples (cages), each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing.}
+  \label{fig:IAPVRNASeq}
+\end{figure}
+
+Summary of what that paper found mortality, micronutrients
+ We found that high quality diets (polyfloral pollen and high quality single-source pollen) have the potential to reduce mortality in the face of infection with Israeli acute paralysis virus (IAPV)
+- There was a significant interaction between diet and virus infection on mortality, with associated differences in bee virus titers, suggesting good diets can help bees keep viral infection levels down
+
 \section{Methods}
+
+RNA-seq
+RNA prep
+Illumina sequencing at DNA facility
+What version of genome mapped to
+What mapping program did he use
+What percent of reads mapped to the genome
+Statistics (DESeq, EdgerR, visualizations, overlaps fisher, GO)
 

### 2018-04-08 Manuscript/chapter4.tex
Intro and methods draft

--- 2018-04-07 Manuscript/chapter4.tex
@@
 
-\chapter{Gene expression responses to interactive stressors of diet quality and viral infection in \textit{Apis mellifera}}
+\chapter{Gene expression responses to diet quality and viral infection in \textit{Apis mellifera}}
 
@@
 Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honeybees.
-In particular, honeybees are confronted with less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets (\citealt{diverse}, \citealt{diverse2}, \citealt{diverse3}).
+In particular, honeybees are confronted with less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets (\citealt{diverse}, \citealt{diverse2}, \citealt{alaux}).
 Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas (\citealt{undeveloped}), and beekeepers rank poor nutrition as one of the main reasons for colony losses (\citealt{bkLoss}).
@@
 
+Consequently, in our prior study, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut).
+Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}).
+For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV.
+This experimental design resulted in four treatment groups that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.
+
+Mortality rates of honeybees 72 hour post-inoculation differed (or did not?) among the treatment groups (mixed model ANOVA across all treatment groups, df=@@@, @@@; F=@@@; p<@@@).
+The effect of virus treatment (mixed model ANOVA, df=@@@, @@@; F=@@@; p<@@@), diet treatment (mixed model ANOVA, df=@@@,@@@; F=@@@; p<@@@), and interaction between the two factors (mixed model ANOVA, df=@@@, @@@; F=@@@, p=@@@) did (or did not?) differ.
+The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05).
+Without virus exposure, there was only an intermediate reduction in mortality rate for bees fed Castanea pollen?
+(Tukey HSD, p>0.05).
+However, with virus exposure, there was a significant reduction in mortality rate for beeds fed Castanea pollen (Tukey HSD, p<0.05).
+Overall, we discovered that the higher-quality Castanea diet had the ability to significantly reduce mortality in the presence of IAPV infection compared to the lower-quality Cistus diet (Figure \ref{fig:mortalityAll}).
+
+Comment on (Figure \ref{fig:IAPVAll}) @@@
+
+\begin{figure}[H]
+\centering
+  \begin{framed}
+  \includegraphics[width=0.8\textwidth]{Images/mortalityAll}
+  \end{framed}
+  \caption{Mortality rates between the four treatments. Each treatment consisted of 15 cages each with 35 honeybees.}
+  \label{fig:mortalityAll}
+\end{figure}
+
+Comments on (Figure \ref{fig:IAPVAll}) and (Figure \ref{fig:IAPVRNASeq}). @@@
+
+\begin{figure}[H]
+\centering
+  \begin{framed}
+  \includegraphics[width=0.8\textwidth]{Images/IAPVAll}
+  \end{framed}
+  \caption{IAPV titers between the four treatments. Each treatment contains 15 samples (cages).}
+  \label{fig:IAPVAll}
+\end{figure}
+
+Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms.
+Transcriptomics is one means to achieve this goal.
+As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees.
+As far as we know, there are little to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and litte to no studies investigating gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense in honeybees.
+
+In this study, we examine how monofloral diets and viral innoculation influence gene expression patterns in honeybees by focusing on four treatment groups (Cistus diet without IAPV exposure, Castanea diet without IAPV exposure, Cistus diet with IAPV exposure, and Castanea diet with IAPV exposure).
+We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section).
+We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the interactive effect of the two factors on gene expression patterns.
+
+We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues (\citealt{galbraith}).
+While our study examined honeybees from polyandrous colonies, the Galbraith study examined honeybees from single-drone colonies.
+As a consequence, our honeybees will have on average about 75\% genetic variance, and the honeybees from the Galbraith study will have about 25\% genetic variance (\citealt{sisters}).
+We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates.
+At the same time, our honeybees will be more likely to display the health benefits gained from increased genotypic variation within colonies, including decreased parasitic load (\citealt{multParasite}), increased tolerance to environmental changes (\citealt{divHyp2}), and increasead colony performance (\citealt{geneticDiverse}, \citealt{geneticDiverse2}).
+Given that honeybees are naturally very polyandrous (\citealt{patriline}), our honeybees may also reflect more realistic environmental and genetic simulations.
+Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data.
+We wish to examine how the gene expression effects of IAPV innoculation compare between these two studies that used such different experimental designs.
+Namely, we use visualization techniques to assess the signal:to:noise ratios between these two datasets, and differential gene expression (DEG) analyses to assess any overlaps between these two datasets.
+
 \section{Methods}
 
+The procedures we used to prepare virus inoculum, infect and feed caged honeybees, and quantify IAPV can be reviewed in our previous work (\citealt{adamInt}).
+The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report (\citealt{adamInt}).
+
+\subsection{RNA extraction}
+
+Fifteen cages per treatment were originally sampled.
+Six live honeybees from each cage were randomly selected 36 hours post inoculation and placed into tubes.
+Tubes were kept on dry ice and then transferred into a -80C freezer until processing.
+Eight cages were randomly selected from the original 15 cages, and 2 honeybees per cage were randomly selected from the original six live honeybees per cage.
+RNA from each pool of two honeybees were extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment.
+Samples were suspended in water to 200-400 ng/$\mu$l.
+All samples were then tested on a Bioanalyzer at the DNA core facility to ensure quality (RIN>8).
+
+\subsection{Gene expression}
+
+Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing).
+Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}).
+We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs).
+We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs).
+We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honeybees (\citealt{galbraith}).
+We tested each DEG analysis using recommended parameters with DESeq2 (\citealt{deseq2}), edgeR (\citealt{edger}), and LimmaVoom (\citealt{limma}).
+In all cases, we used a false discovery rate (FDR) threshold of 0.05 (\citealt{benjamini}).
+Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).
+
+\subsection{Visualization}
+
+We used @@@ visualization tools from @@@ and visual inference techniques to assess the signal:to:noise ratio in the datasets and to assess the suitability of the DEG calls.
+
+\subsection{Gene Ontology}
+
+DEGs were uploaded as a background list to DAVID Bioinformatics Resources 6.7 (\citealt{davidBio}, \citealt{davidBio2}).
+The overrepresented gene ontology (GO) terms of DEGs were identified using the BEEBASE\_ID identifier.
+To fine-tune the GO term list, only significant terms (FDR < 0.05) and those correlating to Biological Processes were considered.
+The refined GO term list was then imported into REVIGO (\citealt{revigo}), which uses semantic similarity measures to cluster long lists of GO terms.
+
 \section{Results}
+
+When statistically analyzing the subset of this data that was used for RNA-sequencing analysis, we found that the mortality rates across diet quality and virus exposure at least numerically retained the same trends (Figure \ref{fig:mortalityRNASeq}).
+
+Comments on (Figure \ref{fig:mortalityRNASeq}) and (Figure \ref{fig:IAPVRNASeq}). @@@
+
+\begin{figure}[H]
+\centering
+  \begin{framed}
+  \includegraphics[width=0.8\textwidth]{Images/mortalityRNASeq}
+  \end{framed}
+  \caption{Mortality rates between the four treatments.
+Each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing out of its original 15 samples (cages).
+Two honeybees were randomly selected from each sample (cage) to create a pooled RNA-seqeuncing sample.}
+  \label{fig:mortalityRNASeq}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+  \begin{framed}
+  \includegraphics[width=0.8\textwidth]{Images/IAPVRNASeq}
+  \end{framed}
+  \caption{IAPV titers between the four treatments.
+Each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing out of its original 15 samples (cages).
+Two honeybees were randomly selected from each sample (cage) to create a pooled RNA-seqeuncing sample.}
+  \label{fig:IAPVRNASeq}
+\end{figure}
 

### 2018-04-08 Manuscript/chapter4.tex
Draft to review during meeting on 20180409

@@
 \usepackage{lipsum}
+
+\usepackage{lineno} %% LINE COUNTS
+\linenumbers %% LINE COUNTS
+
 \tcbset{enhanced}
@@
 We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues (\citealt{galbraith}).
-While our study examined honeybees from polyandrous colonies, the Galbraith study examined honeybees from single-drone colonies.
-As a consequence, our honeybees will have on average about 75\% genetic variance, and the honeybees from the Galbraith study will have about 25\% genetic variance (\citealt{sisters}).
+While our study examines honeybees from polyandrous colonies, the Galbraith study examined honeybees from single-drone colonies.
+As a consequence, our honeybees will have an average of about 75\% genetic variance, and the honeybees from the Galbraith study will have an average of about 25\% genetic variance (\citealt{sisters}).
 We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates.
-At the same time, our honeybees will be more likely to display the health benefits gained from increased genotypic variation within colonies, including decreased parasitic load (\citealt{multParasite}), increased tolerance to environmental changes (\citealt{divHyp2}), and increasead colony performance (\citealt{geneticDiverse}, \citealt{geneticDiverse2}).
+At the same time, our honeybees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load (\citealt{multParasite}), increased tolerance to environmental changes (\citealt{divHyp2}), and increasead colony performance (\citealt{geneticDiverse}, \citealt{geneticDiverse2}).
 Given that honeybees are naturally very polyandrous (\citealt{patriline}), our honeybees may also reflect more realistic environmental and genetic simulations.
 Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data.
-We wish to examine how the gene expression effects of IAPV innoculation compare between these two studies that used such different experimental designs.
-Namely, we use visualization techniques to assess the signal:to:noise ratios between these two datasets, and differential gene expression (DEG) analyses to assess any overlaps between these two datasets.
+We wish to explore how the gene expression effects of IAPV innoculation compare between these two studies that used such different experimental designs.
+To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
+It is our hope that this aspect of our study may shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honeybees.
 
@@
 
+@@@ Which body tissue was used for RNA extraction?
+
 \subsection{Gene expression}
@@
 Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).
+
+@@@ What percent of reads mapped?
 

### 2018-04-09 VirusHoneyBee/DESeq2/Method1/makeDataMetrics.Rmd
Examine overlap between Galbraith supplemental DEGs and our DESEq DEGs of their data

@@
     degLength <- length(which((res@listData)$padj <0.05))
-    allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength))
 
@@
     dataMetrics[[paste0(uTreat[i], "_", uTreat[j])]] <- metrics
+
+    t1Length <- length(which(metrics[which(metrics$padj <0.05),]$log2FoldChange>0))
+    t2Length <- length(which(metrics[which(metrics$padj <0.05),]$log2FoldChange<0))
+
+    allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength, Treatment1Larger = t1Length, Treatment2Larger = t2Length))
   }

### 2018-04-09 Overlaps/Overlaps.R

@@ -1,5 +1,6 @@
 library(VennDiagram)
 library(readr)
+library(readxl)
 
 #draw.triple.venn(65, 75, 85, 35, 15, 25, 5, c("First", "Second", "Third"))
 
@@ -166,6 +167,33 @@ invisible(dev.off())
 # 11825 | 7 | 3317 | 13
 fisher.test(matrix(c(N, intgEB-EBcrossArea, inttEB-EBcrossArea, EBcrossArea), nrow=2), alternative="greater")
 
+############### Venn diagrams for Supplemental Galbraith versus Toth per Method ############## 
+##############################################################################################
+
+# Keep 753 DEGs from Galbraith paper
+gDSupp = read_excel("../VirusHoneyBee/GalbraithSuppFile.xlsx")
+gDSupp = as.data.frame(gDSupp)
+
+keepColNms <- gDSupp[3,]
+keepColNms <- as.data.frame(keepColNms)
+colnames(keepColNms) <- NULL
+keepColNms <- unlist(keepColNms)
+
+keepDEGs <- gDSupp[4:(4+753-1),]
+keepDEGs <- as.data.frame(keepDEGs)
+colnames(keepDEGs) <- keepColNms
+
+# Extract 753 IDs from Galbraith paper
+gDSupp = keepDEGs$id
+gDSupp = unname(sapply(gDSupp, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+
+gtDSupp = intersect(gDSupp, gD2)
+draw.pairwise.venn(area1=length(gD), area2=length(gDSupp), cross.area = length(gtDSupp), category=c("Galbraith Us", "Galbraith Supp"))
+invisible(dev.off())
+
+###################### Save GO Terms ##################### 
+##########################################################
+  
 # Save GO 
 gD = sapply(gD, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1])
 gE = sapply(gE, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1])

### 2018-04-09 Manuscript/chapter4.tex
Examine overlap between Galbraith supplemental DEGs and our DESEq DEGs of their data
Incorporate all of Toth's comments

@@
 Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}).
-We recently used laboratory cages and nucleus hive experiments to investigate how these two factors interact (\citealt{adamInt}).
-As part of our experiment, we specifically studied the interactive effects of IAPV infection and monofloral diet quality on titers and honeybee mortality rates.
+We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show a significant interaction between diet quality and virus infection.
+Specifically, high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen (\citealt{adamInt}).
+
+Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms.
+Transcriptomics is one means to achieve this goal.
+As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees.
+As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense in honeybees.
+
+In this study, we examine how monofloral diets and viral innoculation influence gene expression patterns in honeybees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure).
+We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section).
+We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the interactive effect of the two factors on gene expression patterns.
+
+We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues (\citealt{galbraith}).
+As RNA-sequencing data can be highly noisy, this comparison allowed us to characterize how repeatable and robust our RNA-seq results were in comparison to previous studies.
+Importantly, we use an in-depth data visualization approach to explore and validate our data, and suggest such an approach can be useful for cross-study comparisons of RNA-sequencing data in the future.
+
+\section{Methods}
+
+Details of the procedures we used to prepare virus inoculum, infect and feed caged honeybees, and quantify IAPV can be reviewed in our previous work (\citealt{adamInt}).
+The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report (\citealt{adamInt}).
+
+\subsection{Design of diet experiment}
 
@@
 For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV.
-This experimental design resulted in four treatment groups that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.
+This experimental design resulted in four treatment groups (Cistus pollen without IAPV exposure, Castanea pollen without IAPV exposure, Cistus pollen with IAPV exposure, and Castanea pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.
 
-Mortality rates of honeybees 72 hour post-inoculation differed (or did not?) among the treatment groups (mixed model ANOVA across all treatment groups, df=@@@, @@@; F=@@@; p<@@@).
-The effect of virus treatment (mixed model ANOVA, df=@@@, @@@; F=@@@; p<@@@), diet treatment (mixed model ANOVA, df=@@@,@@@; F=@@@; p<@@@), and interaction between the two factors (mixed model ANOVA, df=@@@, @@@; F=@@@, p=@@@) did (or did not?) differ.
-The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05).
-Without virus exposure, there was only an intermediate reduction in mortality rate for bees fed Castanea pollen?
-(Tukey HSD, p>0.05).
-However, with virus exposure, there was a significant reduction in mortality rate for beeds fed Castanea pollen (Tukey HSD, p<0.05).
-Overall, we discovered that the higher-quality Castanea diet had the ability to significantly reduce mortality in the presence of IAPV infection compared to the lower-quality Cistus diet (Figure \ref{fig:mortalityAll}).
+\subsection{RNA extraction}
 
-Comment on (Figure \ref{fig:IAPVAll}) @@@
+Fifteen cages per treatment were originally sampled.
+Six live honeybees from each cage were randomly selected 36 hours post inoculation and placed into tubes.
+Tubes were kept on dry ice and then transferred into a -80C freezer until processing.
+Eight cages were randomly selected from the original 15 cages, and 2 honeybees per cage were randomly selected from the original six live honeybees per cage.
+Whole body RNA from each pool of two honeybees were extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment.
+Samples were suspended in water to 200-400 ng/$\mu$l.
+All samples were then tested on a Bioanalyzer at the DNA core facility to ensure quality (RIN>8).
 
-\begin{figure}[H]
-\centering
-  \begin{framed}
-  \includegraphics[width=0.8\textwidth]{Images/mortalityAll}
-  \end{framed}
-  \caption{Mortality rates between the four treatments. Each treatment consisted of 15 cages each with 35 honeybees.}
-  \label{fig:mortalityAll}
-\end{figure}
+\subsection{Gene expression}
 
-Comments on (Figure \ref{fig:IAPVAll}) and (Figure \ref{fig:IAPVRNASeq}). @@@
+Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing).
+A standard Illumina mRNA library was prepared by the DNA facility.
+Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}).
+We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs).
+We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs).
+We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honeybees (\citealt{galbraith}).
+We tested each DEG analysis using recommended parameters with DESeq2 (\citealt{deseq2}), edgeR (\citealt{edger}), and LimmaVoom (\citealt{limma}).
+In all cases, we used a false discovery rate (FDR) threshold of 0.05 (\citealt{benjamini}).
+Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).
 
-\begin{figure}[H]
-\centering
-  \begin{framed}
-  \includegraphics[width=0.8\textwidth]{Images/IAPVAll}
-  \end{framed}
-  \caption{IAPV titers between the four treatments. Each treatment contains 15 samples (cages).}
-  \label{fig:IAPVAll}
-\end{figure}
+@@@ What percent of reads mapped?
+@@@ Total number of raw reads
+@@@ How many lanes
+@@@ How many samples per lane
 
-Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms.
-Transcriptomics is one means to achieve this goal.
-As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees.
-As far as we know, there are little to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and litte to no studies investigating gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense in honeybees.
-
-In this study, we examine how monofloral diets and viral innoculation influence gene expression patterns in honeybees by focusing on four treatment groups (Cistus diet without IAPV exposure, Castanea diet without IAPV exposure, Cistus diet with IAPV exposure, and Castanea diet with IAPV exposure).
-We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section).
-We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the interactive effect of the two factors on gene expression patterns.
+\subsection{Comparison to previous studies on transcriptomic response to viral infection}
 
@@
 It is our hope that this aspect of our study may shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honeybees.
-
-\section{Methods}
-
-The procedures we used to prepare virus inoculum, infect and feed caged honeybees, and quantify IAPV can be reviewed in our previous work (\citealt{adamInt}).
-The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report (\citealt{adamInt}).
-
-\subsection{RNA extraction}
-
-Fifteen cages per treatment were originally sampled.
-Six live honeybees from each cage were randomly selected 36 hours post inoculation and placed into tubes.
-Tubes were kept on dry ice and then transferred into a -80C freezer until processing.
-Eight cages were randomly selected from the original 15 cages, and 2 honeybees per cage were randomly selected from the original six live honeybees per cage.
-RNA from each pool of two honeybees were extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment.
-Samples were suspended in water to 200-400 ng/$\mu$l.
-All samples were then tested on a Bioanalyzer at the DNA core facility to ensure quality (RIN>8).
-
-@@@ Which body tissue was used for RNA extraction?
-
-\subsection{Gene expression}
-
-Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing).
-Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}).
-We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs).
-We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs).
-We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honeybees (\citealt{galbraith}).
-We tested each DEG analysis using recommended parameters with DESeq2 (\citealt{deseq2}), edgeR (\citealt{edger}), and LimmaVoom (\citealt{limma}).
-In all cases, we used a false discovery rate (FDR) threshold of 0.05 (\citealt{benjamini}).
-Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).
-
-@@@ What percent of reads mapped?
 
@@
 
-When statistically analyzing the subset of this data that was used for RNA-sequencing analysis, we found that the mortality rates across diet quality and virus exposure at least numerically retained the same trends (Figure \ref{fig:mortalityRNASeq}).
+\subsection{Phenotypic results}
 
-Comments on (Figure \ref{fig:mortalityRNASeq}) and (Figure \ref{fig:IAPVRNASeq}). @@@
+We reanalyzed our previously published dataset with a subset more relevant to our RNA-sequencing approaches in the current study that have a more focused question regarding diet quality.
+We briefly show it again here to inform the RNA-seq comparison because we reduced the number of treatments (from eight to four) from the original published data (\citealt{adamInt}).
+When statistically analyzing the subset of this data that was used for RNA-sequencing analysis, we found that the mortality rates across diet quality and virus exposure at least numerically retained the same trends (Figure \ref{fig:4panels}).
+
+Mortality rates of honeybees 72 hour post-inoculation differed (or did not?) among the treatment groups (mixed model ANOVA across all treatment groups, df=@@@, @@@; F=@@@; p<@@@).
+The effect of virus treatment (mixed model ANOVA, df=@@@, @@@; F=@@@; p<@@@), diet treatment (mixed model ANOVA, df=@@@,@@@; F=@@@; p<@@@), and interaction between the two factors (mixed model ANOVA, df=@@@, @@@; F=@@@, p=@@@) did (or did not?) differ.
+The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05).
+Without virus exposure, there was only an intermediate reduction in mortality rate for bees fed Castanea pollen?
+(Tukey HSD, p>0.05).
+However, with virus exposure, there was a significant reduction in mortality rate for beeds fed Castanea pollen (Tukey HSD, p<0.05).
+Overall, we discovered that the higher-quality Castanea diet had the ability to significantly reduce mortality in the presence of IAPV infection compared to the lower-quality Cistus diet (Figure \ref{fig:4panels}).
+
+Comment on (Figure \ref{fig:4panels}) @@@
 
@@
   \begin{framed}
-  \includegraphics[width=0.8\textwidth]{Images/mortalityRNASeq}
+  \includegraphics[width=0.8\textwidth]{Images/4panels}
   \end{framed}
-  \caption{Mortality rates between the four treatments.
-Each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing out of its original 15 samples (cages).
-Two honeybees were randomly selected from each sample (cage) to create a pooled RNA-seqeuncing sample.}
-  \label{fig:mortalityRNASeq}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-  \begin{framed}
-  \includegraphics[width=0.8\textwidth]{Images/IAPVRNASeq}
-  \end{framed}
-  \caption{IAPV titers between the four treatments.
-Each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing out of its original 15 samples (cages).
-Two honeybees were randomly selected from each sample (cage) to create a pooled RNA-seqeuncing sample.}
-  \label{fig:IAPVRNASeq}
+  \caption{Mortality rates for all cages, mortality rates for subset of honeybees used for RNA-seq, IAPV titers for all cages, IAPV titers for subset of honeybees used for RNA-seq subset.}
+  \label{fig:4panels}
 \end{figure}

### 2018-04-09 NC_NR_VC_VR/DESeq2/Interaction/makeDataMetrics-Test.Rmd
Incorporate all of Toth's comments

--- 2018-03-29 VirusHoneyBee/DESeq2/Method1/makeDataMetrics.Rmd
@@
 ---
-title: "Honey Bees - DESeq"
+title: "Honey Bees - DESeq Interaction"
 author: Lindsay Rutter
@@
 library(DESeq2)
+library(gplots)
 library(RColorBrewer)
@@
 library(ggplot2)
-library(gplots)
 
@@
 
-dds = DESeqDataSetFromMatrix(countData = data, colData = coldata, design = ~ treatment)
+dds = DESeqDataSetFromMatrix(countData = data, colData = coldata, design = ~ virus + diet + virus*diet)
+
 dds <- DESeq(dds)
+
+# Intercept, virus_V_vs_N, diet_R_vs_C, virusV.dietR
+resultsNames(dds)
+
+res <- results(dds, name="Intercept")
+length(which((res@listData)$padj <0.05))
+
+res <- results(dds, name="virus_V_vs_N")
+length(which((res@listData)$padj <0.05))
+
+res <- results(dds, name="diet_R_vs_C")
+length(which((res@listData)$padj <0.05))
+
+res <- results(dds, name="virusV.dietR")
+length(which((res@listData)$padj <0.05))
+
 ```
-
-We can examine pairwise combinations of all samples
-
-```{r}
-uTreat <- unique(unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1])))
-allPairs = data.frame(Treatment1 = factor(), Treatment2 = factor(), NumberDEG = numeric())
-dataMetrics <- list()
-
-for (i in 1:(length(uTreat)-1)){
-  for (j in (i+1):length(uTreat)){
-    res <- results(dds, contrast=c("treatment",uTreat[i],uTreat[j]))
-    hist(res$pvalue, breaks=20, col="grey", main=paste("pvalue for", uTreat[i], "and", uTreat[j]))
-    degLength <- length(which((res@listData)$padj <0.05))
-    allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength))
-
-    metrics = as.data.frame(res@listData)
-    metrics = cbind(ID = res@rownames, metrics)
-    metrics$ID = as.character(metrics$ID)
-    dataMetrics[[paste0(uTreat[i], "_", uTreat[j])]] <- metrics
-  }
-}
-allPairs <- allPairs[order(allPairs$NumberDEG),]
-allPairs
-saveRDS(dataMetrics, file="dataMetrics.Rds")
-```
-
-```{r}
-plotDispEsts(dds, ylim = c(1e-6, 1e1))
-```
-
-Taking rlog transformation for EDA.
-
-```{r}
-rld <- rlog(dds)
-rldDF <- as.data.frame(assay(rld))
-saveRDS(rldDF, file="dataRLD.Rds")
-```
-
-```{r}
-sampleDists <- dist(t(assay(rld)))
-sampleDistMatrix <- as.matrix(sampleDists)
-rownames(sampleDistMatrix) <- paste(rld$treatment, rld$patient, sep="-")
-colnames(sampleDistMatrix) <- NULL
-colours = colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
-heatmap.2(sampleDistMatrix, trace="none", col=colours)
-```
-
-```{r}
-print(plotPCA(rld, intgroup="treatment") + geom_label(aes(label = name), label.size=0.001))
-```

### 2018-04-09 Manuscript/chapter4.tex
Added transcriptomic review

@@
 
-Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms.
+Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality.
+For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability/vigor (resilience).
 Transcriptomics is one means to achieve this goal.
+Transcriptomic analysis can help to identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection.
+This information, heretofore lacking in the literature, can help us to better understand how good nutrition may be able to serve as a "buffer" against other stressors (\citealt{AdamTothReview}).
 As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees.

### 2018-04-11 N_V/EdgeR/edgeR-btwnLane/makeDataMetrics.Rmd
Standard scatterplots Gailbraith data

@@
 ---
-title: "Honey Bees - edgeR"
+title: "Honey Bees - edgeR-btwnLane"
 author: Lindsay Rutter
@@
 minLib <- min(y$samples$lib.size)
-keep <- rowSums(cpm(y)>4) >= 6
+nRep = length(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])) / length(unique(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])))
+keep <- rowSums(cpm(y)>round(minLib/1000000)) >= nRep/2
 # Number of genes 15,314--> 8,672

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
Standard scatterplots Gailbraith data

@@
 minLib <- min(y$samples$lib.size)
-keep <- rowSums(cpm(y)>6) >= 3
+nRep = length(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])) / length(unique(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])))
+keep <- rowSums(cpm(y)>round(minLib/1000000)) >= nRep/2
 # Number of genes 15,314--> 8,672

### 2018-04-11 N_V/LineupsDESeq2/Log/Lineup.R

@@ -1,0 +1,15 @@
+library(edgeR)
+library(ggplot2)
+library(GGally)
+library(EDASeq)
+library(utils)
+library(data.table)
+library(bigPint)
+
+thisPath <- getwd()
+
+beeCounts <- readRDS("../../data/data.Rds")
+beeCounts <- cbind(ID=rownames(beeCounts), beeCounts)
+beeCounts$ID <- as.character(beeCounts$ID)
+
+plotPermutationsD(beeCounts, nPerm = 20, topThresh = 50, option="log", outDir = getwd())

### 2018-04-11 N_V/LineupsDESeq2/Raw/Lineup.R

@@ -1,0 +1,14 @@
+library(edgeR)
+library(ggplot2)
+library(GGally)
+library(EDASeq)
+library(utils)
+library(data.table)
+
+thisPath <- getwd()
+
+beeCounts <- readRDS("../../data/data.Rds")
+beeCounts <- cbind(ID=rownames(beeCounts), beeCounts)
+beeCounts$ID <- as.character(beeCounts$ID)
+
+plotPermutationsD(beeCounts, nPerm = 20, topThresh = 50, outDir = getwd())

### 2018-04-11 N_V/LineupsDESeq2/Standardize/Lineup.R

@@ -1,0 +1,14 @@
+library(edgeR)
+library(ggplot2)
+library(GGally)
+library(EDASeq)
+library(utils)
+library(data.table)
+
+thisPath <- getwd()
+
+beeCounts <- readRDS("../../data/data.Rds")
+beeCounts <- cbind(ID=rownames(beeCounts), beeCounts)
+beeCounts$ID <- as.character(beeCounts$ID)
+
+plotPermutationsD(beeCounts, nPerm = 20, topThresh = 50, option="standardize", outDir = getwd())

### 2018-04-11 VirusHoneyBee/DESeq2/Method1/scatMatStand/scatMatStand.R

@@ -1,0 +1,78 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- length(which(metricsAll[["C_T"]]$padj<0.05))
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "padj", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith DESeq2 Significant Genes (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
+  data = as.data.frame(data)
+  # Add mean and standard deviation for each row/gene
[…]

### 2018-04-11 VirusHoneyBee/DESeq2/Method1/scatMatStand/scatMatStandTop213.R

@@ -1,0 +1,82 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- 213
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+metricsAll[["C_T"]] <- metricsAll[["C_T"]][with(metricsAll[["C_T"]], order(padj)),]
+metricsAll[["C_T"]]$padj = 1
+metricsAll[["C_T"]]$padj[1:nGenes] = 0.01
+
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "padj", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith DESeq2 Significant Genes Top 213 (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
[…]

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR-btwnLane/scatMatStand/scatMatStand.R

@@ -1,0 +1,78 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- length(which(metricsAll[["C_T"]]$FDR<0.05))
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "FDR", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith DESeq2 Significant Genes (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
+  data = as.data.frame(data)
+  # Add mean and standard deviation for each row/gene
[…]

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR-btwnLane/scatMatStand/scatMatStandBottom.R

@@ -1,0 +1,82 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- 730
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+metricsAll[["C_T"]]$FDR = 1
+lastI = nrow(metricsAll[["C_T"]])
+metricsAll[["C_T"]]$FDR[(lastI-730+1):lastI] = 0.01
+
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "FDR", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith EdgeR-Btwn Bottom Significant Genes (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
+  data = as.data.frame(data)
[…]

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR-btwnLane/scatMatStand/scatMatStandTop.R

@@ -1,0 +1,81 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- 730
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+metricsAll[["C_T"]]$FDR = 1
+metricsAll[["C_T"]]$FDR[1:730] = 0.01
+
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "FDR", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith EdgeR-Btwn Top Significant Genes (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
+  data = as.data.frame(data)
+  # Add mean and standard deviation for each row/gene
[…]

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR-btwnLane/scatMatStand/scatMatStandTop213.R

@@ -1,0 +1,81 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- 213
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+metricsAll[["C_T"]]$FDR = 1
+metricsAll[["C_T"]]$FDR[1:nGenes] = 0.01
+
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "FDR", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith EdgeR-Btwn Top Significant Genes (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
+  data = as.data.frame(data)
+  # Add mean and standard deviation for each row/gene
[…]

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR/scatMatStand/scatMatStand.R

@@ -1,0 +1,78 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- length(which(metricsAll[["C_T"]]$FDR<0.05))
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "FDR", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith EdgeR Significant Genes (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
+  data = as.data.frame(data)
+  # Add mean and standard deviation for each row/gene
[…]

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR/scatMatStand/scatMatStandTop213.R

@@ -1,0 +1,81 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- 213
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+metricsAll[["C_T"]]$FDR = 1
+metricsAll[["C_T"]]$FDR[1:nGenes] = 0.01
+
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "FDR", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith EdgeR Significant Genes Top 213 (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
+  data = as.data.frame(data)
+  # Add mean and standard deviation for each row/gene
[…]

### 2018-04-11 VirusHoneyBee/LimmaVoom/scatMatStand/scatMatStand.R

@@ -1,0 +1,78 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+data <- as.data.frame(readRDS("../../data/data.Rds"))
+metricsAll <- readRDS("../dataMetrics.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- length(which(metricsAll[["C_T"]]$adj.P.Val<0.05))
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "adj.P.Val", threshVal = 0.05) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith Limma Significant Genes (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
+  data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
+  data = as.data.frame(data)
+  # Add mean and standard deviation for each row/gene
[…]

### 2018-04-11 VirusHoneyBee/LineupsDESeq2/Log/Lineups.R

@@ -1,0 +1,15 @@
+library(edgeR)
+library(ggplot2)
+library(GGally)
+library(EDASeq)
+library(utils)
+library(data.table)
+library(bigPint)
+
+thisPath <- getwd()
+
+beeCounts <- readRDS("../../data/data.Rds")
+beeCounts <- cbind(ID=rownames(beeCounts), beeCounts)
+beeCounts$ID <- as.character(beeCounts$ID)
+
+plotPermutationsD(beeCounts, nPerm = 10, topThresh = 50, option="log", outDir = getwd())

### 2018-04-11 VirusHoneyBee/LineupsDESeq2/Raw/Lineups.R

@@ -1,0 +1,15 @@
+library(edgeR)
+library(ggplot2)
+library(GGally)
+library(EDASeq)
+library(utils)
+library(data.table)
+library(bigPint)
+
+thisPath <- getwd()
+
+beeCounts <- readRDS("../../data/data.Rds")
+beeCounts <- cbind(ID=rownames(beeCounts), beeCounts)
+beeCounts$ID <- as.character(beeCounts$ID)
+
+plotPermutationsD(beeCounts, nPerm = 10, topThresh = 50, outDir = getwd())

### 2018-04-11 VirusHoneyBee/LineupsDESeq2/Standardize/Lineups.R

@@ -1,0 +1,15 @@
+library(edgeR)
+library(ggplot2)
+library(GGally)
+library(EDASeq)
+library(utils)
+library(data.table)
+library(bigPint)
+
+thisPath <- getwd()
+
+beeCounts <- readRDS("../../data/data.Rds")
+beeCounts <- cbind(ID=rownames(beeCounts), beeCounts)
+beeCounts$ID <- as.character(beeCounts$ID)
+
+plotPermutationsD(beeCounts, nPerm = 10, topThresh = 50, option="standardize", outDir = getwd())

### 2018-04-11 VirusHoneyBee/LimmaVoom/ClusterStandard/Clustering_data_FDR_05.R
Fix LimmaVoom StandardCluster

@@ -90,7 +90,8 @@ pair1 <- strsplit(currPair, "_")[[1]][1]
 pair2 <- strsplit(currPair, "_")[[1]][2]
 
 metrics <- metricsAll[[currPair]]
-data <- as.data.frame(readRDS("../../data/data.Rds"))
+data <- readRDS("../data.Rds")
+data <- as.data.frame(data[[1]])
 data <- data[,which(sapply(colnames(data), function(x) unlist(strsplit(x,"[.]"))[1]) %in% c(pair1, pair2))]
 data<- cbind(ID = rownames(data), data)
 data$ID <- as.character(data$ID)
@@ -129,7 +130,13 @@ datas[nID,1:nColumns] <- 0
 boxDat <- melt(datas, id.vars="ID")
 colnames(boxDat) <- c("ID", "Sample", "Count")
 
-sigDatas = datas[which(metricsAll[["C_T"]]$adj.P.Val<0.05),]
+#### WRONG!!!!!
+#sigDatas = datas[which(metricsAll[["C_T"]]$adj.P.Val<0.05),]
+
+sigID = metricsAll[["C_T"]][which(metricsAll[["C_T"]]$adj.P.Val<0.05),]$ID
+sigDatas = datas[which(rownames(datas) %in% sigID),]
+#sigDatas = sigDatas[,1:6]
+
 dendo = sigDatas
 rownames(dendo) = NULL
 d = dist(as.matrix(dendo))

### 2018-04-11 VirusHoneyBee/LimmaVoom/scatMatStand/scatMatStandCluster1.R

@@ -1,0 +1,84 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(data.table)
+
+colList = scales::hue_pal()(4+1)
+
+data <- readRDS("../data.Rds")
+data <- as.data.frame(data[[1]])
+metricsAll <- readRDS("../dataMetrics.Rds")
+clust1ID <- readRDS("../ClusterStandard/sigDatas.Rds")
+
+data$ID <- rownames(data)
+data <- data[,c(7,1:6)]
+
+# Filter, normalize, and standardize the data so each gene has mean=0 and stdev=1
+res <- filterStandardizeKL(data)
+# Fitered data standardized
+datas <- res[["fulls"]]
+
+# Indices of the NAN rows
+nID <- which(is.nan(datas$C.1))
+# Set these filtered values that have all same values for samples to 0
+datas[nID,1:6] <- 0
+
+yMin = min(datas[,1:6])
+yMax = max(datas[,1:6])
+
+datas <- datas[,c(7,1:6)]
+nGenes <- length(clust1ID)
+
+metricsAll[["C_T"]]$ID = as.factor(as.character(metricsAll[["C_T"]]$ID))
+metricsAll[["C_T"]] = metricsAll[["C_T"]][which(metricsAll[["C_T"]]$ID %in% clust1ID),]
+
+ret <- plotDEG(data = datas, dataMetrics = metricsAll, option="scatterPoints", threshVar = "adj.P.Val", threshVal = 1) # degPointColor = colList[j]
+ret[["C_T"]] + xlab("Standardized Count") + ylab("Standardized Count") + ggtitle(paste("Galbraith Limma Significant Genes (n=", format(nGenes, big.mark=",", scientific=FALSE), ")",sep="")) + theme(plot.title = element_text(hjust = 0.5, size=11), axis.text=element_text(size=11), axis.title=element_text(size=12), strip.text = element_text(size = 10))
+
+filterStandardizeKL <- function(data){
+  
+  data_Rownames <- data$ID
+  data = data[,-1]
+  rownames(data) <- data_Rownames
[…]

### 2018-04-11 VirusHoneyBee/DESeq2/Method1/makeDataMetrics.Rmd
Fix Standard Cluster Limma and edgeR

@@
 allPairs
+saveRDS(data, file="data.Rds")
 saveRDS(dataMetrics, file="dataMetrics.Rds")

### 2018-04-11 VirusHoneyBee/DESeq2/Method1/ClusterStandard/Clustering_data_FDR_05.R

@@ -90,7 +90,7 @@ pair1 <- strsplit(currPair, "_")[[1]][1]
 pair2 <- strsplit(currPair, "_")[[1]][2]
 
 metrics <- metricsAll[[currPair]]
-data <- as.data.frame(readRDS("../../../data/data.Rds"))
+data <- as.data.frame(readRDS("../data.Rds"))
 data <- data[,which(sapply(colnames(data), function(x) unlist(strsplit(x,"[.]"))[1]) %in% c(pair1, pair2))]
 data<- cbind(ID = rownames(data), data)
 data$ID <- as.character(data$ID)

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR-btwnLane/ClusterStandard/Clustering_data_FDR_05.R

@@ -76,8 +76,8 @@ getPCP <- function(nC){
   ggBP
   invisible(dev.off())
 
-   jpeg(file = paste(getwd(), "/", outDir, "/", plotName, "_", nC, ".jpg", sep=""), width=1000, height=700)
-  p = do.call("grid.arrange", c(plot_clusters, ncol=ceiling(nC/2)))
+   jpeg(file = paste(getwd(), "/", outDir, "/", plotName, "_", nC, ".jpg", sep=""), width=1000, height=1400)
+  p = do.call("grid.arrange", c(plot_clusters, ncol=2))
    invisible(dev.off())
 }
   
@@ -90,7 +90,7 @@ pair1 <- strsplit(currPair, "_")[[1]][1]
 pair2 <- strsplit(currPair, "_")[[1]][2]
 
 metrics <- metricsAll[[currPair]]
-data <- as.data.frame(readRDS("../../../data/data.Rds"))
+data <- as.data.frame(readRDS("../data.Rds"))
 data <- data[,which(sapply(colnames(data), function(x) unlist(strsplit(x,"[.]"))[1]) %in% c(pair1, pair2))]
 data<- cbind(ID = rownames(data), data)
 data$ID <- as.character(data$ID)
@@ -129,7 +129,11 @@ datas[nID,1:nColumns] <- 0
 boxDat <- melt(datas, id.vars="ID")
 colnames(boxDat) <- c("ID", "Sample", "Count")
 
-sigDatas = datas[which(metricsAll[["C_T"]]$FDR<0.05),]
+#### WRONG!!!!!!!!
+#sigDatas = datas[which(metricsAll[["C_T"]]$FDR<0.05),]
+sigID = metricsAll[["C_T"]][which(metricsAll[["C_T"]]$FDR<0.05),]$ID
+sigDatas = datas[which(rownames(datas) %in% sigID),]
+
 dendo = sigDatas
 rownames(dendo) = NULL
 d = dist(as.matrix(dendo))

### 2018-04-11 VirusHoneyBee/EdgeR/edgeR/ClusterStandard/Clustering_data_FDR_05.R

@@ -129,7 +129,12 @@ datas[nID,1:nColumns] <- 0
 boxDat <- melt(datas, id.vars="ID")
 colnames(boxDat) <- c("ID", "Sample", "Count")
 
-sigDatas = datas[which(metricsAll[["C_T"]]$FDR<0.05),]
+# WRONG!!!!!!!!!!!
+###sigDatas = datas[which(metricsAll[["C_T"]]$FDR<0.05),]
+
+sigID = metricsAll[["C_T"]][which(metricsAll[["C_T"]]$FDR<0.05),]$ID
+sigDatas = datas[which(rownames(datas) %in% sigID),]
+
 dendo = sigDatas
 rownames(dendo) = NULL
 d = dist(as.matrix(dendo))

### 2018-04-12 C_R/DESeq2/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

--- 2018-04-11 VirusHoneyBee/DESeq2/Method1/makeDataMetrics.Rmd
@@
 library(gplots)
+library(GGally)
 
@@
 
-data <- readRDS("../../data/data.Rds")
+data <- readRDS("../data/data.Rds")
 data <- as.matrix(data)
 
-coldata = data.frame(row.names = colnames(data), virus = unlist(lapply(colnames(data), function (x) substring(unlist(strsplit(x, "[.]"))[1],1,1))), diet = unlist(lapply(colnames(data), function (x) substring(unlist(strsplit(x, "[.]"))[1],2,2))), treatment = unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1])))
+coldata = data.frame(row.names = colnames(data), treatment = unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1])))
 
@@
 ```
+
+```{r}
+ggparcoord(data.frame(log(as.data.frame(assay(rld))/colMeans(as.data.frame(assay(rld))))), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
+```
+
+```{r}
+dds <- estimateSizeFactors(dds)
+normDat <- counts(dds, normalized=TRUE)
+normDat <- as.data.frame(normDat)
+ggparcoord(data.frame(log(normDat)/colMeans(normDat)), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
+```

### 2018-04-12 C_R/EdgeR/edgeR-btwnLane/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

--- 2018-04-11 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
@@
 ---
-title: "Honey Bees - edgeR"
+title: "Honey Bees - edgeR-btwnlane"
 author: Lindsay Rutter
@@
 beeCounts <- readRDS("../../data/data.Rds")
-
 beeCounts <- as.matrix(beeCounts)
@@
 ```{r}
-y <- calcNormFactors(y)
+y[[1]] <- betweenLaneNormalization(y[[1]], which="full", round=FALSE)
 ```
@@
 ```{r}
-ggparcoord(data.frame(log(y[[1]]/colMeans(y[[1]]))), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
+ggparcoord(data.frame(log(y[[1]]/colMeans(y[[1]]))), columns=1:24, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
 ```
@@
 ```{r}
-allGroups <- c(rep("C",3), rep("T",3))
+allGroups <- c(rep("C",12), rep("R",12))
 y$samples$group <- allGroups
@@
 ```{r}
-Group = factor(c(rep("C",3), rep("T",3)))
-#targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
+Group = factor(c(rep("C",12), rep("R",12)))
+targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
 design <- model.matrix(~0+Group, data=y$samples)

### 2018-04-12 C_R/EdgeR/edgeR/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

--- 2018-04-11 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
@@
 ```{r}
-Group = factor(c(rep("C",3), rep("T",3)))
-#targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
+Group = factor(c(rep("C",12), rep("R",12)))
 design <- model.matrix(~0+Group, data=y$samples)

### 2018-04-12 C_R/LimmaVoom/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

--- 2018-03-29 VirusHoneyBee/LimmaVoom/makeDataMetrics.Rmd
@@
 
-group <- factor(c(rep("C",3), rep("T",3)))
+group <- factor(c(rep("C",12), rep("R",12)))
 x$samples$group <- group
@@
 cpm <- cpm(x)
-keep.exprs <- rowSums(cpm>1)>=4 # tried filtering up to 24 and not much difference
+nRep = length(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])) / length(unique(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])))
+keep.exprs <- rowSums(cpm>1)>=nRep # tried filtering up to 24 and not much difference
 x <- x[keep.exprs,, keep.lib.sizes=FALSE] # 15,314 to 10,654
@@
 ```{r}
-ggparcoord(data.frame(log(x[[1]]/colMeans(x[[1]]))), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
+ggparcoord(data.frame(log(x[[1]]/colMeans(x[[1]]))), columns=1:24, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
 ```
@@
 contr.matrix <- makeContrasts(
-   C_T = C-T,
+   C_R = C-R,
    levels = colnames(design))

### 2018-04-12 Manuscript/chapter4.tex
Standard scatterplots Gailbraith data
Redo 2 level factor comparisons D, E, L

--- 2018-04-09 Manuscript/chapter4.tex
@@
 
-Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms.
+Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality.
+For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (resilience).
 Transcriptomics is one means to achieve this goal.
+Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection.
+This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a "buffer" against other stressors (\citealt{AdamTothReview}).
 As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees.
-As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense in honeybees.
+As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honeybee gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense.
 
@@
 
-\subsection{Design of diet experiment}
+\subsection{Design of two-factor experiment}
 
-There are several reasons why, in this part of our previous study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
+There are several reasons why we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
 First, when assessing diet diversity, a sugar diet is often used as a control.
@@
 Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010).
-Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced wtih less diversified diets (monoculture) (\citealt{landscape1}, \citealt{brodschneider}).
+Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced with less diversified diets (monoculture) (\citealt{landscape1}, \citealt{brodschneider}).
 As a result, there is a need to better understand how monofloral diets affect honeybee health as a step toward mitigating the negative impact of human activity on the honeybee population.
 
-Consequently, in our prior study, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut).
+Consequently, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut).
 Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}).
@@
 Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).
+The \pkg{eulerr} shiny application was used to construct Venn diagram overlap images (\citealt{euler}).
 

### 2018-04-12 N_V/EdgeR/edgeR-btwnLane/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

@@
 ---
-title: "Honey Bees - edgeR-btwnLane"
+title: "Honey Bees - edgeR"
 author: Lindsay Rutter

### 2018-04-12 N_V/EdgeR/edgeR/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

@@
 Group = factor(c(rep("N",12), rep("V",12)))
-targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
 design <- model.matrix(~0+Group, data=y$samples)

### 2018-04-12 N_V/LimmaVoom/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

@@
 x$samples$group <- group
-lane <- as.factor(c("L12","L12","L12","L12","L34","L34","L12","L12","L34","L34","L34","L34","L12","L12","L12","L12","L34","L34","L12","L12","L34","L34","L34","L34"))
-x$samples$lane <- lane
 ```

### 2018-04-12 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

@@
 Group = factor(c(rep("C",3), rep("T",3)))
-#targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
 design <- model.matrix(~0+Group, data=y$samples)

### 2018-04-12 C_R/Lineups/Log/Lineup.R

@@ -1,0 +1,14 @@
+library(edgeR)
+library(ggplot2)
+library(GGally)
+library(EDASeq)
+library(utils)
+library(data.table)
+
+thisPath <- getwd()
+
+beeCounts <- readRDS("../../data/data.Rds")
+beeCounts <- cbind(ID=rownames(beeCounts), beeCounts)
+beeCounts$ID <- as.character(beeCounts$ID)
+
+plotPermutations(beeCounts, nPerm = 20, topThresh = 50, option="log", outDir = getwd())

### 2018-04-12 C_R/data/makeData.R

@@ -1,0 +1,13 @@
+beeCounts <-read.delim(file="AllLaneCount.txt",row.names=1,stringsAsFactors = FALSE)
+colnames(beeCounts) <- c("NC.1", "NC.2", "NR.1", "VR.1", "NS.1", "VP.1", "NS.2", "VR.2", "NP.1", "VP.2", "VC.1", "NP.2", "VP.3", "NP.3", "VS.1", "VS.2", "VC.2", "NC.3", "VP.4", "NC.4", "NR.2", "VC.3", "VC.4", "NP.4", "VR.3", "NC.5", "VS.3", "NP.5", "VC.5", "VS.4", "NS.3", "VS.5", "VP.5", "NR.3", "NR.4", "VC.6", "NS.4", "NC.6", "NP.6", "VR.4", "NR.5", "NR.6", "NS.5", "VP.6", "NS.6", "VR.5", "VR.6", "VS.6")
+data <- beeCounts[ , order(names(beeCounts))]
+
+treatment = unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1]))
+iTreatment = which(treatment %in% c("NC","NR","VC","VR"))
+
+data = data[,iTreatment]
+data=data[,c(1:6,13:18,7:12,19:24)]
+
+colnames(data) = c(c(paste0("C.",1:12)), c(paste0("R.",1:12)))
+
+saveRDS(data, file="data.Rds")

### 2018-04-12 N_V/DESeq2/ClusterStandard/Clustering_data_FDR_05.R

@@ -1,0 +1,147 @@
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(matrixStats)
+
+getPCP <- function(nC){
+  
+  set.seed(1)
+  colList = scales::hue_pal()(nC+1)
+  k = cutree(hc, k=nC)
+  
+  yMin = min(sigDatas[,1:nColumns])
+  yMax = max(sigDatas[,1:nColumns])
+  
+  ###########################
+  
+  sbsDF <- data.frame()
+  for (i in 1:nC){
+    x = as.data.frame(sigDatas[which(k==i),])
+    xNames = rownames(x)
+    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+    sbsDF = rbind(sbsDF, data.frame(Cluster = paste("Cluster", i), FDR = xFDR))
+  }
+  
+  plot_clusters = lapply(1:nC, function(i){
+    x = as.data.frame(sigDatas[which(k==i),])
+    nGenes = nrow(x)
+    x$cluster = "color"
+    x$cluster2 = factor(x$cluster)
+    xNames = rownames(x)
+    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+    scatMatMetrics = list()
+    scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
+    scatMatMetrics[[currPair]]$FDR = 10e-10
+    scatMatMetrics[[currPair]]$ID = as.factor(as.character(scatMatMetrics[[currPair]]$ID))
+    
+    plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
+    
+    fileName = paste(getwd(), "/", outDir, "/", currPair, "_SM_", nC, "_", i, ".jpg", sep="")
+    plotDEG(data = plotDatas, dataMetrics = scatMatMetrics, option="scatterPoints", threshVar = "padj", threshVal = 0.05, degPointColor = colList[i+1], fileName=fileName)
+    
+    x$ID = xNames
+    
+    pcpDat <- melt(x[,c(1:(nColumns+1))], id.vars="ID")
+    colnames(pcpDat) <- c("ID", "Sample", "Count")
[…]

### 2018-04-12 VirusHoneyBee/DESeq2/Method3/ClusterStandard/Clustering_data_FDR_05.R

@@ -1,147 +1,0 @@
-library(rtracklayer)
-library(Rsamtools)
-library(grid)
-library(GenomicAlignments)
-library(ggplot2)
-library(GGally)
-library(edgeR)
-library(stringr)
-library(EDASeq)
-library(dplyr)
-library(matrixStats)
-library(gridExtra)
-library(reshape2)
-library(scales)
-library(bigPint)
-library(matrixStats)
-
-getPCP <- function(nC){
-  
-  set.seed(1)
-  colList = scales::hue_pal()(nC+1)
-  k = cutree(hc, k=nC)
-  
-  yMin = min(sigDatas[,1:nColumns])
-  yMax = max(sigDatas[,1:nColumns])
-  
-  ###########################
-  
-  sbsDF <- data.frame()
-  for (i in 1:nC){
-    x = as.data.frame(sigDatas[which(k==i),])
-    xNames = rownames(x)
-    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
-    sbsDF = rbind(sbsDF, data.frame(Cluster = paste("Cluster", i), FDR = xFDR))
-  }
-  
-  plot_clusters = lapply(1:nC, function(i){
-    x = as.data.frame(sigDatas[which(k==i),])
-    nGenes = nrow(x)
-    x$cluster = "color"
-    x$cluster2 = factor(x$cluster)
-    xNames = rownames(x)
-    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
-    scatMatMetrics = list()
-    scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
-    scatMatMetrics[[currPair]]$FDR = 10e-10
-    scatMatMetrics[[currPair]]$ID = as.factor(as.character(scatMatMetrics[[currPair]]$ID))
-    
-    plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
-    
-    fileName = paste(getwd(), "/", outDir, "/", currPair, "_SM_", nC, "_", i, ".jpg", sep="")
-    plotDEG(data = plotDatas, dataMetrics = scatMatMetrics, option="scatterPoints", threshVar = "padj", threshVal = 0.05, degPointColor = colList[i+1], fileName=fileName)
-    
-    x$ID = xNames
-    
-    pcpDat <- melt(x[,c(1:(nColumns+1))], id.vars="ID")
-    colnames(pcpDat) <- c("ID", "Sample", "Count")
[…]

### 2018-04-13 C_R/EdgeR/edgeR/makeDataMetrics.Rmd
Updated up and down regulated DEG counts

@@
     df <- y[[1]][rownames(y[[1]]) %in% rownames(lrt5), c(which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[i])), which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[j])))]
-    firstLarger <- length(which(apply(df, 1, function(x) mean(x[1:3]) > mean(x[4:6]))))
-    secondLarger <- length(which(apply(df, 1, function(x) mean(x[1:3]) < mean(x[4:6]))))
+    firstLarger <-  length(which(lrt5$logFC>0))
+    secondLarger <-  length(which(lrt5$logFC<0))
 

### 2018-04-13 NC_NR_VC_VR/EdgeR/edgeR/makeDataMetrics.Rmd
Updated up and down regulated DEG counts

--- 2018-04-13 C_R/EdgeR/edgeR/makeDataMetrics.Rmd
@@
 minLib <- min(y$samples$lib.size)
-nRep = length(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])) / length(unique(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])))
-keep <- rowSums(cpm(y)>round(minLib/1000000)) >= nRep/2
+keep <- rowSums(cpm(y)>4) >= 6
 # Number of genes 15,314--> 8,672
@@
 ```{r}
-allGroups <- c(rep("C",12), rep("R",12))
+allGroups <- c(rep("NC",6), rep("NR",6), rep("VC",6), rep("VR",6))
 y$samples$group <- allGroups
@@
 ```{r}
-Group = factor(c(rep("C",12), rep("R",12)))
+Group = factor(c(rep("NC",6), rep("NR",6), rep("VC",6), rep("VR",6)))
+targets <- data.frame(rownames=paste0("Sample",1:24), Virus = c(rep("N",12), rep("V",12)), Diet = c(rep("C",6), rep("R",6), rep("C",6), rep("R",6)), Group = Group)
 design <- model.matrix(~0+Group, data=y$samples)

### 2018-04-13 N_V/EdgeR/edgeR/makeDataMetrics.Rmd
Updated up and down regulated DEG counts

@@
     df <- y[[1]][rownames(y[[1]]) %in% rownames(lrt5), c(which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[i])), which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[j])))]
-    firstLarger <- length(which(apply(df, 1, function(x) mean(x[1:3]) > mean(x[4:6]))))
-    secondLarger <- length(which(apply(df, 1, function(x) mean(x[1:3]) < mean(x[4:6]))))
+    firstLarger <- length(which(lrt5$logFC>0))
+    secondLarger <- length(which(lrt5$logFC<0))
 

### 2018-04-13 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
Updated up and down regulated DEG counts

@@
     df <- y[[1]][rownames(y[[1]]) %in% rownames(lrt5), c(which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[i])), which(sapply(colnames(y[[1]]), function(x) strsplit(x, "[.]")[[1]][1]) %in% (colnames(fit)[j])))]
-    firstLarger <- length(which(apply(df, 1, function(x) mean(x[1:3]) > mean(x[4:6]))))
-    secondLarger <- length(which(apply(df, 1, function(x) mean(x[1:3]) < mean(x[4:6]))))
+    firstLarger <- length(which(lrt5$logFC>0)) 
+    secondLarger <- length(which(lrt5$logFC<0))
 

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method2/Clusters-St/Clusters_C.R

@@ -1,149 +1,0 @@
-library(bigPint)
-library(data.table)
-library(EDASeq)
-library(plyr)
-library(gridExtra)
-
-outDir = getwd()
-data <- readRDS("../../../data/data.Rds")
-data <- data[,c(1:6, 13:18, 7:12, 19:24)]
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-
-metrics <- readRDS("../dataMetrics.Rds")[["C_R"]]
-sigMets = metrics[which(metrics$padj<0.05),]
-sigC <- sigMets[which(sigMets$log2FoldChange>0),]
-
-
-# File output information
-plotName = "C"
-logSoy = data
-logSoy[,-1] <- log(data[,-1]+1)
-
-
-
-
-
-
-
-RowSD = function(x) {
-  sqrt(rowSums((x - rowMeans(x))^2)/(dim(x)[2] - 1))
-}
-
-data_Rownames <- data$ID
-data = data[,-1]
-rownames(data) <- data_Rownames
-data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
-data = as.data.frame(data)
-# Add mean and standard deviation for each row/gene
-data = mutate(data, mean = (NC.1+NC.2+NC.3+NC.4+NC.5+NC.6+NR.1+NR.2+NR.3+NR.4+NR.5+NR.6+VC.1+VC.2+VC.3+VC.4+VC.5+VC.6+VR.1+VR.2+VR.3+VR.4+VR.5+VR.6)/6, stdev = RowSD(cbind(NC.1,NC.2,NC.3,NC.4,NC.5,NC.6,NR.1,NR.2,NR.3,NR.4,NR.5,NR.6,VC.1,VC.2,VC.3,VC.4,VC.5,VC.6,VR.1,VR.2,VR.3,VR.4,VR.5,VR.6)))
-rownames(data)=data_Rownames
-data$ID <- data_Rownames
-
-dataqps <- t(apply(as.matrix(data[,1:24]), 1, scale))
-dataqps <- as.data.frame(dataqps)
-colnames(dataqps) <- colnames(data[,1:24])
-dataqps$ID <- rownames(dataqps)
-
-# Comine the filtered and remaining data
-fulls <- dataqps
-boxDat <- melt(fulls, id.vars="ID")
-colnames(boxDat) <- c("ID", "Sample", "Count")
-
[…]

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method2/Clusters-St/Clusters_CR.R

@@ -1,111 +1,0 @@
-library(rtracklayer)
-library(Rsamtools)
-library(grid)
-library(GenomicAlignments)
-library(ggplot2)
-library(GGally)
-library(edgeR)
-library(stringr)
-library(EDASeq)
-library(dplyr)
-library(matrixStats)
-library(gridExtra)
-library(reshape2)
-library(scales)
-library(bigPint)
-library(data.table)
-library(dplyr)
-
-outDir = getwd()
-data <- readRDS("../../../data/data.Rds")
-data <- data[,c(1:6, 13:18, 7:12, 19:24)]
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-
-metrics <- readRDS("../dataMetrics.Rds")[["C_R"]]
-sigMets = metrics[which(metrics$padj<0.05),]
-sigC <- sigMets[which(sigMets$log2FoldChange<0),]
-sigR <- sigMets[which(sigMets$log2FoldChange>0),]
-
-
-# File output information
-plotName = "C_R"
-logSoy = data
-logSoy[,-1] <- log(data[,-1]+1)
-
-
-
-RowSD = function(x) {
-  sqrt(rowSums((x - rowMeans(x))^2)/(dim(x)[2] - 1))
-}
-
-data_Rownames <- data$ID
-data = data[,-1]
-rownames(data) <- data_Rownames
-data <- betweenLaneNormalization(as.matrix(data), which="full", round=FALSE)
-data = as.data.frame(data)
-# Add mean and standard deviation for each row/gene
-data = mutate(data, mean = (NC.1+NC.2+NC.3+NC.4+NC.5+NC.6+NR.1+NR.2+NR.3+NR.4+NR.5+NR.6+VC.1+VC.2+VC.3+VC.4+VC.5+VC.6+VR.1+VR.2+VR.3+VR.4+VR.5+VR.6)/6, stdev = RowSD(cbind(NC.1,NC.2,NC.3,NC.4,NC.5,NC.6,NR.1,NR.2,NR.3,NR.4,NR.5,NR.6,VC.1,VC.2,VC.3,VC.4,VC.5,VC.6,VR.1,VR.2,VR.3,VR.4,VR.5,VR.6)))
-rownames(data)=data_Rownames
-data$ID <- data_Rownames
-
-dataqps <- t(apply(as.matrix(data[,1:24]), 1, scale))
[…]

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method2/Clusters/Clusters.R

@@ -1,22 +1,0 @@
-library(bigPint)
-library(data.table)
-
-outDir = getwd()
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-data <- readRDS("../dataRLD.Rds")
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-
-colnames(data) <- c("ID","N.1","N.2","N.3","N.4","N.5","N.6","N.7","N.8","N.9","N.10","N.11","N.12","V.1","V.2","V.3","V.4","V.5","V.6","V.7","V.8","V.9","V.10","V.11","V.12")
-
-for (nC in 2:6){
-  plotClusters(data=data, dataMetrics=dataMetrics, nC, threshVar="padj", threshVal=0.05, outDir=outDir)  
-}
-
-colnames(data) <- c("ID","C.1","C.2","C.3","C.4","C.5","C.6","R.1","R.2","R.3","R.4","R.5","R.6","C.7","C.8","C.9","C.10","C.11","C.12","R.7","R.8","R.9","R.10","R.11","R.12")
-data <- data[,c(1:7,14:19,8:13,20:25)]
-for (nC in 2:15){
-  plotClusters(data=data, dataMetrics=dataMetrics, nC, threshVar="padj", threshVal=0.05, outDir=outDir)  
-}

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method2/PCP/PCP.R

@@ -1,16 +1,0 @@
-library(bigPint)
-
-outDir = getwd()
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-data <- readRDS("../dataRLD.Rds")
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-
-colnames(data) <- c("ID","N.1","N.2","N.3","N.4","N.5","N.6","N.7","N.8","N.9","N.10","N.11","N.12","V.1","V.2","V.3","V.4","V.5","V.6","V.7","V.8","V.9","V.10","V.11","V.12")
-plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "padj", threshVal = 0.05, lineSize=0.5, option = "parallelCoord")
-
-colnames(data) <- c("ID","C.1","C.2","C.3","C.4","C.5","C.6","R.1","R.2","R.3","R.4","R.5","R.6","C.7","C.8","C.9","C.10","C.11","C.12","R.7","R.8","R.9","R.10","R.11","R.12")
-data <- data[,c(1:7,14:19,8:13,20:25)]
-plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "padj", threshVal = 0.05, lineSize=0.1, option = "parallelCoord")

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method2/goAnalysis.R

@@ -1,1980 +1,0 @@
-metrics <- readRDS("dataMetrics.Rds")
-
-library(topGO)
-library(ALL)
-data(ALL)
-data(geneList)
-
-virusID = metrics[["N_V"]][which(metrics[["N_V"]]$padj <0.05),]$ID
-virusFDR =  metrics[["N_V"]][which(metrics[["N_V"]]$padj <0.05),]$padj
-virusList = metrics[["N_V"]][which(metrics[["N_V"]]$padj <0.05),]$padj
-names(virusList) = virusID
-
-sum(topDiffGenes(virusList))
-
-sampleGOdata <- new("topGOdata", description = "Simple session", ontology = "BP", allGenes = virusList,
-                    geneSel = topDiffGenes, nodeSize = 10, annot = annFUN.db, affyLib = affyLib)
-
-
-dietID = metrics[["C_R"]][which(metrics[["C_R"]]$padj <0.05),]$ID
-dietFDR =  metrics[["C_R"]][which(metrics[["C_R"]]$padj <0.05),]$padj
-dietList = metrics[["C_R"]][which(metrics[["C_R"]]$padj <0.05),]$padj
-names(dietList) = dietID
-
-
-# Virus
-GB40006
-GB40280
-GB41545
-GB41739
-GB42046
-GB42302
-GB42313
-GB42507
-GB42607
-GB42908
-GB43779
-GB43783
-GB43784
-GB45455
-GB45462
-GB45704
-GB47381
-GB47407
-GB47486
-GB48755
-GB49034
-GB50151
-GB50178
-GB50401
-GB50550
-GB50940
-GB51305
-GB52281
-GB52441
-GB52449
-GB53352
-GB53833
-GB55188
-GB55218
-
-#Diet
-GB40007
-GB40008
-GB40010
-GB40018
-GB40057
-GB40061
-GB40074
-GB40075
-GB40077
-GB40095
-GB40102
-GB40108
-GB40109
-GB40112
-GB40114
-GB40129
-GB40134
-GB40136
-GB40166
-GB40168
-GB40207
-GB40221
-GB40234
-GB40263
-GB40265
-GB40270
-GB40278
-GB40280
-GB40306
-GB40307
-GB40318
-GB40338
-GB40341
-GB40364
[…]

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method3/Clusters/Clusters.R

@@ -1,16 +1,0 @@
-library(bigPint)
-
-outDir = getwd()
-dataMetrics <- readRDS("../dataFolder/dataMetrics.Rds")
-
-for (i in 3:length(names(dataMetrics))){
-  currPair = names(dataMetrics)[i]
-  data <- readRDS(paste0("../dataFolder/dataRLD_", currPair,".Rds"))
-  setDT(data, keep.rownames = TRUE)[]
-  colnames(data)[1] = "ID"
-  data <- as.data.frame(data)
-  for (nC in 2:15){  
-    plotClusters(data=data, dataMetrics=dataMetrics, nC, threshVar="padj", threshVal=0.05, outDir=outDir)  
-  }
-}
-  

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method3/HexSM/HexSM.R

@@ -1,13 +1,0 @@
-library(bigPint)
-
-outDir = getwd()
-dataMetrics <- readRDS("../dataFolder/dataMetrics.Rds")
-
-for (i in 1:length(names(dataMetrics))){
-  currPair = names(dataMetrics)[i]
-  data <- readRDS(paste0("../dataFolder/dataRLD_", currPair,".Rds"))
-  setDT(data, keep.rownames = TRUE)[]
-  colnames(data)[1] = "ID"
-  data <- as.data.frame(data)
-  plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, pointSize = 0.5, xbins = 10, threshVar = "padj", threshVal = 0.05, option = "scatterHexagon")
-}

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method3/PCP/PCP.R

@@ -1,13 +1,0 @@
-library(bigPint)
-
-outDir = getwd()
-dataMetrics <- readRDS("../dataFolder/dataMetrics.Rds")
-
-for (i in 1:length(names(dataMetrics))){
-  currPair = names(dataMetrics)[i]
-  data <- readRDS(paste0("../dataFolder/dataRLD_", currPair,".Rds"))
-  setDT(data, keep.rownames = TRUE)[]
-  colnames(data)[1] = "ID"
-  data <- as.data.frame(data)
-  plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "padj", threshVal = 0.05, option = "parallelCoord")
-}

### 2018-04-13 NC_NR_VC_VR/DESeq2/Method3/PISM/PISM.R

@@ -1,14 +1,0 @@
-# Points that are DEG but not outside PI are plotted blue, points that are DEG and outside PI are plotted red, and points that are not DEG but are outside PI are plotted grey.
-library(bigPint)
-
-outDir = getwd()
-dataMetrics <- readRDS("../dataFolder/dataMetrics.Rds")
-
-for (i in 1:length(names(dataMetrics))){
-  currPair = names(dataMetrics)[i]
-  data <- readRDS(paste0("../dataFolder/dataRLD_", currPair,".Rds"))
-  setDT(data, keep.rownames = TRUE)[]
-  colnames(data)[1] = "ID"
-  data <- as.data.frame(data)
-  plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "padj", threshVal = 0.05, option = "scatterPrediction")
-}

### 2018-04-13 NC_NR_VC_VR/EdgeR/edgeR-Virus/Clusters/Clusters.R

@@ -1,16 +1,0 @@
-library(bigPint)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data)
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-for (nC in 2:6){
-  plotClusters(data=data, dataMetrics=dataMetrics, nC=nC, threshVar="FDR", threshVal=0.05, outDir=outDir)  
-}

### 2018-04-13 NC_NR_VC_VR/EdgeR/edgeR-Virus/PCP/PCP.R

@@ -1,15 +1,0 @@
-library(bigPint)
-library(data.table)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data)
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "FDR", threshVal = 0.05, option = "parallelCoord")

### 2018-04-13 NC_NR_VC_VR/EdgeR/edgeR-btwnLane-Diet/Clusters/Clusters.R

@@ -1,17 +1,0 @@
-library(bigPint)
-library(data.table)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data)
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-for (nC in 2:15){
-  plotClusters(data=data, dataMetrics=dataMetrics, nC=nC, threshVar="FDR", threshVal=0.05, outDir=outDir)  
-}

### 2018-04-13 NC_NR_VC_VR/EdgeR/edgeR-btwnLane-Diet/PCP/PCP.R

@@ -1,15 +1,0 @@
-library(bigPint)
-library(data.table)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data)
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "FDR", threshVal = 0.05, option = "parallelCoord", lineSize=0.5)

### 2018-04-13 NC_NR_VC_VR/EdgeR/edgeR-btwnLane-Virus/Clusters/Clusters.R

@@ -1,17 +1,0 @@
-library(bigPint)
-library(data.table)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data)
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-for (nC in 2:6){
-  plotClusters(data=data, dataMetrics=dataMetrics, nC=nC, threshVar="FDR", threshVal=0.05, outDir=outDir)  
-}

### 2018-04-13 NC_NR_VC_VR/EdgeR/edgeR-btwnLane-Virus/PCP/PCP.R

@@ -1,15 +1,0 @@
-library(bigPint)
-library(data.table)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data)
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "FDR", threshVal = 0.05, option = "parallelCoord", lineSize=1)

### 2018-04-13 NC_NR_VC_VR/LimmaVoom/LimmaVoom-Diet/glimma-plots/js/MDS-Plot.js

@@ -1,11 +1,0 @@
-glimma.layout.setupGrid(d3.select(".container"), "md", [1,2]);
-glimma.storage.chartData.push([{"dim1":-0.5785,"dim2":-0.5729,"dim3":0.3079,"dim4":0.2408,"dim5":0.323,"dim6":0.565,"dim7":0.4807,"dim8":-0.0078,"label":"C_L12_C.1","group":"C","lane":"L12","_row":"C.1"},{"dim1":-0.4699,"dim2":-0.6111,"dim3":1.0111,"dim4":0.0118,"dim5":0.3594,"dim6":0.5259,"dim7":0.1534,"dim8":-0.1405,"label":"C_L12_C.2","group":"C","lane":"L12","_row":"C.2"},{"dim1":1.3685,"dim2":0.704,"dim3":-0.0679,"dim4":-0.8467,"dim5":0.1285,"dim6":-0.2571,"dim7":-0.6146,"dim8":0.5193,"label":"C_L12_C.3","group":"C","lane":"L12","_row":"C.3"},{"dim1":1.0066,"dim2":0.1337,"dim3":0.8773,"dim4":0.5844,"dim5":-0.8252,"dim6":0.0035,"dim7":-0.1251,"dim8":-0.3423,"label":"C_L12_C.4","group":"C","lane":"L12","_row":"C.4"},{"dim1":-0.3029,"dim2":-1.3418,"dim3":0.9045,"dim4":-1.1013,"dim5":-0.8739,"dim6":-0.8797,"dim7":0.5775,"dim8":0.0723,"label":"C_L34_C.5","group":"C","lane":"L34","_row":"C.5"},{"dim1":1.8127,"dim2":0.0397,"dim3":-0.4713,"dim4":-0.0989,"dim5":0.1187,"dim6":-0.8769,"dim7":-0.0316,"dim8":-0.2095,"label":"C_L34_C.6","group":"C","lane":"L34","_row":"C.6"},{"dim1":-1.7596,"dim2":0.7449,"dim3":0.2599,"dim4":-0.9654,"dim5":-0.6353,"dim6":0.4407,"dim7":-0.6375,"dim8":-0.5142,"label":"C_L12_C.7","group":"C","lane":"L12","_row":"C.7"},{"dim1":-0.0092,"dim2":-0.359,"dim3":-0.2708,"dim4":0.1841,"dim5":0.3851,"dim6":-0.5818,"dim7":0.4227,"dim8
[…]

### 2018-04-13 NC_NR_VC_VR/LimmaVoom/LimmaVoom-Virus/Clusters/Clusters.R

@@ -1,17 +1,0 @@
-library(bigPint)
-library(data.table)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data[[1]])
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-for (nC in 2:15){
-  plotClusters(data=data, dataMetrics=dataMetrics, nC=nC, threshVar="adj.P.Val", threshVal=0.05, outDir=outDir)  
-}

### 2018-04-13 NC_NR_VC_VR/LimmaVoom/LimmaVoom-Virus/PCP/PCP.R

@@ -1,15 +1,0 @@
-library(bigPint)
-library(data.table)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data[[1]])
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "adj.P.Val", threshVal = 0.05, option = "parallelCoord")

### 2018-04-13 NC_NR_VC_VR/LimmaVoom/LimmaVoom-Virus/PISM/PISM.R

@@ -1,15 +1,0 @@
-# Points that are DEG but not outside PI are plotted blue, points that are DEG and outside PI are plotted red, and points that are not DEG but are outside PI are plotted grey.
-library(bigPint)
-
-outDir = getwd()
-data <- readRDS("../data.Rds")
-
-data <- as.data.frame(data[[1]])
-setDT(data, keep.rownames = TRUE)[]
-colnames(data)[1] = "ID"
-data <- as.data.frame(data)
-data[,-1] <- log(data[,-1]+1)
-
-dataMetrics <- readRDS("../dataMetrics.Rds")
-
-plotDEG(data = data, dataMetrics = dataMetrics, outDir = outDir, threshVar = "adj.P.Val", threshVal = 0.05, option = "scatterPrediction")

### 2018-04-13 NC_NR_VC_VR/LimmaVoom/LimmaVoom-Virus/glimma-plots/js/MDS-Plot.js

@@ -1,11 +1,0 @@
-glimma.layout.setupGrid(d3.select(".container"), "md", [1,2]);
-glimma.storage.chartData.push([{"dim1":-0.5785,"dim2":-0.5729,"dim3":0.3079,"dim4":-0.2408,"dim5":0.323,"dim6":-0.565,"dim7":0.4807,"dim8":-0.0078,"label":"N_L12_N.1","group":"N","lane":"L12","_row":"N.1"},{"dim1":-0.4699,"dim2":-0.6111,"dim3":1.0111,"dim4":-0.0118,"dim5":0.3594,"dim6":-0.5259,"dim7":0.1534,"dim8":-0.1405,"label":"N_L12_N.2","group":"N","lane":"L12","_row":"N.2"},{"dim1":1.3685,"dim2":0.704,"dim3":-0.0679,"dim4":0.8467,"dim5":0.1285,"dim6":0.2571,"dim7":-0.6146,"dim8":0.5193,"label":"N_L12_N.3","group":"N","lane":"L12","_row":"N.3"},{"dim1":1.0066,"dim2":0.1337,"dim3":0.8773,"dim4":-0.5844,"dim5":-0.8252,"dim6":-0.0035,"dim7":-0.1251,"dim8":-0.3423,"label":"N_L12_N.4","group":"N","lane":"L12","_row":"N.4"},{"dim1":-0.3029,"dim2":-1.3418,"dim3":0.9045,"dim4":1.1013,"dim5":-0.8739,"dim6":0.8797,"dim7":0.5775,"dim8":0.0723,"label":"N_L34_N.5","group":"N","lane":"L34","_row":"N.5"},{"dim1":1.8127,"dim2":0.0397,"dim3":-0.4713,"dim4":0.0989,"dim5":0.1187,"dim6":0.8769,"dim7":-0.0316,"dim8":-0.2095,"label":"N_L34_N.6","group":"N","lane":"L34","_row":"N.6"},{"dim1":0.0686,"dim2":-0.1947,"dim3":-0.6728,"dim4":0.0265,"dim5":-0.257,"dim6":-0.6119,"dim7":1.3398,"dim8":0.1803,"label":"N_L12_N.7","group":"N","lane":"L12","_row":"N.7"},{"dim1":-0.951,"dim2":-1.0481,"dim3":-1.985,"dim4":-0.1172,"dim5":-0.3374,"dim6":-0.1603,"dim7":-0.7133,"dim
[…]

### 2018-04-13 NC_NR_VC_VR/LimmaVoom/glimma-plots/js/MDS-Plot.js

@@ -1,8 +1,8 @@
 glimma.layout.setupGrid(d3.select(".container"), "md", [1,2]);
-glimma.storage.chartData.push([{"dim1":-0.5785,"dim2":-0.5729,"dim3":0.3079,"dim4":-0.2408,"dim5":0.323,"dim6":-0.565,"dim7":0.4807,"dim8":-0.0078,"label":"NC_L12_NC.1","group":"NC","lane":"L12","_row":"NC.1"},{"dim1":-0.4699,"dim2":-0.6111,"dim3":1.0111,"dim4":-0.0118,"dim5":0.3594,"dim6":-0.5259,"dim7":0.1534,"dim8":-0.1405,"label":"NC_L12_NC.2","group":"NC","lane":"L12","_row":"NC.2"},{"dim1":1.3685,"dim2":0.704,"dim3":-0.0679,"dim4":0.8467,"dim5":0.1285,"dim6":0.2571,"dim7":-0.6146,"dim8":0.5193,"label":"NC_L12_NC.3","group":"NC","lane":"L12","_row":"NC.3"},{"dim1":1.0066,"dim2":0.1337,"dim3":0.8773,"dim4":-0.5844,"dim5":-0.8252,"dim6":-0.0035,"dim7":-0.1251,"dim8":-0.3423,"label":"NC_L12_NC.4","group":"NC","lane":"L12","_row":"NC.4"},{"dim1":-0.3029,"dim2":-1.3418,"dim3":0.9045,"dim4":1.1013,"dim5":-0.8739,"dim6":0.8797,"dim7":0.5775,"dim8":0.0723,"label":"NC_L34_NC.5","group":"NC","lane":"L34","_row":"NC.5"},{"dim1":1.8127,"dim2":0.0397,"dim3":-0.4713,"dim4":0.0989,"dim5":0.1187,"dim6":0.8769,"dim7":-0.0316,"dim8":-0.2095,"label":"NC_L34_NC.6","group":"NC","lane":"L34","_row":"NC.6"},{"dim1":0.0686,"dim2":-0.1947,"dim3":-0.6728,"dim4":0.0265,"dim5":-0.257,"dim6":-0.6119,"dim7":1.3398,"dim8":0.1803,"label":"NR_L12_NR.1","group":"NR","lane":"L12","_row":"NR.1"},{"dim1":-0.951,"dim2":-1.0481,"dim3":-1.985,"dim4":-0.1172,"dim5":
[…]

### 2018-04-13 C_R/DESeq2/makeDataMetrics.Rmd
Count RDS files diet and virus

@@
 
+    RDC <- metrics[which(metrics[which(metrics$padj <0.05),]$log2FoldChange>0),]$ID
+    RDR <- metrics[which(metrics[which(metrics$padj <0.05),]$log2FoldChange<0),]$ID
+    RD_DIET <- metrics[which(metrics$padj <0.05),]$ID
+    saveRDS(RDC, file="RDC.Rds")
+    saveRDS(RDR, file="RDR.Rds")
+    saveRDS(RD_DIET, file="RD_DIET.Rds")
+
     allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength, Treatment1Larger = t1Length, Treatment2Larger = t2Length))

### 2018-04-13 N_V/DESeq2/makeDataMetrics.Rmd
Count RDS files diet and virus

--- 2018-04-12 C_R/DESeq2/makeDataMetrics.Rmd
@@
 
+    RDV <- metrics[which(metrics[which(metrics$padj <0.05),]$log2FoldChange>0),]$ID
+    RDN <- metrics[which(metrics[which(metrics$padj <0.05),]$log2FoldChange<0),]$ID
+    RD_VIRUS <- metrics[which(metrics$padj <0.05),]$ID
+    saveRDS(RDV, file="RDV.Rds")
+    saveRDS(RDN, file="RDN.Rds")
+    saveRDS(RD_VIRUS, file="RD_VIRUS.Rds")
+
     allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength, Treatment1Larger = t1Length, Treatment2Larger = t2Length))

### 2018-04-13 VirusHoneyBee/DESeq2/makeDataMetrics.Rmd
Finsh 27 sets of DEG counts

--- 2018-04-12 C_R/DESeq2/makeDataMetrics.Rmd
@@
 
+    GDC1 <- metrics[which(metrics[which(metrics$padj <0.05),]$log2FoldChange>0),]$ID
+    GDV1 <- metrics[which(metrics[which(metrics$padj <0.05),]$log2FoldChange<0),]$ID
+    GD_VIRUS1 <- metrics[which(metrics$padj <0.05),]$ID
+
+    GDV <- unname(sapply(GDV1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+    GDC <- unname(sapply(GDC1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+    GD_VIRUS <- unname(sapply(GD_VIRUS1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+
+    saveRDS(GDV, file="GDV.Rds")
+    saveRDS(GDC, file="GDC.Rds")
+    saveRDS(GD_VIRUS, file="GD_VIRUS.Rds")
+
     allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength, Treatment1Larger = t1Length, Treatment2Larger = t2Length))

### 2018-04-13 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
Finsh 27 sets of DEG counts

@@
 
+    GEC1 <- rownames(lrt5[which(lrt5$logFC>0),])
+    GEV1 <- rownames(lrt5[which(lrt5$logFC<0),])
+    GE_VIRUS1 <- rownames(lrt5)
+
+    GEC <- unname(sapply(GEC1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+    GEV <- unname(sapply(GEV1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+    GE_VIRUS <- unname(sapply(GE_VIRUS1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+
+    saveRDS(GEC, file="GEC.Rds")
+    saveRDS(GEV, file="GEV.Rds")
+    saveRDS(GE_VIRUS, file="GE_VIRUS.Rds")
+
     lrtLength <- nrow(lrt5)

### 2018-04-13 VirusHoneyBee/LimmaVoom/makeDataMetrics.Rmd
Finsh 27 sets of DEG counts

@@
 summary(decideTests(efit))
+
+countDF <- as.data.frame(decideTests(efit)@.Data)
+GLC1 <- rownames(countDF)[which(countDF==1)]
+GLV1 <- rownames(countDF)[which(countDF==-1)]
+GL_VIRUS1 <- c(GLC1, GLV1)
+
+GLC <- unname(sapply(GLC1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+GLV <- unname(sapply(GLV1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+GL_VIRUS <- unname(sapply(GL_VIRUS1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+
+saveRDS(GLC, file="GLC.Rds")
+saveRDS(GLV, file="GLV.Rds")
+saveRDS(GL_VIRUS, file="GL_VIRUS.Rds")
+
 ```

### 2018-04-13 C_R/DESeq2/makeDataMetrics.Rmd
Venn updates

--- 2018-04-12 C_R/DESeq2/makeDataMetrics.Rmd
@@
 
+    sigMetrics <- metrics[which(metrics$padj <0.05),]
+    RDC <- sigMetrics[which(sigMetrics$log2FoldChange>0),]$ID
+    RDR <- sigMetrics[which(sigMetrics$log2FoldChange<0),]$ID
+    RD_DIET_TOTAL <- metrics[which(metrics$padj <0.05),]$ID
+
+    saveRDS(RDC, file="RDC.Rds")
+    saveRDS(RDR, file="RDR.Rds")
+    saveRDS(RD_DIET_TOTAL, file="RD_DIET_TOTAL.Rds")
+
     allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength, Treatment1Larger = t1Length, Treatment2Larger = t2Length))

### 2018-04-13 C_R/LimmaVoom/makeDataMetrics.Rmd
Count RDS files diet and virus
Venn updates

@@
 summary(decideTests(efit))
+countDF <- as.data.frame(decideTests(efit)@.Data)
+RLC <- rownames(countDF)[which(countDF==1)]
+RLR <- rownames(countDF)[which(countDF==-1)]
+RL_DIET_TOTAL <- c(RLC, RLR)
+
+saveRDS(RLC, file="RLC.Rds")
+saveRDS(RLR, file="RLR.Rds")
+saveRDS(RL_DIET_TOTAL, file="RL_DIET_TOTAL.Rds")
 ```

### 2018-04-13 N_V/DESeq2/makeDataMetrics.Rmd
Venn updates

--- 2018-04-12 C_R/DESeq2/makeDataMetrics.Rmd
@@
 
+    sigMetrics <- metrics[which(metrics$padj <0.05),]
+    RDN <- sigMetrics[which(sigMetrics$log2FoldChange>0),]$ID
+    RDV <- sigMetrics[which(sigMetrics$log2FoldChange<0),]$ID
+    RD_VIRUS_TOTAL <- metrics[which(metrics$padj <0.05),]$ID
+
+    saveRDS(RDV, file="RDV.Rds")
+    saveRDS(RDN, file="RDN.Rds")
+    saveRDS(RD_VIRUS_TOTAL, file="RD_VIRUS_TOTAL.Rds")
+
     allPairs = rbind(allPairs, data.frame(Treatment1 = uTreat[i], Treatment2 = uTreat[j], NumberDEG = degLength, Treatment1Larger = t1Length, Treatment2Larger = t2Length))

### 2018-04-13 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
Venn updates

--- 2018-04-13 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
@@
 
+    GEC1 <- rownames(lrt5[which(lrt5$logFC>0),])
+    GEV1 <- rownames(lrt5[which(lrt5$logFC<0),])
+    GE_VIRUS1 <- rownames(lrt5)
+
+    GEC <- unname(sapply(GEC1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+    GEV <- unname(sapply(GEV1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+    GE_TOTAL <- unname(sapply(GE_TOTAL1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+
+    saveRDS(GEC, file="GEC.Rds")
+    saveRDS(GEV, file="GEV.Rds")
+    saveRDS(GE_TOTAL, file="GE_TOTAL.Rds")
+
     lrtLength <- nrow(lrt5)

### 2018-04-13 VirusHoneyBee/LimmaVoom/makeDataMetrics.Rmd
Venn updates

--- 2018-04-12 VirusHoneyBee/LimmaVoom/makeDataMetrics.Rmd
@@
 summary(decideTests(efit))
+
+countDF <- as.data.frame(decideTests(efit)@.Data)
+GLC1 <- rownames(countDF)[which(countDF==1)]
+GLV1 <- rownames(countDF)[which(countDF==-1)]
+GL_TOTAL1 <- c(GLC1, GLV1)
+
+GLC <- unname(sapply(GLC1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+GLV <- unname(sapply(GLV1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+GL_TOTAL <- unname(sapply(GL_TOTAL1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+
+saveRDS(GLC, file="GLC.Rds")
+saveRDS(GLV, file="GLV.Rds")
+saveRDS(GL_TOTAL, file="GL_TOTAL.Rds")
+
 ```

### 2018-04-13 Overlaps/Overlaps.R

@@ -2,173 +2,169 @@ library(VennDiagram)
 library(readr)
 library(readxl)
 
-#draw.triple.venn(65, 75, 85, 35, 15, 25, 5, c("First", "Second", "Third"))
+#######################################################
+# Read in DEG files
+RDC <- readRDS("../C_R/DESeq2/RDC.Rds")
+RDR <- readRDS("../C_R/DESeq2/RDR.Rds")
+RD_DIET_TOTAL <- readRDS("../C_R/DESeq2/RD_DIET_TOTAL.Rds")
+REC <- readRDS("../C_R/EdgeR/edgeR/REC.Rds")
+RER <- readRDS("../C_R/EdgeR/edgeR/RER.Rds")
+RE_DIET_TOTAL <- readRDS("../C_R/EdgeR/edgeR/RE_DIET_TOTAL.Rds")
+RLC <- readRDS("../C_R/LimmaVoom/RLC.Rds")
+RLR <- readRDS("../C_R/LimmaVoom/RLR.Rds")
+RL_DIET_TOTAL <- readRDS("../C_R/LimmaVoom/RL_DIET_TOTAL.Rds")
+RDV <- readRDS("../N_V/DESeq2/RDV.Rds")
+RDN <- readRDS("../N_V/DESeq2/RDN.Rds")
+RD_VIRUS_TOTAL <- readRDS("../N_V/DESeq2/RD_VIRUS_TOTAL.Rds")
+REV <- readRDS("../N_V/EdgeR/edgeR/REV.Rds")
+REN <- readRDS("../N_V/EdgeR/edgeR/REN.Rds")
+RE_VIRUS_TOTAL <- readRDS("../N_V/EdgeR/edgeR/RE_VIRUS_TOTAL.Rds")
+GDV <- readRDS("../VirusHoneyBee/DESeq2/GDV.Rds")
+GDC <- readRDS("../VirusHoneyBee/DESeq2/GDC.Rds")
+GD_TOTAL <- readRDS("../VirusHoneyBee/DESeq2/GD_TOTAL.Rds")
+GEV <- readRDS("../VirusHoneyBee/EdgeR/edgeR/GEV.Rds")
+GEC <- readRDS("../VirusHoneyBee/EdgeR/edgeR/GEC.Rds")
+GE_TOTAL <- readRDS("../VirusHoneyBee/EdgeR/edgeR/GE_TOTAL.Rds")
+GLV <- readRDS("../VirusHoneyBee/LimmaVoom/GLV.Rds")
[…]

### 2018-04-16 NC_NR_VC_VR/DESeq2/Method1/makeDataMetrics.Rmd
Updated up and down regulated DEG counts
Grep linux scripts for GOs

--- 2018-04-09 VirusHoneyBee/DESeq2/Method1/makeDataMetrics.Rmd
@@
 
-    t1Length <- length(which(metrics[which(metrics$padj <0.05),]$log2FoldChange>0))
-    t2Length <- length(which(metrics[which(metrics$padj <0.05),]$log2FoldChange<0))
+    sigMet <- metrics[which(metrics$padj <0.05),]
+    t1Length <- length(which(sigMet$log2FoldChange>0))
+    t2Length <- length(which(sigMet$log2FoldChange<0))
+
+    saveRDS(sigMet[which(sigMet$log2FoldChange>0),]$ID, file=paste0(uTreat[i], "_", uTreat[j], "_", uTreat[i], ".Rds"))
+    saveRDS(sigMet[which(sigMet$log2FoldChange<0),]$ID, file=paste0(uTreat[i], "_", uTreat[j], "_", uTreat[j], ".Rds"))
 

### 2018-04-16 Overlaps/GO/ReadingGO.R

@@ -1,2 +1,4 @@
-dat <- readRDS("gD.rds")
+dat <- readRDS("RDC.rds")
 cat(dat,sep="\n")
+
+#grep -o 'GO:[0-9]*' fileName

### 2018-04-16 Overlaps/Overlaps.R

@@ -327,9 +327,8 @@ invisible(dev.off())
 
 ############### Rutter vs. Galbriath DESeq (overlaps between up and down) ############## 
 
-# One overlap!
 length(intersect(RDV, GDC))
-length(intersect(RDN, GDV)) # One overlap!
+length(intersect(RDN, GDV)) # One overlap! ("GB51305")
 
 
 
@@ -360,9 +359,8 @@ invisible(dev.off())
 
 ############### Rutter vs. Galbriath DESeq (overlaps between up and down) ############## 
 
-# One overlap!
 length(intersect(REV, GDC))
-length(intersect(GEV, REN)) # One overlap!
+length(intersect(GEV, REN)) # One overlap! ("GB51305")
 
 
 

### 2018-04-19 Manuscript/chapter4.tex
Add results

@@
 The \pkg{eulerr} shiny application was used to construct Venn diagram overlap images (\citealt{euler}).
+In the main section of our paper and in subsequent analyses, we focus on the DEG results from DESeq2 (\citealt{deseq2}) as this pipeline was also used in the Galbraith study (\citealt{galbraith}).
 
@@
 The refined GO term list was then imported into REVIGO (\citealt{revigo}), which uses semantic similarity measures to cluster long lists of GO terms.
+@@@ Pathways analysis @@@
 
@@
 
+\subsection{Main effect results}
+
+We observed a substantially larger number of DEGs in our diet main effect (n = 1914) than in our virus main effect (n = 43) (Table \ref{tbl:mainEffectDEGs}A and B).
+In the diet factor, there were more Castanea DEGs (n = 1033) than Rockrose DEGs (n = 881).
+In the virus factor, there were more virus-exposed DEGs (n = 38) than control DEGs (n = 5).
+While these reported DEGs numbers are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Table \ref{tbl:mainEffectDEGs}A and B).
+Pathway
+
 \section{Discussion}
-
-% SYNTAX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\ref{sec:ggenealogy}
-% \code{ggparcoord}
-% \Figure \ref{fig:porcupine}
-
-% \begin{figure}[H]
-% \centering
-%     \begin{framed}
-%     \includegraphics[width=0.8\textwidth]{porcupine}
-%     \end{framed}
-%     \caption{Three}
-%     \label{fig:porcupine}
-% \end{figure}
-
-% \section{Literature review}
-% \label{sec:litReview}
-
-% \pkg{R} statistical programming language allows for tools to be distributed and modified at ease, encourages cross-platform collaboration, and provides a foundation for effective and aesthetic data visualization from the grammar of graphics.
-There are several useful \pkg{R} packages that offer tools for analyzing and visualizing genealogical datasets.
-Here, we introduce these packages, and emphasize the shortcomings for which our package \pkg{ggenealogy} brings to this collection of work.
-
-%\clearpage
-
-%\begin{enumerate}
-%\item The data can be viewed at \url{http://shiny.soybase.org/CNV/}.
-%\end{enumerate}
 
@@
 
+\section{Appendix}
+
+% \begin{table}
+%   \caption{A stunning table}
+%   \label{tbl:excel-table}
+%   \includegraphics[width=\linewidth]{excel-table}
+% \end{table}
+
+\begin{table}[H]
+  \includegraphics[width=0.65\textwidth]{Images/mainEffectDEGs}
+  \caption{Number of DEGs across three analysis pipelines for (A) the diet effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study.}
+  \label{tbl:mainEffectDEGs}
+\end{table}
+
 \bibliography{chapter4}

### 2018-04-19 N_V/DESeq2/ClusterStandard/Clustering_data_FDR_05.R

@@ -40,6 +40,8 @@ getPCP <- function(nC){
     x$cluster = "color"
     x$cluster2 = factor(x$cluster)
     xNames = rownames(x)
+    
+    saveRDS(xNames, file=paste0("Sig_", nC, "_", i, ".Rds"))
     xFDR = metrics[which(metrics$ID %in% xNames),]$padj
     scatMatMetrics = list()
     scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
@@ -90,7 +92,7 @@ pair1 <- strsplit(currPair, "_")[[1]][1]
 pair2 <- strsplit(currPair, "_")[[1]][2]
 
 metrics <- metricsAll[[currPair]]
-data <- as.data.frame(readRDS("../../../data/data.Rds"))
+data <- as.data.frame(readRDS("../../data/data.Rds"))
 data <- data[,which(sapply(colnames(data), function(x) unlist(strsplit(x,"[.]"))[1]) %in% c(pair1, pair2))]
 data<- cbind(ID = rownames(data), data)
 data$ID <- as.character(data$ID)
@@ -134,6 +136,8 @@ dendo = sigDatas
 rownames(dendo) = NULL
 d = dist(as.matrix(dendo))
 hc = hclust(d, method="ward.D")
+
+saveRDS(sigDatas$ID, file="Sig.Rds")
 
 plotName = currPair
 outDir = "Clustering_data_FDR_05"

### 2018-04-19 N_V/DESeq2/Dashboard/litreSig/app.R

@@ -1,0 +1,283 @@
+library(shinydashboard)
+library(shiny)
+library(plotly)
+library(hexbin)
+library(htmlwidgets)
+library(dplyr)
+library(tidyr)
+library(data.table)
+library(RColorBrewer)
+library(Hmisc)
+library(shinycssloaders)
+
+options(spinner.color.background="#F5F5F5")
+
+dat <- readRDS("data.Rds")
+dat$ID <- rownames(dat)
+dat <- dat[,c(25,1:24)]
+
+dat[,-1] <- log(dat[,-1]+1)
+datCol <- colnames(dat)[-which(colnames(dat) %in% "ID")]
+myPairs <- unique(sapply(datCol, function(x) unlist(strsplit(x,"[.]"))[1]))
+load("Sigmetrics.rda")
+myMetrics <- colnames(metrics[[1]])[-which(colnames(metrics[[1]]) %in% "ID")]
+values <- reactiveValues(x=0, selPair=NULL, selMetric=NULL, selOrder=NULL)
+
+sidebar <- dashboardSidebar(
+  width = 180,
+  hr(),
+  sidebarMenu(id="tabs",
+    shinydashboard::menuItem("Application", tabName="hexPlot"),
+    shinydashboard::menuItem("About", tabName = "about", selected=TRUE)
+  )
+)
+
+body <- dashboardBody(
+  tabItems(
+    tabItem(tabName = "hexPlot",
+      fluidRow(
+        column(width = 4, 
+         box(width = NULL, status = "primary", title = "Plot metrics", solidHeader = TRUE,
+           selectizeInput("selPair", "Treatment pairs:", choices = myPairs, multiple = TRUE, options = list(maxItems = 2)),
+           selectInput("selMetric", "Metric:", choices = myMetrics, selected = "FDR"),
+           selectInput("selOrder", "Metric order:", choices = c("Increasing", "Decreasing")),
+           numericInput("binSize", "Hexagon size:", value = 10, min = 1),
+           numericInput("pointSize", "Point size:", value = 8, min = 1),
+           actionButton("goButton", "Plot gene!"))),
+        column(width = 8,
+           box(width = NULL, withSpinner(plotlyOutput("hexPlot")), collapsible = FALSE, background = "black", title = "Litre plot", status = "primary", solidHeader = TRUE))),
+    
+      fluidRow(
+        column(width = 12,
[…]

### 2018-04-19 N_V/DESeq2/Dashboard/litreSigStand/app.R

@@ -1,0 +1,305 @@
+library(shinydashboard)
+library(shiny)
+library(plotly)
+library(hexbin)
+library(htmlwidgets)
+library(dplyr)
+library(tidyr)
+library(data.table)
+library(RColorBrewer)
+library(Hmisc)
+library(shinycssloaders)
+
+options(spinner.color.background="#F5F5F5")
+
+dat <- readRDS("data.Rds")
+dat$ID <- rownames(dat)
+dat <- dat[,c(25,1:24)]
+
+
+
+# Standardize in this application
+RowSD = function(x) {
+  sqrt(rowSums((x - rowMeans(x))^2)/(dim(x)[2] - 1))
+}
+dat_Rownames <- dat$ID
+dat = dat[,-1]
+rownames(dat) <- dat_Rownames
+dat <- betweenLaneNormalization(as.matrix(dat), which="full", round=FALSE)
+dat = as.data.frame(dat)
+dat = mutate(dat, mean = (N.1+N.2+N.3+N.4+N.5+N.6+N.7+N.8+N.9+N.10+N.11+N.12+V.1+V.2+V.3+V.4+V.5+V.6+V.7+V.8+V.9+V.10+V.11+V.12)/24, stdev = RowSD(cbind(N.1,N.2,N.3,N.4,N.5,N.6,N.7,N.8,N.9,N.10,N.11,N.12,V.1,V.2,V.3,V.4,V.5,V.6,V.7,V.8,V.9,V.10,V.11,V.12)))
+rownames(dat)=dat_Rownames
+dat$ID <- dat_Rownames
+datqps <- t(apply(as.matrix(dat[,1:24]), 1, scale))
+datqps <- as.data.frame(datqps)
+colnames(datqps) <- colnames(dat[,1:24])
+datqps$ID <- rownames(datqps)
+nID <- which(is.nan(datqps$N.1))
+datqps[nID,1:24] <- 0
+dat <- datqps[,c(25,1:24)]
+
+
+datCol <- colnames(dat)[-which(colnames(dat) %in% "ID")]
+myPairs <- unique(sapply(datCol, function(x) unlist(strsplit(x,"[.]"))[1]))
+load("Sigmetrics.rda")
+myMetrics <- colnames(metrics[[1]])[-which(colnames(metrics[[1]]) %in% "ID")]
+values <- reactiveValues(x=0, selPair=NULL, selMetric=NULL, selOrder=NULL)
+
+sidebar <- dashboardSidebar(
[…]

### 2018-04-19 N_V/DESeq2/Dashboard/makeClusterData/makeDataCluster1.R

@@ -1,0 +1,16 @@
+library(bigPint)
+
+ID <- readRDS("../../ClusterStandard/Sig.Rds")
+#load("../../Bioinformatics/Pictures/FilterNotSig/soybean_ir_noFilt_metrics.rda")
+
+metricsAll <- readRDS("../../dataMetrics.Rds")
+
+allMetrics = metricsAll[["N_V"]]
+
+metricsCluster <- allMetrics[which(allMetrics$ID %in% ID),]
+metricsCluster$ID <- as.character(metricsCluster$ID)
+
+metrics <- list()
+metrics[["N_V"]] <- metricsCluster
+
+save(metrics, file = "Sigmetrics.rda")

### 2018-04-19 VirusHoneyBee/DESeq2/ClusterStandard/Clustering_data_FDR_05.R

@@ -52,6 +52,7 @@ getPCP <- function(nC){
     plotDEG(data = plotDatas, dataMetrics = scatMatMetrics, option="scatterPoints", threshVar = "padj", threshVal = 0.05, degPointColor = colList[i+1], fileName=fileName)
     
     x$ID = xNames
+    saveRDS(xNames, file=paste0("Sig_", nC,"_", i,".Rds"))
     
     pcpDat <- melt(x[,c(1:(nColumns+1))], id.vars="ID")
     colnames(pcpDat) <- c("ID", "Sample", "Count")
@@ -130,6 +131,9 @@ boxDat <- melt(datas, id.vars="ID")
 colnames(boxDat) <- c("ID", "Sample", "Count")
 
 sigDatas = datas[which(metricsAll[["C_T"]]$padj<0.05),]
+
+saveRDS(sigDatas$ID, file="Sig.Rds")
+
 dendo = sigDatas
 rownames(dendo) = NULL
 d = dist(as.matrix(dendo))

### 2018-04-19 VirusHoneyBee/DESeq2/Dashboard/litreCluster1/app.R

@@ -1,0 +1,283 @@
+library(shinydashboard)
+library(shiny)
+library(plotly)
+library(hexbin)
+library(htmlwidgets)
+library(dplyr)
+library(tidyr)
+library(data.table)
+library(RColorBrewer)
+library(Hmisc)
+library(shinycssloaders)
+
+options(spinner.color.background="#F5F5F5")
+
+dat <- readRDS("data.Rds")
+dat$ID <- rownames(dat)
+dat <- dat[,c(7,1:6)]
+
+dat[,-1] <- log(dat[,-1]+1)
+datCol <- colnames(dat)[-which(colnames(dat) %in% "ID")]
+myPairs <- unique(sapply(datCol, function(x) unlist(strsplit(x,"[.]"))[1]))
+load("SigCluster1.rda")
+myMetrics <- colnames(metrics[[1]])[-which(colnames(metrics[[1]]) %in% "ID")]
+values <- reactiveValues(x=0, selPair=NULL, selMetric=NULL, selOrder=NULL)
+
+sidebar <- dashboardSidebar(
+  width = 180,
+  hr(),
+  sidebarMenu(id="tabs",
+    shinydashboard::menuItem("Application", tabName="hexPlot"),
+    shinydashboard::menuItem("About", tabName = "about", selected=TRUE)
+  )
+)
+
+body <- dashboardBody(
+  tabItems(
+    tabItem(tabName = "hexPlot",
+      fluidRow(
+        column(width = 4, 
+         box(width = NULL, status = "primary", title = "Plot metrics", solidHeader = TRUE,
+           selectizeInput("selPair", "Treatment pairs:", choices = myPairs, multiple = TRUE, options = list(maxItems = 2)),
+           selectInput("selMetric", "Metric:", choices = myMetrics, selected = "FDR"),
+           selectInput("selOrder", "Metric order:", choices = c("Increasing", "Decreasing")),
+           numericInput("binSize", "Hexagon size:", value = 10, min = 1),
+           numericInput("pointSize", "Point size:", value = 8, min = 1),
+           actionButton("goButton", "Plot gene!"))),
+        column(width = 8,
+           box(width = NULL, withSpinner(plotlyOutput("hexPlot")), collapsible = FALSE, background = "black", title = "Litre plot", status = "primary", solidHeader = TRUE))),
+    
+      fluidRow(
+        column(width = 12,
[…]

### 2018-04-19 VirusHoneyBee/DESeq2/Dashboard/litreCluster1Stand/app.R

@@ -1,0 +1,304 @@
+library(shinydashboard)
+library(shiny)
+library(plotly)
+library(hexbin)
+library(htmlwidgets)
+library(dplyr)
+library(tidyr)
+library(data.table)
+library(RColorBrewer)
+library(Hmisc)
+library(shinycssloaders)
+
+options(spinner.color.background="#F5F5F5")
+
+dat <- readRDS("data.Rds")
+dat$ID <- rownames(dat)
+dat <- dat[,c(7,1:6)]
+
+# Standardize in this application
+RowSD = function(x) {
+  sqrt(rowSums((x - rowMeans(x))^2)/(dim(x)[2] - 1))
+}
+dat_Rownames <- dat$ID
+dat = dat[,-1]
+rownames(dat) <- dat_Rownames
+dat <- betweenLaneNormalization(as.matrix(dat), which="full", round=FALSE)
+dat = as.data.frame(dat)
+dat = mutate(dat, mean = (C.1+C.2+C.3+T.1+T.2+T.3)/6, stdev = RowSD(cbind(C.1,C.2,C.3,T.1,T.2,T.3)))
+rownames(dat)=dat_Rownames
+dat$ID <- dat_Rownames
+datqps <- t(apply(as.matrix(dat[,1:6]), 1, scale))
+datqps <- as.data.frame(datqps)
+colnames(datqps) <- colnames(dat[,1:6])
+datqps$ID <- rownames(datqps)
+nID <- which(is.nan(datqps$V.1))
+datqps[nID,1:6] <- 0
+dat <- datqps[,c(7,1:6)]
+
+
+
+datCol <- colnames(dat)[-which(colnames(dat) %in% "ID")]
+myPairs <- unique(sapply(datCol, function(x) unlist(strsplit(x,"[.]"))[1]))
+load("SigCluster1.rda")
+myMetrics <- colnames(metrics[[1]])[-which(colnames(metrics[[1]]) %in% "ID")]
+values <- reactiveValues(x=0, selPair=NULL, selMetric=NULL, selOrder=NULL)
+
+sidebar <- dashboardSidebar(
+  width = 180,
+  hr(),
+  sidebarMenu(id="tabs",
+    shinydashboard::menuItem("Application", tabName="hexPlot"),
+    shinydashboard::menuItem("About", tabName = "about", selected=TRUE)
+  )
+)
+
+body <- dashboardBody(
+  tabItems(
+    tabItem(tabName = "hexPlot",
+      fluidRow(
+        column(width = 4, 
[…]

### 2018-04-19 VirusHoneyBee/DESeq2/Dashboard/litreSig/app.R

@@ -1,0 +1,283 @@
+library(shinydashboard)
+library(shiny)
+library(plotly)
+library(hexbin)
+library(htmlwidgets)
+library(dplyr)
+library(tidyr)
+library(data.table)
+library(RColorBrewer)
+library(Hmisc)
+library(shinycssloaders)
+
+options(spinner.color.background="#F5F5F5")
+
+dat <- readRDS("data.Rds")
+dat$ID <- rownames(dat)
+dat <- dat[,c(7,1:6)]
+
+dat[,-1] <- log(dat[,-1]+1)
+datCol <- colnames(dat)[-which(colnames(dat) %in% "ID")]
+myPairs <- unique(sapply(datCol, function(x) unlist(strsplit(x,"[.]"))[1]))
+load("Sigmetrics.rda")
+myMetrics <- colnames(metrics[[1]])[-which(colnames(metrics[[1]]) %in% "ID")]
+values <- reactiveValues(x=0, selPair=NULL, selMetric=NULL, selOrder=NULL)
+
+sidebar <- dashboardSidebar(
+  width = 180,
+  hr(),
+  sidebarMenu(id="tabs",
+    shinydashboard::menuItem("Application", tabName="hexPlot"),
+    shinydashboard::menuItem("About", tabName = "about", selected=TRUE)
+  )
+)
+
+body <- dashboardBody(
+  tabItems(
+    tabItem(tabName = "hexPlot",
+      fluidRow(
+        column(width = 4, 
+         box(width = NULL, status = "primary", title = "Plot metrics", solidHeader = TRUE,
+           selectizeInput("selPair", "Treatment pairs:", choices = myPairs, multiple = TRUE, options = list(maxItems = 2)),
+           selectInput("selMetric", "Metric:", choices = myMetrics, selected = "FDR"),
+           selectInput("selOrder", "Metric order:", choices = c("Increasing", "Decreasing")),
+           numericInput("binSize", "Hexagon size:", value = 10, min = 1),
+           numericInput("pointSize", "Point size:", value = 8, min = 1),
+           actionButton("goButton", "Plot gene!"))),
+        column(width = 8,
+           box(width = NULL, withSpinner(plotlyOutput("hexPlot")), collapsible = FALSE, background = "black", title = "Litre plot", status = "primary", solidHeader = TRUE))),
+    
+      fluidRow(
+        column(width = 12,
[…]

### 2018-04-19 VirusHoneyBee/DESeq2/Dashboard/litreSigStand/app.R

@@ -1,0 +1,304 @@
+library(shinydashboard)
+library(shiny)
+library(plotly)
+library(hexbin)
+library(htmlwidgets)
+library(dplyr)
+library(tidyr)
+library(data.table)
+library(RColorBrewer)
+library(Hmisc)
+library(shinycssloaders)
+
+options(spinner.color.background="#F5F5F5")
+
+dat <- readRDS("data.Rds")
+dat$ID <- rownames(dat)
+dat <- dat[,c(7,1:6)]
+
+# Standardize in this application
+RowSD = function(x) {
+  sqrt(rowSums((x - rowMeans(x))^2)/(dim(x)[2] - 1))
+}
+dat_Rownames <- dat$ID
+dat = dat[,-1]
+rownames(dat) <- dat_Rownames
+dat <- betweenLaneNormalization(as.matrix(dat), which="full", round=FALSE)
+dat = as.data.frame(dat)
+dat = mutate(dat, mean = (C.1+C.2+C.3+T.1+T.2+T.3)/6, stdev = RowSD(cbind(C.1,C.2,C.3,T.1,T.2,T.3)))
+rownames(dat)=dat_Rownames
+dat$ID <- dat_Rownames
+datqps <- t(apply(as.matrix(dat[,1:6]), 1, scale))
+datqps <- as.data.frame(datqps)
+colnames(datqps) <- colnames(dat[,1:6])
+datqps$ID <- rownames(datqps)
+nID <- which(is.nan(datqps$V.1))
+datqps[nID,1:6] <- 0
+dat <- datqps[,c(7,1:6)]
+
+
+
+datCol <- colnames(dat)[-which(colnames(dat) %in% "ID")]
+myPairs <- unique(sapply(datCol, function(x) unlist(strsplit(x,"[.]"))[1]))
+load("Sigmetrics.rda")
+myMetrics <- colnames(metrics[[1]])[-which(colnames(metrics[[1]]) %in% "ID")]
+values <- reactiveValues(x=0, selPair=NULL, selMetric=NULL, selOrder=NULL)
+
+sidebar <- dashboardSidebar(
+  width = 180,
+  hr(),
+  sidebarMenu(id="tabs",
+    shinydashboard::menuItem("Application", tabName="hexPlot"),
+    shinydashboard::menuItem("About", tabName = "about", selected=TRUE)
+  )
+)
+
+body <- dashboardBody(
+  tabItems(
+    tabItem(tabName = "hexPlot",
+      fluidRow(
+        column(width = 4, 
[…]

### 2018-04-19 VirusHoneyBee/DESeq2/Dashboard/makeClusterData/makeDataCluster1.R

@@ -1,0 +1,16 @@
+library(bigPint)
+
+ID <- readRDS("../../ClusterStandard/Sig_4_1.Rds")
+#load("../../Bioinformatics/Pictures/FilterNotSig/soybean_ir_noFilt_metrics.rda")
+
+metricsAll <- readRDS("../../dataMetrics.Rds")
+
+allMetrics = metricsAll[["C_T"]]
+
+metricsCluster <- allMetrics[which(allMetrics$ID %in% ID),]
+metricsCluster$ID <- as.character(metricsCluster$ID)
+
+metrics <- list()
+metrics[["C_T"]] <- metricsCluster
+
+save(metrics, file = "SigCluster1.rda")

### 2018-04-19 VirusHoneyBee/DESeq2/Dashboard/makeClusterData/makeDataSig.R

@@ -1,0 +1,16 @@
+library(bigPint)
+
+ID <- readRDS("../../ClusterStandard/Sig.Rds")
+#load("../../Bioinformatics/Pictures/FilterNotSig/soybean_ir_noFilt_metrics.rda")
+
+metricsAll <- readRDS("../../dataMetrics.Rds")
+
+allMetrics = metricsAll[["C_T"]]
+
+metricsCluster <- allMetrics[which(allMetrics$ID %in% ID),]
+metricsCluster$ID <- as.character(metricsCluster$ID)
+
+metrics <- list()
+metrics[["C_T"]] <- metricsCluster
+
+save(metrics, file = "Sigmetrics.rda")

### 2018-04-19 Manuscript/chapter4.tex
Results

@@
 
-\subsection{Main effect results}
+\subsection{Main effect DEG results}
 
@@
 While these reported DEGs numbers are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Table \ref{tbl:mainEffectDEGs}A and B).
-Pathway
+
+Pathway analysis of the Castanea DEGs revealed enriched (Benjamini correction < 0.05) Wnt signaling, hippo signaling, and dorso-ventral axis formation pathways, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Table \ref{tbl:CastaneaPathways}).
+Pathway analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Table \ref{tbl:RockrosePathways}).
+
+@@@ Not enough DEGs to form pathways for virus @@@
+
+\subsection{Interaction DEG results}
+
+No interaction DEGs were observed between the diet and virus factors of the study, in any of the pipelines (DESeq2, edgeR, limma).
+
+\subsection{Pairwise comparison DEG results}
+
+The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 941 (Table \ref{tbl:pairDEGs}).
+Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level.
+Across every pair comparing the Castanea and Rockrose levels, regardless of the virus level, the number of Castanea-related DEGs was higher than the number of Rockrose-related DEGs (Table \ref{tbl:pairDEGs} C, D, E, F).
+For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Table \ref{tbl:pairDEGs} A, B).
+There were no DEGs between the treatment pair controlling for the control level of the virus effect (Table \ref{tbl:pairDEGs} A).
+These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).
+
+\subsection{Comparison with Galbraith study}
+
+Differences between signal to noise ratio
+
+Figure \ref{fig:pcpGalbraith}
+Figure \ref{fig:pcpRutterVirus}
+Figure \ref{fig:pcpRutterDiet}
+
+Overlap between studies Figure \ref{fig:GRVenn}.
+
+27 genes overlapped ("GB40006" "GB41739" "GB42046" "GB42048" "GB42313" "GB43783" "GB43784" "GB45462" "GB45704" "GB47214" "GB47381" "GB47407" "GB48747" "GB48755" "GB49920" "GB50178" "GB50401" "GB50550" "GB50813" "GB50955" "GB51305" "GB52281" "GB52441" "GB52449" "GB53500" "GB53833" "GB55188")
+
+26 genes overlapped virus ("GB40006" "GB41739" "GB42046" "GB42048" "GB42313" "GB43783" "GB43784" "GB45462" "GB45704" "GB47214" "GB47381" "GB47407" "GB48747" "GB48755" "GB49920" "GB50178" "GB50401" "GB50550" "GB50813" "GB50955" "GB52281" "GB52441" "GB52449" "GB53500" "GB53833" "GB55188")
 
@@
 
-% \begin{table}
-%   \caption{A stunning table}
-%   \label{tbl:excel-table}
-%   \includegraphics[width=\linewidth]{excel-table}
-% \end{table}
-
 \begin{table}[H]
@@
 
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/CastaneaPathways}
+  \caption{Pathways related to diet main effect Castanea DEGs.}
+  \label{tbl:CastaneaPathways}
+\end{table}
+
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/CastaneaPathways}
+  \caption{Pathways related to diet main effect Rockrose DEGs.}
+  \label{tbl:RockrosePathways}
+\end{table}
+
+\begin{table}[H]
+  \includegraphics[width=0.65\textwidth]{Images/pairDEGs}
+  \caption{Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor.}
+  \label{tbl:pairDEGs}
+\end{table}
+
+\begin{figure}[H]
+\centering
+  \includegraphics[width=\textwidth]{../VirusHoneyBee/DESeq2/ClusterStandard/Clustering_data_FDR_05/C_T_4.jpg}
+  \caption{Parallel coordinate plots of DEGs between the virus-infected and control groups of the Galbraith study.}
+  \label{fig:pcpGalbraith}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+  \includegraphics[width=\textwidth]{../N_V/DESeq2/ClusterStandard/Clustering_data_FDR_05/N_V_4.jpg}
+  \caption{Parallel coordinate plots of DEGs between the virus-infected and control groups of our study.}
+  \label{fig:pcpRutterVirus}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{../C_R/DESeq2/ClusterStandard/Clustering_data_FDR_05/C_R_6.jpg}
+  \caption{Parallel coordinate plots of DEGs between the Castanea and Rockrose groups of our study.}
+  \label{fig:pcpRutterDiet}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/GRVenn}
+  \caption{Venn diagrams comparing the DEG overlaps of the diet and treatment groups between the Galbraith study and our study.
+Top row left to right: DESeq results for all DEGs, virus DEGs, and control DEGs.
+Bottom row left to right: EdgeR results for all DEGs, virus DEGs, and control DEGs.}
+  \label{fig:GRVenn}
+\end{figure}
+
 \bibliography{chapter4}

### 2018-04-19 Manuscript/chapter4.tex
Mortality ANOVA description

@@
 We briefly show it again here to inform the RNA-seq comparison because we reduced the number of treatments (from eight to four) from the original published data (\citealt{adamInt}).
-When statistically analyzing the subset of this data that was used for RNA-sequencing analysis, we found that the mortality rates across diet quality and virus exposure at least numerically retained the same trends (Figure \ref{fig:4panels}).
 
-Mortality rates of honeybees 72 hour post-inoculation differed (or did not?) among the treatment groups (mixed model ANOVA across all treatment groups, df=@@@, @@@; F=@@@; p<@@@).
-The effect of virus treatment (mixed model ANOVA, df=@@@, @@@; F=@@@; p<@@@), diet treatment (mixed model ANOVA, df=@@@,@@@; F=@@@; p<@@@), and interaction between the two factors (mixed model ANOVA, df=@@@, @@@; F=@@@, p=@@@) did (or did not?) differ.
+Mortality rates of honeybees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df=3, 55; F=10.07; p<2.18e-05).
+The effect of virus treatment (mixed model ANOVA, df=1, 55; F=24.343; p<7.84e-06) and diet treatment (mixed model ANOVA, df=1, 55; F=5.796; p<0.0194) were significant, but the interaction between the two factors (mixed model ANOVA, df=1, 55; F=0.062, p=0.8039) was not significant.
 The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05).
-Without virus exposure, there was only an intermediate reduction in mortality rate for bees fed Castanea pollen?
-(Tukey HSD, p>0.05).
-However, with virus exposure, there was a significant reduction in mortality rate for beeds fed Castanea pollen (Tukey HSD, p<0.05).
-Overall, we discovered that the higher-quality Castanea diet had the ability to significantly reduce mortality in the presence of IAPV infection compared to the lower-quality Cistus diet (Figure \ref{fig:4panels}).
+Without virus exposure, there was only a numerical reduction in mortality rate for bees fed Castanea pollen compared to Rockrose pollen (Tukey HSD, p>0.05).
+Similarly, with virus exposure, there was only a numerical reduction in mortality rate for beeds fed Castanea pollen compared to Rockrose pollen (Tukey HSD, p>0.05).
+Overall, we discovered that the higher-quality Castanea diet had the ability to numerically reduce mortality in the presence of IAPV infection compared to the lower-quality Rockrose diet (Figure \ref{fig:mortality} A).
 
-Comment on (Figure \ref{fig:4panels}) @@@
+IAPV titer volumes of honeybees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df=3, 34; F=6.096; p<0.00196).
+The effect of virus treatment (mixed model ANOVA, df=1, 34; F=15.686; p<0.000362) was significant, but the diet treatment (mixed model ANOVA, df=1, 34; F=1.898; p>0.05) and the interaction between the two factors (mixed model ANOVA, df=1, 34; F=0.702, p>0.05) were not significant.
+Honeybees that were infected with the virus and fed a poor-quality Rockrose diet showed significant increases in IAPV titer volumes compared to honeybees that were not infected with the virus regardless of their diet quality (Tukey HSD, p<0.05).
+Overall, we discovered that the higher-quality Castanea diet had the ability to numerically reduce IAPV titer volume for both infected and non-infected honeybees (Figure \ref{fig:mortality} B).
 
@@
   \begin{framed}
-  \includegraphics[width=0.8\textwidth]{Images/4panels}
+  \includegraphics[width=\textwidth]{Images/mortality}
   \end{framed}
-  \caption{Mortality rates for all cages, mortality rates for subset of honeybees used for RNA-seq, IAPV titers for all cages, IAPV titers for subset of honeybees used for RNA-seq subset.}
-  \label{fig:4panels}
+  \caption{Mortality rates (A) and IAPV titers (B) for the four treatment groups.}
+  \label{fig:mortality}
 \end{figure}

### 2018-04-19 MortalityData/HSD_us.R

@@ -14,7 +14,15 @@ d$Treatment = as.factor(d$Treatment)
 colnames(d)[2] = "Day3Mortality"
 levels(d$Treatment) <- c('NC', 'NR', 'VC', 'VR')
 
+d$Virus = sapply(d$Treatment, function(x) substr(x, 1, 1))
+d$Diet = sapply(d$Treatment, function(x) substr(x, 2, 2))
+d$Virus = as.factor(d$Virus)
+d$Diet = as.factor(d$Diet)
+
 a <- aov(Day3Mortality~Treatment, data=d)
+
+aov.out = aov(Day3Mortality ~ Virus * Diet, data=d)
+
 tHSD <- TukeyHSD(a, ordered = FALSE, conf.level = 0.95)
 
 generate_label_df1 <- function(HSD, flev){
@@ -52,7 +60,14 @@ colnames(d)[2] = "logIAPV"
 levels(d$treatment) <- c('NC', 'NR', 'VC', 'VR')
 colnames(d)[1] <- "Treatment"
 
+
+d$Virus = sapply(d$Treatment, function(x) substr(x, 1, 1))
+d$Diet = sapply(d$Treatment, function(x) substr(x, 2, 2))
+d$Virus = as.factor(d$Virus)
+d$Diet = as.factor(d$Diet)
+
 a <- aov(logIAPV~Treatment, data=d)
+aov.out = aov(logIAPV ~ Virus * Diet, data=d)
 tHSD <- TukeyHSD(a, ordered = FALSE, conf.level = 0.95)
 
 generate_label_df2 <- function(HSD, flev){

### 2018-04-20 Manuscript/chapter4.tex
Add litre plots

@@
 Without virus exposure, there was only a numerical reduction in mortality rate for bees fed Castanea pollen compared to Rockrose pollen (Tukey HSD, p>0.05).
-Similarly, with virus exposure, there was only a numerical reduction in mortality rate for beeds fed Castanea pollen compared to Rockrose pollen (Tukey HSD, p>0.05).
+Similarly, with virus exposure, there was only a numerical reduction in mortality rate for bees fed Castanea pollen compared to Rockrose pollen (Tukey HSD, p>0.05).
 Overall, we discovered that the higher-quality Castanea diet had the ability to numerically reduce mortality in the presence of IAPV infection compared to the lower-quality Rockrose diet (Figure \ref{fig:mortality} A).
@@
 
-Pathway analysis of the Castanea DEGs revealed enriched (Benjamini correction < 0.05) Wnt signaling, hippo signaling, and dorso-ventral axis formation pathways, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Table \ref{tbl:CastaneaPathways}).
+Pathway analysis of the Castanea DEGs revealed enriched (Benjamini correction < 0.05) Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Table \ref{tbl:CastaneaPathways}).
 Pathway analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Table \ref{tbl:RockrosePathways}).
 
-@@@ Not enough DEGs to form pathways for virus @@@
+It was difficult to perform a pathway analysis with so few DEGs (n = 43) in our virus main effect study.
+As a result, we focused on individual genes and their known functionalities (Table \ref{tbl:virusGenes}).
+Of the 43 virus-related DEGs, only 10 of them successfully mapped with DAVID software.
+These genes had implications in the recognition of pathogen-related lipid products and the cleaving of transcripts from viruses, as well as involvement in ubiquitin and proteosome pathways, transcription pathways, apoptopic pathways, oxidoreductase processes, and several more functions (Table \ref{tlb:virusGenes}).
 
@@
 
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/virusGenes}
+  \caption{Known functions of the mapped subset of 43 DEGs in the virus main effect of our study.
+Whether the gene was overrepresented in the virus or non-virus group is also indicated for both our study and the Galbraith study.
+Functionalities were extracted from Flybase, National Center for Biotechnology Information, and The European Bioinformatics Institute databases.}
+  \label{tbl:virusGenes}
+\end{table}
+
 \bibliography{chapter4}

### 2018-04-20 Manuscript/chapter4.tex
Resilience versus resistance

@@
 
+\subsection{Detecting resilience versus resistance}
+
+To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (resilience), we created contrasts of interest (Table \ref{tbl:contrasts}).
+In particular, we assigned ``resistance" genes to be the ones that were upregulated in the Castanea group within the virus infected bees but not upregulated in the Castanea group within the non-infected bees.
+We also assigned "resilience" genes to be the ones that were upregulated in the Castanea group for both the virus infected bees and non-infected bees.
+We then determined how many genes fell into these two categories and analyzed their GO terminologies.
+
 \section{Results}
@@
 
-Differences between signal to noise ratio
+We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset.
+Basic MDS plots are constructed during the DESeq2 analysis pipeline, and we could immediately determine that the Galraith dataset may better separate the infected and uninfected honeybees better than our dataset (Figure \ref{fig:mdsPlots}).
+We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the MDS plots.
+However, through this automatically-generated plot, we can only visualize information at the sample level.
+Wanting to learn more about the data at the gene level, we continued with additional visualization techniques.
 
-Figure \ref{fig:pcpGalbraith}
-Figure \ref{fig:pcpRutterVirus}
-Figure \ref{fig:pcpRutterDiet}
+We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs that came out of the DESeq2 package.
+The background boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG.
+To reduce overlapping of parallel coordinate lines, we often use hierarchical clustering techniques to separate DEGs into common patterns.
+See more information about this plotting method and the ideal visual structure of DEGs in our earlier chapter.
 
-Overlap between studies Figure \ref{fig:GRVenn}.
+We see that the 1,019 DEGs from the Galbraith dataset form pretty clean-looking visual displays (Figure \ref{fig:pcpGalbraith}).
+We do see that the first replicate of the virus group appears somewhat inconsistent with the other virus replicates in Cluster 2, confirming that this trend in the data that we saw in the MDS plot carried through into the DEG results.
+In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure \ref{fig:pcpRutterVirus}).
+The replicates appear strikingly inconsistent and there is not always such a large difference between treatment groups.
+We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study.
 
-27 genes overlapped ("GB40006" "GB41739" "GB42046" "GB42048" "GB42313" "GB43783" "GB43784" "GB45462" "GB45704" "GB47214" "GB47381" "GB47407" "GB48747" "GB48755" "GB49920" "GB50178" "GB50401" "GB50550" "GB50813" "GB50955" "GB51305" "GB52281" "GB52441" "GB52449" "GB53500" "GB53833" "GB55188")
+We also used litre plots to examine the structure of individual DEGs: We see that indeed the individual DEGs from our data (Figure \ref{fig:litreClusterRutter}) show less consistent replications and less differences between the treatment groups compared to the individual DEGs from the Galbraith data (Figure \ref{fig:litreCluster1} and Figure \ref{fig:litreCluster2}).
+For the Galbraith data, we examined individual DEGs from the first cluster (Figure \ref{fig:litreCluster1}) and second cluster (Figure \ref{fig:litreCluster2}) because the second cluster was a bit less ideal due to its inconsistent first replicate of the treatment group.
 
-26 genes overlapped virus ("GB40006" "GB41739" "GB42046" "GB42048" "GB42313" "GB43783" "GB43784" "GB45462" "GB45704" "GB47214" "GB47381" "GB47407" "GB48747" "GB48755" "GB49920" "GB50178" "GB50401" "GB50550" "GB50813" "GB50955" "GB52281" "GB52441" "GB52449" "GB53500" "GB53833" "GB55188")
+Finally, we looked at scatterplot matrices to assess the DEGs.
+We created standardized scatterplot matrices for each of the four clusters (Figure \ref{fig:pcpGalbraith}) of the Galbriath data (Figures \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}).
+We also created standardized scatterplot matrices for our data.
+However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would not be helpful.
+As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Figures \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).
+We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: Deviating more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.
+
+Despite the DEGs from the Galbraith dataset displaying the expected patterns more than those from our dataset, there was significant overlap in the DEGs between the two studies (Figure \ref{fig:GRVenn}).
+
+\subsection{Resilience versus resistance}
+
+Figure \ref{fig:revigo}
 
@@
 
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/mdsPlots}
+  \caption{MDS plots constructed from DESeq2 package for the Galbraith dataset (A) and our dataset (B).}
+  \label{fig:mdsPlots}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/litreCluster2}
+  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 1 (back in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.}
+  \label{fig:litreCluster2}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/litreCluster1}
+  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 2 (back in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.}
+  \label{fig:litreCluster1}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/litreClusterRutter}
+  \caption{Example litre plots of the nine DEGs with the lowest FDR values from the 43 DEGs of our dataset.}
+  \label{fig:litreClusterRutter}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/GalbraithClust1SM}
+  \caption{The DEGs from the first cluster of the Galbraith dataset (back in Figure \ref{fig:pcpGalbraith}) superimposed onto all genes in the form of a scatterplot matrix.
+The data has been standardized.}
+  \label{fig:GalbraithClust1SM}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/GalbraithClust2SM}
+  \caption{The DEGs from the second cluster of the Galbraith dataset (back in Figure \ref{fig:pcpGalbraith}) superimposed onto all genes in the form of a scatterplot matrix.
+The data has been standardized.}
+  \label{fig:GalbraithClust2SM}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/GalbraithClust3SM}
+  \caption{The DEGs from the third cluster of the Galbraith dataset (back in Figure \ref{fig:pcpGalbraith}) superimposed onto all genes in the form of a scatterplot matrix.
+The data has been standardized.}
+  \label{fig:GalbraithClust3SM}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/GalbraithClust4SM}
+  \caption{The DEGs from the fourth cluster of the Galbraith dataset (back in Figure \ref{fig:pcpGalbraith}) superimposed onto all genes in the form of a scatterplot matrix.
+The data has been standardized.}
+  \label{fig:GalbraithClust4SM}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/RutterSM1}
+  \caption{The 43 virus-related DEGs from our dataset superimposed onto all genes in the form of a scatterplot matrix.
+Only replicates 1, 2, and 3 are shown from both treatment groups.
+The data has been standardized.}
+  \label{fig:RutterSM1}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/RutterSM2}
+  \caption{The 43 virus-related DEGs from our dataset superimposed onto all genes in the form of a scatterplot matrix.
+Only replicates 4, 5, and 6 are shown from both treatment groups.
+The data has been standardized.}
+  \label{fig:RutterSM2}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/RutterSM3}
+  \caption{The 43 virus-related DEGs from our dataset superimposed onto all genes in the form of a scatterplot matrix.
+Only replicates 7, 8, and 9 are shown from both treatment groups.
+The data has been standardized.}
+  \label{fig:RutterSM3}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/RutterSM4}
+  \caption{The 43 virus-related DEGs from our dataset superimposed onto all genes in the form of a scatterplot matrix.
+Only replicates 10, 11, and 12 are shown from both treatment groups.
+The data has been standardized.}
+  \label{fig:RutterSM4}
+\end{figure}
+
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/NCNRNCPathways}
+  \caption{Pathways analysis results for the  601 DEGs that were upregulated in the NC treatment in the NC versus NR treatment pair analysis.}
+  \label{tbl:NCNRNCPathways}
+\end{table}
+
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/NCNRNRPathways}
+  \caption{Pathways analysis results for the 340 DEGs that were upregulated in the NR treatment in the NC versus NR treatment pair analysis.}
+  \label{tbl:NCNRNRPathways}
+\end{table}
+
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/VCVRVCPathways}
+  \caption{Pathways analysis results for the 247 DEGs that were upregulated in the VC treatment in the VC versus VR treatment pair analysis.}
+  \label{tbl:VCVRVCPathways}
+\end{table}
+
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/VCVRVRPathways}
+  \caption{Pathways analysis results for the 129 DEGs that were upregulated in the VR treatment in the VC versus VR treatment pair analysis.}
+  \label{tbl:VCVRVRPathways}
+\end{table}
+
+\begin{figure}[H]
+  \includegraphics[width=0.8\textwidth]{Images/revigo}
+  \caption{Gene ontology results for the 122 DEGs related to our ``resilience" hypothesis (A) and for the 125 DEGs related to our ``resistance" hypothesis (B).}
+  \label{fig:revigo}
+\end{figure}
+
+\begin{table}[H]
+\begin{tabular}{ | m{5cm} | m{0.9cm}| m{5.5cm} | m{2cm} | } 
+\hline
+Contrast & DEGs & Interpretation & Results \\ 
+\hline
+V vs N & 43 & Genes that change expression due to virus effect regardless of diet status in bees & Table \ref{tbl:virusGenes} \\ 
+\hline
+NC vs NR & 941 & Genes that change expression due to diet effect in uninfected bees & Tables \ref{tbl:NCNRNCPathways} and \ref{tbl:NCNRNRPathways} \\ 
+\hline
+VC vs VR & 376 & Genes that change expression due to diet effect in infected bees & Tables \ref{tbl:VCVRVCPathways} and \ref{tbl:VCVRVRPathways} \\ 
+\hline
+VC upregulated in VC vs VR overlaped with NC upregulated in NC vs NR & 122 & ``Resilience'' genes that are turned on by good diet regardless of virus infection status in bees & Figure \ref{fig:revigo}A \\
+\hline
+VC upregulated in VC vs VR but NC is not upregulated in NC vs NR & 125 & ``Resistance'' genes that are turned on by good diet only in infected bees & Figure \ref{fig:revigo}B \\
+\hline
+\end{tabular}
+\caption{Contrasts in our study for assessing GO and pathways analysis.}
+  \label{tbl:contrasts}
+\end{table}
+
 \bibliography{chapter4}

### 2018-04-20 N_V/DESeq2/ClusterStandard/Clustering_data_FDR_05.R

@@ -132,6 +132,28 @@ boxDat <- melt(datas, id.vars="ID")
 colnames(boxDat) <- c("ID", "Sample", "Count")
 
 sigDatas = datas[which(metricsAll[["N_V"]]$padj<0.05),]
+
+#### Plot SM ######################## 
+
+xNames = rownames(sigDatas)
+xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+scatMatMetrics = list()
+scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
+scatMatMetrics[[currPair]]$FDR = 10e-10
+scatMatMetrics[[currPair]]$ID = as.factor(as.character(scatMatMetrics[[currPair]]$ID))
+plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
+
+plotDatas2 = plotDatas[,c(1,11:13,23:25)]
+
+logData2 = logData[,c(1,11:13,23:25)]
+
+ret <- plotDEG(data = logData2, dataMetrics = scatMatMetrics, option="scatterPoints", threshVar = "padj", threshVal = 0.05, degPointColor = "lawngreen")
+
+#, fileName=fileName)
+#degPointColor = colList[i+1],
+
+#####################################
+
 dendo = sigDatas
 rownames(dendo) = NULL
 d = dist(as.matrix(dendo))

### 2018-04-20 VirusHoneyBee/DESeq2/ClusterStandard/Clustering_data_FDR_05.R

@@ -19,6 +19,8 @@ getPCP <- function(nC){
   
   set.seed(1)
   colList = scales::hue_pal()(nC+1)
+  colList[2] = "#E9AA0D"
+  colList[3] = "#EA502F"
   k = cutree(hc, k=nC)
   
   yMin = min(sigDatas[,1:nColumns])
@@ -49,7 +51,7 @@ getPCP <- function(nC){
     plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
     
     fileName = paste(getwd(), "/", outDir, "/", currPair, "_SM_", nC, "_", i, ".jpg", sep="")
-    plotDEG(data = plotDatas, dataMetrics = scatMatMetrics, option="scatterPoints", threshVar = "padj", threshVal = 0.05, degPointColor = colList[i+1], fileName=fileName)
+    ret <- plotDEG(data = plotDatas, dataMetrics = scatMatMetrics, option="scatterPoints", threshVar = "padj", threshVal = 0.05, degPointColor = colList[i+1], fileName=fileName)
     
     x$ID = xNames
     saveRDS(xNames, file=paste0("Sig_", nC,"_", i,".Rds"))

### 2018-04-20 VirusHoneyBee/DESeq2/Dashboard/litreCluster1/app.R

@@ -188,7 +188,7 @@ server <- function(input, output, session) {
    y: drawPoints.geneY,
    mode: 'markers',
    marker: {
-   color: '#6E1CC7',
+   color: '#E9AA0D',
    size: drawPoints.pointSize
    },
    hoverinfo: 'none',
@@ -268,7 +268,7 @@ server <- function(input, output, session) {
     y: drawLines,
     mode: 'lines',
     line: {
-    color: '#6E1CC7',
+    color: '#E9AA0D',
     width: 2
     },
     opacity: 0.9,

### 2018-04-20 VirusHoneyBee/DESeq2/Dashboard/litreCluster2/app.R

@@ -1,0 +1,283 @@
+library(shinydashboard)
+library(shiny)
+library(plotly)
+library(hexbin)
+library(htmlwidgets)
+library(dplyr)
+library(tidyr)
+library(data.table)
+library(RColorBrewer)
+library(Hmisc)
+library(shinycssloaders)
+
+options(spinner.color.background="#F5F5F5")
+
+dat <- readRDS("data.Rds")
+dat$ID <- rownames(dat)
+dat <- dat[,c(7,1:6)]
+
+dat[,-1] <- log(dat[,-1]+1)
+datCol <- colnames(dat)[-which(colnames(dat) %in% "ID")]
+myPairs <- unique(sapply(datCol, function(x) unlist(strsplit(x,"[.]"))[1]))
+load("SigCluster2.rda")
+myMetrics <- colnames(metrics[[1]])[-which(colnames(metrics[[1]]) %in% "ID")]
+values <- reactiveValues(x=0, selPair=NULL, selMetric=NULL, selOrder=NULL)
+
+sidebar <- dashboardSidebar(
+  width = 180,
+  hr(),
+  sidebarMenu(id="tabs",
+    shinydashboard::menuItem("Application", tabName="hexPlot"),
+    shinydashboard::menuItem("About", tabName = "about", selected=TRUE)
+  )
+)
+
+body <- dashboardBody(
+  tabItems(
+    tabItem(tabName = "hexPlot",
+      fluidRow(
+        column(width = 4, 
+         box(width = NULL, status = "primary", title = "Plot metrics", solidHeader = TRUE,
+           selectizeInput("selPair", "Treatment pairs:", choices = myPairs, multiple = TRUE, options = list(maxItems = 2)),
+           selectInput("selMetric", "Metric:", choices = myMetrics, selected = "FDR"),
+           selectInput("selOrder", "Metric order:", choices = c("Increasing", "Decreasing")),
+           numericInput("binSize", "Hexagon size:", value = 10, min = 1),
+           numericInput("pointSize", "Point size:", value = 8, min = 1),
+           actionButton("goButton", "Plot gene!"))),
+        column(width = 8,
+           box(width = NULL, withSpinner(plotlyOutput("hexPlot")), collapsible = FALSE, background = "black", title = "Litre plot", status = "primary", solidHeader = TRUE))),
+    
+      fluidRow(
+        column(width = 12,
[…]

### 2018-04-20 VirusHoneyBee/DESeq2/Dashboard/makeClusterData/makeDataCluster2.R

@@ -1,0 +1,16 @@
+library(bigPint)
+
+ID <- readRDS("../../ClusterStandard/Sig_4_2.Rds")
+#load("../../Bioinformatics/Pictures/FilterNotSig/soybean_ir_noFilt_metrics.rda")
+
+metricsAll <- readRDS("../../dataMetrics.Rds")
+
+allMetrics = metricsAll[["C_T"]]
+
+metricsCluster <- allMetrics[which(allMetrics$ID %in% ID),]
+metricsCluster$ID <- as.character(metricsCluster$ID)
+
+metrics <- list()
+metrics[["C_T"]] <- metricsCluster
+
+save(metrics, file = "SigCluster2.rda")

### 2018-04-21 Manuscript/chapter4.tex
Incorporate methods section suggestions from Adam

@@
 
-There are several reasons why we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
+There are several reasons why, in the current study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
 First, when assessing diet diversity, a sugar diet is often used as a control.
@@
 
-Consequently, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut).
-Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}).
+Consequently, for our nutrition factor, we examined two monofloral pollen diets, Rockrose (Cistus) and Chestnut (Chestnut).
+Rockrose pollen is generally considered less nutritious than Chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}).
 For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV.
-This experimental design resulted in four treatment groups (Cistus pollen without IAPV exposure, Castanea pollen without IAPV exposure, Cistus pollen with IAPV exposure, and Castanea pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.
+This experimental design resulted in four treatment groups (Rockrose pollen without IAPV exposure, Chestnut pollen without IAPV exposure, Rockrose pollen with IAPV exposure, and Chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.
 
@@
 Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}).
+There were four lanes of sequencing with 24 samples per lane.
+Each sample was run twice.
+Approximately 75-90\% of reads were mapped to the honeybee genome.
+Each lane produced around 13 million single-end 100 basepair reads.
 We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs).
@@
 
-@@@ What percent of reads mapped?
-@@@ Total number of raw reads
-@@@ How many lanes
-@@@ How many samples per lane
-
 \subsection{Comparison to previous studies on transcriptomic response to viral infection}
 
-We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues (\citealt{galbraith}).
+We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues (\citealt{galbraith}) who also addressed honey bee transcriptomic responses to virus infection.
+
 While our study examines honeybees from polyandrous colonies, the Galbraith study examined honeybees from single-drone colonies.
-As a consequence, our honeybees will have an average of about 75\% genetic variance, and the honeybees from the Galbraith study will have an average of about 25\% genetic variance (\citealt{sisters}).
+As a consequence, the honeybees in our study will be on average 25\% genetically identical, whereas honeybees from the Galbraith study will be on average 75\% genetically identical (\citealt{sisters}).
 We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates.
@@
 
-We used @@@ visualization tools from @@@ and visual inference techniques to assess the signal:to:noise ratio in the datasets and to assess the suitability of the DEG calls.
+We used visualization tools from the DESeq2 package, visualization tools from our work-in-progress bigPint package, and visual inference techniques to assess the signal:to:noise ratio in the datasets and to assess the suitability of the DEG calls.
 
@@
 The overrepresented gene ontology (GO) terms of DEGs were identified using the BEEBASE\_ID identifier.
-To fine-tune the GO term list, only significant terms (FDR < 0.05) and those correlating to Biological Processes were considered.
+To fine-tune the GO term list, only terms correlating to Biological Processes were considered.
 The refined GO term list was then imported into REVIGO (\citealt{revigo}), which uses semantic similarity measures to cluster long lists of GO terms.
-@@@ Pathways analysis @@@
 
@@
 To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (resilience), we created contrasts of interest (Table \ref{tbl:contrasts}).
-In particular, we assigned ``resistance" genes to be the ones that were upregulated in the Castanea group within the virus infected bees but not upregulated in the Castanea group within the non-infected bees.
-We also assigned "resilience" genes to be the ones that were upregulated in the Castanea group for both the virus infected bees and non-infected bees.
+In particular, we assigned ``resistance candidate genes" to be the ones that were upregulated in the Chestnut group within the virus infected bees but not upregulated in the Chestnut group within the non-infected bees.
+We also assigned "resilience candidate genes" to be the ones that were upregulated in the Chestnut group for both the virus infected bees and non-infected bees.
+Our interpretation of these genes is that they represent genes that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not.
 We then determined how many genes fell into these two categories and analyzed their GO terminologies.
@@
 The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05).
-Without virus exposure, there was only a numerical reduction in mortality rate for bees fed Castanea pollen compared to Rockrose pollen (Tukey HSD, p>0.05).
-Similarly, with virus exposure, there was only a numerical reduction in mortality rate for bees fed Castanea pollen compared to Rockrose pollen (Tukey HSD, p>0.05).
-Overall, we discovered that the higher-quality Castanea diet had the ability to numerically reduce mortality in the presence of IAPV infection compared to the lower-quality Rockrose diet (Figure \ref{fig:mortality} A).
+In comparing mortality levels based on pairwise comparisons, we found that bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls.
+However, bees fed Castanea pollen had no significant difference in mortality between virus infected and control groups.
+These results suggest the high quality Castanea diet can ``rescue'' virus induced mortality (Figure \ref{fig:mortality}A).
 
@@
 Honeybees that were infected with the virus and fed a poor-quality Rockrose diet showed significant increases in IAPV titer volumes compared to honeybees that were not infected with the virus regardless of their diet quality (Tukey HSD, p<0.05).
-Overall, we discovered that the higher-quality Castanea diet had the ability to numerically reduce IAPV titer volume for both infected and non-infected honeybees (Figure \ref{fig:mortality} B).
+Overall, we interpreted this effect to mean that Rockrose pollen could not ``rescue'' high virus titers resulting from the inoculation treatment, whereas Castanea pollen could (Figure \ref{fig:mortality}B).
 
@@
   \end{framed}
-  \caption{Mortality rates (A) and IAPV titers (B) for the four treatment groups.}
+  \caption{Mortality rates (A) and IAPV titers (B) for the four treatment groups.
+``N" represent non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen.
+The mortality rate data included 59 samples with 15 replicates per treatment group, except for the ``NC'' group having 14 replicates.
+The IAPV titer data included 38 samples with 10 replicates per treatment group, except for the ``NR'' group having 8 replicates.
+ANOVA values and p-values for the statistical tests are listed in the text of the paper.
+The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
   \label{fig:mortality}
@@
 We observed a substantially larger number of DEGs in our diet main effect (n = 1914) than in our virus main effect (n = 43) (Table \ref{tbl:mainEffectDEGs}A and B).
-In the diet factor, there were more Castanea DEGs (n = 1033) than Rockrose DEGs (n = 881).
-In the virus factor, there were more virus-exposed DEGs (n = 38) than control DEGs (n = 5).
+In the diet factor, there were more Chestnut-upregulated DEGs (n = 1033) than Rockrose-upregulated DEGs (n = 881).
+In the virus factor, there were more virus-upregulated DEGs (n = 38) than control-upregulated DEGs (n = 5).
 While these reported DEGs numbers are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Table \ref{tbl:mainEffectDEGs}A and B).
 
-Pathway analysis of the Castanea DEGs revealed enriched (Benjamini correction < 0.05) Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Table \ref{tbl:CastaneaPathways}).
-Pathway analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Table \ref{tbl:RockrosePathways}).
+GO analysis of the Chestnut-upregulated DEGs revealed the following enriched categories (Benjamini correction < 0.05): Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Table \ref{tbl:ChestnutPathways}).
+GO analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Table \ref{tbl:RockrosePathways}).
 
-It was difficult to perform a pathway analysis with so few DEGs (n = 43) in our virus main effect study.
-As a result, we focused on individual genes and their known functionalities (Table \ref{tbl:virusGenes}).
-Of the 43 virus-related DEGs, only 10 of them successfully mapped with DAVID software.
-These genes had implications in the recognition of pathogen-related lipid products and the cleaving of transcripts from viruses, as well as involvement in ubiquitin and proteosome pathways, transcription pathways, apoptopic pathways, oxidoreductase processes, and several more functions (Table \ref{tlb:virusGenes}).
-
-\subsection{Interaction DEG results}
+With so few DEGs (n = 43) in our virus main effect study, we focused on individual genes and their known functionalities (Table \ref{tbl:virusGenes}).
+Of the 43 virus-related DEGs, only 10 had GO assignments within the DAVID database.
+These genes had implications in the recognition of pathogen-related lipid products and the cleaving of transcripts from viruses, as well as involvement in ubiquitin and proteosome pathways, transcription pathways, apoptopic pathways, oxidoreductase processes, and several more functions (Table \ref{tbl:virusGenes}).
 
@@
 Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level.
-Across every pair comparing the Castanea and Rockrose levels, regardless of the virus level, the number of Castanea-related DEGs was higher than the number of Rockrose-related DEGs (Table \ref{tbl:pairDEGs} C, D, E, F).
+Across every pair comparing the Chestnut and Rockrose levels, regardless of the virus level, the number of Chestnut-upregulated DEGs was higher than the number of Rockrose-upregulated DEGs (Table \ref{tbl:pairDEGs} C, D, E, F).
 For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Table \ref{tbl:pairDEGs} A, B).
@@
 We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset.
-Basic MDS plots are constructed during the DESeq2 analysis pipeline, and we could immediately determine that the Galraith dataset may better separate the infected and uninfected honeybees better than our dataset (Figure \ref{fig:mdsPlots}).
+Basic MDS plots were constructed with the DESeq2 analysis pipeline, and we could immediately determine that the Galbraith dataset may better separate the infected and uninfected honeybees better than our dataset (Figure \ref{fig:mdsPlots}).
 We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the MDS plots.
@@
 
-We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs that came out of the DESeq2 package.
+We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies.
 The background boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG.
 To reduce overlapping of parallel coordinate lines, we often use hierarchical clustering techniques to separate DEGs into common patterns.
-See more information about this plotting method and the ideal visual structure of DEGs in our earlier chapter.
+See more information about this plotting method and the ideal visual structure of DEGs in our earlier chapter @@@.
 
-We see that the 1,019 DEGs from the Galbraith dataset form pretty clean-looking visual displays (Figure \ref{fig:pcpGalbraith}).
+We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays (Figure \ref{fig:pcpGalbraith}).
 We do see that the first replicate of the virus group appears somewhat inconsistent with the other virus replicates in Cluster 2, confirming that this trend in the data that we saw in the MDS plot carried through into the DEG results.
 In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure \ref{fig:pcpRutterVirus}).
-The replicates appear strikingly inconsistent and there is not always such a large difference between treatment groups.
+The replicates appear somewhat inconsistent in their esimated expression levels and there is not always such a large difference between treatment groups.
 We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study.
@@
 We also created standardized scatterplot matrices for our data.
-However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would not be helpful.
+However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data.
 As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Figures \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).
@@
 
-Figure \ref{fig:revigo}
+Within our ``resilience" gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism).
+However, we also discovered gene ontologies related to RNA polyerase II transcription and immune response (Figure \ref{fig:revigo}A).
+Within our ``resistance" gene ontologies, we found functions related to metabolism (such as carboyhydrate metabolism, chitin metabolism, and general metabolism).
+(Figure \ref{fig:revigo}B).
 
@@
 
-% \begin{tabular}{|p{2.6cm}|p{5.65cm}|p{5.65cm}|}
-%  \hline
-%  & \textbf{Drumming Treatment (D)} & \textbf{No Drumming Treatment (N)} \\ 
-%  \hline
-%  \textbf{Restricted} & \textbf{DR} & \textbf{NR} \\ 
-%  \textbf{Nutrition (R)}& Predict most worker-like & Predict some worker-like \\
-%  & gene expression & gene expression \\
-%  \hline
-%  \textbf{Unrestricted} & \textbf{DU} & \textbf{NU} \\ 
-%  \textbf{Nutrition (U)} & Predict some worker-like & Predict least worker-like \\
-%  & gene expression & gene expression \\
-%  \hline
-% \end{tabular}
+Challenges to honey bee health are a growing concern, in particular the combined, interactive effects of nutritional stress and pathogens (Dolezal and Toth 2018).
+In this study, we used RNA-sequencing to probe mechanisms underlying honey bee responses to two effects, diet quality and infection with the major virus of concern, IAPV.
+In general, we found a major nutritional transcriptomic response, with nearly 2,000 transcripts changing in response to diet quality (rockrose/poor diet versus chestnut/good diet).
+The majority of these genes were upregulated in response to high quality diet, and these genes were enriched for functions (Table \ref{tbl:ChestnutPathways}) such as nutrient signaling (insulin resistance) metabolism, and immune response (Notch signaling and JaK-STAT pathways).
+These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into innate immune responses.
+
+Somewhat surprisingly, the transcriptomic response to virus infection in our experiment was fairly limited.
+We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as additional genes related to transcriptional regulation, and muscle contraction.
+The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figures \ref{fig:pcpRutterVirus}, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, \ref{fig:RutterSM4}).
+
+Given the noisy nature of our data, and our desire to hone in on genes with real expression differences, we compared our data to Galbraith et al.
+(2017), a study that also examined bees’ response to viral infection.
+In contrast to our study, Galbraith et al.
+identified a large number of virus responsive trancripts, and generally had less noise in their data (Figures X, X, X).
+To identify the most reliable virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments.
+We found a large, statistically significant overlap, with 26/38 (68%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al.
+(2017, Figure 1.5).
+This result gives us confidence that, although noisy, we were able to uncover consistent, replicable gene expression responses to virus infection.
+
+** paragraph here to talk about the importance and utility of data visualization in RNA-seq data.  For Lindsay to write.
+
+One of the goals of this study was to use our RNA-seq data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of “resistance” or “resilience”.
+We attempted to address this through specific gene expression contrasts (Table 1.10, Figure 1.18), accompanied by GO analysis of the associated gene lists.
+We an approximately equal number of resistance (n=122) and resilience (n=125) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen infection.
+Resilience candidate genes had functions related to carbohydrate metabolism, chitin metabolism, immune response, and regulation of transcription.
+Resistance candidate genes had functions related to several forms of metabolism (chitin, carbohydrate), regulation of transcription, and cell adhesion.
+
+Overall, these data suggest complex transcriptomic responses to multiple stressors in honey bees.
+Diet has the potential for large and profound effects on transcriptional responses in honey bees, and differences in diet may set up the potential for both resistance and resilience to virus infection.
+
+1) Diet effect seemed larger by number of DEGs than virus effect \\
+
+2) No interaction terms in DEGs \\
+
+3) Elaborate more on the GOs that came out of different contrasts \\
+
+4) Possible preliminary evidence for resilience effects, whereby immune-related genes are turned on by good diet in preparation to defend against viral attacks \\
+
+5) Galbraith data was much cleaner as expected due to 75 percent genetic similarity.
+However, we had significant overlap in the virus-related DEGs.
+Possible avenue for future studies to explore as the intersection DEGs between these two studies may represent those that are reliable (well-controlled study) and relevant (study conducted in more realistic environmental settings).
 
@@
 \begin{table}[H]
-  \includegraphics[width=\textwidth]{Images/CastaneaPathways}
-  \caption{Pathways related to diet main effect Castanea DEGs.}
-  \label{tbl:CastaneaPathways}
+  \includegraphics[width=\textwidth]{Images/ChestnutPathways}
+  \caption{Pathways related to diet main effect Chestnut-upregulated DEGs.}
+  \label{tbl:ChestnutPathways}
 \end{table}
@@
 \begin{table}[H]
-  \includegraphics[width=\textwidth]{Images/CastaneaPathways}
-  \caption{Pathways related to diet main effect Rockrose DEGs.}
+  \includegraphics[width=\textwidth]{Images/ChestnutPathways}
+  \caption{Pathways related to diet main effect Rockrose-upregulated DEGs.}
   \label{tbl:RockrosePathways}
@@
   \includegraphics[width=\textwidth]{../VirusHoneyBee/DESeq2/ClusterStandard/Clustering_data_FDR_05/C_T_4.jpg}
-  \caption{Parallel coordinate plots of DEGs between the virus-infected and control groups of the Galbraith study.}
+  \caption{Parallel coordinate plots of the 1,019 DEGs after hiearchical clustering of size four between the virus-infected and control groups of the Galbraith study.
+Here ``C'' represents control, and ``T'' represents treatment of virus.
+Clusters 1, 3, and 4 seem to represent DEGs that were overexpressed in the virus innoculated group, and Cluster 2 seems to represent DEGs that were overexpressed in the control group.
+In general, the DEGs appeared as expected, but there is rather noticeable deviation of the first replicate from the virus-treated sample (``T.1'') from the other virus-treated replicates in Cluster 2.
+Cluster 4 also has some inconsistent replicates across the virus-treated replicates.}
   \label{fig:pcpGalbraith}
@@
   \includegraphics[width=\textwidth]{../N_V/DESeq2/ClusterStandard/Clustering_data_FDR_05/N_V_4.jpg}
-  \caption{Parallel coordinate plots of DEGs between the virus-infected and control groups of our study.}
+  \caption{Parallel coordinate plots of the 43 DEGs after hiearchical clustering of size four between the virus-infected and control groups of our study.
+Here ``N'' represents non-infected control group, and ``V'' represents treatment of virus.
+We see from this plot that the DEG designations for this dataset do not appear as clean compared to what we saw in the Galbraith dataset in Figure \ref{fig:pcpGalbraith}.}
   \label{fig:pcpRutterVirus}
@@
   \includegraphics[width=\textwidth]{../C_R/DESeq2/ClusterStandard/Clustering_data_FDR_05/C_R_6.jpg}
-  \caption{Parallel coordinate plots of DEGs between the Castanea and Rockrose groups of our study.}
+  \caption{Parallel coordinate plots of the 1,914 DEGs after hiearchical clustering of size six between the Chestnut and Rockrose groups of our study.
+Here ``N'' represents non-infected control group, and ``V'' represents treatment of virus.
+We see from this plot that the DEG designations for this dataset do not appear as clean compared to what we saw in the Galbraith dataset in Figure \ref{fig:pcpGalbraith}.}
   \label{fig:pcpRutterDiet}
@@
   \includegraphics[width=\textwidth]{Images/GRVenn}
-  \caption{Venn diagrams comparing the DEG overlaps of the diet and treatment groups between the Galbraith study and our study.
-Top row left to right: DESeq results for all DEGs, virus DEGs, and control DEGs.
-Bottom row left to right: EdgeR results for all DEGs, virus DEGs, and control DEGs.}
+  \caption{Venn diagrams comparing the virus-related DEG overlaps between the Galbraith study (labeled as ``G'') and our study (labeled as ``R'').
+From left to right: Total virus-related DEGs (subplot A), virus-upregulated DEGs (subplot B), control-upregulated DEGs (subplot C).
+Both the total virus-related and virus-upregulated DEGs showed significant overlap between the studies (p-value < 2.2e-16) as per Fisher's Exact Test for Count Data.
+There was one gene that was virus-upregulated in the Galbraith study but control-upregulated in our study.}
   \label{fig:GRVenn}
@@
   \includegraphics[width=\textwidth]{Images/mdsPlots}
-  \caption{MDS plots constructed from DESeq2 package for the Galbraith dataset (A) and our dataset (B).}
+  \caption{MDS plots constructed from DESeq2 package for the Galbraith dataset for non-infected control ``C'' and virus treated ``T'' samples (A) and our dataset for the non-infected control ``N'' and virus treated ``V'' samples (B).
+the x-axis represents the principal component with the most variation and the y-axis repesents the principal component with the second-most variation.}
   \label{fig:mdsPlots}
@@
   \includegraphics[width=\textwidth]{Images/litreCluster2}
-  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 1 (back in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.}
+  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 1 (originally shown in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.
+``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
+Most of the light orange points (representing the nine combinations of samples between treatment groups for a given DEG} deviate from the \textit{x=y} line in a cluster as expected.
   \label{fig:litreCluster2}
@@
   \includegraphics[width=\textwidth]{Images/litreCluster1}
-  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 2 (back in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.}
+  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 2 (originally shown in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.
+``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
+Most of the dark orange points (representing all combinations of samples between treatment groups for a given DEG} deviate from the \textit{x=y} line in a cluster as expected.
+However, they are not as tightly clustered together compared to what we saw in the example litre plots of Cluster 1 (shown in Figure \ref{fig:litreCluster2}).
+As a result, what we see in these litre plots reflects what we saw in the parallel coordinate lines of Figure \ref{fig:pcpGalbraith}: The replicate consistency in the Cluster 2 DEGs is not as clean as that in the Cluster 1 DEGs, but is still relatively clean.
   \label{fig:litreCluster1}
@@
   \includegraphics[width=\textwidth]{Images/litreClusterRutter}
-  \caption{Example litre plots of the nine DEGs with the lowest FDR values from the 43 DEGs of our dataset.}
+  \caption{Example litre plots of the nine DEGs with the lowest FDR values from the 43 DEGs of our dataset.
+``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+Most of the magenta points (representing the 144 combinations of samples between treatment groups for a given DEG) do not reflect the expected pattern as clearly compared to what we saw in the litre plots of the Galbraith data.
+They are not as clustered together (representing replicate inconsistency) and they sometimes overlap the \textit{x=y} line (representing lack of difference between treatment groups).
+This finding reflects what we saw in the messy looking parallel coordinate lines of Figure \ref{fig:pcpGalbraith}}.
   \label{fig:litreClusterRutter}
@@
   \includegraphics[width=\textwidth]{Images/GalbraithClust1SM}
-  \caption{The DEGs from the first cluster of the Galbraith dataset (back in Figure \ref{fig:pcpGalbraith}) superimposed onto all genes in the form of a scatterplot matrix.
-The data has been standardized.}
+  \caption{The 327 DEGs from the first cluster of the Galbraith dataset (shown in Figure \ref{fig:pcpGalbraith}) superimposed as light orange dots onto all genes as black dots in the form of a scatterplot matrix.
+The data has been standardized.
+``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
+We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots.}
   \label{fig:GalbraithClust1SM}
@@
   \includegraphics[width=\textwidth]{Images/GalbraithClust2SM}
-  \caption{The DEGs from the second cluster of the Galbraith dataset (back in Figure \ref{fig:pcpGalbraith}) superimposed onto all genes in the form of a scatterplot matrix.
-The data has been standardized.}
+  \caption{The 365 DEGs from the second cluster of the Galbraith dataset (shown in Figure \ref{fig:pcpGalbraith}) superimposed as dark orange dots onto all genes as black dots in the form of a scatterplot matrix.
+The data has been standardized.
+``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
+We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots.
+We also see again that the first replicate from the virus-treated sample (``T.1'') may be somewhat inconsistent in these DEGs, as its presence in the replicate scatterplots results in the DEGs unexpectedly deviating from the \textit{x=y} line and its presence in the treatment scatterplots results in the DEGs unexpectedly adhering to the \textit{x=y} line.}
   \label{fig:GalbraithClust2SM}
@@
   \includegraphics[width=\textwidth]{Images/GalbraithClust3SM}
-  \caption{The DEGs from the third cluster of the Galbraith dataset (back in Figure \ref{fig:pcpGalbraith}) superimposed onto all genes in the form of a scatterplot matrix.
-The data has been standardized.}
+  \caption{The 224 DEGs from the third cluster of the Galbraith dataset (shown in Figure \ref{fig:pcpGalbraith}) superimposed as turquoise dots onto all genes as black dots in the form of a scatterplot matrix.
+The data has been standardized.
+``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
+We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots.}
   \label{fig:GalbraithClust3SM}
@@
   \includegraphics[width=\textwidth]{Images/GalbraithClust4SM}
-  \caption{The DEGs from the fourth cluster of the Galbraith dataset (back in Figure \ref{fig:pcpGalbraith}) superimposed onto all genes in the form of a scatterplot matrix.
-The data has been standardized.}
+  \caption{The 103 DEGs from the fourth cluster of the Galbraith dataset (shown in Figure \ref{fig:pcpGalbraith}) superimposed as pink dots onto all genes as black dots in the form of a scatterplot matrix.
+The data has been standardized.
+``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
+We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots.
+We also see that the second replicate from the virus-treated sample (``T.2'') may be somewhat inconsistent in these DEGs, as its presence in the replicate scatterplots results in the DEGs unexpectedly deviating from the \textit{x=y} line and its presence in the treatment scatterplots results in the DEGs unexpectedly adhering to the \textit{x=y} line.}
   \label{fig:GalbraithClust4SM}
@@
 Only replicates 1, 2, and 3 are shown from both treatment groups.
-The data has been standardized.}
+The data has been standardized.
+``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+We see that, compared to the scatterplot matrices from the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, somtimes unexpectedly deviating from the \textit{x=y} line in the replicate plots and sometimes unexpectedly adhering to the \textit{x=y} line in the treatment plots.}
   \label{fig:RutterSM1}
@@
 Only replicates 4, 5, and 6 are shown from both treatment groups.
-The data has been standardized.}
+The data has been standardized.
+``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+We see that, compared to the scatterplot matrices from the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, somtimes unexpectedly deviating from the \textit{x=y} line in the replicate plots and sometimes unexpectedly adhering to the \textit{x=y} line in the treatment plots.}
   \label{fig:RutterSM2}
@@
 Only replicates 7, 8, and 9 are shown from both treatment groups.
-The data has been standardized.}
+The data has been standardized.
+``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+We see that, compared to the scatterplot matrices from the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, somtimes unexpectedly deviating from the \textit{x=y} line in the replicate plots and sometimes unexpectedly adhering to the \textit{x=y} line in the treatment plots.}
   \label{fig:RutterSM3}
@@
 Only replicates 10, 11, and 12 are shown from both treatment groups.
-The data has been standardized.}
+The data has been standardized.
+``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+We see that, compared to the scatterplot matrices from the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, somtimes unexpectedly deviating from the \textit{x=y} line in the replicate plots and sometimes unexpectedly adhering to the \textit{x=y} line in the treatment plots.}
   \label{fig:RutterSM4}
@@
   \includegraphics[width=\textwidth]{Images/NCNRNCPathways}
-  \caption{Pathways analysis results for the  601 DEGs that were upregulated in the NC treatment in the NC versus NR treatment pair analysis.}
+  \caption{GO analysis results for the 601 DEGs that were upregulated in the NC treatment in the NC versus NR treatment pair analysis.
+These DEGs represent genes that are upregulated when non-infected honeybees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.}
   \label{tbl:NCNRNCPathways}
@@
   \includegraphics[width=\textwidth]{Images/NCNRNRPathways}
-  \caption{Pathways analysis results for the 340 DEGs that were upregulated in the NR treatment in the NC versus NR treatment pair analysis.}
+  \caption{GO analysis results for the 340 DEGs that were upregulated in the NR treatment in the NC versus NR treatment pair analysis.
+These DEGs represent genes that are upregulated when non-infected honeybees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.}
   \label{tbl:NCNRNRPathways}
@@
   \includegraphics[width=\textwidth]{Images/VCVRVCPathways}
-  \caption{Pathways analysis results for the 247 DEGs that were upregulated in the VC treatment in the VC versus VR treatment pair analysis.}
+  \caption{GO analysis results for the 247 DEGs that were upregulated in the VC treatment in the VC versus VR treatment pair analysis.
+These DEGs represent genes that are upregulated when infected honeybees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.}
   \label{tbl:VCVRVCPathways}
@@
   \includegraphics[width=\textwidth]{Images/VCVRVRPathways}
-  \caption{Pathways analysis results for the 129 DEGs that were upregulated in the VR treatment in the VC versus VR treatment pair analysis.}
+  \caption{GO analysis results for the 129 DEGs that were upregulated in the VR treatment in the VC versus VR treatment pair analysis.
+These DEGs represent genes that are upregulated when infected honeybees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.}
   \label{tbl:VCVRVRPathways}
@@
   \includegraphics[width=0.8\textwidth]{Images/revigo}
-  \caption{Gene ontology results for the 122 DEGs related to our ``resilience" hypothesis (A) and for the 125 DEGs related to our ``resistance" hypothesis (B).}
+  \caption{GO analysis results for the 122 DEGs related to our ``resilience" hypothesis (A) and for the 125 DEGs related to our ``resistance" hypothesis (B).}
   \label{fig:revigo}
@@
 \hline
-VC upregulated in VC vs VR overlaped with NC upregulated in NC vs NR & 122 & ``Resilience'' genes that are turned on by good diet regardless of virus infection status in bees & Figure \ref{fig:revigo}A \\
+VC upregulated in VC vs VR overlapped with NC upregulated in NC vs NR & 122 & ``Resilience'' genes that are turned on by good diet regardless of virus infection status in bees & Figure \ref{fig:revigo}A \\
 \hline

### 2018-04-21 Overlaps/Overlaps.R

@@ -86,6 +86,25 @@ namesG = unname(sapply(dataG$id, function(x) strsplit(strsplit(x, "[|]")[[1]][3]
 N = intersect(namesG, namesR)
 N = length(N)
 
+
+###### Test overlap
+
+# p-value < 2.2e-16, odds ratio 20.10211
+int = length(intersect(GD_TOTAL, RD_VIRUS_TOTAL))
+GRTotal <- matrix(c(N, length(GD_TOTAL)-int, length(RD_VIRUS_TOTAL)-int, int), nrow = 2)
+fisher.test(GRTotal, alternative = "greater")
+
+# p-value < 2.2e-16, odds ratio 33.3018
+int = length(intersect(GDV, RDV))
+GRVirus <- matrix(c(N, length(GDV)-int, length(RDV)-int, int), nrow = 2)
+fisher.test(GRVirus, alternative = "greater")
+
+# p-value < 1, odds ratio 0
+int = length(intersect(GDC, RDN))
+GRControl <- matrix(c(N, length(GDC)-int, length(RDN)-int, int), nrow = 2)
+fisher.test(GRControl, alternative = "greater")
+
+
 ############# Venn diagrams for Galbraith (Total) ############# 
 
 int_GD_GE_TOTAL = intersect(GD_TOTAL, GE_TOTAL)

### 2018-04-21 Manuscript/chapter4.tex
Incorporate discussion section suggestions from Dr. Toth

@@
 We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as additional genes related to transcriptional regulation, and muscle contraction.
-The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figures \ref{fig:pcpRutterVirus}, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, \ref{fig:RutterSM4}).
+The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figures \ref{fig:mdsPlots}B, \ref{fig:pcpRutterVirus}, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, \ref{fig:RutterSM4}).
 
-Given the noisy nature of our data, and our desire to hone in on genes with real expression differences, we compared our data to Galbraith et al.
-(2017), a study that also examined bees’ response to viral infection.
+Given the noisy nature of our data, and our desire to hone in on genes with real expression differences, we compared our data to the Galbraith study (\citealt{galbraith}), which also examined bees response to viral infection.
 In contrast to our study, Galbraith et al.
-identified a large number of virus responsive trancripts, and generally had less noise in their data (Figures X, X, X).
+identified a large number of virus responsive trancripts, and generally had less noise in their data (Figures \ref{fig:mdsPlots}A, \ref{fig:pcpGalbraith}, \ref{fig:litreCluster2}, \ref{fig:litreCluster1}, \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, \ref{fig:GalbraithClust4SM}).
 To identify the most reliable virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments.
-We found a large, statistically significant overlap, with 26/38 (68%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al.
-(2017, Figure 1.5).
-This result gives us confidence that, although noisy, we were able to uncover consistent, replicable gene expression responses to virus infection.
+We found a large, statistically significant (p-value < 2.2e-16) overlap, with 26/38 (68\%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al.
+(Figure \ref{fig:GRVenn}).
+This result gives us confidence that, although noisy, we were able to uncover consistent, replicable gene expression responses to virus infection with our data.
 
@@
 
-One of the goals of this study was to use our RNA-seq data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of “resistance” or “resilience”.
-We attempted to address this through specific gene expression contrasts (Table 1.10, Figure 1.18), accompanied by GO analysis of the associated gene lists.
-We an approximately equal number of resistance (n=122) and resilience (n=125) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen infection.
+One of the goals of this study was to use our RNA-seq data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``resilience''.
+We attempted to address this question through specific gene expression contrasts (Table \ref{tbl:contrasts}, Figure \ref{tbl:NCNRNCPathways}), accompanied by GO analysis of the associated gene lists.
+We found an approximately equal number of resistance (n = 122) and resilience (n = 125) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen infection.
 Resilience candidate genes had functions related to carbohydrate metabolism, chitin metabolism, immune response, and regulation of transcription.
-Resistance candidate genes had functions related to several forms of metabolism (chitin, carbohydrate), regulation of transcription, and cell adhesion.
+Resistance candidate genes had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion.
 
@@
 Diet has the potential for large and profound effects on transcriptional responses in honey bees, and differences in diet may set up the potential for both resistance and resilience to virus infection.
-
-1) Diet effect seemed larger by number of DEGs than virus effect \\
-
-2) No interaction terms in DEGs \\
-
-3) Elaborate more on the GOs that came out of different contrasts \\
-
-4) Possible preliminary evidence for resilience effects, whereby immune-related genes are turned on by good diet in preparation to defend against viral attacks \\
-
-5) Galbraith data was much cleaner as expected due to 75 percent genetic similarity.
-However, we had significant overlap in the virus-related DEGs.
-Possible avenue for future studies to explore as the intersection DEGs between these two studies may represent those that are reliable (well-controlled study) and relevant (study conducted in more realistic environmental settings).
 

### 2018-04-21 Manuscript/chapter4.tex
Final draft for thesis

@@
 We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as additional genes related to transcriptional regulation, and muscle contraction.
-The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figures \ref{fig:mdsPlots}B, \ref{fig:pcpRutterVirus}, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, \ref{fig:RutterSM4}).
+The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figures \ref{fig:mdsPlots}B, \ref{fig:pcpRutterVirus}, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).
 
@@
 In contrast to our study, Galbraith et al.
-identified a large number of virus responsive trancripts, and generally had less noise in their data (Figures \ref{fig:mdsPlots}A, \ref{fig:pcpGalbraith}, \ref{fig:litreCluster2}, \ref{fig:litreCluster1}, \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, \ref{fig:GalbraithClust4SM}).
+identified a large number of virus responsive trancripts, and generally had less noise in their data (Figures \ref{fig:mdsPlots}A, \ref{fig:pcpGalbraith}, \ref{fig:litreCluster2}, \ref{fig:litreCluster1}, \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}).
 To identify the most reliable virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments.
@@
 
-** paragraph here to talk about the importance and utility of data visualization in RNA-seq data.  For Lindsay to write.
+The DESeq2 package comes with certain visualization options that are popular in RNA-sequencing analysis.
+One of these visualization is the multidimensional scaling (MDS) plot, which allows users to visualize the similarity between samples within a dataset.
+We could determine from this plot that indeed the Galbraith data may show more similarity between its replicates and differences between its treatments compared to our data (Figure \ref{fig:mdsPlots}).
+However, the MDS plot only shows us information at the sample level.
+We wanted to investigate how these differences in the signal:to:noise ratios of the datasets would affect the structure of any resulting DEGs.
+As a result, we also used three plotting techniques from the bigPint package: We investigated the 1,019 virus-related DEGs from the Galbraith dataset and the 43 virus-related DEGs from our dataset using parallel coordinate lines, litre plots, and scatterplot matrices.
+To prevent overlapping issues in our plots, we used a hierarchical clustering technique for the parallel coordinate lines to separate the set of DEGs into smaller groups.
+We also needed to examine four subsets of samples from the Galbraith dataset to make effective use of the scatterplot matrices.
+After these tailorizations, we determined that the same patterns we saw in the MDS plots regarding the entire dataset extended down the pipeline analysis into the DEG calls: Even the DEGs from the Galbraith datasaet showed more similarity between their replicates and differences between their treatments compared to those from our data.
+However, the 365 DEGs from the Galbraith data in Cluster 2 of Figure \ref{fig:pcpGalbraith} showed an inconsistent first replicate in the treatment group (``T.1''), which was something we observed in the MDS plot.
+This indicates that this feature also extended down the analysis pipeline into DEG calls.
 
@@
 Diet has the potential for large and profound effects on transcriptional responses in honey bees, and differences in diet may set up the potential for both resistance and resilience to virus infection.
-
-\section{Appendix}
+Moreover, this study in general also demonstrated the possible benefits of examining multiple datasets to address inherently messy biological data.
+For instance, by verifying the subsantial overlap in our DEG lists to those obtained in another study that addressed a similar question but in a more controlled manner, we were able to feel more confident in our DEG calls.
 

### 2018-04-21 Manuscript/chapter4.tex
Final changes before addition to thesis

@@
 Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}).
-We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show a significant interaction between diet quality and virus infection.
+We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show the importance of the combined effects of both diet quality and virus infection.
 Specifically, high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen (\citealt{adamInt}).
@@
 For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (resilience).
-Transcriptomics is one means to achieve this goal.
+Transcriptomics is one means to better understand the mechanistic underpinnings of dietary and viral effects on honey bee health.
 Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection.
@@
 As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees.
-As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honeybee gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense.
+As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honeybee gene expression patterns related to the combined effects of diet in any broad sense and viral innoculation in any broad sense.
 
@@
 We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section).
-We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the interactive effect of the two factors on gene expression patterns.
+We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.
 
@@
 
-Consequently, for our nutrition factor, we examined two monofloral pollen diets, Rockrose (Cistus) and Chestnut (Chestnut).
+Consequently, for our nutrition factor, we examined two monofloral pollen diets, Rockrose (Cistus) and Castanea (Chestnut).
 Rockrose pollen is generally considered less nutritious than Chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}).
@@
 
-\subsection{Detecting resilience versus resistance}
+\begin{figure}[H]
+  \includegraphics[width=0.8\textwidth]{Images/revigo}
+  \caption{GO analysis results for the 122 DEGs related to our ``resilience" hypothesis (A) and for the 125 DEGs related to our ``resistance" hypothesis (B).}
+  \label{fig:revigo}
+\end{figure}
+
+\subsection{Probing resilience versus resistance}
 
@@
 We then determined how many genes fell into these two categories and analyzed their GO terminologies.
+
+\begin{table}[H]
+\begin{tabular}{ | m{4.4cm} | m{0.9cm}| m{5.5cm} | m{2.6cm} | } 
+\hline
+Contrast & DEGs & Interpretation & Results \\ 
+\hline
+V vs N & 43 & Genes that change expression due to virus effect regardless of diet status in bees & Table \ref{tbl:virusGenes} \\ 
+\hline
+NC vs NR & 941 & Genes that change expression due to diet effect in uninfected bees & Supplementary tables \ref{tbl:NCNRNCPathways} and \ref{tbl:NCNRNRPathways} \\ 
+\hline
+VC vs VR & 376 & Genes that change expression due to diet effect in infected bees & Supplementary tables \ref{tbl:VCVRVCPathways} and \ref{tbl:VCVRVRPathways} \\ 
+\hline
+VC upregulated in VC vs VR overlapped with NC upregulated in NC vs NR & 122 & ``Resilience'' genes that are turned on by good diet regardless of virus infection status in bees & Figure \ref{fig:revigo}A \\
+\hline
+VC upregulated in VC vs VR but NC is not upregulated in NC vs NR & 125 & ``Resistance'' genes that are turned on by good diet only in infected bees & Figure \ref{fig:revigo}B \\
+\hline
+\end{tabular}
+\caption{Contrasts in our study for assessing GO and pathways analysis.}
+  \label{tbl:contrasts}
+\end{table}
 
@@
 In comparing mortality levels based on pairwise comparisons, we found that bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls.
-However, bees fed Castanea pollen had no significant difference in mortality between virus infected and control groups.
-These results suggest the high quality Castanea diet can ``rescue'' virus induced mortality (Figure \ref{fig:mortality}A).
+However, bees fed Chestnut pollen had no significant difference in mortality between virus infected and control groups.
+These results suggest the high quality Chestnut diet can ``rescue'' virus induced mortality (Figure \ref{fig:mortality}A).
 
-IAPV titer volumes of honeybees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df=3, 34; F=6.096; p<0.00196).
+IAPV titers of honeybees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df=3, 34; F=6.096; p<0.00196).
 The effect of virus treatment (mixed model ANOVA, df=1, 34; F=15.686; p<0.000362) was significant, but the diet treatment (mixed model ANOVA, df=1, 34; F=1.898; p>0.05) and the interaction between the two factors (mixed model ANOVA, df=1, 34; F=0.702, p>0.05) were not significant.
 Honeybees that were infected with the virus and fed a poor-quality Rockrose diet showed significant increases in IAPV titer volumes compared to honeybees that were not infected with the virus regardless of their diet quality (Tukey HSD, p<0.05).
-Overall, we interpreted this effect to mean that Rockrose pollen could not ``rescue'' high virus titers resulting from the inoculation treatment, whereas Castanea pollen could (Figure \ref{fig:mortality}B).
+Overall, we interpreted this effect to mean that Rockrose pollen could not ``rescue'' high virus titers resulting from the inoculation treatment, whereas Chestnut pollen could (Figure \ref{fig:mortality}B).
 
@@
 
-We observed a substantially larger number of DEGs in our diet main effect (n = 1914) than in our virus main effect (n = 43) (Table \ref{tbl:mainEffectDEGs}A and B).
+We observed a substantially larger number of DEGs in our diet main effect (n = 1914) than in our virus main effect (n = 43) (Supplementary table \ref{tbl:mainEffectDEGs}A and B).
 In the diet factor, there were more Chestnut-upregulated DEGs (n = 1033) than Rockrose-upregulated DEGs (n = 881).
 In the virus factor, there were more virus-upregulated DEGs (n = 38) than control-upregulated DEGs (n = 5).
-While these reported DEGs numbers are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Table \ref{tbl:mainEffectDEGs}A and B).
+While these reported DEGs numbers are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table \ref{tbl:mainEffectDEGs}A and B).
 
-GO analysis of the Chestnut-upregulated DEGs revealed the following enriched categories (Benjamini correction < 0.05): Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Table \ref{tbl:ChestnutPathways}).
-GO analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Table \ref{tbl:RockrosePathways}).
+GO analysis of the Chestnut-upregulated DEGs revealed the following enriched categories (Benjamini correction < 0.05): Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table \ref{tbl:ChestnutPathways}).
+GO analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table \ref{tbl:RockrosePathways}).
 
@@
 
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/virusGenes}
+  \caption{Known functions of the mapped subset of 43 DEGs in the virus main effect of our study.
+Whether the gene was overrepresented in the virus or non-virus group is also indicated for both our study and the Galbraith study.
+Functionalities were extracted from Flybase, National Center for Biotechnology Information, and The European Bioinformatics Institute databases.}
+  \label{tbl:virusGenes}
+\end{table}
+
 No interaction DEGs were observed between the diet and virus factors of the study, in any of the pipelines (DESeq2, edgeR, limma).
@@
 
-The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 941 (Table \ref{tbl:pairDEGs}).
+The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 941 (Supplementary table \ref{tbl:pairDEGs}).
 Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level.
-Across every pair comparing the Chestnut and Rockrose levels, regardless of the virus level, the number of Chestnut-upregulated DEGs was higher than the number of Rockrose-upregulated DEGs (Table \ref{tbl:pairDEGs} C, D, E, F).
-For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Table \ref{tbl:pairDEGs} A, B).
-There were no DEGs between the treatment pair controlling for the control level of the virus effect (Table \ref{tbl:pairDEGs} A).
+Across every pair comparing the Chestnut and Rockrose levels, regardless of the virus level, the number of Chestnut-upregulated DEGs was higher than the number of Rockrose-upregulated DEGs (Supplementary table \ref{tbl:pairDEGs} C, D, E, F).
+For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Supplementary table \ref{tbl:pairDEGs} A, B).
+There were no DEGs between the treatment pair controlling for the control level of the virus effect (Supplementary table \ref{tbl:pairDEGs} A).
 These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).
@@
 We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset.
-Basic MDS plots were constructed with the DESeq2 analysis pipeline, and we could immediately determine that the Galbraith dataset may better separate the infected and uninfected honeybees better than our dataset (Figure \ref{fig:mdsPlots}).
+Basic MDS plots were constructed with the DESeq2 analysis pipeline, and we could immediately determine that the Galbraith dataset may better separate the infected and uninfected honeybees better than our dataset (Supplementary figure \ref{fig:mdsPlots}).
 We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the MDS plots.
@@
 The replicates appear somewhat inconsistent in their esimated expression levels and there is not always such a large difference between treatment groups.
-We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study.
+We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Supplementary figure \ref{fig:pcpRutterDiet}).
 
-We also used litre plots to examine the structure of individual DEGs: We see that indeed the individual DEGs from our data (Figure \ref{fig:litreClusterRutter}) show less consistent replications and less differences between the treatment groups compared to the individual DEGs from the Galbraith data (Figure \ref{fig:litreCluster1} and Figure \ref{fig:litreCluster2}).
-For the Galbraith data, we examined individual DEGs from the first cluster (Figure \ref{fig:litreCluster1}) and second cluster (Figure \ref{fig:litreCluster2}) because the second cluster was a bit less ideal due to its inconsistent first replicate of the treatment group.
+\begin{figure}[H]
+\centering
+% ../VirusHoneyBee/DESeq2/ClusterStandard/Clustering_data_FDR_05/
+  \includegraphics[width=\textwidth]{Images/C_T_4.jpg}
+  \caption{Parallel coordinate plots of the 1,019 DEGs after hiearchical clustering of size four between the virus-infected and control groups of the Galbraith study.
+Here ``C'' represents control, and ``T'' represents treatment of virus.
+Clusters 1, 3, and 4 seem to represent DEGs that were overexpressed in the virus innoculated group, and Cluster 2 seems to represent DEGs that were overexpressed in the control group.
+In general, the DEGs appeared as expected, but there is rather noticeable deviation of the first replicate from the virus-treated sample (``T.1'') from the other virus-treated replicates in Cluster 2.
+Cluster 4 also has some inconsistent replicates across the virus-treated replicates.}
+  \label{fig:pcpGalbraith}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+% /../N_V/DESeq2/ClusterStandard/Clustering_data_FDR_05
+  \includegraphics[width=\textwidth]{Images/N_V_4.jpg}
+  \caption{Parallel coordinate plots of the 43 DEGs after hiearchical clustering of size four between the virus-infected and control groups of our study.
+Here ``N'' represents non-infected control group, and ``V'' represents treatment of virus.
+We see from this plot that the DEG designations for this dataset do not appear as clean compared to what we saw in the Galbraith dataset in Figure \ref{fig:pcpGalbraith}.}
+  \label{fig:pcpRutterVirus}
+\end{figure}
+
+We also used litre plots to examine the structure of individual DEGs: We see that indeed the individual DEGs from our data (Supplementary figure \ref{fig:litreClusterRutter}) show less consistent replications and less differences between the treatment groups compared to the individual DEGs from the Galbraith data (Supplementary figures \ref{fig:litreCluster1} and \ref{fig:litreCluster2}).
+For the Galbraith data, we examined individual DEGs from the first cluster (Supplementary figure \ref{fig:litreCluster1}) and second cluster (Supplementary figure \ref{fig:litreCluster2}) because the second cluster was a bit less ideal due to its inconsistent first replicate of the treatment group.
 
 Finally, we looked at scatterplot matrices to assess the DEGs.
-We created standardized scatterplot matrices for each of the four clusters (Figure \ref{fig:pcpGalbraith}) of the Galbriath data (Figures \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}).
+We created standardized scatterplot matrices for each of the four clusters (Figure \ref{fig:pcpGalbraith}) of the Galbriath data (Supplementary figures \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}).
 We also created standardized scatterplot matrices for our data.
 However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data.
-As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Figures \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).
+As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Supplementary figures \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).
 We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: Deviating more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.
 
-Despite the DEGs from the Galbraith dataset displaying the expected patterns more than those from our dataset, there was significant overlap in the DEGs between the two studies (Figure \ref{fig:GRVenn}).
+Despite the virus-related DEGs (n = 1,019) from the Galbraith dataset displaying the expected patterns more than those from our dataset (n = 43), there was significant overlap (p-value < 2.2e-16) in the DEGs between the two studies (Supplementary figure \ref{fig:GRVenn}).
 
@@
 
-Within our ``resilience" gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism).
+Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 `resilience'' candidate genes and 125 ``resistance'' candidate genes.
+Within our 122 ``resilience" gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism).
 However, we also discovered gene ontologies related to RNA polyerase II transcription and immune response (Figure \ref{fig:revigo}A).
-Within our ``resistance" gene ontologies, we found functions related to metabolism (such as carboyhydrate metabolism, chitin metabolism, and general metabolism).
-(Figure \ref{fig:revigo}B).
+Within our 125 ``resistance" gene ontologies, we found functions related to metabolism (such as carboyhydrate metabolism, chitin metabolism, and general metabolism) (Figure \ref{fig:revigo}B).
 
@@
 In general, we found a major nutritional transcriptomic response, with nearly 2,000 transcripts changing in response to diet quality (rockrose/poor diet versus chestnut/good diet).
-The majority of these genes were upregulated in response to high quality diet, and these genes were enriched for functions (Table \ref{tbl:ChestnutPathways}) such as nutrient signaling (insulin resistance) metabolism, and immune response (Notch signaling and JaK-STAT pathways).
+The majority of these genes were upregulated in response to high quality diet, and these genes were enriched for functions (Supplmentary table \ref{tbl:ChestnutPathways}) such as nutrient signaling (insulin resistance) metabolism, and immune response (Notch signaling and JaK-STAT pathways).
 These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into innate immune responses.
@@
 We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as additional genes related to transcriptional regulation, and muscle contraction.
-The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figures \ref{fig:mdsPlots}B, \ref{fig:pcpRutterVirus}, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).
+The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure \ref{fig:pcpRutterVirus} and Supplementary figures \ref{fig:mdsPlots}B, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).
 
@@
 In contrast to our study, Galbraith et al.
-identified a large number of virus responsive trancripts, and generally had less noise in their data (Figures \ref{fig:mdsPlots}A, \ref{fig:pcpGalbraith}, \ref{fig:litreCluster2}, \ref{fig:litreCluster1}, \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}).
+identified a large number of virus responsive trancripts, and generally had less noise in their data (Figure \ref{fig:pcpGalbraith} and Supplementary figures \ref{fig:mdsPlots}A, \ref{fig:litreCluster1}, \ref{fig:litreCluster2}, \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}).
 To identify the most reliable virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments.
 We found a large, statistically significant (p-value < 2.2e-16) overlap, with 26/38 (68\%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al.
-(Figure \ref{fig:GRVenn}).
+(Supplementary figure \ref{fig:GRVenn}).
 This result gives us confidence that, although noisy, we were able to uncover consistent, replicable gene expression responses to virus infection with our data.
 
-The DESeq2 package comes with certain visualization options that are popular in RNA-sequencing analysis.
+Data visualization is a useful method to identify noise and robustness in RNA-seq data.
+In this study, we used extensive data visualization to improve the interpretation of our RNA-seq results.
+For example, the DESeq2 package comes with certain visualization options that are popular in RNA-sequencing analysis.
 One of these visualization is the multidimensional scaling (MDS) plot, which allows users to visualize the similarity between samples within a dataset.
-We could determine from this plot that indeed the Galbraith data may show more similarity between its replicates and differences between its treatments compared to our data (Figure \ref{fig:mdsPlots}).
+We could determine from this plot that indeed the Galbraith data may show more similarity between its replicates and differences between its treatments compared to our data (Supplementary figure \ref{fig:mdsPlots}).
 However, the MDS plot only shows us information at the sample level.
@@
 This indicates that this feature also extended down the analysis pipeline into DEG calls.
+We believe these visualization applications can be useful for future researchers analyzing RNAs-sequencing data to quickly and effectively ensure that the DEG calls look reliable or at least overlap with DEG calls from similar studies that look reliable.
+We believe this type of visualization exploration can be especially crucial when studying complex organisms that do not have genetic identicalness or similarity between replicates and/or when using experiments that may lack rigid design control.
 
 One of the goals of this study was to use our RNA-seq data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``resilience''.
-We attempted to address this question through specific gene expression contrasts (Table \ref{tbl:contrasts}, Figure \ref{tbl:NCNRNCPathways}), accompanied by GO analysis of the associated gene lists.
-We found an approximately equal number of resistance (n = 122) and resilience (n = 125) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen infection.
+We attempted to address this question through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists.
+We found an approximately equal number of resistance (n = 125) and resilience (n = 122) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen infection.
 Resilience candidate genes had functions related to carbohydrate metabolism, chitin metabolism, immune response, and regulation of transcription.
@@
 Diet has the potential for large and profound effects on transcriptional responses in honey bees, and differences in diet may set up the potential for both resistance and resilience to virus infection.
-Moreover, this study in general also demonstrated the possible benefits of examining multiple datasets to address inherently messy biological data.
-For instance, by verifying the subsantial overlap in our DEG lists to those obtained in another study that addressed a similar question but in a more controlled manner, we were able to feel more confident in our DEG calls.
+Moreover, this study in general also demonstrated the possible benefits of using data visualizations and multiple datasets to address inherently messy biological data.
+For instance, by verifying the subsantial overlap in our DEG lists to those obtained in another study that addressed a similar question but in a more controlled manner, we were able to place much higher confidence in the differential gene expression results from our otherwise noisy data.
+We hope these results underline the need for researchers to use data visualization techniques to understand and interpret RNA-sequencing datasets.
+
+\section{Appendix}
 
@@
   \includegraphics[width=0.65\textwidth]{Images/mainEffectDEGs}
-  \caption{Number of DEGs across three analysis pipelines for (A) the diet effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study.}
+  \caption{Number of DEGs across three analysis pipelines for (A) the diet effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study.
+For the diet effects, ``C'' represents Chestnut diet and ``R'' represents Rockrose diet.
+For the virus effects, ``V'' represents virus-innoculated and ``C'' represents control non-innoculated.
+Green cells represent the level that showed a larger number of DEGs.}
   \label{tbl:mainEffectDEGs}
@@
 \begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/NCNRNCPathways}
+  \caption{GO analysis results for the 601 DEGs that were upregulated in the NC treatment in the NC versus NR treatment pair analysis.
+These DEGs represent genes that are upregulated when non-infected honeybees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.}
+  \label{tbl:NCNRNCPathways}
+\end{table}
+
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/NCNRNRPathways}
+  \caption{GO analysis results for the 340 DEGs that were upregulated in the NR treatment in the NC versus NR treatment pair analysis.
+These DEGs represent genes that are upregulated when non-infected honeybees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.}
+  \label{tbl:NCNRNRPathways}
+\end{table}
+
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/VCVRVCPathways}
+  \caption{GO analysis results for the 247 DEGs that were upregulated in the VC treatment in the VC versus VR treatment pair analysis.
+These DEGs represent genes that are upregulated when infected honeybees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.}
+  \label{tbl:VCVRVCPathways}
+\end{table}
+
+\begin{table}[H]
+  \includegraphics[width=\textwidth]{Images/VCVRVRPathways}
+  \caption{GO analysis results for the 129 DEGs that were upregulated in the VR treatment in the VC versus VR treatment pair analysis.
+These DEGs represent genes that are upregulated when infected honeybees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.}
+  \label{tbl:VCVRVRPathways}
+\end{table}
+
+\begin{table}[H]
   \includegraphics[width=0.65\textwidth]{Images/pairDEGs}
-  \caption{Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor.}
+  \caption{Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor.
+``C'' represents Chestnut diet, ``R'' represents Rockrose diet, ``V'' represents virus-innoculated, and ``N'' represents control non-innoculated.
+Green cells represent the level that showed a larger number of DEGs.}
   \label{tbl:pairDEGs}
@@
 \begin{figure}[H]
-\centering
-  \includegraphics[width=\textwidth]{../VirusHoneyBee/DESeq2/ClusterStandard/Clustering_data_FDR_05/C_T_4.jpg}
-  \caption{Parallel coordinate plots of the 1,019 DEGs after hiearchical clustering of size four between the virus-infected and control groups of the Galbraith study.
-Here ``C'' represents control, and ``T'' represents treatment of virus.
-Clusters 1, 3, and 4 seem to represent DEGs that were overexpressed in the virus innoculated group, and Cluster 2 seems to represent DEGs that were overexpressed in the control group.
-In general, the DEGs appeared as expected, but there is rather noticeable deviation of the first replicate from the virus-treated sample (``T.1'') from the other virus-treated replicates in Cluster 2.
-Cluster 4 also has some inconsistent replicates across the virus-treated replicates.}
-  \label{fig:pcpGalbraith}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-  \includegraphics[width=\textwidth]{../N_V/DESeq2/ClusterStandard/Clustering_data_FDR_05/N_V_4.jpg}
-  \caption{Parallel coordinate plots of the 43 DEGs after hiearchical clustering of size four between the virus-infected and control groups of our study.
-Here ``N'' represents non-infected control group, and ``V'' represents treatment of virus.
-We see from this plot that the DEG designations for this dataset do not appear as clean compared to what we saw in the Galbraith dataset in Figure \ref{fig:pcpGalbraith}.}
-  \label{fig:pcpRutterVirus}
+  \includegraphics[width=\textwidth]{Images/mdsPlots}
+  \caption{MDS plots constructed from DESeq2 package for the Galbraith dataset for non-infected control ``C'' and virus treated ``T'' samples (A) and our dataset for the non-infected control ``N'' and virus treated ``V'' samples (B).
+the x-axis represents the principal component with the most variation and the y-axis repesents the principal component with the second-most variation.}
+  \label{fig:mdsPlots}
 \end{figure}
@@
 \begin{figure}[H]
-  \includegraphics[width=\textwidth]{Images/GRVenn}
-  \caption{Venn diagrams comparing the virus-related DEG overlaps between the Galbraith study (labeled as ``G'') and our study (labeled as ``R'').
-From left to right: Total virus-related DEGs (subplot A), virus-upregulated DEGs (subplot B), control-upregulated DEGs (subplot C).
-Both the total virus-related and virus-upregulated DEGs showed significant overlap between the studies (p-value < 2.2e-16) as per Fisher's Exact Test for Count Data.
-There was one gene that was virus-upregulated in the Galbraith study but control-upregulated in our study.}
-  \label{fig:GRVenn}
-\end{figure}
-
-\begin{table}[H]
-  \includegraphics[width=\textwidth]{Images/virusGenes}
-  \caption{Known functions of the mapped subset of 43 DEGs in the virus main effect of our study.
-Whether the gene was overrepresented in the virus or non-virus group is also indicated for both our study and the Galbraith study.
-Functionalities were extracted from Flybase, National Center for Biotechnology Information, and The European Bioinformatics Institute databases.}
-  \label{tbl:virusGenes}
-\end{table}
-
-\begin{figure}[H]
-  \includegraphics[width=\textwidth]{Images/mdsPlots}
-  \caption{MDS plots constructed from DESeq2 package for the Galbraith dataset for non-infected control ``C'' and virus treated ``T'' samples (A) and our dataset for the non-infected control ``N'' and virus treated ``V'' samples (B).
-the x-axis represents the principal component with the most variation and the y-axis repesents the principal component with the second-most variation.}
-  \label{fig:mdsPlots}
-\end{figure}
-
-\begin{figure}[H]
-  \includegraphics[width=\textwidth]{Images/litreCluster2}
-  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 1 (originally shown in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.
-``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
-Most of the light orange points (representing the nine combinations of samples between treatment groups for a given DEG} deviate from the \textit{x=y} line in a cluster as expected.
-  \label{fig:litreCluster2}
-\end{figure}
-
-\begin{figure}[H]
-  \includegraphics[width=\textwidth]{Images/litreCluster1}
-  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 2 (originally shown in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.
-``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
-Most of the dark orange points (representing all combinations of samples between treatment groups for a given DEG} deviate from the \textit{x=y} line in a cluster as expected.
-However, they are not as tightly clustered together compared to what we saw in the example litre plots of Cluster 1 (shown in Figure \ref{fig:litreCluster2}).
-As a result, what we see in these litre plots reflects what we saw in the parallel coordinate lines of Figure \ref{fig:pcpGalbraith}: The replicate consistency in the Cluster 2 DEGs is not as clean as that in the Cluster 1 DEGs, but is still relatively clean.
-  \label{fig:litreCluster1}
-\end{figure}
-
-\begin{figure}[H]
   \includegraphics[width=\textwidth]{Images/litreClusterRutter}
@@
   \label{fig:litreClusterRutter}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/litreCluster1}
+  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 1 (originally shown in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.
+``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
+Most of the light orange points (representing the nine combinations of samples between treatment groups for a given DEG) deviate from the \textit{x=y} line in a cluster as expected.}
+  \label{fig:litreCluster1}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{Images/litreCluster2}
+  \caption{Example litre plots of the nine DEGs with the lowest FDR values from Cluster 2 (originally shown in Figure \ref{fig:pcpGalbraith}) of the Galbraith dataset.
+``C'' represents non-infected control samples and ``T'' represents virus-treated samples.
+Most of the dark orange points (representing all combinations of samples between treatment groups for a given DEG) deviate from the \textit{x=y} line in a cluster as expected.
+However, they are not as tightly clustered together compared to what we saw in the example litre plots of Cluster 1 (shown in Supplementary figure \ref{fig:litreCluster1}).
+As a result, what we see in these litre plots reflects what we saw in the parallel coordinate lines of Figure \ref{fig:pcpGalbraith}: The replicate consistency in the Cluster 2 DEGs is not as clean as that in the Cluster 1 DEGs, but is still relatively clean.}
+  \label{fig:litreCluster2}
 \end{figure}
@@
 
-\begin{table}[H]
-  \includegraphics[width=\textwidth]{Images/NCNRNCPathways}
-  \caption{GO analysis results for the 601 DEGs that were upregulated in the NC treatment in the NC versus NR treatment pair analysis.
-These DEGs represent genes that are upregulated when non-infected honeybees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.}
-  \label{tbl:NCNRNCPathways}
-\end{table}
-
-\begin{table}[H]
-  \includegraphics[width=\textwidth]{Images/NCNRNRPathways}
-  \caption{GO analysis results for the 340 DEGs that were upregulated in the NR treatment in the NC versus NR treatment pair analysis.
-These DEGs represent genes that are upregulated when non-infected honeybees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.}
-  \label{tbl:NCNRNRPathways}
-\end{table}
-
-\begin{table}[H]
-  \includegraphics[width=\textwidth]{Images/VCVRVCPathways}
-  \caption{GO analysis results for the 247 DEGs that were upregulated in the VC treatment in the VC versus VR treatment pair analysis.
-These DEGs represent genes that are upregulated when infected honeybees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.}
-  \label{tbl:VCVRVCPathways}
-\end{table}
-
-\begin{table}[H]
-  \includegraphics[width=\textwidth]{Images/VCVRVRPathways}
-  \caption{GO analysis results for the 129 DEGs that were upregulated in the VR treatment in the VC versus VR treatment pair analysis.
-These DEGs represent genes that are upregulated when infected honeybees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.}
-  \label{tbl:VCVRVRPathways}
-\end{table}
-
 \begin{figure}[H]
-  \includegraphics[width=0.8\textwidth]{Images/revigo}
-  \caption{GO analysis results for the 122 DEGs related to our ``resilience" hypothesis (A) and for the 125 DEGs related to our ``resistance" hypothesis (B).}
-  \label{fig:revigo}
+  \includegraphics[width=\textwidth]{Images/GRVenn}
+  \caption{Venn diagrams comparing the virus-related DEG overlaps between the Galbraith study (labeled as ``G'') and our study (labeled as ``R'').
+From left to right: Total virus-related DEGs (subplot A), virus-upregulated DEGs (subplot B), control-upregulated DEGs (subplot C).
+Both the total virus-related and virus-upregulated DEGs showed significant overlap between the studies (p-value < 2.2e-16) as per Fisher's Exact Test for Count Data.
+There was one gene that was virus-upregulated in the Galbraith study but control-upregulated in our study.}
+  \label{fig:GRVenn}
 \end{figure}
-
-\begin{table}[H]
-\begin{tabular}{ | m{5cm} | m{0.9cm}| m{5.5cm} | m{2cm} | } 
-\hline
-Contrast & DEGs & Interpretation & Results \\ 
-\hline
-V vs N & 43 & Genes that change expression due to virus effect regardless of diet status in bees & Table \ref{tbl:virusGenes} \\ 
-\hline
-NC vs NR & 941 & Genes that change expression due to diet effect in uninfected bees & Tables \ref{tbl:NCNRNCPathways} and \ref{tbl:NCNRNRPathways} \\ 
-\hline
-VC vs VR & 376 & Genes that change expression due to diet effect in infected bees & Tables \ref{tbl:VCVRVCPathways} and \ref{tbl:VCVRVRPathways} \\ 
-\hline
-VC upregulated in VC vs VR overlapped with NC upregulated in NC vs NR & 122 & ``Resilience'' genes that are turned on by good diet regardless of virus infection status in bees & Figure \ref{fig:revigo}A \\
-\hline
-VC upregulated in VC vs VR but NC is not upregulated in NC vs NR & 125 & ``Resistance'' genes that are turned on by good diet only in infected bees & Figure \ref{fig:revigo}B \\
-\hline
-\end{tabular}
-\caption{Contrasts in our study for assessing GO and pathways analysis.}
-  \label{tbl:contrasts}
-\end{table}
 

### 2018-09-04 Galbraith/DESeq2/makeDataMetrics.Rmd
Remove old files

--- 2018-04-13 C_R/DESeq2/makeDataMetrics.Rmd
@@
 
-    RDC <- metrics[which(metrics[which(metrics$padj <0.05),]$log2FoldChange>0),]$ID
-    RDR <- metrics[which(metrics[which(metrics$padj <0.05),]$log2FoldChange<0),]$ID
-    RD_DIET <- metrics[which(metrics$padj <0.05),]$ID
-    saveRDS(RDC, file="RDC.Rds")
-    saveRDS(RDR, file="RDR.Rds")
-    saveRDS(RD_DIET, file="RD_DIET.Rds")
+    sigMetrics <- metrics[which(metrics$padj <0.05),]
+    GDC1 <- sigMetrics[which(sigMetrics$log2FoldChange>0),]$ID
+    GDV1 <- sigMetrics[which(sigMetrics$log2FoldChange<0),]$ID
+    GD_TOTAL1 <- metrics[which(metrics$padj <0.05),]$ID
+
+    GDV <- unname(sapply(GDV1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+    GDC <- unname(sapply(GDC1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+    GD_TOTAL <- unname(sapply(GD_TOTAL1, function(x) strsplit(strsplit(x, "[|]")[[1]][3], "[-]")[[1]][1]))
+
+    saveRDS(GDV, file="GDV.Rds")
+    saveRDS(GDC, file="GDC.Rds")
+    saveRDS(GD_TOTAL, file="GD_TOTAL.Rds")
 

### 2018-09-04 Cotyledon/Lineups/Log/Lineups.R

@@ -1,13 +1,0 @@
-library(edgeR)
-library(ggplot2)
-library(GGally)
-library(EDASeq)
-library(utils)
-library(data.table)
-library(bigPint)
-
-thisPath <- getwd()
-
-data(soybean_cn)
-
-plotPermutations(soybean_cn, nPerm = 10, topThresh = 50, option = "log", outDir = getwd())

### 2018-09-04 Cotyledon/Lineups/Raw/Lineups.R

@@ -1,13 +1,0 @@
-library(edgeR)
-library(ggplot2)
-library(GGally)
-library(EDASeq)
-library(utils)
-library(data.table)
-library(bigPint)
-
-thisPath <- getwd()
-
-data(soybean_cn)
-
-plotPermutations(soybean_cn, nPerm = 10, topThresh = 20, outDir = getwd())

### 2018-09-04 Cotyledon/Lineups/Standardized/Lineups.R

@@ -1,13 +1,0 @@
-library(edgeR)
-library(ggplot2)
-library(GGally)
-library(EDASeq)
-library(utils)
-library(data.table)
-library(bigPint)
-
-thisPath <- getwd()
-
-data(soybean_cn)
-
-plotPermutations(soybean_cn, nPerm = 10, topThresh = 50, option = "standardize", outDir = getwd())

### 2018-09-04 Galbraith/DESeq2/ClusterStandard/makeFigure3.R

@@ -1,0 +1,157 @@
+# Run in /Users/suzu/HoneyBeePaper/Galbraith/DESeq2
+
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(matrixStats)
+
+getPCP <- function(nC){
+  
+  set.seed(1)
+  colList = scales::hue_pal()(nC+1)
+  colList[2] = "#E9AA0D"
+  colList[3] = "#EA502F"
+  k = cutree(hc, k=nC)
+  
+  yMin = min(sigDatas[,1:nColumns])
+  yMax = max(sigDatas[,1:nColumns])
+  
+  ###########################
+  
+  sbsDF <- data.frame()
+  for (i in 1:nC){
+    x = as.data.frame(sigDatas[which(k==i),])
+    xNames = rownames(x)
+    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+    sbsDF = rbind(sbsDF, data.frame(Cluster = paste("Cluster", i), FDR = xFDR))
+  }
+  
+  plot_clusters = lapply(1:nC, function(i){
+    j = rev(order(table(k)))[i]
+    x = as.data.frame(sigDatas[which(k==j),])
+    nGenes = nrow(x)
+    x$cluster = "color"
+    x$cluster2 = factor(x$cluster)
+    xNames = rownames(x)
+    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+    scatMatMetrics = list()
+    scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
+    scatMatMetrics[[currPair]]$FDR = 10e-10
+    scatMatMetrics[[currPair]]$ID = as.factor(as.character(scatMatMetrics[[currPair]]$ID))
+    
+    plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
+    plotData = data[,c(9,1:6)]
+    plotData[,c(2:7)] <- log(plotData[,c(2:7)] +1)
+    
+    fileName = paste(getwd(), "/", outDir, "/", currPair, "_SM_", nC, "_", i, ".jpg", sep="")
[…]

### 2018-09-04 Overlaps/Overlaps.R

@@ -19,15 +19,15 @@ RD_VIRUS_TOTAL <- readRDS("../N_V/DESeq2/RD_VIRUS_TOTAL.Rds")
 REV <- readRDS("../N_V/EdgeR/edgeR/REV.Rds")
 REN <- readRDS("../N_V/EdgeR/edgeR/REN.Rds")
 RE_VIRUS_TOTAL <- readRDS("../N_V/EdgeR/edgeR/RE_VIRUS_TOTAL.Rds")
-GDV <- readRDS("../VirusHoneyBee/DESeq2/GDV.Rds")
-GDC <- readRDS("../VirusHoneyBee/DESeq2/GDC.Rds")
-GD_TOTAL <- readRDS("../VirusHoneyBee/DESeq2/GD_TOTAL.Rds")
-GEV <- readRDS("../VirusHoneyBee/EdgeR/edgeR/GEV.Rds")
-GEC <- readRDS("../VirusHoneyBee/EdgeR/edgeR/GEC.Rds")
-GE_TOTAL <- readRDS("../VirusHoneyBee/EdgeR/edgeR/GE_TOTAL.Rds")
-GLV <- readRDS("../VirusHoneyBee/LimmaVoom/GLV.Rds")
-GLC <- readRDS("../VirusHoneyBee/LimmaVoom/GLC.Rds")
-GL_TOTAL <- readRDS("../VirusHoneyBee/LimmaVoom/GL_TOTAL.Rds")
+GDV <- readRDS("../Galbraith/DESeq2/GDV.Rds")
+GDC <- readRDS("../Galbraith/DESeq2/GDC.Rds")
+GD_TOTAL <- readRDS("../Galbraith/DESeq2/GD_TOTAL.Rds")
+GEV <- readRDS("../Galbraith/EdgeR/edgeR/GEV.Rds")
+GEC <- readRDS("../Galbraith/EdgeR/edgeR/GEC.Rds")
+GE_TOTAL <- readRDS("../Galbraith/EdgeR/edgeR/GE_TOTAL.Rds")
+GLV <- readRDS("../Galbraith/LimmaVoom/GLV.Rds")
+GLC <- readRDS("../Galbraith/LimmaVoom/GLC.Rds")
+GL_TOTAL <- readRDS("../Galbraith/LimmaVoom/GL_TOTAL.Rds")
 
 #######################################################
 
@@ -78,7 +78,7 @@ saveRDS(VCVRminInt, file="VCVRminInt.Rds")
 # Overlap has 11825
 
[…]

### 2018-09-04 Soybean/Lineups/Raw/Lineups.R

@@ -1,13 +1,0 @@
-library(edgeR)
-library(ggplot2)
-library(GGally)
-library(EDASeq)
-library(utils)
-library(data.table)
-library(bigPint)
-
-thisPath <- getwd()
-
-data(soybean_ir)
-
-plotPermutations(soybean_ir, nPerm = 10, topThresh = 50, outDir = getwd())

### 2018-09-04 VirusHoneyBee/DESeq2/ClusterStandard/makeFigure3.R

@@ -1,157 +1,0 @@
-# Run in /Users/suzu/HoneyBeePaper/VirusHoneyBee/DESeq2
-
-library(rtracklayer)
-library(Rsamtools)
-library(grid)
-library(GenomicAlignments)
-library(ggplot2)
-library(GGally)
-library(edgeR)
-library(stringr)
-library(EDASeq)
-library(dplyr)
-library(matrixStats)
-library(gridExtra)
-library(reshape2)
-library(scales)
-library(bigPint)
-library(matrixStats)
-
-getPCP <- function(nC){
-  
-  set.seed(1)
-  colList = scales::hue_pal()(nC+1)
-  colList[2] = "#E9AA0D"
-  colList[3] = "#EA502F"
-  k = cutree(hc, k=nC)
-  
-  yMin = min(sigDatas[,1:nColumns])
-  yMax = max(sigDatas[,1:nColumns])
-  
-  ###########################
-  
-  sbsDF <- data.frame()
-  for (i in 1:nC){
-    x = as.data.frame(sigDatas[which(k==i),])
-    xNames = rownames(x)
-    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
-    sbsDF = rbind(sbsDF, data.frame(Cluster = paste("Cluster", i), FDR = xFDR))
-  }
-  
-  plot_clusters = lapply(1:nC, function(i){
-    j = rev(order(table(k)))[i]
-    x = as.data.frame(sigDatas[which(k==j),])
-    nGenes = nrow(x)
-    x$cluster = "color"
-    x$cluster2 = factor(x$cluster)
-    xNames = rownames(x)
-    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
-    scatMatMetrics = list()
-    scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
-    scatMatMetrics[[currPair]]$FDR = 10e-10
-    scatMatMetrics[[currPair]]$ID = as.factor(as.character(scatMatMetrics[[currPair]]$ID))
-    
-    plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
-    plotData = data[,c(9,1:6)]
-    plotData[,c(2:7)] <- log(plotData[,c(2:7)] +1)
-    
-    fileName = paste(getwd(), "/", outDir, "/", currPair, "_SM_", nC, "_", i, ".jpg", sep="")
[…]

### 2018-09-04 Manuscript/Text/manuscript.tex
Create README

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}
\usepackage{setspace} % Lindsay added for double-spacing
\usepackage{lineno} % Lindsay added for line numbering

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{makecell} % Lindsay added for line breaks in tables

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

\title{Gene expression responses to diet quality and viral infection in Apis mellifera}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   email={lrutter@iastate.edu}   % email address
]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
\author[
   addressref={aff2},
   email={dicook@monash.edu}
]{\inits{DC}\fnm{Dianne} \snm{Cook}}
\author[
   addressref={aff3,aff4},
   email={amytoth@iastate.edu}
]{\inits{AT}\fnm{Amy L.} \snm{Toth}}
\author[
   addressref={aff5},
   corref={aff5},
   email={adolezal@illinois.edu}
]{\inits{AD}\fnm{Adam} \snm{Dolezal}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{Bioinformatics and Computational Biology Program, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}
\address[id=aff2]{%
  \orgname{Econometrics and Business Statistics, Monash University},
  \city{Clayton},   
  \state{VIC}
  \postcode{3800},
  \cny{Australia}
}

\address[id=aff3]{%
  \orgname{Department of Entomology, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}

\address[id=aff4]{%
  \orgname{Department of Ecology, Evolution, and Organismal Biology, Iowa State University},
  \city{Ames},   
  \state{IA}
  \postcode{50011},
  \cny{USA}
}

\address[id=aff5]{%
  \orgname{Department of Entomology, University of Illinois at Urbana-Champaign},
  \city{Urbana},   
  \state{IL}
  \postcode{61801},
  \cny{USA}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract}
\parttitle{Background}
Parts of Europe and the United States have witnessed dramatic losses in commercially managed honey bees over the past decade to what is considered an unsustainable extent. The large-scale loss of honey bees has considerable implications for the agricultural economy because honey bees are one of the leading pollinators of numerous crops. Honey bee declines have been associated with several interactive factors. Poor nutrition and viral infection are two environmental stressors that pose heightened dangers to honey bee health. In this study, we used RNA-sequencing to examine how monofloral diets and Israeli Acute Paralysis Virus inoculation influence gene expression patterns in honey bees.

\parttitle{Results}
We found a considerable nutritional response, with almost 2,000 transcripts changing with diet quality. The majority of these genes were over-represented for nutrient signaling (insulin resistance) and immune response (Notch signaling and JaK-STAT pathways). Somewhat unexpectedly, the transcriptomic response to viral infection was fairly limited. We only found 43 transcripts to be differentially expressed, some with known immune functions (argonaute-2), transcriptional regulation, and muscle contraction. We created contrasts to determine if any protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance). A similar number of resistance and tolerance candidate genes were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection. We also compared the virus main effect in our study (polyandrous colonies) to that obtained in a previous study (single-drone colonies) and verified significant overlap in differential expression despite visualization methods showing differences in the noisiness levels between these two datasets.

\parttitle{Conclusions}
Through transcriptional contrasts and functional enrichment analysis, we add to evidence of feedbacks between diet and disease in honey bees. We also show that comparing results derived from polyandrous colonies (which are typically more natural) and single-drone colonies (which usually yield more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently less artificial and less noisy. Altogether, we hope our results underline the merits of using data visualization techniques and multiple datasets to interpret RNA-sequencing datasets.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{Honey bee}
\kwd{RNA-sequencing}
\kwd{Israeli acute paralysis virus}
\kwd{Monofloral pollen}
\kwd{Visualization}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%

\begin{linenumbers} % Lindsay added
\begin{doublespacing} % Lindsay added

\section*{Background}

Commercially managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}. More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}. As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications regarding agricultural sustainability \cite{ccd4}.

Honey bee declines have been associated with several factors, including pesticide use, parasites, pathogens, habitat loss, and poor nutrition \cite{factors, factors2}. Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honey bees in interactive fashions as the environment changes \cite{interacting}. Nutrition and viral infection are two broad factors that pose heightened dangers to honey bee health in response to recent environmental changes.

Pollen is the main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}. At the individual level, pollen supplies most of the nutrients necessary for physiological development \cite{brodschneider} and is believed to have considerable impact on longevity \cite{longevity}. At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen \cite{jelly, jelly2}. Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honey bees. In particular, honey bees are confronted with less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}. Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}. Understanding how undiversified diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.

Viral infection was a comparatively minor problem in honey bees until the last century when Varroa destructor (an ectoparasitic mite) spread worldwide \cite{miteSpread}. This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits cocktails of viruses, and supports replication of certain viruses \cite{miteVirus, miteVirus2, miteVirus3}. More than 20 honey bee viruses have been identified \cite{numVirus}. One of these viruses that has been linked to honey bee decline is Israeli Acute Paralysis Virus (IAPV). A positive-sense RNA virus of the Dicistroviridae family \cite{fam}, IAPV causes infected honey bees to display shivering wings, decreased locomotion, muscle spams, and paralysis, and 80\% of caged infected adult honey bees die prematurely \cite{symptoms}. IAPV has demonstrated higher infectious capacities than other honey bee viruses in certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}. Its role in the rising phenomenon of ``Colony Collapse Disorder'' (in which the majority of worker bees disappear from a hive) remains unclear: It has been implicated in some studies \cite{iapvCCD, iapvCCD2} but not in other studies \cite{ccd1, fam, iapvCCD3}. Nonetheless, it is clear that IAPV reduces colony strength and survival.

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honey bees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honey bees \cite{intNV, intNV2, intNV3}. We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show the importance of the combined effects of both diet quality and virus infection. Specifically, high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen \cite{adamInt}.

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality. For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on vigor (tolerance) \cite{resTol1}. Transcriptomics is one means to better understand the mechanistic underpinnings of dietary and viral effects on honey bee health. Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection. This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a ``buffer'' against other stressors \cite{AdamTothReview}. As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and IAPV infection effects \cite{galbraith} in honey bees. As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense.

In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. As RNA-sequencing data can be highly noisy, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. Importantly, we use an in-depth data visualization approach to explore and corroborate our data, and suggest such an approach can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

\section*{Results}

\subsection*{Phenotypic results}

We reanalyzed our previously published dataset with a subset more relevant to our RNA-sequencing approaches in the current study that have a more focused question regarding diet quality. We briefly show it again here to inform the RNA-sequencing comparison because we reduced the number of treatments (from eight to four) from the original published data \cite{adamInt} as a means to focus on diet quality effects.

As shown in Figure 1, mortality rates of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.30e-05). The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 1.00e-05) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant. We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus. Namely, bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Tukey HSD, p $<$ 1.18e-03), and bees fed Chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Tukey HSD, p $<$ 4.80e-03) (Figure 1).

As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 1.96e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p $<$ 5.44e-03). However, bees fed Chestnut pollen did not have significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p = 1.11e-01). Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not do so (Figure 2).

\subsection*{Main effect DEG results}

We observed a substantially larger number of DEGs in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1). In the diet factor, there were more Chestnut-upregulated DEGs (\textit{n} = 1033) than Rockrose-upregulated DEGs (\textit{n} = 881). In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5). While these reported DEG counts are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table 1, Additional file 1 and Additional file 18).

GO analysis of the Chestnut-upregulated DEGs revealed the following over-represented categories: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1). GO analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).

With so few DEGs (\textit{n} = 43) in our virus main effect comparison, we focused on individual genes and their known functionalities rather than GO over-representation (Table \ref{tbl:virusGenes}). Of the 43 virus-related DEGs, only 10 had GO assignments within the DAVID database. These genes had putative roles in the recognition of pathogen-related lipid products and the cleaving of transcripts from viruses, as well as involvement in ubiquitin and proteosome pathways, transcription pathways, apoptotic pathways, oxidoreductase processes, and several more functions (Table \ref{tbl:virusGenes}).

No interaction DEGs were observed between the diet and virus factors of the study, in any of the pipelines (DESeq2, edgeR, and limma).

\subsection*{Pairwise comparison of DEG results}

The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 955 (Supplementary table 8, Additional file 1). Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level. Across every pair comparing the Chestnut and Rockrose levels, regardless of the virus level, the number of Chestnut-upregulated DEGs was higher than the number of Rockrose-upregulated DEGs (Supplementary table 8 C, D, E, F, Additional file 1). For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Supplementary table 8 A and B, Additional file 1). There were no DEGs between the treatment pair controlling for the Chestnut level of the virus effect (Supplementary table 8A, Additional file 1). These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).

\subsection*{Prior study comparison results}

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may better separate the infected and uninfected honey bees better than our dataset (Additional file 2). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with additional visualization techniques.

We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies. The background boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG. To reduce overlapping of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns. See more information about this plotting method and the ideal visual structure of DEGs from the bigPint package.

We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays (Figure 3). We do see that the first replicate of the virus group appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that this trend in the data that we saw in the PCA plot carried through into the DEG results. In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4). The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Additional file 3).

We also used litre plots to examine the structure of individual DEGs: We see that indeed the individual virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the individual virus DEGs from the Galbraith data (Additional files 5 and 6). For the Galbraith data, we examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3). We verify this trend again in the litre plots with the DEG points in the first cluster showing less tight cluster patterns (Additional files 5 and 6).

Finally, we looked at scatterplot matrices to assess the DEGs. We created standardized scatterplot matrices for each of the four clusters (Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: Deviating more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.

Despite the virus-related DEGs (\textit{n} = 1,019) from the Galbraith dataset displaying the expected patterns more than those from our dataset (\textit{n} = 43), there was significant overlap (p-value $<$ 2.2e-16) in the DEGs between the two studies, with 26/38 (68\%) of virus-upregulated DEGs from our study also showing virus-upregulated response in the Galbraith study (Figure 6).

\subsection*{Tolerance versus resistance results}

Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate genes and 125 ``resistance'' candidate genes. We again used parallel coordinate lines superimposed onto boxplots to visualize these candidate genes. To reduce overlapping of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns. Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate genes (Additional files 15 and 16). However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate genes, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group. Indeed, we see this pattern in the associated parallel coordinate plots (Additional file 15). Likewise, based on the contrasts we created to obtain the `resistance'' candidate genes, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups. We do generally see these expected patterns in the associated parallel coordinate plots: While there are large outliers in the ``NC'' group, the ``NR'' replicates are no longer typically below a standardized count of zero (Additional file 16). The genes in Cluster 3 may follow the expected pattern the most distinctively (Additional file 16).

Within our 122 ``tolerance'' gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism). However, we also discovered gene ontologies related to RNA polymerase II transcription, immune response, and regulation of response to reactive oxygen species (Figure 5A). Within our 125 ``resistance'' gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, chitin metabolism, oligosaccharide biosynthesis, and general metabolism) (Figure 5B).

\subsection*{Post hoc analysis results}

In general, the R-squared values between gene read counts and physiological measurements were low (R-squared < 0.1). However, some DEGs showed slightly larger R-squared values than the remaining non-DEG data. One prominent example of this includes the first and second cluster of the virus-related DEGs (Additional file 19I). The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those of the rest of the data. As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate genes, and virus-related DEGs) and physiological measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests. The p-values and Bonferroni correction values for each of these tests is provided in Supplementary table 9, Additional file 1. An overall trend emerges to suggest that DEGs may have significantly larger correlation with the physiological measurements compared to non-DEGs. Nonetheless, almost all genes, whether differentially expressed or not, showed low R-squared values. It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between physiological measurements (such as low viral titers versus high viral titers, or low mortality rate versus high mortality rate).

\section*{Discussion}

Challenges to honey bee health are a growing concern, in particular the combined, interactive effects of nutritional stress and pathogens (Dolezal and Toth 2018). In this study, we used RNA-sequencing to probe mechanisms underlying honey bee responses to two effects, diet quality and infection with the prominent virus of concern, IAPV. In general, we found a major nutritional transcriptomic response, with nearly 2,000 transcripts changing in response to diet quality (rockrose/poor diet versus chestnut/good diet). The majority of these genes were upregulated in response to high quality diet, and these genes were over-represented for functions (Supplementary table 2, Additional file 1) such as nutrient signaling metabolism (insulin resistance) and immune response (Notch signaling and JaK-STAT pathways). These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into innate immune responses.

Somewhat surprisingly, the transcriptomic response to virus infection in our experiment was fairly limited. We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as genes related to transcriptional regulation and muscle contraction. The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14).

Given the noisy nature of our data, and our desire to hone in on genes with real expression differences, we compared our data to the Galbraith study \cite{galbraith}, which also examined bees response to IAPV infection. In contrast to our study, Galbraith et al. identified a large number of virus responsive transcripts, and generally had less noise in their data (Figure 3 and Additional files 2A, 5, 6, 7, 8, 9, and 10). To identify the most consistent virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments. We found a large, statistically significant (p-value $<$ 2.2e-16) overlap, with 26/38 (68\%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al. (Figure 6). This result gives us confidence that, although noisy, we were able to uncover reliable, replicable gene expression responses to virus infection with our data.

Data visualization is a useful method to identify noise and robustness in RNA-sequencing data. In this study, we used extensive data visualization to improve the interpretation of our RNA-sequencing results. For example, the DESeq2 package comes with certain visualization options that are popular in RNA-sequencing analysis. One of these visualization is the principal component analysis (PCA) plot, which allows users to visualize the similarity between samples within a dataset. We could determine from this plot that indeed the Galbraith data may show more similarity between its replicates and differences between its treatments compared to our data (Additional file 2). However, the PCA plot only shows us information at the sample level. We wanted to investigate how these differences in the signal:to:noise ratios of the datasets would affect the structure of any resulting DEGs. As a result, we also used three plotting techniques from the bigPint package: We investigated the 1,019 virus-related DEGs from the Galbraith dataset and the 43 virus-related DEGs from our dataset using parallel coordinate lines, scatterplot matrices, and litre plots. To prevent overplotting issues in our graphics, we used a hierarchical clustering technique for the parallel coordinate lines to separate the set of DEGs into smaller groups. We also needed to examine four subsets of samples from our dataset to make effective use of the scatterplot matrices. After these tailorizations, we determined that the same patterns we saw in the PCA plots regarding the entire dataset extended down the pipeline analysis into the DEG calls: Even the DEGs from the Galbraith dataset showed more similarity between their replicates and differences between their treatments compared to those from our data. However, the 365 DEGs from the Galbraith data in Cluster 1 of Figure 3 showed an inconsistent first replicate in the treatment group (``V.1''), which was something we observed in the PCA plot. This indicates that this feature also extended down the analysis pipeline into DEG calls. Despite the differences in signal between these two datasets, there was substantial overlap in the resulting DEGs. We believe these visualization applications can be useful for future researchers analyzing RNA-sequencing data to quickly and effectively ensure that the DEG calls look reliable or at least overlap with DEG calls from similar studies that look reliable. We also expect this type of visualization exploration can be especially crucial when studying complex organisms that do not have genetic identicalness or similarity between replicates and/or when using experiments that may lack rigid design control.

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate genes had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion. Tolerance candidate genes had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. Moreover, circulating haemocytes in insects encapsulate and nodulate pathogens by forming a barrier between the pathogen and the host tissues. This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}. These possible immunological defense mechanisms within our ``tolerance'' candidate genes and metabolic processes within our ``resistance'' candidate genes may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

There were several limitations in this study that could be improved upon in future studies. For instance, our comparison between the Galbraith data (single-drone colonies) and our data (polyandrous colonies) was limited by numerous extraneous variables between these studies. In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on one-day old worker honey bees that were fed sugar and MegaBee diet, whereas our study focused on adult worker honey bees that were fed monofloral diets. Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group. Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}. Our comparative visualization assessment between these two datasets was also somewhat limited because the virus effect in the Galbraith study used three replicates for each level, whereas the virus effect in our study used twelve replicates for each level that were actually further subdivided into six replicates for each diet level. Hence the apparent reduction in noise observed in the Galbraith data compared to our data in the PCA plots, parallel coordinate plots, scatterplot matrices, and litre plots may be an inadvertent product of the smaller number of replicates used and the lack of a secondary treatment group rather than solely the reduction in genetic variability through the single-drone colony design itself. With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between polyandrous colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.

In addition, this study used a whole body RNA-sequencing approach. In future related studies, it may be informative to use tissue-specific methods. Recent evidence has suggested that RNA-sequencing approaches toward composite structures in honey bees leads to false negatives, implying that genes strongly differentially expressed in particular structures may not reach significance within the composite structure. On a similar note, recent studies have also found that within a composite extraction, structures therein may contain opposite patterns of differential expression. We can provide more detailed answers to our original transcriptomic questions if we were to repeat this same experimental design only now at a more refined tissue level \cite{tissueLevel}. Another future direction related to this work would be to integrate multiple omics datasets to investigate monofloral diet quality and IAPV infection in honey bees. Indeed, previous studies in honey bees have found that multiple omics datasets do not always align in a clear-cut manner, and hence may broaden our understanding of the molecular mechanisms being explored \cite{galbraith}.

\section*{Conclusions}

To the best of our knowledge, there are few to no studies investigating honey bee gene expression specifically related to monofloral diets, and few to no studies examining honey bee gene expression related to the combined effects of diet in any general sense and viral inoculation in any general sense. It also remains unknown whether the protective effects of good diet in honey bees is due to direct effects on immune function (resistance) or indirect effects of energy availability on vigor and health (tolerance). We attempted to address these unresolved areas by conducting a two-factor RNA-sequencing study that examined how monofloral diets and IAPV inoculation influence gene expression patterns in honey bees. Overall, our data suggest complex transcriptomic responses to multiple stressors in honey bees. Diet has the capacity for large and profound effects on gene expression and may set up the potential for both resistance and tolerance to viral infection, adding to previous evidence of possible feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

Moreover, this study also demonstrated the benefits of using data visualizations and multiple datasets to address inherently messy biological data. For instance, by verifying the substantial overlap in our DEG lists to those obtained in another study that addressed a similar question using specimens with less genetic variability, we were able to place much higher confidence in the differential gene expression results from our otherwise noisy data. We also suggested that comparing results derived from polyandrous colony designs (which are usually more natural) and single-drone colony designs (which usually have more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently more realistic and less noisy. Altogether, we hope our results underline the merits of using data visualization techniques and multiple datasets to understand and interpret RNA-sequencing datasets.

\section*{Methods}

Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt}. The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report \cite{adamInt}.

\subsection*{Design of two-factor experiment}

There are several reasons why, in the current study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets). First, when assessing diet diversity, a sugar diet is often used as a control. However, such an experimental design does not reflect real-world conditions for honey bees as they rarely face a total lack of pollen \cite{DiPasquale}. Second, in studies that compared honey bee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet \cite{DiPasquale}. Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010). Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced with less diversified diets (monoculture) \cite{landscape1, brodschneider}. As a result, there is a need to better understand how monofloral diets affect honey bee health as a step toward mitigating the negative impact of human activity on the honey bee population.

Consequently, for our nutrition factor, we examined two monofloral pollen diets, Rockrose (Cistus) and Castanea (Chestnut). Rockrose pollen is generally considered less nutritious than Chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron \cite{DiPasquale, adamInt}. For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV. This experimental design resulted in four treatment groups (Rockrose pollen without IAPV exposure, Chestnut pollen without IAPV exposure, Rockrose pollen with IAPV exposure, and Chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.

\subsection*{RNA extraction}

Fifteen cages per treatment were originally sampled. Six live honey bees from each cage were randomly selected 36 hours post inoculation and placed into tubes \cite{carrillo}. Tubes were kept on dry ice and then transferred into a -80C freezer until processing. Eight cages were randomly selected from the original 15 cages, and 2 honey bees per cage were randomly selected from the original six live honey bees per cage. Whole body RNA from each pool of two honey bees were extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment. Samples were suspended in water to 200-400 ng/$\mu$l. All samples were then tested on a Bioanalyzer at the DNA core facility to ensure quality (RIN $>$ 8).

\subsection*{Gene expression}

Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing). A standard Illumina mRNA library was prepared by the DNA facility. Reads were aligned to the BeeBase Version 3.2 genome \cite{hbGenome} from the Hymenoptera Genome Database \cite{hymenopteraDB} using the programs GMAP and GSNAP \cite{gsnap}. There were four lanes of sequencing with 24 samples per lane. Each sample was run twice. Approximately 75-90\% of reads were mapped to the honey bee genome. Each lane produced around 13 million single-end 100 basepair reads. We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs). We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs). We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honey bees \cite{galbraith}. We tested each DEG analysis using recommended parameters with DESeq2 \cite{deseq2}, edgeR \cite{edger}, and LimmaVoom \cite{limma}. In all cases, we used a false discovery rate (FDR) threshold of 0.05 \cite{benjamini}. Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines). The eulerr shiny application was used to construct Venn diagram overlap images \cite{euler}. In the main section of our chapter and in subsequent analyses, we focus on the DEG results from DESeq2 \cite{deseq2} as this pipeline was also used in the Galbraith study \cite{galbraith}.

\subsection*{Comparison to prior studies on transcriptomic response to viral infection}

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith} who also addressed honey bee transcriptomic responses to virus infection. We applied the same downstream bioinformatics analyses between our count table and the count table provided in the Galbraith paper. When we applied our bioinformatics pipeline to the Galbraith count table, we obtained different differential expression counts compared to the results published in the Galbraith paper. However, there was substantial overlap and we considered this justification to use the differential expression list we obtained in order to keep the downstream bioinformatics analyses as similar as possible between the two datasets (Additional file 17).

While our study examines honey bees from polyandrous colonies, the Galbraith study examined honey bees from single-drone colonies. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our honey bees may also reflect more realistic environmental and genetic simulations. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs. To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. It is our hope that this aspect of our study may shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees.

\subsection*{Visualization}

We used an array of visualization tools as part of our analysis. We first used popular tools (like the PCA plot) from the DESeq2 package. After that, we used multivariate visualization tools from our work-in-progress package called bigPint. Specifically, we used parallel coordinate plots, litre plots, and scatterplot matrices to assess the variability between the replicates and the treatments in our data and in the DEG outputs from applied models. We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines.

\subsection*{Gene ontology}

DEGs were uploaded as a background list to DAVID Bioinformatics Resources 6.7 \cite{davidBio, davidBio2}. The overrepresented gene ontology (GO) terms of DEGs were determined using the BEEBASE\_ID identifier option (honey bee gene model) in the DAVID software. To fine-tune the GO term list, only terms correlating to Biological Processes were considered. The refined GO term list was then imported into REVIGO \cite{revigo}, which uses semantic similarity measures to cluster long lists of GO terms.

\subsection*{Probing tolerance versus resistance}

To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (tolerance), we created contrasts of interest (Table \ref{tbl:contrasts}). In particular, we assigned ``resistance candidate genes'' to be the ones that were upregulated in the Chestnut group within the virus infected bees but not upregulated in the Chestnut group within the non-infected bees. We also assigned ``tolerance candidate genes'' to be the ones that were upregulated in the Chestnut group for both the virus infected bees and non-infected bees. Our interpretation of these genes is that they represent those that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not. We then determined how many genes fell into these two categories and analyzed their GO terminologies.

\subsection*{Post hoc analysis}

We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates. On account of this observation, we wanted to explore whether our DEG read counts correlated with physiological metrics, including IAPV titers, Schmallenberg Virus (SBV) titers, and mortality rates. For this process, we considered virus main effect DEGs, ``resistance candidate'' genes, and ``tolerance candidate'' genes. We grouped each of these DEG sets into four hierarchical clusters. For each gene in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the physiological metrics across its 24 samples. We then used the Kruskal–Wallis test to determine if the distribution of the R-squared values in the DEG clusters significantly differed from those in the non-DEG genes.

\end{doublespacing} % Lindsay added

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Ethics approval and consent to participate}
  Not applicable.

\section*{Consent for publication}
  Not applicable.

\section*{Availability of data and materials}
  The datasets generated and/or analysed during the current study are available in the [NAME] repository, [PERSISTENT WEB LINK TO DATASETS]. Include our data, Galbraith data, scripts to reproduce tables and figures.

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Funding}
  This work was supported by the United States Department of Agriculture, Agriculture and Food Research Initiative (USDA-AFRI) 2011-04894.

\section*{Author's contributions}
  Each author must write their authorsip contributions.
    %All authors read and approved the final manuscript.

\section*{Acknowledgements}
  Who/what should we mention here? \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc_honeybee}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files
\newpage
\section*{Figures}

\begin{figure}[h!]
\caption{\csentence{Mortality rates for the four treatment groups, two virus groups, and two diet groups.}
Left to right: Mortality rates for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The mortality rate data included 59 samples with 15 replicates per treatment group, except for the ``NC'' group having 14 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the chapter. The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups.}
Left to right: IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The IAPV titer data included 38 samples with 10 replicates per treatment group, except for the ``NR'' group having 8 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the chapter. The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{Parallel coordinate plots of the 1,019 DEGs after hierarchical clustering of size four between the virus-infected and control groups of the Galbraith data \cite{galbraith}.}
Parallel coordinate plots of the 1,019 DEGs after hierarchical clustering of size four between the virus-infected and control groups of the Galbraith study. ``N'' represents non-inoculation, ``V'' represents viral inoculation. Clusters 1, 2, and 4 seem to represent DEGs that were overexpressed in the virus inoculated group, and Cluster 3 seems to represent DEGs that were overexpressed in the non-inoculated control group. In general, the DEGs appeared as expected, but there is rather noticeable deviation of the first replicate from the virus-treated sample (``V.1'') from the other virus-treated replicates in Cluster 1.}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{Parallel coordinate plots of the 43 DEGs after hierarchical clustering of size four between the virus-infected and control groups of our study.}
Parallel coordinate plots of the 43 DEGs after hierarchical clustering of size four between the virus-infected and control groups of our study. ``N'' represents non-infected control group, and ``V'' represents treatment of virus. The vertical red line indicates the distinction between treatment groups. We see from this plot that the DEG designations for this dataset do not appear as clean compared to what we saw in the Galbraith dataset in Figure 3.}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{Gene ontology analysis results for the 122 DEGs related to our ``tolerance'' hypothesis and for the 125 DEGs related to our ``resistance'' hypothesis.}
GO analysis results for the 122 DEGs related to our ``tolerance'' hypothesis (A) and for the 125 DEGs related to our ``resistance'' hypothesis (B).}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{Venn diagrams comparing the virus-related DEG overlaps between our dataset and the Galbraith dataset.}
Venn diagrams comparing the virus-related DEG overlaps between the Galbraith study (labeled as ``G'') and our study (labeled as ``R''). From left to right: Total virus-related DEGs (subplot A), virus-upregulated DEGs (subplot B), control-upregulated DEGs (subplot C). Both the total virus-related and virus-upregulated DEGs showed significant overlap between the studies (p-value $<$ 2.2e-16) as per Fisher's Exact Test for Count Data. There was one gene that was virus-upregulated in the Galbraith study but control-upregulated in our study.}
\end{figure}

\newpage
\section*{Tables}

\begin{table}[h!]
\begin{tabular}{ccccc}
\hline
BeeBase ID & Gene Name & Known functions & Us & Galbraith \\ 
\hline
GB41545 & \makecell{MD-2-related \\ lipid-recognition \\ protein-like} & \makecell{Implicated in lipid recognition, \\ particularly in the recognition of \\ pathogen related products} & N & - \\
\hline
GB50955 & \makecell{Protein \\ argonaute-2} & \makecell{Interacts with small interfering RNAs \\ to form RNA-induced silencing \\ complexes which target and cleave \\ transcripts that are mostly from \\ viruses and transposons} & V & V \\
\hline
GB48755 & \makecell{UBA-like \\ domain-containing \\ protein 2} & \makecell{Found in diverse proteins involved \\ in ubiquitin/proteasome \\ pathways} & V & V \\
\hline
GB47407 & Histone H4 & \makecell{Capable of affecting transcription, \\ DNA repair, and DNA replication \\ when post-transcriptionally modified} & V & V \\
\hline
GB42313 & \makecell{Leishmanolysin-like \\ peptidase} & \makecell{Encodes a protein involved in cell \\ migration and invasion; implicated in \\ mitotic progression in D. melanogaster} & V & V \\
\hline
GB50813 & \makecell{Rho guanine \\ nucleotide \\ exchange factor 11} & \makecell{Implicated in regulation of apoptopic \\ processes, cell growth, signal \\ transduction, and transcription} & V & V \\
\hline
GB54503 & \makecell{Thioredoxin \\ domain-containing \\ protein} & \makecell{Serves as a general protein \\ disulphide oxidoreductase} & N & - \\
\hline
GB53500 & \makecell{Transcriptional \\ regulator Myc-B} & \makecell{Regulator gene that codes for a \\ transcription factor} & V & V \\
\hline
GB51305 & Tropomyosin-like & \makecell{Related to protein involved in muscle \\ contraction} & N & N \\
\hline
GB50178 & \makecell{Cilia and \\ flagella-associated \\ protein 61-like} & \makecell{Induces components required for \\ wild-type motility and \\ stable assembly of motile cilia} & V & V \\
\hline
\end{tabular}
\caption{Known functions of the mapped subset of 43 DEGs in the virus main effect of our study. Whether the gene was overrepresented in the virus or non-virus group is also indicated for both our study and the Galbraith study. Functionalities were extracted from Flybase, National Center for Biotechnology Information and The European Bioinformatics Institute databases.}
  \label{tbl:virusGenes}
\end{table}

\begin{table}[h!]
\begin{tabular}{cccc}
\hline
Contrast & DEGs & Interpretation & Results \\ 
\hline
V (all) vs N (all) & 43 & \makecell{Genes that change expression \\ due to virus effect regardless \\ of diet status in bees} & Table \ref{tbl:virusGenes} \\
\hline
NC vs NR & 941 & \makecell{Genes that change expression \\ due to diet effect in \\ uninfected bees} & \makecell{Supplementary \\ tables 4 and 5, \\ Additional file 1} \\
\hline
VC vs VR & 376 & \makecell{Genes that change expression \\ due to diet effect in \\ infected bees} & \makecell{Supplementary \\ tables 6 and 7, \\ Additional file 1} \\
\hline
\makecell{VC upregulated in VC vs VR, and \\ NC upregulated in NC vs NR} & 122 & \makecell{``Tolerance'' genes that turn \\ on by good diet regardless of \\ virus infection status in bees} & Figure 5A \\
\hline
\makecell{VC upregulated in VC vs VR, but \\ NC not upregulated in NC vs NR} & 125 & \makecell{``Resistance'' genes that turn \\ on by good diet only in \\ infected bees} & Figure 5B \\
\hline
\end{tabular}
\caption{Contrasts in our study for assessing GO and pathways analysis.}
  \label{tbl:contrasts}
\end{table}

\newpage
\section*{Additional Files}

  \subsection*{Additional file 1 --- Supplementary tables.}
    \textbf{Table 1:} Number of DEGs across three analysis pipelines for (A) the diet main effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study. For the diet effects, ``C'' represents Chestnut diet and ``R'' represents Rockrose diet. For the virus effects, ``N'' represents control non-inoculated and ``V'' represents virus-inoculated.
    \textbf{Table 2:} Pathways related to the 1,033 DEGs that were upregulated in the Chestnut treatment from the diet main effect.
    \textbf{Table 3:} Pathways related to the 881 DEGs that were upregulated in the Rockrose treatment from the diet main effect.
    \textbf{Table 4:} GO analysis results for the 601 DEGs that were upregulated in the NC treatment from the NC versus NR treatment pair analysis. These DEGs represent genes that are upregulated when non-infected honey bees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.
    \textbf{Table 5:} GO analysis results for the 340 DEGs that were upregulated in the NR treatment from the NC versus NR treatment pair analysis. These DEGs represent genes that are upregulated when non-infected honey bees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.
    \textbf{Table 6:} GO analysis results for the 247 DEGs that were upregulated in the VC treatment from the VC versus VR treatment pair analysis. These DEGs represent genes that are upregulated when infected honey bees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.
    \textbf{Table 7:} GO analysis results for the 129 DEGs that were upregulated in the VR treatment from the VC versus VR treatment pair analysis. These DEGs represent genes that are upregulated when infected honey bees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.
    \textbf{Table 8:} Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor. ``C'' represents Chestnut diet, ``R'' represents Rockrose diet, ``V'' represents virus-inoculated, and ``N'' represents control non-inoculated.
    \textbf{Table 9:} Kruskal-Wallis p-value and Bonferroni corrections for the 36 combinations of DEG lists, physiological metrics, and cluster number. (XLS).

  \subsection*{Additional file 2 --- PCA plots for the Galbraith dataset and for our dataset.}
    PCA plots for the Galbraith dataset (A) and for our dataset (B). ``V'' represents virus-inoculated, and ``N'' represents control non-inoculated. The x-axis represents the principal component with the most variation and the y-axis represents the principal component with the second-most variation (PNG).

  \subsection*{Additional file 3 --- Parallel coordinate lines of the diet-related DEGs of our dataset.}
    Parallel coordinate plots of the 1,914 DEGs after hierarchical clustering of size six between the Chestnut and Rockrose groups of our study. Here ``C'' represents Chestnut samples, and ``R'' represents Rockrose samples. The vertical red line indicates the distinction between treatment groups. We see from this plot that the DEG designations for this dataset do not appear as clean compared to what we saw in the Galbraith dataset in Figure 3 (PNG).

  \subsection*{Additional file 4 --- Example litre plots from the virus-related DEGs of our dataset.}
    Example litre plots of the nine DEGs with the lowest FDR values from the 43 virus-related DEGs of our dataset. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. Most of the magenta points (representing the 144 combinations of samples between treatment groups for a given DEG) do not reflect the expected pattern as clearly compared to what we saw in the litre plots of the Galbraith data. They are not as clustered together (representing replicate inconsistency) and they sometimes cross the \textit{x=y} line (representing lack of difference between treatment groups). This finding reflects what we saw in the messy looking parallel coordinate lines of Figure 4 (PNG).

  \subsection*{Additional file 5 --- Example litre plots of DEGs from Cluster 1 of the Galbraith dataset.}
    Example litre plots of the nine DEGs with the lowest FDR values from the 365 DEGs in Cluster 1 (originally shown in Figure 3) of the Galbraith dataset. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. Most of the light orange points (representing the nine combinations of samples between treatment groups for a given DEG) deviate from the \textit{x=y} line in a tight bundle as expected (PNG).

  \subsection*{Additional file 6 --- Example litre plots of DEGs from Cluster 2 of the Galbraith dataset.}
    Example litre plots of the nine DEGs with the lowest FDR values from the 327 DEGs in Cluster 2 (originally shown in Figure 3) of the Galbraith dataset. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. Most of the dark orange points (representing the nine combinations of samples between treatment groups for a given DEG) deviate from the \textit{x=y} line in a compact clump as expected. However, they are not as tightly bunched together compared to what we saw in the example litre plots of Cluster 1 (shown in Additional file 5). As a result, what we see in these litre plots reflects what we saw in the parallel coordinate lines of Figure 3: The replicate consistency in the Cluster 1 DEGs is not as clean as that in the Cluster 2 DEGs, but is still relatively clean (PNG).

  \subsection*{Additional file 7 --- Scatterplot matrix of DEGs from Cluster 1 of the Galbraith dataset.}
    The 365 DEGs from the first cluster of the Galbraith dataset (originally shown in Figure 3) superimposed as light orange dots onto all genes as black dots in the form of a scatterplot matrix. The data has been standardized. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots. However, we do see that sample ``V.1'' may be somewhat inconsistent in these DEGs, as its presence in the replicate scatterplots shows DEGs deviating from the \textit{x=y} line more than expected and its presence in the treatment scatterplots shows DEGs adhering to the \textit{x=y} line more than expected. This inconsistent sample was something we observed in Figure 3 (PNG).

  \subsection*{Additional file 8 --- Scatterplot matrix of DEGs from Cluster 2 of the Galbraith dataset.}
    The 327 DEGs from the second cluster of the Galbraith dataset (originally shown in Figure 3) superimposed as dark orange dots onto all genes as black dots in the form of a scatterplot matrix. The data has been standardized. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots (PNG).

  \subsection*{Additional file 9 --- Scatterplot matrix of DEGs from Cluster 3 of the Galbraith dataset.}
    The 224 DEGs from the third cluster of the Galbraith dataset (originally shown in Figure 3) superimposed as turquoise dots onto all genes as black dots in the form of a scatterplot matrix. The data has been standardized. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots (PNG).

  \subsection*{Additional file 10 --- Scatterplot matrix of DEGs from Cluster 4 of the Galbraith dataset.} 
    The 103 DEGs from the fourth cluster of the Galbraith dataset (originally shown in Figure 3) superimposed as pink dots onto all genes as black dots in the form of a scatterplot matrix. The data has been standardized. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots. We also see that the second replicate from the virus-treated sample (``V.2'') may be somewhat inconsistent in these DEGs, as its presence in the replicate scatterplots results in the DEGs unexpectedly deviating from the \textit{x=y} line and its presence in the treatment scatterplots results in the DEGs unexpectedly adhering to the \textit{x=y} line (PNG).

  \subsection*{Additional file 11 --- Scatterplot matrix of virus-related DEGs from our dataset, showing only replicates 1, 2, and 3.} 
    The 43 virus-related DEGs from our dataset superimposed as magenta dots onto all genes in the form of a scatterplot matrix. Only replicates 1, 2, and 3 are shown from both treatment groups. The data has been standardized. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. We see that, compared to the scatterplot matrices from certain clusters of the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, sometimes unexpectedly deviating from the \textit{x=y} line in the replicate plots and sometimes unexpectedly adhering to the \textit{x=y} line in the treatment plots (PNG).

  \subsection*{Additional file 12 --- Scatterplot matrix of virus-related DEGs from our dataset, showing only replicates 4, 5, and 6.}     
    The 43 virus-related DEGs from our dataset superimposed as magenta dots onto all genes in the form of a scatterplot matrix. Only replicates 4, 5, and 6 are shown from both treatment groups. The data has been standardized. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. We see that, compared to the scatterplot matrices from certain clusters of the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, and most of them unexpectedly adhere to the \textit{x=y} line in the treatment plots (PNG).

  \subsection*{Additional file 13 --- Scatterplot matrix of virus-related DEGs from our dataset, showing only replicates 7, 8, and 9.}    
    The 43 virus-related DEGs from our dataset superimposed as magenta dots onto all genes in the form of a scatterplot matrix. Only replicates 7, 8, and 9 are shown from both treatment groups. The data has been standardized. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. We see that, compared to the scatterplot matrices from certain clusters of the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, sometimes unexpectedly deviating from the \textit{x=y} line in the replicate plots and sometimes unexpectedly adhering to the \textit{x=y} line in the treatment plots (PNG).

  \subsection*{Additional file 14 --- Scatterplot matrix of virus-related DEGs from our dataset, showing only replicates 10, 11, and 12.}    
    The 43 virus-related DEGs from our dataset superimposed onto all genes in the form of a scatterplot matrix. Only replicates 10, 11, and 12 are shown from both treatment groups. The data has been standardized. ``N'' represents non-infected control samples and ``V'' represents virus-treated samples. We see that, compared to the scatterplot matrices from certain clusters of the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, and most of them unexpectedly deviate from the \textit{x=y} line in the virus-related replicate plots (PNG).

   \subsection*{Additional file 15 --- Parallel coordinate plots of the ``tolerance'' candidate genes.}    
    Parallel coordinate plots of the 122 DEGs after hierarchical clustering of size four between the ``tolerance'' candidate genes. Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet. The vertical red line indicates the distinction between treatment groups. We see there is considerable noise in the data (non-consistent replicate values), but that the general patterns of the DEGs follow what we expect based on our ``tolerance'' contrast (PNG).

   \subsection*{Additional file 16 --- Parallel coordinate plots of the ``resistance'' candidate genes.}    
    Parallel coordinate plots of the 125 DEGs after hierarchical clustering of size four between the ``resistance'' candidate genes. Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet. The vertical red line indicates the distinction between treatment groups. We see there is considerable noise in the data (non-consistent replicate values), but that the general patterns of the DEGs follow what we expect based on our ``resistance'' contrasts (PNG).

   \subsection*{Additional file 17 --- Venn diagrams comparing the virus-related DEG overlaps in the Galbraith data using our pipeline and the pipeline used by Galbraith et al.}
  Venn diagrams comparing the virus-related DEG overlaps of the Galbraith data from the DESeq2 bioinformatics pipelines used in the Galbraith study (labeled as ``G.O.'') and the DESeq2 bioinformatics pipelines used in our study (labeled as ``G.R''). While we were not able to fully replicate the DEG list published in the Galbraith study, our DEG list maintained significant overlaps with their DEG list. From left to right: Total virus-related DEGs (subplot A), virus-upregulated DEGs (subplot B), control-upregulated DEGs (subplot C) (PNG).

   \subsection*{Additional file 18 --- Venn diagrams of main effect DEG overlaps across DESeq2, edgeR, and limma}
  Venn diagrams comparing DEG overlaps across DESeq2, edgeR, and limma for our diet main effect (top row), our virus main effect (middle row), and the Galbraith virus main effect (bottom row). Within a given subplot, ``D'' represents DESeq2, ``E'' represents edgeR, and ``L'' represents limma. From left to right on top row: Total diet-related DEGs (subplot A), Castanea-upregulated DEGs (subplot B), Rockrose-upregulated DEGs (subplot C). From left to right on middle row: Total virus-related DEGs (subplot D), virus-upregulated DEGs (subplot E), control-upregulated DEGs in our data (subplot F). From left to right on bottom row: Total virus-related DEGs (subplot G), virus-upregulated DEGs (subplot H), control-upregulated DEGs in the Galbraith data (subplot I) (PNG). With the exception of the limma pipeline resulting in zero DEGs in our virus main effect analysis, we found significant overlaps between DEG lists across the different pipelines (DESeq2, edgeR, and limma). In general, DESeq2 resulted in the largest number of DEGs and limma resulted in the least number of DEGs (PNG).

  \subsection*{Additional file 19 --- Analysis of correlation between DEG read counts and physiological metrics}
  Distribution of R-squared values for DEG cluster read counts and physiological metrics. Columns left to right: SBV titers, mortality rates, and IAPV titers. Rows top to bottom: Tolerance candidate DEGs, resistance candidate DEGs, and virus-related DEGs. Each subplot includes five boxplots which represent the R-squared value distributions for four DEG clusters and all remaining non-DEGs in the data. The top number above each boxplot represents the number of genes included. The first four boxplots also include a bottom number, which represents the Kruskal–Wallis p-value of the comparison of the R-squared distribution of the cluster and the R-squared distribution of the non-DEG data (PNG).

\end{backmatter}
\end{linenumbers} % Lindsay added
\end{document}

### 2018-09-04 Manuscript/MakeFiguresFiles/makeFigure1and2.R

@@ -1,0 +1,77 @@
+library(plyr)
+library(ggplot2)
+library(multcompView)
+library(readr)
+library(multcomp)
+library(nlme)
+
+d <- read_csv("day3Mortality.csv")
+d = as.data.frame(d)
+attr(d, "spec") <- NULL
+# Note: There are 14 NC; 15 VR, VC, NR
+d = d[which(d$Treatment %in% c("None Chestnut", "None Rockrose", "Virus Chestnut", "Virus Rockrose")),]
+d$Treatment
+
+d$Treatment = as.factor(d$Treatment)
+colnames(d)[3] = "Day3Mortality"
+levels(d$Treatment) <- c('NC', 'NR', 'VC', 'VR')
+
+d$Virus = sapply(d$Treatment, function(x) substr(x, 1, 1))
+d$Diet = sapply(d$Treatment, function(x) substr(x, 2, 2))
+d$Virus = as.factor(d$Virus)
+d$Diet = as.factor(d$Diet)
+
+mortcomp = lme(Day3Mortality ~ Diet*Virus, data=d, random = ~1|Experiment) 
+mortcomp2 = lme(Day3Mortality ~ Treatment, data=d, random = ~1|Experiment) 
+anova(mortcomp)
+summary(glht(mortcomp2, linfct=mcp(Treatment="Tukey")))
+
+# Mortality (All)
+labelDF = data.frame(plot.labels=c("NR","VC","VR","NC"), labels = c("ab","bc","c","a"), V1 = c(0.22, 0.479, 0.793, 0.164))
+plotMortality = ggplot(d, aes(x=Treatment, y=Day3Mortality)) + geom_boxplot(fill="palegreen2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) + ylab("Day 3 Mortality Rate") + theme_gray() + ylim(0,0.8) + theme(text=element_text(size=20))
+
+# Mortality (N vs V)
+labelDF = data.frame(plot.labels=c("N","V"), labels = c("a","b"), V1 = c(0.22, 0.52))
+plotMortality2 = ggplot(d, aes(x=Virus, y=Day3Mortality)) + geom_boxplot(fill="palegreen2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) + theme_gray() + ylim(0,0.8) +theme(axis.title.y=element_blank(), text=element_text(size=20))
[…]

### 2018-09-04 Manuscript/MakeFiguresFiles/makeFigure4.R

@@ -1,0 +1,157 @@
+# Run in /Users/suzu/HoneyBeePaper/N_V/DESeq2/ClusterStandard
+
+library(rtracklayer)
+library(Rsamtools)
+library(grid)
+library(GenomicAlignments)
+library(ggplot2)
+library(GGally)
+library(edgeR)
+library(stringr)
+library(EDASeq)
+library(dplyr)
+library(matrixStats)
+library(gridExtra)
+library(reshape2)
+library(scales)
+library(bigPint)
+library(matrixStats)
+
+getPCP <- function(nC){
+  set.seed(1)
+  myCols = scales::hue_pal()(nC+1)
+  colList = c(myCols[1], "#C11B8D", myCols[2:4])
+  k = cutree(hc, k=nC)
+  
+  yMin = min(sigDatas[,1:nColumns])
+  yMax = max(sigDatas[,1:nColumns])
+  
+  ###########################
+  
+  sbsDF <- data.frame()
+  for (i in 1:nC){
+    x = as.data.frame(sigDatas[which(k==i),])
+    xNames = rownames(x)
+    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+    sbsDF = rbind(sbsDF, data.frame(Cluster = paste("Cluster", i), FDR = xFDR))
+  }
+  
+  plot_clusters = lapply(1:nC, function(i){
+    j = rev(order(table(k)))[i]
+    x = as.data.frame(sigDatas[which(k==j),])
+    nGenes = nrow(x)
+    x$cluster = "color"
+    x$cluster2 = factor(x$cluster)
+    xNames = rownames(x)
+    
+    saveRDS(xNames, file=paste0("Sig_", nC, "_", i, ".Rds"))
+    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
+    scatMatMetrics = list()
+    scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
+    scatMatMetrics[[currPair]]$FDR = 10e-10
+    scatMatMetrics[[currPair]]$ID = as.factor(as.character(scatMatMetrics[[currPair]]$ID))
+    
+    plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
+    
+    fileName = paste(getwd(), "/", outDir, "/", currPair, "_SM_", nC, "_", i, ".jpg", sep="")
[…]

### 2018-09-04 MortalityData/HSD.R

@@ -1,55 +1,0 @@
-library(agricolae)
-data(sweetpotato)
-# Analysis of variance model
-model<-aov(yield~virus, data=sweetpotato)
-out <- HSD.test(model,"virus", group=TRUE,console=TRUE,
-                main="Yield of sweetpotato\nDealt with different virus")
-#stargraph
-# Variation range: max and min
-plot(out)
-#endgraph
-out<-HSD.test(model,"virus", group=FALSE)
-print(out$comparison)
-# Old version HSD.test()
-df<-df.residual(model)
-MSerror<-deviance(model)/df
-with(sweetpotato,HSD.test(yield,virus,df,MSerror, group=TRUE,console=TRUE, main="Yield of sweetpotato. Dealt with different virus"))
-
-
-
-
-
-
-library(plyr)
-library(ggplot2)
-library(multcompView)
-
-set.seed(0)
-lev <- gl(3, 10)
-y <- c(rnorm(10), rnorm(10) + 0.1, rnorm(10) + 3)
-d <- data.frame(lev=lev, y=y)
-
-a <- aov(y~lev, data=d)
-tHSD <- TukeyHSD(a, ordered = FALSE, conf.level = 0.95)
-
-p_base <- ggplot(d, aes(x=lev, y=y)) + geom_boxplot() + geom_text(data = generate_label_df(tHSD, 'lev'), aes(x = plot.labels, y = V1, label = labels))
-
-generate_label_df <- function(HSD, flev){
-  # Extract labels and factor levels from Tukey post-hoc 
-  Tukey.levels <- HSD[[flev]][,4]
-  Tukey.labels <- multcompLetters(Tukey.levels)['Letters']
-  plot.labels <- names(Tukey.labels[['Letters']])
-  
-  # Get highest quantile for Tukey's 5 number summary and add a bit of space to buffer between    
-  # upper quantile and label placement
-  boxplot.df <- ddply(d, flev, function (x) max(fivenum(x$y)) + 0.2)
-  
-  # Create a data frame out of the factor levels and Tukey's homogenous group letters
-  plot.levels <- data.frame(plot.labels, labels = Tukey.labels[['Letters']],
-                            stringsAsFactors = FALSE)
-  
-  # Merge it with the labels
-  labels.df <- merge(plot.levels, boxplot.df, by.x = 'plot.labels', by.y = flev, sort = FALSE)
-  
-  return(labels.df)
-}

### 2018-09-04 MortalityData/HSD_us.R

@@ -1,98 +1,0 @@
-library(plyr)
-library(ggplot2)
-library(multcompView)
-library(readr)
-
-d <- read_csv("day3Mortality.csv")
-d = as.data.frame(d)
-attr(d, "spec") <- NULL
-# Note: There are 14 NC; 15 VR, VC, NR
-d = d[which(d$Treatment %in% c("None Chestnut", "None Rockrose", "Virus Chestnut", "Virus Rockrose")),]
-d$Treatment
-
-d$Treatment = as.factor(d$Treatment)
-colnames(d)[3] = "Day3Mortality"
-levels(d$Treatment) <- c('NC', 'NR', 'VC', 'VR')
-
-d$Virus = sapply(d$Treatment, function(x) substr(x, 1, 1))
-d$Diet = sapply(d$Treatment, function(x) substr(x, 2, 2))
-d$Virus = as.factor(d$Virus)
-d$Diet = as.factor(d$Diet)
-
-a <- aov(Day3Mortality~Treatment, data=d)
-
-aov.out = aov(Day3Mortality ~ Virus * Diet, data=d)
-
-tHSD <- TukeyHSD(a, ordered = TRUE, conf.level = 0.95)
-#tHSD <- TukeyHSD(aov.out, ordered = FALSE, conf.level = 0.95)
-
-generate_label_df1 <- function(HSD, flev){
-  # Extract labels and factor levels from Tukey post-hoc 
-  Tukey.levels <- HSD[[flev]][,4]
-  Tukey.labels <- multcompLetters(Tukey.levels)['Letters']
-  plot.labels <- names(Tukey.labels[['Letters']])
-  
-  # Get highest quantile for Tukey's 5 number summary and add a bit of space to buffer between    
-  # upper quantile and label placement
-  boxplot.df <- ddply(d, flev, function (x) max(fivenum(x$Day3Mortality)) + 0.05)
-  boxplot.df[2,]$V1 = 0.22
-  
-  # Create a data frame out of the factor levels and Tukey's homogenous group letters
-  plot.levels <- data.frame(plot.labels, labels = Tukey.labels[['Letters']],
-                            stringsAsFactors = FALSE)
-  
-  # Merge it with the labels
-  labels.df <- merge(plot.levels, boxplot.df, by.x = 'plot.labels', by.y = flev, sort = FALSE)
-  
-  return(labels.df)
-}
-
[…]

### 2018-09-05 Manuscript/Text/manuscript.tex
First draft revision
Cook changes

@@
 We created contrasts to determine if any protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance).
-A similar number of resistance and tolerance candidate genes were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection.
+A similar number of resistance and tolerance candidate DEGs were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection.
 We also compared the virus main effect in our study (polyandrous colonies) to that obtained in a previous study (single-drone colonies) and verified significant overlap in differential expression despite visualization methods showing differences in the noisiness levels between these two datasets.
@@
 We also show that comparing results derived from polyandrous colonies (which are typically more natural) and single-drone colonies (which usually yield more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently less artificial and less noisy.
-Altogether, we hope our results underline the merits of using data visualization techniques and multiple datasets to interpret RNA-sequencing datasets.
+Altogether, we hope this work underlines possible merits of using data visualization techniques and multiple datasets when interpreting RNA-sequencing studies.
 
@@
 The background boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG.
-To reduce overlapping of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns.
-See more information about this plotting method and the ideal visual structure of DEGs from the bigPint package.
+To reduce overplotting of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.
 
@@
 
-Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate genes and 125 ``resistance'' candidate genes.
-We again used parallel coordinate lines superimposed onto boxplots to visualize these candidate genes.
-To reduce overlapping of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns.
-Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate genes (Additional files 15 and 16).
-However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate genes, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group.
+Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate DEGs and 125 ``resistance'' candidate DEGs.
+We again used parallel coordinate lines superimposed onto boxplots to visualize these candidate DEGs.
+To reduce overplotting of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns.
+Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate DEGs (Additional files 15 and 16).
+However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate DEGs, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group.
 Indeed, we see this pattern in the associated parallel coordinate plots (Additional file 15).
-Likewise, based on the contrasts we created to obtain the `resistance'' candidate genes, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups.
+Likewise, based on the contrasts we created to obtain the `resistance'' candidate DEGs, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups.
 We do generally see these expected patterns in the associated parallel coordinate plots: While there are large outliers in the ``NC'' group, the ``NR'' replicates are no longer typically below a standardized count of zero (Additional file 16).
@@
 
-In general, the R-squared values between gene read counts and physiological measurements were low (R-squared < 0.1).
-However, some DEGs showed slightly larger R-squared values than the remaining non-DEG data.
+In general, the R-squared values between gene read counts and physiological measurements were low (R-squared $<$ 0.1).
+However, some DEG clusters showed slightly larger R-squared values than the non-DEG group (the rest of the data).
 One prominent example of this includes the first and second cluster of the virus-related DEGs (Additional file 19I).
 The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those of the rest of the data.
-As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate genes, and virus-related DEGs) and physiological measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests.
-The p-values and Bonferroni correction values for each of these tests is provided in Supplementary table 9, Additional file 1.
+The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1.
 An overall trend emerges to suggest that DEGs may have significantly larger correlation with the physiological measurements compared to non-DEGs.
-Nonetheless, almost all genes, whether differentially expressed or not, showed low R-squared values.
-It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between physiological measurements (such as low viral titers versus high viral titers, or low mortality rate versus high mortality rate).
+It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between physiological measurements.
 
@@
 We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists.
-We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality.
-Resistance candidate genes had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion.
-Tolerance candidate genes had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A).
+We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality.
+Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion.
+Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A).
 Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D.
@@
 This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}.
-These possible immunological defense mechanisms within our ``tolerance'' candidate genes and metabolic processes within our ``resistance'' candidate genes may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.
+These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.
 
@@
 Second, in studies that compared honey bee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet \cite{DiPasquale}.
-Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010).
-Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced with less diversified diets (monoculture) \cite{landscape1, brodschneider}.
-As a result, there is a need to better understand how monofloral diets affect honey bee health as a step toward mitigating the negative impact of human activity on the honey bee population.
+Third, as was previously mentioned, honey bees are now confronted with less diverse sources of pollen.
+As a result, there is a need to better understand how monofloral diets affect honey bee health.
 
@@
 Each lane produced around 13 million single-end 100 basepair reads.
+
 We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs).
@@
 The eulerr shiny application was used to construct Venn diagram overlap images \cite{euler}.
-In the main section of our chapter and in subsequent analyses, we focus on the DEG results from DESeq2 \cite{deseq2} as this pipeline was also used in the Galbraith study \cite{galbraith}.
+In the end, we focused on the DEG results from DESeq2 \cite{deseq2} as this pipeline was also used in the Galbraith study \cite{galbraith}.
+We used the independent filtering process built into the DESeq2 software that mitigates multiple comparison corrections on genes with no power rather than defining one filtering threshold.
 
@@
 
-We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith} who also addressed honey bee transcriptomic responses to virus infection.
+We compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith} who also addressed honey bee transcriptomic responses to virus infection.
 We applied the same downstream bioinformatics analyses between our count table and the count table provided in the Galbraith paper.
@@
 We used an array of visualization tools as part of our analysis.
-We first used popular tools (like the PCA plot) from the DESeq2 package.
-After that, we used multivariate visualization tools from our work-in-progress package called bigPint.
-Specifically, we used parallel coordinate plots, litre plots, and scatterplot matrices to assess the variability between the replicates and the treatments in our data and in the DEG outputs from applied models.
-We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines.
+We first used popular tools like the PCA plot \cite{pca} from the DESeq2 package.
+After that, we used multivariate visualization tools from our work-in-progress R package called bigPint.
+Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds} (cite bigPint too)) to assess the variability between the replicates and the treatments in our data.
+We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines \cite{jds} (cite bigPint too).
+
+We also used statistical graphics to better understand patterns in our DEGs.
+However, in cases of large DEG lists, these visualization tools had overplotting problems (where multiple objects are drawn on top of one another, making it impossible to detect individual values).
+To remedy this problem, we first standardized each DEG to have a mean of zero and standard deviation of unity \cite{Chandrasekhar, deSouto}.
+Then, we performed hierarchical clustering on the standardized DEGs using Ward's linkage.
+This process divided large DEG lists into smaller clusters of similar patterns, which allowed us to more efficiently visualize the different types of patterns within large DEG lists.
 
@@
 To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (tolerance), we created contrasts of interest (Table \ref{tbl:contrasts}).
-In particular, we assigned ``resistance candidate genes'' to be the ones that were upregulated in the Chestnut group within the virus infected bees but not upregulated in the Chestnut group within the non-infected bees.
-We also assigned ``tolerance candidate genes'' to be the ones that were upregulated in the Chestnut group for both the virus infected bees and non-infected bees.
+In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the Chestnut group within the virus infected bees but not upregulated in the Chestnut group within the non-infected bees.
+We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the Chestnut group for both the virus infected bees and non-infected bees.
 Our interpretation of these genes is that they represent those that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not.
@@
 We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates.
-On account of this observation, we wanted to explore whether our DEG read counts correlated with physiological metrics, including IAPV titers, Schmallenberg Virus (SBV) titers, and mortality rates.
-For this process, we considered virus main effect DEGs, ``resistance candidate'' genes, and ``tolerance candidate'' genes.
-We grouped each of these DEG sets into four hierarchical clusters.
-For each gene in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the physiological metrics across its 24 samples.
-We then used the Kruskal–Wallis test to determine if the distribution of the R-squared values in the DEG clusters significantly differed from those in the non-DEG genes.
+Hence, we wanted to explore whether our DEG read counts correlated with physiological metrics, including IAPV titers, Schmallenberg Virus (SBV) titers, and mortality rates.
+For this process, we considered virus main effect DEGs (Figure 4), ``tolerance candidate'' DEGs (Additional file 15), and ``resistance candidate'' DEGs (Additional file 16).
+For each DEG in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the physiological metrics across its 24 samples.
+We then used the Kruskal–Wallis test to determine if the distribution of the R-squared values in any of the DEG clusters significantly differed from those in the non-DEG genes (the rest of the data).
+As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate DEGs, and virus-related DEGs) and physiological measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests.
 
@@
   The datasets generated and/or analysed during the current study are available in the [NAME] repository, [PERSISTENT WEB LINK TO DATASETS].
-Include our data, Galbraith data, scripts to reproduce tables and figures.
+Include our data, Galbraith data, scripts to reproduce tables and figures (on GitHub).
 
@@
 The mortality rate data included 59 samples with 15 replicates per treatment group, except for the ``NC'' group having 14 replicates.
-ANOVA values and p-values for the statistical tests are listed in the text of the chapter.
+ANOVA values and p-values for the statistical tests are listed in the text of the paper.
 The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
@@
 The IAPV titer data included 38 samples with 10 replicates per treatment group, except for the ``NR'' group having 8 replicates.
-ANOVA values and p-values for the statistical tests are listed in the text of the chapter.
+ANOVA values and p-values for the statistical tests are listed in the text of the paper.
 The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
@@
 
-   \subsection*{Additional file 15 --- Parallel coordinate plots of the ``tolerance'' candidate genes.}    
-    Parallel coordinate plots of the 122 DEGs after hierarchical clustering of size four between the ``tolerance'' candidate genes.
+   \subsection*{Additional file 15 --- Parallel coordinate plots of the ``tolerance'' candidate DEGs.}    
+    Parallel coordinate plots of the 122 DEGs after hierarchical clustering of size four between the ``tolerance'' candidate DEGs.
 Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet.
@@
 
-   \subsection*{Additional file 16 --- Parallel coordinate plots of the ``resistance'' candidate genes.}    
-    Parallel coordinate plots of the 125 DEGs after hierarchical clustering of size four between the ``resistance'' candidate genes.
+   \subsection*{Additional file 16 --- Parallel coordinate plots of the ``resistance'' candidate DEGs.}    
+    Parallel coordinate plots of the 125 DEGs after hierarchical clustering of size four between the ``resistance'' candidate DEGs.
 Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet.

### 2018-09-09 Physiology/Tukey.R
Tukey and Fisher Mortality and IAPV

@@ -71,9 +71,10 @@ plotIAPV = ggplot(d, aes(x=Treatment, y=logIAPV)) + geom_boxplot(fill="paleturqu
 # IAPV (N vs V)
 labelDF = data.frame(plot.labels=c("N","V"), labels = c("a","b"), V1 = c(4.3, 8.7))
 plotIAPV2 = ggplot(d, aes(x=Virus, y=logIAPV)) + geom_boxplot(fill="paleturquoise2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) +theme_gray() + theme(axis.title.y=element_blank(), text=element_text(size=20)) +ylim(2,9)
+# summary(glht(mortcomp, linfct=mcp(Virus="Tukey")))
 
 # IAPV (R vs C)
 labelDF = data.frame(plot.labels=c("C","R"), labels = c("a","a"), V1 = c(4.3, 8.7))
 plotIAPV3 = ggplot(d, aes(x=Diet, y=logIAPV)) + geom_boxplot(fill="paleturquoise2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) +theme_gray() + theme(axis.title.y=element_blank(), text=element_text(size=20)) +ylim(2,9)
-
+# summary(glht(mortcomp, linfct=mcp(Virus="Tukey")))
 

### 2018-09-09 Manuscript/Text/manuscript.tex
Ethics statement
Tukey versus Bonferroni

@@
 We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus.
-Namely, bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Tukey HSD, p $<$ 1.18e-03), and bees fed Chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Tukey HSD, p $<$ 4.80e-03) (Figure 1).
+Namely, bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Tukey HSD, p $<$ 1.31e-03), and bees fed Chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Tukey HSD, p $<$ 4.78e-03) (Figure 1).
 
@@
 The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant.
-We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p $<$ 5.44e-03).
+We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p $<$ 5.65e-03).
 However, bees fed Chestnut pollen did not have significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p = 1.11e-01).
@@
 \section*{Ethics approval and consent to participate}
-  Not applicable.
+  All honey bees used in this work were sampled in the United States, and no ethical use approval is required for this species in this country.
 

### 2018-09-09 Physiology/Bonf.R

@@ -1,11 +1,4 @@
 library(agricolae)
-data(sweetpotato)
-model<-aov(yield~virus, data=sweetpotato)
-out <- LSD.test(model,"virus", p.adj="bonferroni")
-# Variation range: max and min
-plot(out)
-
-
 
 d <- read_csv("day3Mortality.csv")
 d = as.data.frame(d)
@@ -29,7 +22,7 @@ out <- LSD.test(model,"Treatment", p.adj="bonferroni")
 # Variation range: max and min
 plot(out, ylab="Mortality")
 # Used to print the actual p-values
-pvals <- LSD.test(model,"Treatment", p.adj="bonferroni", group=FALSE)
+anova(model)
 
 # Mortality (N vs V)
 model<-aov(Day3Mortality~Virus, data=d)
@@ -37,7 +30,7 @@ out <- LSD.test(model,"Virus", p.adj="bonferroni")
 # Variation range: max and min
 plot(out, ylab="Mortality")
 # Used to print the actual p-values
-pvals <- LSD.test(model,"Virus", p.adj="bonferroni", group=FALSE)
+anova(model)
 
 # Mortality (R vs C)
 model<-aov(Day3Mortality~Diet, data=d)
@@ -45,7 +38,7 @@ out <- LSD.test(model,"Diet", p.adj="bonferroni")
 # Variation range: max and min
 plot(out, ylab="Mortality")
 # Used to print the actual p-values
-pvals <- LSD.test(model,"Diet", p.adj="bonferroni", group=FALSE)
+anova(model)
 
 ###### Plot IAPV Values ########
 
@@ -72,7 +65,7 @@ out <- LSD.test(model,"Treatment", p.adj="bonferroni")
 # Variation range: max and min
 plot(out, ylab="Log IAPV")
 # Used to print the actual p-values
-pvals <- LSD.test(model,"Treatment", p.adj="bonferroni", group=FALSE)
+anova(model)
 
 # IAPV (N vs V)
 model<-aov(logIAPV~Virus, data=d)
@@ -80,7 +73,7 @@ out <- LSD.test(model,"Virus", p.adj="bonferroni")
 # Variation range: max and min
 plot(out, ylab="Log IAPV")
 # Used to print the actual p-values
-pvals <- LSD.test(model,"Virus", p.adj="bonferroni", group=FALSE)
+anova(model)
 
 # IAPV (R vs C)
 model<-aov(logIAPV~Diet, data=d)
[…]

### 2018-09-09 Physiology/Tukey.R

@@ -24,6 +24,7 @@ d$Diet = as.factor(d$Diet)
 mortcomp = lme(Day3Mortality ~ Diet*Virus, data=d, random = ~1|Experiment) 
 mortcomp2 = lme(Day3Mortality ~ Treatment, data=d, random = ~1|Experiment) 
 anova(mortcomp)
+anova(mortcomp2)
 summary(glht(mortcomp2, linfct=mcp(Treatment="Tukey")))
 
 # Mortality (All)
@@ -62,6 +63,7 @@ d$Diet = as.factor(d$Diet)
 mortcomp = lme(logIAPV ~ Diet*Virus, data=d, random = ~1|Experiment) 
 mortcomp2 = lme(logIAPV ~ Treatment, data=d, random = ~1|Experiment)
 anova(mortcomp)
+anova(mortcomp2)
 summary(glht(mortcomp2, linfct=mcp(Treatment="Tukey")))
 
 # IAPV (All)

### 2018-09-15 Manuscript/Text/manuscript.tex
Adam's edits

@@
 We created contrasts to determine if any protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance).
-A similar number of resistance and tolerance candidate DEGs were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection.
+A similar number of resistance and tolerance candidate differentially expressed genes were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection.
 We also compared the virus main effect in our study (polyandrous colonies) to that obtained in a previous study (single-drone colonies) and verified significant overlap in differential expression despite visualization methods showing differences in the noisiness levels between these two datasets.
@@
 
-Pollen is the main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}.
+Pollen is a main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}.
 At the individual level, pollen supplies most of the nutrients necessary for physiological development \cite{brodschneider} and is believed to have considerable impact on longevity \cite{longevity}.
@@
 Viral infection was a comparatively minor problem in honey bees until the last century when Varroa destructor (an ectoparasitic mite) spread worldwide \cite{miteSpread}.
-This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits cocktails of viruses, and supports replication of certain viruses \cite{miteVirus, miteVirus2, miteVirus3}.
+This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of certain viruses \cite{miteVirus, miteVirus2, miteVirus3}.
 More than 20 honey bee viruses have been identified \cite{numVirus}.
 One of these viruses that has been linked to honey bee decline is Israeli Acute Paralysis Virus (IAPV).
-A positive-sense RNA virus of the Dicistroviridae family \cite{fam}, IAPV causes infected honey bees to display shivering wings, decreased locomotion, muscle spams, and paralysis, and 80\% of caged infected adult honey bees die prematurely \cite{symptoms}.
+A positive-sense RNA virus of the Dicistroviridae family \cite{fam}, IAPV causes infected honey bees to display shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}.
 IAPV has demonstrated higher infectious capacities than other honey bee viruses in certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.
@@
 This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a ``buffer'' against other stressors \cite{AdamTothReview}.
-As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and IAPV infection effects \cite{galbraith} in honey bees.
+As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees.
 As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense.
@@
 We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}.
-As RNA-sequencing data can be highly noisy, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies.
+As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies.
 Importantly, we use an in-depth data visualization approach to explore and corroborate our data, and suggest such an approach can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.
@@
 
-We reanalyzed our previously published dataset with a subset more relevant to our RNA-sequencing approaches in the current study that have a more focused question regarding diet quality.
-We briefly show it again here to inform the RNA-sequencing comparison because we reduced the number of treatments (from eight to four) from the original published data \cite{adamInt} as a means to focus on diet quality effects.
+We reanalyzed our previously published dataset with a subset that focuses on diet quality and is more relevant to the current study.
+We briefly show it again here to inform the RNA-sequencing comparison because we reduced the number of treatments from the original published data (from eight to four) \cite{adamInt} as a means to focus on diet quality effects.
 
@@
 
-We observed a substantially larger number of DEGs in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1).
+We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1).
 In the diet factor, there were more Chestnut-upregulated DEGs (\textit{n} = 1033) than Rockrose-upregulated DEGs (\textit{n} = 881).
@@
 We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays (Figure 3).
-We do see that the first replicate of the virus group appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that this trend in the data that we saw in the PCA plot carried through into the DEG results.
+We do see that the first replicate of the virus group (V.1) appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that this trend in the data that we saw in the PCA plot carried through into the DEG results.
 In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4).
@@
 
-In general, the R-squared values between gene read counts and physiological measurements were low (R-squared $<$ 0.1).
+In general, the R-squared values between gene read counts and pathogen response measurements were low (R-squared $<$ 0.1).
 However, some DEG clusters showed slightly larger R-squared values than the non-DEG group (the rest of the data).
@@
 The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1.
-An overall trend emerges to suggest that DEGs may have significantly larger correlation with the physiological measurements compared to non-DEGs.
-It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between physiological measurements.
+An overall trend emerges to suggest that DEGs may have significantly larger correlation with the pathogen response measurements compared to non-DEGs.
+It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between pathogen response measurements.
 
@@
 
-Somewhat surprisingly, the transcriptomic response to virus infection in our experiment was fairly limited.
+While many other insect systems show relatively low transcriptional responses to virus infection, previous work on honey bees has revealed hundreds of DEGs \cite{galbraith}.
+However, the transcriptomic response to virus infection in our experiment was fairly limited.
 We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as genes related to transcriptional regulation and muscle contraction.
@@
 For instance, our comparison between the Galbraith data (single-drone colonies) and our data (polyandrous colonies) was limited by numerous extraneous variables between these studies.
-In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on one-day old worker honey bees that were fed sugar and MegaBee diet, whereas our study focused on adult worker honey bees that were fed monofloral diets.
+In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on one-day old worker honey bees that were fed sugar and artificial pollen diet, whereas our study focused on adult worker honey bees that were fed bee-collected monofloral diets.
 Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group.
@@
 
-Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt}.
+Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}.
 The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report \cite{adamInt}.
@@
 
-Fifteen cages per treatment were originally sampled.
-Six live honey bees from each cage were randomly selected 36 hours post inoculation and placed into tubes \cite{carrillo}.
+Fifteen cages per treatment were originally produced for monitoring of mortality.
+From these, six live honey bees were randomly selected from each cage 36 hours post inoculation and placed into tubes \cite{carrillo}.
 Tubes were kept on dry ice and then transferred into a -80C freezer until processing.
-Eight cages were randomly selected from the original 15 cages, and 2 honey bees per cage were randomly selected from the original six live honey bees per cage.
-Whole body RNA from each pool of two honey bees were extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment.
+From the fifteen possible cages, eight were randomly selected for RNA-sequencing.
+From these eight cages, two of the honey bees per cage were randomly selected from the original six live honey bees per cage.
+These two bees were combined to form a pooled sample representing the cage.
+Whole body RNA from each pool was extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment.
 Samples were suspended in water to 200-400 ng/$\mu$l.
@@
 We compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith} who also addressed honey bee transcriptomic responses to virus infection.
-We applied the same downstream bioinformatics analyses between our count table and the count table provided in the Galbraith paper.
-When we applied our bioinformatics pipeline to the Galbraith count table, we obtained different differential expression counts compared to the results published in the Galbraith paper.
+We applied the same downstream bioinformatics analyses between our count table and the count table provided in the Galbraith study.
+When we applied our bioinformatics pipeline to the Galbraith count table, we obtained different differential expression counts compared to the results published in the Galbraith study.
 However, there was substantial overlap and we considered this justification to use the differential expression list we obtained in order to keep the downstream bioinformatics analyses as similar as possible between the two datasets (Additional file 17).
@@
 As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}.
+We note that the difference between these studies is even greater than this as we used naturally mated honey bees from 15 different colonies.
 We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates.
@@
 Then, we performed hierarchical clustering on the standardized DEGs using Ward's linkage.
-This process divided large DEG lists into smaller clusters of similar patterns, which allowed us to more efficiently visualize the different types of patterns within large DEG lists.
+This process divided large DEG lists into smaller clusters of similar patterns, which allowed us to more efficiently visualize the different types of patterns within large DEG lists (see Figures 3 and 4 for examples).
 
@@
 In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the Chestnut group within the virus infected bees but not upregulated in the Chestnut group within the non-infected bees.
+Our interpretation of these genes is that they represent those that are only activated in infected bees that are fed a high quality diet.
 We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the Chestnut group for both the virus infected bees and non-infected bees.
@@
 We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates.
-Hence, we wanted to explore whether our DEG read counts correlated with physiological metrics, including IAPV titers, Schmallenberg Virus (SBV) titers, and mortality rates.
+Hence, we wanted to explore whether our DEG read counts correlated with pathogen response metrics, including IAPV titers, Schmallenberg Virus (SBV) titers, and mortality rates.
 For this process, we considered virus main effect DEGs (Figure 4), ``tolerance candidate'' DEGs (Additional file 15), and ``resistance candidate'' DEGs (Additional file 16).
-For each DEG in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the physiological metrics across its 24 samples.
+For each DEG in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the pathogen response metrics across its 24 samples.
 We then used the Kruskal–Wallis test to determine if the distribution of the R-squared values in any of the DEG clusters significantly differed from those in the non-DEG genes (the rest of the data).
-As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate DEGs, and virus-related DEGs) and physiological measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests.
+As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate DEGs, and virus-related DEGs) and pathogen response measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests.
 
@@
 \section*{Author's contributions}
-  Each author must write their authorsip contributions.
+  LR performed the bioinformatic and statistical analyses, produced the figures and tables, and drafted the manuscript.
+AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq.
     %All authors read and approved the final manuscript.
@@
 \section*{Acknowledgements}
-  Who/what should we mention here? \ldots
+  We would like to thank Giselle Narvaez for assisting with cage experiments.
 
@@
 ``C'' represents Chestnut diet, ``R'' represents Rockrose diet, ``V'' represents virus-inoculated, and ``N'' represents control non-inoculated.
-    \textbf{Table 9:} Kruskal-Wallis p-value and Bonferroni corrections for the 36 combinations of DEG lists, physiological metrics, and cluster number. (XLS).
+    \textbf{Table 9:} Kruskal-Wallis p-value and Bonferroni corrections for the 36 combinations of DEG lists, pathogen response metrics, and cluster number. (XLS).
 
@@
 
-  \subsection*{Additional file 19 --- Analysis of correlation between DEG read counts and physiological metrics}
-  Distribution of R-squared values for DEG cluster read counts and physiological metrics.
+  \subsection*{Additional file 19 --- Analysis of correlation between DEG read counts and pathogen response metrics}
+  Distribution of R-squared values for DEG cluster read counts and pathogen response metrics.
 Columns left to right: SBV titers, mortality rates, and IAPV titers.

### 2018-09-16 Manuscript/Text/manuscript.tex
LME and pairwise BH

@@
 
-As shown in Figure 1, mortality rates of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.30e-05).
-The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 1.00e-05) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant.
+As shown in Figure 1, mortality rates of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.34e-05).
+The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 7.04e-06) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant.
 We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus.
-Namely, bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Tukey HSD, p $<$ 1.31e-03), and bees fed Chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Tukey HSD, p $<$ 4.78e-03) (Figure 1).
+Namely, bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed Chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).
 
-As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 1.96e-03).
+As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03).
 The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant.
-We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p $<$ 5.65e-03).
-However, bees fed Chestnut pollen did not have significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p = 1.11e-01).
+We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03).
+However, bees fed Chestnut pollen did not have significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02).
 Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not do so (Figure 2).
@@
 As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}.
-We note that the difference between these studies is even greater than this as we used naturally mated honey bees from 15 different colonies.
+We note that the difference between these studies may be even greater than this as we used naturally mated honey bees from 15 different colonies.
 We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates.

### 2018-09-16 PathogenResponse/BH.R

@@ -1,0 +1,108 @@
+library(plyr)
+library(ggplot2)
+library(multcompView)
+library(readr)
+library(multcomp)
+library(nlme)
+
+################# Mortality ################# 
+
+d <- read_csv("day3Mortality.csv")
+d = as.data.frame(d)
+attr(d, "spec") <- NULL
+# Note: There are 14 NC; 15 VR, VC, NR
+d = d[which(d$Treatment %in% c("None Chestnut", "None Rockrose", "Virus Chestnut", "Virus Rockrose")),]
+d$Treatment
+
+d$Treatment = as.factor(d$Treatment)
+colnames(d)[3] = "Day3Mortality"
+levels(d$Treatment) <- c('NC', 'NR', 'VC', 'VR')
+
+d$Virus = sapply(d$Treatment, function(x) substr(x, 1, 1))
+d$Diet = sapply(d$Treatment, function(x) substr(x, 2, 2))
+d$Virus = as.factor(d$Virus)
+d$Diet = as.factor(d$Diet)
+
+# Main effects and interactive terms
+lmeMortMain <- lme(Day3Mortality ~ Diet*Virus, data = d, random = ~1|Experiment)
+anova(lmeMortMain)[["p-value"]]
+
+# Treatment
+lmeMortTreatment <- lme(Day3Mortality ~ Treatment, data = d, random = ~1|Experiment)
+anova(lmeMortTreatment)[["p-value"]]
+
+# Pairwise (with BH correction)
+d$Treatment <- relevel(d$Treatment, ref = "NC")
+lmemort1 = lme(Day3Mortality ~ Treatment, data=d, random = ~1|Experiment) 
+pval1 = summary(lmemort1)$tTable[,5][2:4]
+d$Treatment <- relevel(d$Treatment, ref = "NR")
+lmemort2 = lme(Day3Mortality ~ Treatment, data=d, random = ~1|Experiment) 
+pval2 = summary(lmemort2)$tTable[,5][3:4]
+d$Treatment <- relevel(d$Treatment, ref = "VC")
+lmemort3 = lme(Day3Mortality ~ Treatment, data=d, random = ~1|Experiment) 
+pval3 = summary(lmemort3)$tTable[,5][4]
+mortpval = c(pval1, pval2, pval3)
+p.adjust(mortpval, "BH")
+
+# Plot Mortality (All)
[…]

### 2018-09-16 Physiology/Bonf.R

@@ -1,85 +1,0 @@
-library(agricolae)
-
-d <- read_csv("day3Mortality.csv")
-d = as.data.frame(d)
-attr(d, "spec") <- NULL
-# Note: There are 14 NC; 15 VR, VC, NR
-d = d[which(d$Treatment %in% c("None Chestnut", "None Rockrose", "Virus Chestnut", "Virus Rockrose")),]
-d$Treatment
-
-d$Treatment = as.factor(d$Treatment)
-colnames(d)[3] = "Day3Mortality"
-levels(d$Treatment) <- c('NC', 'NR', 'VC', 'VR')
-
-d$Virus = sapply(d$Treatment, function(x) substr(x, 1, 1))
-d$Diet = sapply(d$Treatment, function(x) substr(x, 2, 2))
-d$Virus = as.factor(d$Virus)
-d$Diet = as.factor(d$Diet)
-
-# Mortality (All)
-model<-aov(Day3Mortality~Treatment, data=d)
-out <- LSD.test(model,"Treatment", p.adj="bonferroni")
-# Variation range: max and min
-plot(out, ylab="Mortality")
-# Used to print the actual p-values
-anova(model)
-
-# Mortality (N vs V)
-model<-aov(Day3Mortality~Virus, data=d)
-out <- LSD.test(model,"Virus", p.adj="bonferroni")
-# Variation range: max and min
-plot(out, ylab="Mortality")
-# Used to print the actual p-values
-anova(model)
-
-# Mortality (R vs C)
-model<-aov(Day3Mortality~Diet, data=d)
-out <- LSD.test(model,"Diet", p.adj="bonferroni")
-# Variation range: max and min
-plot(out, ylab="Mortality")
-# Used to print the actual p-values
-anova(model)
-
-###### Plot IAPV Values ########
-
-d <- read_delim("logIAPV.csv", "\t", escape_double = FALSE, trim_ws = TRUE)
-d = as.data.frame(d)
-attr(d, "spec") <- NULL
-# Note: There are 14 NC; 15 VR, VC, NR
-
-d = d[which(d$treatment %in% c("None Chestnut", "None Rockrose", "Virus Chestnut", "Virus Rockrose")),]
-
-d$treatment = as.factor(d$treatment)
-colnames(d)[3] = "logIAPV"
-levels(d$treatment) <- c('NC', 'NR', 'VC', 'VR')
-colnames(d)[1] <- "Treatment"
-
[…]

### 2018-09-17 Manuscript/Text/manuscript.tex
Adam additional edits

@@
 
-Data visualization is a useful method to identify noise and robustness in RNA-sequencing data.
+Data visualization is a useful method to identify noise and robustness in RNA-sequencing data \cite{edger}.
 In this study, we used extensive data visualization to improve the interpretation of our RNA-sequencing results.
@@
 This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}.
+In insects, melanization has been shown to be toxic to parasites, bacteria, fungi, and lately viruses \cite{melanization}.
 These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.
@@
 
+\subsection*{Pathogen response}
+
 Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}.
-The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report \cite{adamInt}.
+A linear mixed effects model was used to relate the mortality rates and IAPV titers to the main and interaction effects of the diet and virus factors.
+The model was fitted to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''.
+Post-hoc pairwise comparisons of the four (diet and virus combination) treatment groups were performed and Benjamini-Hochberg adjusted p-values were calculated to limit familywise Type I error rates \cite{bh}.
 
@@
 ANOVA values and p-values for the statistical tests are listed in the text of the paper.
-The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
+The letters above the bars represent significant differences with a confidence level of 95\%.}
 \end{figure}

### 2018-09-24 Manuscript/Text/manuscript.tex
Bonning corrections

@@
 
-\title{Gene expression responses to diet quality and viral infection in Apis mellifera}
+\title{Transcriptomic responses to diet quality and viral infection in Apis mellifera}
 
@@
 ]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
+\author[
+   addressref={aff6},
+   email={bbonning@ufl.edu}
+]{\inits{BB}\fnm{Bryony C.} \snm{Bonning}}
 \author[
@@
 
+\address[id=aff6]{%
+  \orgname{Department of Entomology and Nematology, University of Florida},
+  \city{Gainesville},   
+  \state{FL}
+  \postcode{32611},
+  \cny{USA}
+}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Poor nutrition and viral infection are two environmental stressors that pose heightened dangers to honey bee health.
-In this study, we used RNA-sequencing to examine how monofloral diets and Israeli Acute Paralysis Virus inoculation influence gene expression patterns in honey bees.
+In this study, we used RNA-sequencing to examine how monofloral diets and Israeli acute paralysis virus inoculation influence gene expression patterns in honey bees.
 
@@
 More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}.
-As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications regarding agricultural sustainability \cite{ccd4}.
+As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications for agricultural sustainability \cite{ccd4}.
 
 Honey bee declines have been associated with several factors, including pesticide use, parasites, pathogens, habitat loss, and poor nutrition \cite{factors, factors2}.
-Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honey bees in interactive fashions as the environment changes \cite{interacting}.
+Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honey bees in an interactive fashion as the environment changes \cite{interacting}.
 Nutrition and viral infection are two broad factors that pose heightened dangers to honey bee health in response to recent environmental changes.
@@
 Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honey bees.
-In particular, honey bees are confronted with less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}.
+In particular, honey bees are confronted with a less diverse selection of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}.
 Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}.
@@
 
-Viral infection was a comparatively minor problem in honey bees until the last century when Varroa destructor (an ectoparasitic mite) spread worldwide \cite{miteSpread}.
-This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of certain viruses \cite{miteVirus, miteVirus2, miteVirus3}.
+Viral infection was a comparatively minor problem in honey bees until the last century when \textit{Varroa destructor} (an ectoparasitic mite) spread worldwide \cite{miteSpread}.
+This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}.
 More than 20 honey bee viruses have been identified \cite{numVirus}.
-One of these viruses that has been linked to honey bee decline is Israeli Acute Paralysis Virus (IAPV).
-A positive-sense RNA virus of the Dicistroviridae family \cite{fam}, IAPV causes infected honey bees to display shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}.
-IAPV has demonstrated higher infectious capacities than other honey bee viruses in certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.
+One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV).
+A positive-sense RNA virus of the family Dicistroviridae \cite{fam}, IAPV infection causes shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}.
+IAPV has demonstrated higher infectious capacities than other honey bee viruses under certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.
 Its role in the rising phenomenon of ``Colony Collapse Disorder'' (in which the majority of worker bees disappear from a hive) remains unclear: It has been implicated in some studies \cite{iapvCCD, iapvCCD2} but not in other studies \cite{ccd1, fam, iapvCCD3}.
@@
 We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show the importance of the combined effects of both diet quality and virus infection.
-Specifically, high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen \cite{adamInt}.
+Specifically, ingestion by honey bees of high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen \cite{adamInt}.
 
@@
 This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a ``buffer'' against other stressors \cite{AdamTothReview}.
+
 As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees.
+Honey bee transcriptomic studies have found that pollen nutrition upregulates genes involved in macromolecule metabolism, longetivity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
+Insect gene expression studies have implicated RNA silencing, autophagy, JAK/STAT, Toll, and IMD as antiviral pathways for a range of viral infections, including dicistrovirus infections \cite{galbraith, Avadhanula, fewDEGs2, kemp, costa}.
+Transcriptional pausing is also believed to be imperative for early antiviral immunity in many insects \cite{resTol11}.
+See \cite{cherryReview} for a review of known antiviral mechanisms in insect models.
+
 As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense.
-
 In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure).
@@
 The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant.
-We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03).
-However, bees fed Chestnut pollen did not have significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02).
-Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not do so (Figure 2).
+We compared IAPV titers  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03).
+However, bees fed Chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02).
+Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not (Figure 2).
 
@@
 We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1).
-In the diet factor, there were more Chestnut-upregulated DEGs (\textit{n} = 1033) than Rockrose-upregulated DEGs (\textit{n} = 881).
+In the diet factor, more DEGs were upregulated in the more-nutritious Chestnut group (\textit{n} = 1,033) than in the less-nutritious Rockrose group (\textit{n} = 881).
 In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5).
@@
 We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset.
-Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may better separate the infected and uninfected honey bees better than our dataset (Additional file 2).
+Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level.
+Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2).
 We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots.
@@
 We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies.
-The background boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG.
+The background side-by-side boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG.
 To reduce overplotting of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.
@@
 Finally, we looked at scatterplot matrices to assess the DEGs.
-We created standardized scatterplot matrices for each of the four clusters (Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10).
+We created standardized scatterplot matrices for each of the four clusters (from Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10).
 We also created standardized scatterplot matrices for our data.
@@
 As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14).
-We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: Deviating more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.
+We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: They deviated more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.
 
@@
 Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate DEGs and 125 ``resistance'' candidate DEGs.
-We again used parallel coordinate lines superimposed onto boxplots to visualize these candidate DEGs.
+We again used parallel coordinate lines superimposed onto side-by-side boxplots to visualize these candidate DEGs.
 To reduce overplotting of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns.
@@
 However, some DEG clusters showed slightly larger R-squared values than the non-DEG group (the rest of the data).
-One prominent example of this includes the first and second cluster of the virus-related DEGs (Additional file 19I).
-The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those of the rest of the data.
+One prominent example of this includes the first and second cluster of the virus-related DEGs and their correlation with IAPV titers (Additional file 19I).
+The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those in the rest of the data.
 The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1.
@@
 
-While many other insect systems show relatively low transcriptional responses to virus infection, previous work on honey bees has revealed hundreds of DEGs \cite{galbraith}.
-However, the transcriptomic response to virus infection in our experiment was fairly limited.
+While some insect systems have shown relatively low transcriptional responses to dicistrovirus infection \cite{fewDEGs, fewDEGs2}, previous work on honey bees has revealed many hundreds of DEGs \cite{galbraith}.
+Despite this, the transcriptomic response to virus infection in our experiment was fairly limited.
 We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as genes related to transcriptional regulation and muscle contraction.
@@
 We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality.
-Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion.
-Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A).
+Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion (Figure 5B).
+Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription (Figure 5A).
 Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D.
 melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}.
-Moreover, circulating haemocytes in insects encapsulate and nodulate pathogens by forming a barrier between the pathogen and the host tissues.
-This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}.
-In insects, melanization has been shown to be toxic to parasites, bacteria, fungi, and lately viruses \cite{melanization}.
 These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.
@@
 In future related studies, it may be informative to use tissue-specific methods.
-Recent evidence has suggested that RNA-sequencing approaches toward composite structures in honey bees leads to false negatives, implying that genes strongly differentially expressed in particular structures may not reach significance within the composite structure.
-On a similar note, recent studies have also found that within a composite extraction, structures therein may contain opposite patterns of differential expression.
-We can provide more detailed answers to our original transcriptomic questions if we were to repeat this same experimental design only now at a more refined tissue level \cite{tissueLevel}.
+Previous work has shown that even though IAPV replication occurs in all honey bee tissues, it localizes more in gut and nerve tissues and in the hypopharyngeal glands.
+Likewise, the highest IAPV titers have been observed in gut tissues \cite{winter}.
+Recent evidence has suggested that RNA-sequencing approaches toward composite structures in honey bees leads to false negatives, implying that genes strongly differentially expressed in particular structures may not reach significance within the composite structure \cite{tissueLevel}.
+These studies have also found that within a composite extraction, structures therein may contain opposite patterns of differential expression.
+We can provide more detailed answers to our original transcriptomic questions if we were to repeat this same experimental design only now at a more refined tissue level.
 Another future direction related to this work would be to integrate multiple omics datasets to investigate monofloral diet quality and IAPV infection in honey bees.
@@
 The model was fitted to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''.
+A random (intercept) effect for experimental setup was included in the model.
 Post-hoc pairwise comparisons of the four (diet and virus combination) treatment groups were performed and Benjamini-Hochberg adjusted p-values were calculated to limit familywise Type I error rates \cite{bh}.
@@
 Samples were suspended in water to 200-400 ng/$\mu$l.
-All samples were then tested on a Bioanalyzer at the DNA core facility to ensure quality (RIN $>$ 8).
+All samples were then tested on a Bioanalyzer at the Iowa State University DNA Facility to ensure quality (RIN $>$ 8).
 
@@
 We used an array of visualization tools as part of our analysis.
-We first used popular tools like the PCA plot \cite{pca} from the DESeq2 package.
-After that, we used multivariate visualization tools from our work-in-progress R package called bigPint.
+We first used well-known tools like the PCA plot \cite{pca} from the DESeq2 package.
+After that, we used lesser-known multivariate visualization tools from our work-in-progress R package called bigPint.
 Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds} (cite bigPint too)) to assess the variability between the replicates and the treatments in our data.
@@
   LR performed the bioinformatic and statistical analyses, produced the figures and tables, and drafted the manuscript.
+BB conceptualized the study and critically revised the manuscript.
 AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq.
@@
 \begin{figure}[h!]
-\caption{\csentence{IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups.}
-Left to right: IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups.
+\caption{\csentence{IAPV titers for the four treatment groups, two virus groups, and two diet groups.}
+Left to right: IAPV titers for the four treatment groups, two virus groups, and two diet groups.
 ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen.
@@
 \caption{\csentence{Gene ontology analysis results for the 122 DEGs related to our ``tolerance'' hypothesis and for the 125 DEGs related to our ``resistance'' hypothesis.}
-GO analysis results for the 122 DEGs related to our ``tolerance'' hypothesis (A) and for the 125 DEGs related to our ``resistance'' hypothesis (B).}
+GO analysis results for the 122 DEGs related to our ``tolerance'' hypothesis (A) and for the 125 DEGs related to our ``resistance'' hypothesis (B).
+The color and size of the circles both represent the number of genes in that ontology.
+The x-axis and y-axis are organized by SimRel, a semantic similarity metric \cite{semantic}.}
 \end{figure}

### 2018-09-24 PathogenResponse/BH.R

@@ -25,11 +25,13 @@ d$Diet = as.factor(d$Diet)
 
 # Main effects and interactive terms
 lmeMortMain <- lme(Day3Mortality ~ Diet*Virus, data = d, random = ~1|Experiment)
-anova(lmeMortMain)[["p-value"]]
+anova(lmeMortMain)
+anova(lmeMortMain)[["p-value"]] # Obtain more precise p-values
 
 # Treatment
 lmeMortTreatment <- lme(Day3Mortality ~ Treatment, data = d, random = ~1|Experiment)
-anova(lmeMortTreatment)[["p-value"]]
+anova(lmeMortTreatment)
+anova(lmeMortTreatment)[["p-value"]] # Obtain more precise p-values
 
 # Pairwise (with BH correction)
 d$Treatment <- relevel(d$Treatment, ref = "NC")
@@ -42,7 +44,7 @@ d$Treatment <- relevel(d$Treatment, ref = "VC")
 lmemort3 = lme(Day3Mortality ~ Treatment, data=d, random = ~1|Experiment) 
 pval3 = summary(lmemort3)$tTable[,5][4]
 mortpval = c(pval1, pval2, pval3)
-p.adjust(mortpval, "BH")
+p.adjust(mortpval, "BH") # Obtain p-values for NC_NR, NC_VC, NC_VR, NR_VC, NR_VR, VC_VR
 
 # Plot Mortality (All)
 labelDF = data.frame(plot.labels=c("NR","VC","VR","NC"), labels = c("ab","bc","c","a"), V1 = c(0.22, 0.479, 0.793, 0.164))
@@ -75,11 +77,13 @@ d$Diet = as.factor(d$Diet)
 
 # Main effects and interactive terms
 lmeMortMain <- lme(logIAPV ~ Diet*Virus, data = d, random = ~1|Experiment)
-anova(lmeMortMain)[["p-value"]]
+anova(lmeMortMain)
+anova(lmeMortMain)[["p-value"]] # Obtain more precise p-values
 
 # Treatment
 lmeMortTreatment <- lme(logIAPV ~ Treatment, data = d, random = ~1|Experiment)
-anova(lmeMortTreatment)[["p-value"]]
+anova(lmeMortTreatment)
+anova(lmeMortTreatment)[["p-value"]] # Obtain more precise p-values
 
 # Pairwise (with BH correction)
[…]

### 2018-10-23 Manuscript/Text/manuscript.tex
Jimena corrections

@@
 The majority of these genes were over-represented for nutrient signaling (insulin resistance) and immune response (Notch signaling and JaK-STAT pathways).
-Somewhat unexpectedly, the transcriptomic response to viral infection was fairly limited.
+In our experimental conditions, the transcriptomic response to viral infection was fairly limited.
 We only found 43 transcripts to be differentially expressed, some with known immune functions (argonaute-2), transcriptional regulation, and muscle contraction.
@@
 We also show that comparing results derived from polyandrous colonies (which are typically more natural) and single-drone colonies (which usually yield more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently less artificial and less noisy.
-Altogether, we hope this work underlines possible merits of using data visualization techniques and multiple datasets when interpreting RNA-sequencing studies.
+Altogether, this work underlines the merits of using data visualization techniques and multiple datasets to interpret RNA-sequencing studies.
 
@@
 
-Viral infection was a comparatively minor problem in honey bees until the last century when \textit{Varroa destructor} (an ectoparasitic mite) spread worldwide \cite{miteSpread}.
+Viral infection was a comparatively minor problem in honey bees until the last century when the ectoparasitic varroa mite (\textit{Varroa destructor}) spread worldwide \cite{miteSpread}.
 This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}.
 More than 20 honey bee viruses have been identified \cite{numVirus}.
-One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV).
-A positive-sense RNA virus of the family Dicistroviridae \cite{fam}, IAPV infection causes shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}.
+One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV), a positive-sense RNA virus of the family Dicistroviridae \cite{fam}.
+IAPV infection causes shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}.
 IAPV has demonstrated higher infectious capacities than other honey bee viruses under certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.
@@
 
-Although there is growing interest in how viruses and diet quality affect the health and sustainability of honey bees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honey bees \cite{intNV, intNV2, intNV3}.
+Although there is growing interest in how viruses and diet quality affect the health and sustainability of honey bees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honey bees \cite{intNV, intNV2, intNV3, interactingDV, interactingDV2}.
 We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show the importance of the combined effects of both diet quality and virus infection.
@@
 
-Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality.
+Following up on these findings, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality.
 For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on vigor (tolerance) \cite{resTol1}.
@@
 As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees.
-Honey bee transcriptomic studies have found that pollen nutrition upregulates genes involved in macromolecule metabolism, longetivity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
-Insect gene expression studies have implicated RNA silencing, autophagy, JAK/STAT, Toll, and IMD as antiviral pathways for a range of viral infections, including dicistrovirus infections \cite{galbraith, Avadhanula, fewDEGs2, kemp, costa}.
-Transcriptional pausing is also believed to be imperative for early antiviral immunity in many insects \cite{resTol11}.
-See \cite{cherryReview} for a review of known antiviral mechanisms in insect models.
+Honey bee transcriptomic studies have found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
 
-As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense.
+On the other hand, gene expression studies in insect models such as Drosophila and mosquito have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway and transcriptional pausing \cite{cherryReview, swevers}.
+It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}.
+In fact, every virus-host interaction has its own particularities derived of the diverse ways of replication and infection cycle evolved by different viruses.
+In this sense, an intricate set of pro- and anti-virus host factors are involved in each case \cite{cherryReview, swevers}.
+For example, dicistroviruses Drosophila C virus (DCV) and cricket paralysis virus (CrPV) infection requires ribosomal proteins RpS6, RpL19 and RACK1, but such proteins are dispensable during infection by the rhabdovirus vesicular stomatitis virus (VSV) in Drosophila.
+In contrast, the autophagy pathway has proven to be an effective antiviral strategy against VSV, but has no role limiting DCV or other positive-sense RNA viruses \cite{swevers}.
+For an extensive up-to-date review of known antiviral and proviral factors for specific viruses in Drosophila see \cite{swevers}.
+Model insect studies have shed light on how honey bees respond to viruses because the function of many genes relies on the characterization of orthologue genes.
+Nonetheless, recent transcriptomic studies in honey bees have improved our understanding of compatible and incompatible virus-bee interactions.
+Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g.
+RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}.
+Added to these, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken}.
+This path has been studied recently in more depth using transcriptomics, which found known immune factors but also new ones for which indispensability in antiviral response was determined by silencing \cite{Brutscher}.
+Similar to Drosophila examples, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed.
+This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs.
+deformed wing virus (DWV) \cite{Ryabov}.
+Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.
+In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level.
+Chen et al.
+2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}.
+Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood.
+The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}.
+IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}.
+These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors.
+The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy to mention that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.
+
+As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense \cite{interactingDV}.
 In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure).
+For our diet factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.).
+Rockrose pollen is generally considered less nutritious than chestnut pollen because it contains smaller amounts of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}.
 We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section).
@@
 We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}.
+While our study examines honey bees from polyandrous colonies, the Galbraith study examined honey bees from single-drone colonies.
+As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}.
+We note that the difference between these studies may be even greater than this as we used naturally mated honey bees from 15 different colonies.
+We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates.
+At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}.
+Given that honey bees are naturally very polyandrous \cite{patriline}, our honey bees may also reflect more realistic environmental and genetic simulations.
+Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data.
+We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs.
+To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
 As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies.
-Importantly, we use an in-depth data visualization approach to explore and corroborate our data, and suggest such an approach can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.
+It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees.
+We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.
 
@@
 
-\subsection*{Phenotypic results}
+\subsection*{Pathogen response results}
 
@@
 We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus.
-Namely, bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed Chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).
+Namely, bees fed rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).
 
@@
 The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant.
-We compared IAPV titers  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03).
-However, bees fed Chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02).
-Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not (Figure 2).
+We compared IAPV titers  based on pairwise comparisons: Bees fed rockrose pollen had significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03).
+However, bees fed chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02).
+Overall, we interpreted these findings to mean that high-quality chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).
 
@@
 We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1).
-In the diet factor, more DEGs were upregulated in the more-nutritious Chestnut group (\textit{n} = 1,033) than in the less-nutritious Rockrose group (\textit{n} = 881).
+In the diet factor, more DEGs were upregulated in the more-nutritious chestnut group (\textit{n} = 1,033) than in the less-nutritious rockrose group (\textit{n} = 881).
 In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5).
@@
 
-GO analysis of the Chestnut-upregulated DEGs revealed the following over-represented categories: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1).
-GO analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).
+GO analysis of the chestnut-upregulated DEGs revealed the following over-represented categories: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1).
+GO analysis of the rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).
 
@@
 Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level.
-Across every pair comparing the Chestnut and Rockrose levels, regardless of the virus level, the number of Chestnut-upregulated DEGs was higher than the number of Rockrose-upregulated DEGs (Supplementary table 8 C, D, E, F, Additional file 1).
+Across every pair comparing the chestnut and rockrose levels, regardless of the virus level, the number of chestnut-upregulated DEGs was higher than the number of rockrose-upregulated DEGs (Supplementary table 8 C, D, E, F, Additional file 1).
 For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Supplementary table 8 A and B, Additional file 1).
-There were no DEGs between the treatment pair controlling for the Chestnut level of the virus effect (Supplementary table 8A, Additional file 1).
+There were no DEGs between the treatment pair controlling for the chestnut level of the virus effect (Supplementary table 8A, Additional file 1).
 These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).
@@
 
-Challenges to honey bee health are a growing concern, in particular the combined, interactive effects of nutritional stress and pathogens (Dolezal and Toth 2018).
+Challenges to honey bee health are a growing concern, in particular the combined, interactive effects of nutritional stress and pathogens \cite{AdamTothReview}.
 In this study, we used RNA-sequencing to probe mechanisms underlying honey bee responses to two effects, diet quality and infection with the prominent virus of concern, IAPV.
@@
 While some insect systems have shown relatively low transcriptional responses to dicistrovirus infection \cite{fewDEGs, fewDEGs2}, previous work on honey bees has revealed many hundreds of DEGs \cite{galbraith}.
-Despite this, the transcriptomic response to virus infection in our experiment was fairly limited.
+Discrepancies between datasets may be due to noise and complexity of the honey bee microbiome.
+The transcriptomic response to virus infection in our experiment was fairly limited.
 We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as genes related to transcriptional regulation and muscle contraction.
@@
 
-There are several reasons why, in the current study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
+For our nutrition factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.).
+Rockrose pollen is generally considered less nutritious than chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}.
+For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV.
+This experimental design resulted in four treatment groups (rockrose pollen without IAPV exposure, chestnut pollen without IAPV exposure, rockrose pollen with IAPV exposure, and chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.
+
+There are several reasons why our design focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
 First, when assessing diet diversity, a sugar diet is often used as a control.
@@
 As a result, there is a need to better understand how monofloral diets affect honey bee health.
-
-Consequently, for our nutrition factor, we examined two monofloral pollen diets, Rockrose (Cistus) and Castanea (Chestnut).
-Rockrose pollen is generally considered less nutritious than Chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron \cite{DiPasquale, adamInt}.
-For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV.
-This experimental design resulted in four treatment groups (Rockrose pollen without IAPV exposure, Chestnut pollen without IAPV exposure, Rockrose pollen with IAPV exposure, and Chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.
 
@@
 
-While our study examines honey bees from polyandrous colonies, the Galbraith study examined honey bees from single-drone colonies.
-As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}.
-We note that the difference between these studies may be even greater than this as we used naturally mated honey bees from 15 different colonies.
-We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates.
-At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}.
-Given that honey bees are naturally very polyandrous \cite{patriline}, our honey bees may also reflect more realistic environmental and genetic simulations.
-Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data.
-We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs.
-To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
-It is our hope that this aspect of our study may shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees.
+We used honey bees from polyandrous colonies, whereas Galbraith et al.
+\cite{galbraith} used honey bees from single-drone colonies.
+In light of this, we should expect the Galbraith et al.
+dataset to contain lower genetic variation between its replicates and higher signal:to:noise ratios than our dataset.
+We use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
 
@@
 We used an array of visualization tools as part of our analysis.
-We first used well-known tools like the PCA plot \cite{pca} from the DESeq2 package.
-After that, we used lesser-known multivariate visualization tools from our work-in-progress R package called bigPint.
+We used the PCA plot \cite{pca} from the DESeq2 package, a well-known and established tool.
+Along with that, we used lesser-known multivariate visualization tools from our work-in-progress R package called bigPint.
 Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds} (cite bigPint too)) to assess the variability between the replicates and the treatments in our data.
@@
 
-We also used statistical graphics to better understand patterns in our DEGs.
+Furthermore, we used statistical graphics to better understand patterns in our DEGs.
 However, in cases of large DEG lists, these visualization tools had overplotting problems (where multiple objects are drawn on top of one another, making it impossible to detect individual values).
@@
 To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (tolerance), we created contrasts of interest (Table \ref{tbl:contrasts}).
-In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the Chestnut group within the virus infected bees but not upregulated in the Chestnut group within the non-infected bees.
+In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the chestnut group within the virus infected bees but not upregulated in the chestnut group within the non-infected bees.
 Our interpretation of these genes is that they represent those that are only activated in infected bees that are fed a high quality diet.
-We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the Chestnut group for both the virus infected bees and non-infected bees.
+We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the chestnut group for both the virus infected bees and non-infected bees.
 Our interpretation of these genes is that they represent those that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not.
@@
 We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates.
-Hence, we wanted to explore whether our DEG read counts correlated with pathogen response metrics, including IAPV titers, Schmallenberg Virus (SBV) titers, and mortality rates.
+Hence, we wanted to explore whether our DEG read counts correlated with pathogen response metrics, including IAPV titers, sacbrood bee virus (SBV) titers, and mortality rates.
 For this process, we considered virus main effect DEGs (Figure 4), ``tolerance candidate'' DEGs (Additional file 15), and ``resistance candidate'' DEGs (Additional file 16).
@@
 Left to right: Mortality rates for the four treatment groups, two virus groups, and two diet groups.
-``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen.
+``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents chestnut pollen, and ``R'' represents rockrose pollen.
 The mortality rate data included 59 samples with 15 replicates per treatment group, except for the ``NC'' group having 14 replicates.
@@
 Left to right: IAPV titers for the four treatment groups, two virus groups, and two diet groups.
-``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen.
+``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents chestnut pollen, and ``R'' represents rockrose pollen.
 The IAPV titer data included 38 samples with 10 replicates per treatment group, except for the ``NR'' group having 8 replicates.
@@
     \textbf{Table 1:} Number of DEGs across three analysis pipelines for (A) the diet main effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study.
-For the diet effects, ``C'' represents Chestnut diet and ``R'' represents Rockrose diet.
+For the diet effects, ``C'' represents chestnut diet and ``R'' represents rockrose diet.
 For the virus effects, ``N'' represents control non-inoculated and ``V'' represents virus-inoculated.
-    \textbf{Table 2:} Pathways related to the 1,033 DEGs that were upregulated in the Chestnut treatment from the diet main effect.
-    \textbf{Table 3:} Pathways related to the 881 DEGs that were upregulated in the Rockrose treatment from the diet main effect.
+    \textbf{Table 2:} Pathways related to the 1,033 DEGs that were upregulated in the chestnut treatment from the diet main effect.
+    \textbf{Table 3:} Pathways related to the 881 DEGs that were upregulated in the rockrose treatment from the diet main effect.
     \textbf{Table 4:} GO analysis results for the 601 DEGs that were upregulated in the NC treatment from the NC versus NR treatment pair analysis.
-These DEGs represent genes that are upregulated when non-infected honey bees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.
+These DEGs represent genes that are upregulated when non-infected honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
     \textbf{Table 5:} GO analysis results for the 340 DEGs that were upregulated in the NR treatment from the NC versus NR treatment pair analysis.
-These DEGs represent genes that are upregulated when non-infected honey bees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.
+These DEGs represent genes that are upregulated when non-infected honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
     \textbf{Table 6:} GO analysis results for the 247 DEGs that were upregulated in the VC treatment from the VC versus VR treatment pair analysis.
-These DEGs represent genes that are upregulated when infected honey bees are given high quality Chestnut pollen compared to being given low quality Rockrose pollen.
+These DEGs represent genes that are upregulated when infected honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
     \textbf{Table 7:} GO analysis results for the 129 DEGs that were upregulated in the VR treatment from the VC versus VR treatment pair analysis.
-These DEGs represent genes that are upregulated when infected honey bees are given low quality Rockrose pollen compared to being given high quality Chestnut pollen.
+These DEGs represent genes that are upregulated when infected honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
     \textbf{Table 8:} Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor.
-``C'' represents Chestnut diet, ``R'' represents Rockrose diet, ``V'' represents virus-inoculated, and ``N'' represents control non-inoculated.
+``C'' represents chestnut diet, ``R'' represents rockrose diet, ``V'' represents virus-inoculated, and ``N'' represents control non-inoculated.
     \textbf{Table 9:} Kruskal-Wallis p-value and Bonferroni corrections for the 36 combinations of DEG lists, pathogen response metrics, and cluster number. (XLS).
@@
   \subsection*{Additional file 3 --- Parallel coordinate lines of the diet-related DEGs of our dataset.}
-    Parallel coordinate plots of the 1,914 DEGs after hierarchical clustering of size six between the Chestnut and Rockrose groups of our study.
-Here ``C'' represents Chestnut samples, and ``R'' represents Rockrose samples.
+    Parallel coordinate plots of the 1,914 DEGs after hierarchical clustering of size six between the chestnut and rockrose groups of our study.
+Here ``C'' represents chestnut samples, and ``R'' represents rockrose samples.
 The vertical red line indicates the distinction between treatment groups.
@@
     Parallel coordinate plots of the 122 DEGs after hierarchical clustering of size four between the ``tolerance'' candidate DEGs.
-Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet.
+Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality chestnut diet, and ``R'' represents low-quality rockrose diet.
 The vertical red line indicates the distinction between treatment groups.
@@
     Parallel coordinate plots of the 125 DEGs after hierarchical clustering of size four between the ``resistance'' candidate DEGs.
-Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet.
+Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality chestnut diet, and ``R'' represents low-quality rockrose diet.
 The vertical red line indicates the distinction between treatment groups.
@@
 Within a given subplot, ``D'' represents DESeq2, ``E'' represents edgeR, and ``L'' represents limma.
-From left to right on top row: Total diet-related DEGs (subplot A), Castanea-upregulated DEGs (subplot B), Rockrose-upregulated DEGs (subplot C).
+From left to right on top row: Total diet-related DEGs (subplot A), chestnut-upregulated DEGs (subplot B), rockrose-upregulated DEGs (subplot C).
 From left to right on middle row: Total virus-related DEGs (subplot D), virus-upregulated DEGs (subplot E), control-upregulated DEGs in our data (subplot F).

### 2018-10-26 Manuscript/Text/manuscript.tex
Added diet transcriptomics in model organisms paragraph

@@
 
-As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees.
-Honey bee transcriptomic studies have found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
-
-On the other hand, gene expression studies in insect models such as Drosophila and mosquito have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway and transcriptional pausing \cite{cherryReview, swevers}.
+There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have also been several such studies in model organisms.
+For example, diet effects on transcriptomics have been studied in insect models such as Drosophila.
+For instance, one recent transcriptomic study in Drosophila melanogaster reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}.
+CHK kinase-like protein is believed to respond to DNA damage by arresting the cell cycle.
+This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake.
+Another recent Drosophila study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}.
+No differential expression was found in Drosophila melanogaster, a species known to thrive in all three dietary conditions.
+However, Drosophila mojavensisa and Drosophila arizonae showed substantial differential expression between the dietary conditions: Genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets.
+Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}.
+One transcriptomic study in honey bees found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
+
+Likewise, studies on the transcriptomic effects of virus infection have been performed in insect models such as Drosophila and mosquito.
+These studies have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway, and transcriptional pausing \cite{cherryReview, swevers}.
 It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}.
@@
 Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group.
+There are also differences in the hours post inoculation and possible differences in the inoculation amount between the studies.
 Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}.

### 2018-10-26 Manuscript/Text/manuscript.tex
Added bigPint explanations

@@
 There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have also been several such studies in model organisms.
-For example, diet effects on transcriptomics have been studied in insect models such as Drosophila.
-For instance, one recent transcriptomic study in Drosophila melanogaster reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}.
+For example, diet effects on transcriptomics have been studied in the insect model Drosophila.
+One recent transcriptomic study in Drosophila melanogaster reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}.
 CHK kinase-like protein is believed to respond to DNA damage by arresting the cell cycle.
@@
 Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}.
-One transcriptomic study in honey bees found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
+While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
 
@@
 These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors.
-The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy to mention that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.
+The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.
 
@@
 However, through this automatically-generated plot, we can only visualize information at the sample level.
-Wanting to learn more about the data at the gene level, we continued with additional visualization techniques.
-
-We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies.
-The background side-by-side boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG.
-To reduce overplotting of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.
-
-We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays (Figure 3).
-We do see that the first replicate of the virus group (V.1) appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that this trend in the data that we saw in the PCA plot carried through into the DEG results.
+Wanting to learn more about the data at the gene level, we continued with new visualization techniques that we recently published in our R package bigPint (https://lrutter.github.io/bigPint).
+
+We used parallel coordinate lines superimposed onto side-by-side boxplots to visualize the DEGs associated with virus infection in the two studies.
+The background side-by-side boxplot represents the distribution of \textit{all} genes in the data, and each parallel coordinate line represents one DEG.
+In a parallel coordinate line, connections between samples with positive correlations should be flat, while connections between samples with negative correlations should be crossed.
+We expect DEGs to show more variability between treatments than between replicates.
+This means the parallel coordinate lines should be flat between replicates but crossed between treatments.
+However, overplotting problems would obscure our visualization if we were to plot all DEGs onto the same side-by-side boxplot.
+As a result, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.
+
+We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments (Figure 3).
+We do see that the first replicate of the virus group (V.1) appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that the trend we saw in the PCA plot carried through into the DEG results.
+Cluster 4 reveals somewhat inconsistent replicates in the virus group, although most virus standardized read counts (group V) remain consistently larger than most control standardized read counts (group N).
 In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4).
-The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large difference between treatment groups.
+The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large (or even consistent) difference between treatment groups.
 We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Additional file 3).
 
-We also used litre plots to examine the structure of individual DEGs: We see that indeed the individual virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the individual virus DEGs from the Galbraith data (Additional files 5 and 6).
-For the Galbraith data, we examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3).
-We verify this trend again in the litre plots with the DEG points in the first cluster showing less tight cluster patterns (Additional files 5 and 6).
-
-Finally, we looked at scatterplot matrices to assess the DEGs.
+We next used repLIcate TREatment (``litre'') plots, which we recently developed and published in our bigPint software package.
+Litre plots allow users to visualize one DEG onto the Cartesian coordinates of one scatterplot matrix.
+In the litre plot, each gene in the data is plotted once for every combination of replicates between treatment groups.
+For example, there are nine ways to pair a replicate from one treatment group with a replicate from the other treatment group in the Galbraith dataset (N.1 and V.1, N.1 and V.2, N.1 and V.3, N.2 and V.1, N.2.
+and V.2, N.2 and V.3, N.3 and V.1, N.3 and V.2, and N.3 and V.3).
+Hence, each gene in the Galbraith dataset is plotted as nine points in the litre plot.
+With 11,825 genes in the Galbriath data, 106,425 points would need to be plotted.
+Our dataset is even more dramatic: There are 144 ways to pair a replicate from one treatment group with a replicate from the other treatment group, and with 15,314 genes in our data, we would need to plot 2,205,216 points.
+For either dataset, plotting all these points would reduce the speed of interactive functionality and cause overplotting problems.
+As a result, we use hexagon bins to summarize this massive information.
+Once the background of hexagons has been drawn to reveal the distribution of all between-treatment sample pair combinations for \textit{all} genes, the user can superimpose all between-treatment sample pair combinations for one gene of interest.
+
+Additional file 4 shows nine example litre plots for our dataset.
+The hexagon background is the same for all nine litre plots because it simply shows the distribution of all between-treatment sample pair combinations for \textit{all} genes in our dataset.
+In each litre plot, there are 144 magenta points superimposed that show all between-treatment sample pair combinations for one DEG of interest.
+Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith datset.
+We examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) of the Galbriath data because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3).
+Notice that, as previously explained, we now show each DEG as nine points for the Galbraith dataset.
+We see that indeed the virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the virus DEGs from the Galbraith data (Additional files 5 and 6).
+We also observe that, in the Galbraith dataset, the DEG points in the first cluster show less tight cluster patterns than the DEG points in the second cluster (Additional files 5 and 6), an observation we saw previously in the parallel coordinate plots (Figure 3).
+
+Finally, we used scatterplot matrices from the bigPint software to further assess the DEGs.
+A scatterplot matrix is another effective multivariate visualization tool that plots read count distributions across all genes and samples.
+Specifically, it represents every gene in the dataset as a black point in each scatterplot.
+DEGs can be superimposed as colored points to assess their patterns against the full dataset.
+We expect DEGs to mostly fall along the \textit{x=y} line in replicate scatterplots (denoting replicate consistency) but deviate from the \textit{x=y} line in treatment scatterplots (denoting significant treatment changes).
+The \textit{x=y} line is shown in red in our plots.
+
 We created standardized scatterplot matrices for each of the four clusters (from Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10).

### 2018-11-04 Manuscript/Text/manuscript.tex
Dr. Toth corrections

@@
    email={adolezal@illinois.edu}
-]{\inits{AD}\fnm{Adam} \snm{Dolezal}}
+]{\inits{AD}\fnm{Adam G.} \snm{Dolezal}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Parts of Europe and the United States have witnessed dramatic losses in commercially managed honey bees over the past decade to what is considered an unsustainable extent.
-The large-scale loss of honey bees has considerable implications for the agricultural economy because honey bees are one of the leading pollinators of numerous crops.
-Honey bee declines have been associated with several interactive factors.
-Poor nutrition and viral infection are two environmental stressors that pose heightened dangers to honey bee health.
-In this study, we used RNA-sequencing to examine how monofloral diets and Israeli acute paralysis virus inoculation influence gene expression patterns in honey bees.
+The large-scale loss of bees has considerable implications for the agricultural economy because bees are one of the leading pollinators of numerous crops.
+Bee declines have been associated with several interactive factors.
+Recent studies suggest nutritional and pathogen stress can interactively contribute to bee physiological declines, but the molecular mechanisms underlying interactive effects remain unknown.
+In this study, we provide insight into this question by using RNA-sequencing to examine how monofloral diets and Israeli acute paralysis virus inoculation influence gene expression patterns in bees.
 
@@
 We only found 43 transcripts to be differentially expressed, some with known immune functions (argonaute-2), transcriptional regulation, and muscle contraction.
-We created contrasts to determine if any protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance).
+We created contrasts to explore whether protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance).
 A similar number of resistance and tolerance candidate differentially expressed genes were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection.
-We also compared the virus main effect in our study (polyandrous colonies) to that obtained in a previous study (single-drone colonies) and verified significant overlap in differential expression despite visualization methods showing differences in the noisiness levels between these two datasets.
 
 \parttitle{Conclusions}
-Through transcriptional contrasts and functional enrichment analysis, we add to evidence of feedbacks between diet and disease in honey bees.
-We also show that comparing results derived from polyandrous colonies (which are typically more natural) and single-drone colonies (which usually yield more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently less artificial and less noisy.
-Altogether, this work underlines the merits of using data visualization techniques and multiple datasets to interpret RNA-sequencing studies.
+Through transcriptional contrasts and functional enrichment analysis, we contribute to our understanding of the mechanisms underlying feedbacks between nutrition and disease in bees.
+We also show that comparing results derived from combined analyses across multiple RNA-seq studies may allow researchers to identify transcriptomic patterns in bees that are concurrently less artificial and less noisy.
+This work underlines the merits of using data visualization techniques and multiple datasets to interpret RNA-sequencing studies.
 
@@
 Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}.
-Understanding how undiversified diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.
+Understanding how low diversity diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.
 
@@
 One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV), a positive-sense RNA virus of the family Dicistroviridae \cite{fam}.
-IAPV infection causes shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}.
+IAPV infection causes shivering wings, decreased locomotion, muscle spasms, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}.
 IAPV has demonstrated higher infectious capacities than other honey bee viruses under certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.
-Its role in the rising phenomenon of ``Colony Collapse Disorder'' (in which the majority of worker bees disappear from a hive) remains unclear: It has been implicated in some studies \cite{iapvCCD, iapvCCD2} but not in other studies \cite{ccd1, fam, iapvCCD3}.
-Nonetheless, it is clear that IAPV reduces colony strength and survival.
 
@@
 For example, diet effects on transcriptomics have been studied in the insect model Drosophila.
-One recent transcriptomic study in Drosophila melanogaster reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}.
-CHK kinase-like protein is believed to respond to DNA damage by arresting the cell cycle.
+One recent transcriptomic study in \textit{Drosophila melanogaster} reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}.
 This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake.
-Another recent Drosophila study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}.
-No differential expression was found in Drosophila melanogaster, a species known to thrive in all three dietary conditions.
-However, Drosophila mojavensisa and Drosophila arizonae showed substantial differential expression between the dietary conditions: Genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets.
+Another recent study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}.
+\textit{Drosophila mojavensis} and \textit{Drosophila arizonae} showed substantial differential expression between the dietary conditions: genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets.
 Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}.
@@
 
-Likewise, studies on the transcriptomic effects of virus infection have been performed in insect models such as Drosophila and mosquito.
+There have been numerous studies on the transcriptomic effects of virus infection in model organisms like fruit flies and mosquitoes that can provide a useful framework for interpreting virus responses in honey bees.
 These studies have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway, and transcriptional pausing \cite{cherryReview, swevers}.
 It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}.
-In fact, every virus-host interaction has its own particularities derived of the diverse ways of replication and infection cycle evolved by different viruses.
-In this sense, an intricate set of pro- and anti-virus host factors are involved in each case \cite{cherryReview, swevers}.
-For example, dicistroviruses Drosophila C virus (DCV) and cricket paralysis virus (CrPV) infection requires ribosomal proteins RpS6, RpL19 and RACK1, but such proteins are dispensable during infection by the rhabdovirus vesicular stomatitis virus (VSV) in Drosophila.
-In contrast, the autophagy pathway has proven to be an effective antiviral strategy against VSV, but has no role limiting DCV or other positive-sense RNA viruses \cite{swevers}.
-For an extensive up-to-date review of known antiviral and proviral factors for specific viruses in Drosophila see \cite{swevers}.
-Model insect studies have shed light on how honey bees respond to viruses because the function of many genes relies on the characterization of orthologue genes.
-Nonetheless, recent transcriptomic studies in honey bees have improved our understanding of compatible and incompatible virus-bee interactions.
+In fact, every virus-host interaction has its own particularities derived from the diverse methods of replication and infection cycle evolved by different viruses.
+An intricate set of pro- and anti-virus host factors such as ribosomal proteins and autophagy pathways are involved, but the response depends on the virus species, as has been elucidated in Drosophila \cite{cherryReview, swevers}.
+Model insect studies can inform studies of honey bee virus response, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms.
+Recent transcriptomic studies in honey bees have improved our understanding of putative molecular mechanisms of honey bee-virus interactions.
+
 Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g.
 RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}.
-Added to these, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken}.
-This path has been studied recently in more depth using transcriptomics, which found known immune factors but also new ones for which indispensability in antiviral response was determined by silencing \cite{Brutscher}.
-Similar to Drosophila examples, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed.
+In addition, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken, Brutscher}.
+As in Drosophila, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed.
 This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs.
@@
 We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}.
-While our study examines honey bees from polyandrous colonies, the Galbraith study examined honey bees from single-drone colonies.
+While our study examines honey bees from naturally mated colonies, the Galbraith study examined honey bees from single-drone colonies.
 As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}.
@@
 At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}.
-Given that honey bees are naturally very polyandrous \cite{patriline}, our honey bees may also reflect more realistic environmental and genetic simulations.
+Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally mated honey bees may also reflect more realistic environmental and genetic simulations.
 Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data.
@@
 
-\subsection*{Pathogen response results}
+\subsection*{Mortality and virus titers}
 
 We reanalyzed our previously published dataset with a subset that focuses on diet quality and is more relevant to the current study.
-We briefly show it again here to inform the RNA-sequencing comparison because we reduced the number of treatments from the original published data (from eight to four) \cite{adamInt} as a means to focus on diet quality effects.
-
-As shown in Figure 1, mortality rates of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.34e-05).
+We show the data subset here to inform the RNA-sequencing comparison because we reduced the number of treatments from the original published data (from eight to four) \cite{adamInt} as a means to focus on diet quality effects.
+
+As shown in Figure 1, mortality rates of honey bees 72 hours post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.34e-05).
 The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 7.04e-06) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant.
 We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus.
-Namely, bees fed rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).
-
-As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03).
+Bees fed rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).
+
+As shown in Figure 2, IAPV titers of honey bees 72 hours post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03).
 The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant.
@@
 However, bees fed chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02).
-Overall, we interpreted these findings to mean that high-quality chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).
-
-\subsection*{Main effect DEG results}
+Overall, we interpreted these findings to mean that high-quality chestnut pollen could partially ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).
+
+\subsection*{Transcriptomic responses to virus infection and diet}
 
@@
 
-GO analysis of the chestnut-upregulated DEGs revealed the following over-represented categories: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1).
+GO analysis of the chestnut-upregulated DEGs revealed the following over-represented biological functions: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1).
 GO analysis of the rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).
@@
 
-\subsection*{Pairwise comparison of DEG results}
-
 The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 955 (Supplementary table 8, Additional file 1).
-Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level.
+Again, diet level appeared to have greater influence on the number of DEGs than the virus level.
 Across every pair comparing the chestnut and rockrose levels, regardless of the virus level, the number of chestnut-upregulated DEGs was higher than the number of rockrose-upregulated DEGs (Supplementary table 8 C, D, E, F, Additional file 1).
-For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Supplementary table 8 A and B, Additional file 1).
-There were no DEGs between the treatment pair controlling for the chestnut level of the virus effect (Supplementary table 8A, Additional file 1).
+Virus-treated bees showed equal to or more upregulated genes relative to controls, under both diet treatments (Supplementary table 8 A and B, Additional file 1).
 These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).
 
-\subsection*{Prior study comparison results}
+\subsection*{Transcriptomic data visualization and comparison to a previous study}
 
@@
 However, through this automatically-generated plot, we can only visualize information at the sample level.
-Wanting to learn more about the data at the gene level, we continued with new visualization techniques that we recently published in our R package bigPint (https://lrutter.github.io/bigPint).
+Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online (https://lrutter.github.io/bigPint) and are in preparation for publication.
 
@@
 
-\subsection*{Tolerance versus resistance results}
+\subsection*{Tolerance versus resistance}
 
 Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate DEGs and 125 ``resistance'' candidate DEGs.
-We again used parallel coordinate lines superimposed onto side-by-side boxplots to visualize these candidate DEGs.
+Within our 122 ``tolerance'' gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism).
+However, we also discovered gene ontologies related to RNA polymerase II transcription, immune response, and regulation of response to reactive oxygen species (Figure 5A).
+Within our 125 ``resistance'' gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, chitin metabolism, oligosaccharide biosynthesis, and general metabolism) (Figure 5B).
+
+To visually explore gene expression patterns related to tolerance and resistance, we used hierarchical clustering to separate candidate DEGs into common patterns, and then visualized these clusters using parallel coordinate lines superimposed onto side-by-side boxplots.
 To reduce overplotting of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns.
@@
 
-Within our 122 ``tolerance'' gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism).
-However, we also discovered gene ontologies related to RNA polymerase II transcription, immune response, and regulation of response to reactive oxygen species (Figure 5A).
-Within our 125 ``resistance'' gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, chitin metabolism, oligosaccharide biosynthesis, and general metabolism) (Figure 5B).
-
-\subsection*{Post hoc analysis results}
-
-In general, the R-squared values between gene read counts and pathogen response measurements were low (R-squared $<$ 0.1).
-However, some DEG clusters showed slightly larger R-squared values than the non-DEG group (the rest of the data).
-One prominent example of this includes the first and second cluster of the virus-related DEGs and their correlation with IAPV titers (Additional file 19I).
-The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those in the rest of the data.
+\subsection*{Post hoc analysis}
+
+To better understand sources of transcriptomic noise, we explored whether pathogen response measurements (virus titers and mortality), which varied widely across samples, were correlated with observed patterns in gene expression.
+
+The R-squared values between gene read counts and pathogen response measurements were generally low (R-squared $<$ 0.1) across our dataset (Supplementary table 9, Additional file 1).
+We further explored whether clusters of DEGs showed higher correlations with pathogen response measurements than non-DEGs (the latter serving as a control, where we do not expect a correlation).
+A Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those in the rest of the data.
 The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1.
 An overall trend emerges to suggest that DEGs may have significantly larger correlation with the pathogen response measurements compared to non-DEGs.
-It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between pathogen response measurements.
 
@@
 We believe these visualization applications can be useful for future researchers analyzing RNA-sequencing data to quickly and effectively ensure that the DEG calls look reliable or at least overlap with DEG calls from similar studies that look reliable.
-We also expect this type of visualization exploration can be especially crucial when studying complex organisms that do not have genetic identicalness or similarity between replicates and/or when using experiments that may lack rigid design control.
+We also expect this type of visualization exploration can be especially crucial when studying wild populations with high levels of genetic and environmental variation between replicates and/or when using experiments that may lack rigid design control.
 
@@
 There were several limitations in this study that could be improved upon in future studies.
-For instance, our comparison between the Galbraith data (single-drone colonies) and our data (polyandrous colonies) was limited by numerous extraneous variables between these studies.
+For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally mated colonies) was limited by numerous extraneous variables between these studies.
 In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on one-day old worker honey bees that were fed sugar and artificial pollen diet, whereas our study focused on adult worker honey bees that were fed bee-collected monofloral diets.
@@
 Hence the apparent reduction in noise observed in the Galbraith data compared to our data in the PCA plots, parallel coordinate plots, scatterplot matrices, and litre plots may be an inadvertent product of the smaller number of replicates used and the lack of a secondary treatment group rather than solely the reduction in genetic variability through the single-drone colony design itself.
-With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between polyandrous colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.
+With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between naturally mated colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.
 
@@
 For instance, by verifying the substantial overlap in our DEG lists to those obtained in another study that addressed a similar question using specimens with less genetic variability, we were able to place much higher confidence in the differential gene expression results from our otherwise noisy data.
-We also suggested that comparing results derived from polyandrous colony designs (which are usually more natural) and single-drone colony designs (which usually have more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently more realistic and less noisy.
+We also suggested that comparing results derived from multiple studies varying in level of genetic and environmental variability may allow researchers to identify transcriptomic patterns that are concurrently more realistic and less noisy.
 Altogether, we hope our results underline the merits of using data visualization techniques and multiple datasets to understand and interpret RNA-sequencing datasets.
@@
 
-\subsection*{Pathogen response}
+\subsection*{Mortality and virus titers}
 
@@
 
-Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing).
+Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing 2500 in rapid run mode; Category: Single End 100 cycle sequencing).
 A standard Illumina mRNA library was prepared by the DNA facility.
@@
 
-We used honey bees from polyandrous colonies, whereas Galbraith et al.
+We used honey bees from naturally mated colonies, whereas Galbraith et al.
 \cite{galbraith} used honey bees from single-drone colonies.
@@
 Along with that, we used lesser-known multivariate visualization tools from our work-in-progress R package called bigPint.
-Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds} (cite bigPint too)) to assess the variability between the replicates and the treatments in our data.
-We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines \cite{jds} (cite bigPint too).
+Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds}) to assess the variability between the replicates and the treatments in our data.
+We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines \cite{jds}.
 

### 2018-11-05 Manuscript/Text/manuscript.tex
Dolezal and Rutter changes

@@
 
-There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have also been several such studies in model organisms.
-For example, diet effects on transcriptomics have been studied in the insect model Drosophila.
-One recent transcriptomic study in \textit{Drosophila melanogaster} reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}.
-This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake.
-Another recent study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}.
-\textit{Drosophila mojavensis} and \textit{Drosophila arizonae} showed substantial differential expression between the dietary conditions: genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets.
-Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}.
+There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have been several such studies in model organisms.
+Model insect studies can inform studies of honey bee transcriptomic responses, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms.
+Previous Drosophila studies that examined various diet effects have found gene expression changes related to immunity, metabolism, cell cycle activity, DNA binding, transcription, and insulin signaling \cite{sugarFat, sugarProtein, jHormone, alaux2}.
 While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
-
-There have been numerous studies on the transcriptomic effects of virus infection in model organisms like fruit flies and mosquitoes that can provide a useful framework for interpreting virus responses in honey bees.
-These studies have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway, and transcriptional pausing \cite{cherryReview, swevers}.
-It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}.
-In fact, every virus-host interaction has its own particularities derived from the diverse methods of replication and infection cycle evolved by different viruses.
-An intricate set of pro- and anti-virus host factors such as ribosomal proteins and autophagy pathways are involved, but the response depends on the virus species, as has been elucidated in Drosophila \cite{cherryReview, swevers}.
-Model insect studies can inform studies of honey bee virus response, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms.
-Recent transcriptomic studies in honey bees have improved our understanding of putative molecular mechanisms of honey bee-virus interactions.
-
-Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g.
-RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}.
-In addition, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken, Brutscher}.
-As in Drosophila, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed.
-This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs.
-deformed wing virus (DWV) \cite{Ryabov}.
-Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.
-In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level.
-Chen et al.
-2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}.
-Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood.
-The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}.
-IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}.
-These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors.
-The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.
+Numerous studies on the transcriptomic effects of virus infection in model organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}.
+Virus-bee systems have also revealed key factors of these antiviral conserved defense pathways, such as RNA silencing, Toll pathways, IMD pathways, JAK/STAT pathways, autophagy, and endocytosis \cite{McMenamin}.
 
@@
 We further explored whether clusters of DEGs showed higher correlations with pathogen response measurements than non-DEGs (the latter serving as a control, where we do not expect a correlation).
-A Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those in the rest of the data.
+A Kruskal–Wallis test was used to determine if R-squared distributions of DEG clusters significantly differed from those in the rest of the data.
 The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1.
@@
 In general, we found a major nutritional transcriptomic response, with nearly 2,000 transcripts changing in response to diet quality (rockrose/poor diet versus chestnut/good diet).
-The majority of these genes were upregulated in response to high quality diet, and these genes were over-represented for functions (Supplementary table 2, Additional file 1) such as nutrient signaling metabolism (insulin resistance) and immune response (Notch signaling and JaK-STAT pathways).
+The majority of these genes were upregulated in response to high quality diet, and these genes were over-represented for functions such as nutrient signaling metabolism (insulin resistance), immune response (Notch signaling and JaK-STAT pathways), and carbon metabolism (Supplementary table 2, Additional file 1).
 These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into innate immune responses.
+
+One of the few studies that has investigated transcriptomic response to nutrition in honey bees similarly found that pollen upregulates genes related to macromolecule metabolism, insulin pathways, and TOR pathways \cite{alaux2}.
+Diet effects on transcriptomics have been more extensively studied in the insect model Drosophila.
+One recent transcriptomic study in \textit{Drosophila melanogaster} reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}.
+This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake.
+Another recent study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}.
+\textit{Drosophila mojavensis} and \textit{Drosophila arizonae} showed substantial differential expression between the dietary conditions: genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets.
+Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}.
+As we saw in our study, these studies generally suggest that diet differences may relate to gene expression changes in metabolism and immune responses in honey bees.
 
@@
 The transcriptomic response to virus infection in our experiment was fairly limited.
-We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as genes related to transcriptional regulation and muscle contraction.
+We found only 43 differentially expressed transcripts, some with known immune functions such as a gene with similarity to MD-2 lipid recognition protein and argonaute-2, a protein that plays a central role in RNA silencing (Table \ref{tbl:virusGenes}).
+We also found genes related to transcriptional regulation and muscle contraction.
 The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14).
+
+There have been numerous studies on the transcriptomic effects of virus infection in model organisms like fruit flies and mosquitoes that can provide a useful framework for interpreting virus responses in honey bees.
+These studies have shown that RNA silencing is a major antiviral strategy, along with transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7-autophagy pathways \cite{cherryReview, swevers}.
+Recent transcriptomic studies in honey bees have shown similar hallmarks of these same antiviral defense mechanisms \cite{McMenamin}.
+It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}.
+In fact, every virus-host interaction has its own particularities derived from the diverse methods of replication and infection cycle evolved by different viruses.
+An intricate set of pro- and anti-virus host factors such as ribosomal proteins and autophagy pathways are involved, but the response depends on the virus species, as has been elucidated in Drosophila \cite{cherryReview, swevers}.
+In addition, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken, Brutscher}.
+In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level.
+Chen et al.
+2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}.
+Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood.
+The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}.
+IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}.
+These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors.
+The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.
 
@@
 Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}.
-Our comparative visualization assessment between these two datasets was also somewhat limited because the virus effect in the Galbraith study used three replicates for each level, whereas the virus effect in our study used twelve replicates for each level that were actually further subdivided into six replicates for each diet level.
+The different factors between these two studies may be critical because particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed.
+This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs.
+deformed wing virus (DWV) \cite{Ryabov}.
+Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.
+
+Moreover, our comparative visualization assessment between these two datasets was also somewhat limited because the virus effect in the Galbraith study used three replicates for each level, whereas the virus effect in our study used twelve replicates for each level that were actually further subdivided into six replicates for each diet level.
 Hence the apparent reduction in noise observed in the Galbraith data compared to our data in the PCA plots, parallel coordinate plots, scatterplot matrices, and litre plots may be an inadvertent product of the smaller number of replicates used and the lack of a secondary treatment group rather than solely the reduction in genetic variability through the single-drone colony design itself.
@@
 AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq.
+DC advised on statistical analyses and visualisation.
     %All authors read and approved the final manuscript.

### 2018-11-05 Manuscript/Text/manuscript.tex
Add GEO submission number

@@
 While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
-Numerous studies on the transcriptomic effects of virus infection in model organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}.
-Virus-bee systems have also revealed key factors of these antiviral conserved defense pathways, such as RNA silencing, Toll pathways, IMD pathways, JAK/STAT pathways, autophagy, and endocytosis \cite{McMenamin}.
+Numerous studies on the transcriptomic effects of virus infection in model insect organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}.
+Virus-bee systems have also revealed key factors of these antiviral conserved defense pathways \cite{McMenamin}.
 
@@
 Hence, each gene in the Galbraith dataset is plotted as nine points in the litre plot.
-With 11,825 genes in the Galbriath data, 106,425 points would need to be plotted.
+With 11,825 genes in the Galbraith data, 106,425 points would need to be plotted.
 Our dataset is even more dramatic: There are 144 ways to pair a replicate from one treatment group with a replicate from the other treatment group, and with 15,314 genes in our data, we would need to plot 2,205,216 points.
@@
 In each litre plot, there are 144 magenta points superimposed that show all between-treatment sample pair combinations for one DEG of interest.
-Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith datset.
-We examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) of the Galbriath data because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3).
+Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith dataset.
+We examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) of the Galbraith data because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3).
 Notice that, as previously explained, we now show each DEG as nine points for the Galbraith dataset.
@@
 Another recent study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}.
-\textit{Drosophila mojavensis} and \textit{Drosophila arizonae} showed substantial differential expression between the dietary conditions: genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets.
+\textit{Drosophila mojavensis} and \textit{Drosophila arizonae} showed substantial differential expression between the dietary conditions: genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in response to low sugar high protein diets.
 Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}.
@@
 There have been numerous studies on the transcriptomic effects of virus infection in model organisms like fruit flies and mosquitoes that can provide a useful framework for interpreting virus responses in honey bees.
-These studies have shown that RNA silencing is a major antiviral strategy, along with transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7-autophagy pathways \cite{cherryReview, swevers}.
-Recent transcriptomic studies in honey bees have shown similar hallmarks of these same antiviral defense mechanisms \cite{McMenamin}.
+These studies have showed that RNA silencing is a major antiviral strategy, along with transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7-autophagy pathways \cite{cherryReview, swevers}.
+Recent transcriptomic studies in honey bees have shown similar hallmarks of these same antiviral defense mechanisms, including RNA silencing, Toll pathways, IMD pathways, JAK/STAT pathways, autophagy, and endocytosis \cite{McMenamin}.
 It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}.
@@
 \section*{Availability of data and materials}
-  The datasets generated and/or analysed during the current study are available in the [NAME] repository, [PERSISTENT WEB LINK TO DATASETS].
-Include our data, Galbraith data, scripts to reproduce tables and figures (on GitHub).
+  The data discussed in this publication have been deposited in NCBI's Gene Expression Omnibus \cite{geo} and are accessible through GEO Series accession number GSE121885 (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE121885).
+The scripts to reproduce analyses and figures in this publication are available online (https://github.com/lrutter/HoneyBeePaper).
 
@@
 AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq.
-DC advised on statistical analyses and visualisation.
+DC advised on statistical analyses and visualization.
     %All authors read and approved the final manuscript.

### 2018-11-06 Manuscript/Text/manuscript.tex
Adam additional changes

@@
 \author[
-   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-   email={lrutter@iastate.edu}   % email address
+   addressref={aff1}, 
+   email={lrutter@iastate.edu} 
 ]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
 \author[
-   addressref={aff6},
+   addressref={aff2}, 
+   email={jcarrillo@cicese.mx} 
+]{\inits{LR}\fnm{Jimena} \snm{Carrillo-Tripp}}
+\author[
+   addressref={aff3},
    email={bbonning@ufl.edu}
@@
 \author[
-   addressref={aff2},
+   addressref={aff4},
    email={dicook@monash.edu}
@@
 \author[
-   addressref={aff3,aff4},
+   addressref={aff5,aff6},
    email={amytoth@iastate.edu}
@@
 \author[
-   addressref={aff5},
-   corref={aff5},
+   addressref={aff7},
+   corref={aff7},
    email={adolezal@illinois.edu}
@@
 }
+
 \address[id=aff2]{%
+  \orgname{Department of Microbiology, Center for Scientific Research and Higher Education of Ensenada},
+  \city{Ensenada},   
+  \state{Baja California}
+  \postcode{22860},
+  \cny{Mexico}
+}
+
+\address[id=aff3]{%
+  \orgname{Department of Entomology and Nematology, University of Florida},
+  \city{Gainesville},   
+  \state{FL}
+  \postcode{32611},
+  \cny{USA}
+}
+
+\address[id=aff4]{%
   \orgname{Econometrics and Business Statistics, Monash University},
@@
 
-\address[id=aff3]{%
+\address[id=aff5]{%
   \orgname{Department of Entomology, Iowa State University},
@@
 
-\address[id=aff4]{%
+\address[id=aff6]{%
   \orgname{Department of Ecology, Evolution, and Organismal Biology, Iowa State University},
@@
 
-\address[id=aff5]{%
+\address[id=aff7]{%
   \orgname{Department of Entomology, University of Illinois at Urbana-Champaign},
@@
   \postcode{61801},
-  \cny{USA}
-}
-
-\address[id=aff6]{%
-  \orgname{Department of Entomology and Nematology, University of Florida},
-  \city{Gainesville},   
-  \state{FL}
-  \postcode{32611},
   \cny{USA}
@@
 
-Commercially managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}.
+Managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}.
 More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}.
@@
 Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honey bees in an interactive fashion as the environment changes \cite{interacting}.
-Nutrition and viral infection are two broad factors that pose heightened dangers to honey bee health in response to recent environmental changes.
+Nutrition and viral infection are two factors that pose heightened dangers to honey bee health in response to recent environmental changes.
+Interactions between nutrition and viral infection may create feedbacks that impact bee health through several mechanisms \cite{adamInt, AdamTothReview}.
 
@@
 We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}.
-While our study examines honey bees from naturally mated colonies, the Galbraith study examined honey bees from single-drone colonies.
+While our study examines honey bees derived from naturally-mated queens, the Galbraith study examined honey bees derived from single-drone inseminated queens.
 As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}.
-We note that the difference between these studies may be even greater than this as we used naturally mated honey bees from 15 different colonies.
+We note that the difference between these studies may be even greater than this as we used honey bees from 15 different colonies, i.e.
+from 15 different, naturally-mated queens.
 We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates.
 At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}.
-Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally mated honey bees may also reflect more realistic environmental and genetic simulations.
+Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally-mated honey bees may also reflect more realistic environmental and genetic conditions.
 Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data.
-We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs.
+We therefore wish to explore how the gene expression effects of IAPV inoculation compare between these two studies using different experimental designs.
 To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
@@
 
-Given the noisy nature of our data, and our desire to hone in on genes with real expression differences, we compared our data to the Galbraith study \cite{galbraith}, which also examined bees response to IAPV infection.
+Given the noisy nature of our data, and our desire to home in on genes with real expression differences, we compared our data to the Galbraith study \cite{galbraith}, which also examined bees response to IAPV infection.
 In contrast to our study, Galbraith et al.
@@
 There were several limitations in this study that could be improved upon in future studies.
-For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally mated colonies) was limited by numerous extraneous variables between these studies.
-In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on one-day old worker honey bees that were fed sugar and artificial pollen diet, whereas our study focused on adult worker honey bees that were fed bee-collected monofloral diets.
+For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally-mated colonies) was limited by numerous extraneous variables between these studies.
+In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on worker honey bees that were fed sugar and artificial pollen diets, whereas our study focused on worker honey bees that were fed bee-collected monofloral diets.
 Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group.
@@
 Hence the apparent reduction in noise observed in the Galbraith data compared to our data in the PCA plots, parallel coordinate plots, scatterplot matrices, and litre plots may be an inadvertent product of the smaller number of replicates used and the lack of a secondary treatment group rather than solely the reduction in genetic variability through the single-drone colony design itself.
-With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between naturally mated colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.
+With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between naturally-mated colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.
 
@@
 
-We used honey bees from naturally mated colonies, whereas Galbraith et al.
+We used honey bees from naturally-mated colonies, whereas Galbraith et al.
 \cite{galbraith} used honey bees from single-drone colonies.
@@
 AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq.
+JCT contributed to experimental design and laboratory experiments.
 DC advised on statistical analyses and visualization.

### 2018-11-14 Manuscript/Text/manuscript.tex
Toth additional revisions

@@
 
-Managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}.
+Managed honey bees have undergone health declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}.
 More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}.
@@
 Numerous studies on the transcriptomic effects of virus infection in model insect organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}.
-Virus-bee systems have also revealed key factors of these antiviral conserved defense pathways \cite{McMenamin}.
-
-As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense \cite{interactingDV}.
+Studies of virus-bee systems have revealed some of the antiviral defense pathways known in model organisms are conserved and also related to bee antiviral immune responses \cite{McMenamin}.
+
+To our knowledge, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense \cite{interactingDV}.
 In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure).
@@
 
-We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}.
+Because RNA-seq data can be noisy and subject to high levels of inter-experiment variation, we further sought to validate our transcriptomic data via comparison to a previous RNA-seq study on honey bee responses to viral infection.
+To do this, we compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}.
 While our study examines honey bees derived from naturally-mated queens, the Galbraith study examined honey bees derived from single-drone inseminated queens.
@@
 Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data.
-We therefore wish to explore how the gene expression effects of IAPV inoculation compare between these two studies using different experimental designs.
-To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
+To achieve this comparison, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
 As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies.
@@
 Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2).
-We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots.
-However, through this automatically-generated plot, we can only visualize information at the sample level.
-Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online (https://lrutter.github.io/bigPint) and are in preparation for publication.
+Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online (https://lrutter.github.io/bigPint).
+To understand how to interpret the visualizations used in this section, please read (https://lrutter.github.io/bigPint/articles/plotIntro.html).
 
@@
 However, overplotting problems would obscure our visualization if we were to plot all DEGs onto the same side-by-side boxplot.
-As a result, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.
-
-We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments (Figure 3).
-We do see that the first replicate of the virus group (V.1) appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that the trend we saw in the PCA plot carried through into the DEG results.
-Cluster 4 reveals somewhat inconsistent replicates in the virus group, although most virus standardized read counts (group V) remain consistently larger than most control standardized read counts (group N).
+Therefore, we graphed clustered subsets of the DEGs (based on hierarchical clustering).
+
+The 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments (Figure 3).
 In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4).
@@
 
-We next used repLIcate TREatment (``litre'') plots, which we recently developed and published in our bigPint software package.
+We next used repLIcate TREatment (``litre'') plots, which we recently developed for our bigPint software package.
 Litre plots allow users to visualize one DEG onto the Cartesian coordinates of one scatterplot matrix.
 In the litre plot, each gene in the data is plotted once for every combination of replicates between treatment groups.
-For example, there are nine ways to pair a replicate from one treatment group with a replicate from the other treatment group in the Galbraith dataset (N.1 and V.1, N.1 and V.2, N.1 and V.3, N.2 and V.1, N.2.
-and V.2, N.2 and V.3, N.3 and V.1, N.3 and V.2, and N.3 and V.3).
-Hence, each gene in the Galbraith dataset is plotted as nine points in the litre plot.
-With 11,825 genes in the Galbraith data, 106,425 points would need to be plotted.
-Our dataset is even more dramatic: There are 144 ways to pair a replicate from one treatment group with a replicate from the other treatment group, and with 15,314 genes in our data, we would need to plot 2,205,216 points.
-For either dataset, plotting all these points would reduce the speed of interactive functionality and cause overplotting problems.
-As a result, we use hexagon bins to summarize this massive information.
+We use hexagon bins to summarize this massive information.
 Once the background of hexagons has been drawn to reveal the distribution of all between-treatment sample pair combinations for \textit{all} genes, the user can superimpose all between-treatment sample pair combinations for one gene of interest.
 
-Additional file 4 shows nine example litre plots for our dataset.
-The hexagon background is the same for all nine litre plots because it simply shows the distribution of all between-treatment sample pair combinations for \textit{all} genes in our dataset.
-In each litre plot, there are 144 magenta points superimposed that show all between-treatment sample pair combinations for one DEG of interest.
-Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith dataset.
-We examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) of the Galbraith data because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3).
-Notice that, as previously explained, we now show each DEG as nine points for the Galbraith dataset.
+Additional file 4 shows nine example litre plots for our dataset; each litre plot shows the 144 between-treatment sample pair combinations for one DEG of interest.
+Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith dataset; each litre plot shows the nine between-treatment sample pair combinations for one DEG of interest.
 We see that indeed the virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the virus DEGs from the Galbraith data (Additional files 5 and 6).

### 2018-11-14 Manuscript/MakeFiguresFiles/makeFigure1.R
Recreate figure 3

@@ -5,7 +5,7 @@ library(readr)
 library(multcomp)
 library(nlme)
 
-d <- read_csv("day3Mortality.csv")
+d <- read_csv("../../PathogenResponse/day3Mortality.csv")
 d = as.data.frame(d)
 attr(d, "spec") <- NULL
 # Note: There are 14 NC; 15 VR, VC, NR
@@ -38,40 +38,7 @@ plotMortality2 = ggplot(d, aes(x=Virus, y=Day3Mortality)) + geom_boxplot(fill="p
 labelDF = data.frame(plot.labels=c("C","R"), labels = c("a","b"), V1 = c(0.48, 0.54))
 plotMortality3 = ggplot(d, aes(x=Diet, y=Day3Mortality)) + geom_boxplot(fill="palegreen2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) + theme_gray() + ylim(0,0.8)+theme(axis.title.y=element_blank(), text=element_text(size=20))
 
-###### Plot IAPV Values ########
-
-d <- read_csv("logIAPV.csv")
-d = as.data.frame(d)
-attr(d, "spec") <- NULL
-# Note: There are 14 NC; 15 VR, VC, NR
-
-d = d[which(d$treatment %in% c("None Chestnut", "None Rockrose", "Virus Chestnut", "Virus Rockrose")),]
-
-d$treatment = as.factor(d$treatment)
-colnames(d)[3] = "logIAPV"
-levels(d$treatment) <- c('NC', 'NR', 'VC', 'VR')
-colnames(d)[1] <- "Treatment"
-
-d$Virus = sapply(d$Treatment, function(x) substr(x, 1, 1))
-d$Diet = sapply(d$Treatment, function(x) substr(x, 2, 2))
-d$Virus = as.factor(d$Virus)
-d$Diet = as.factor(d$Diet)
-
-mortcomp = lme(logIAPV ~ Diet*Virus, data=d, random = ~1|Experiment) 
-mortcomp2 = lme(logIAPV ~ Treatment, data=d, random = ~1|Experiment)
-anova(mortcomp)
-summary(glht(mortcomp2, linfct=mcp(Treatment="Tukey")))
-
-# IAPV (All)
-labelDF = data.frame(plot.labels=c("NR","VC","VR","NC"), labels = c("a","ab","b","a"), V1 = c(4.25, 5.35, 8.57, 3.85))
[…]

### 2018-11-14 Manuscript/MakeFiguresFiles/makeFigure2.R

@@ -1,0 +1,45 @@
+library(plyr)
+library(ggplot2)
+library(multcompView)
+library(readr)
+library(multcomp)
+library(nlme)
+
+d <- read_csv("../../PathogenResponse/logIAPV.csv")
+d = as.data.frame(d)
+attr(d, "spec") <- NULL
+# Note: There are 14 NC; 15 VR, VC, NR
+
+d = d[which(d$treatment %in% c("None Chestnut", "None Rockrose", "Virus Chestnut", "Virus Rockrose")),]
+
+d$treatment = as.factor(d$treatment)
+colnames(d)[3] = "logIAPV"
+levels(d$treatment) <- c('NC', 'NR', 'VC', 'VR')
+colnames(d)[1] <- "Treatment"
+
+d$Virus = sapply(d$Treatment, function(x) substr(x, 1, 1))
+d$Diet = sapply(d$Treatment, function(x) substr(x, 2, 2))
+d$Virus = as.factor(d$Virus)
+d$Diet = as.factor(d$Diet)
+
+mortcomp = lme(logIAPV ~ Diet*Virus, data=d, random = ~1|Experiment) 
+mortcomp2 = lme(logIAPV ~ Treatment, data=d, random = ~1|Experiment)
+anova(mortcomp)
+summary(glht(mortcomp2, linfct=mcp(Treatment="Tukey")))
+
+# IAPV (All)
+labelDF = data.frame(plot.labels=c("NR","VC","VR","NC"), labels = c("a","ab","b","a"), V1 = c(4.25, 5.35, 8.57, 3.85))
+plotIAPV = ggplot(d, aes(x=Treatment, y=logIAPV)) + geom_boxplot(fill="paleturquoise2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) + ylab("Log IAPV Titer") +theme_gray() + theme(text=element_text(size=20)) + ylim(2,9)
+
+# IAPV (N vs V)
+labelDF = data.frame(plot.labels=c("N","V"), labels = c("a","b"), V1 = c(4.3, 8.7))
[…]

### 2018-11-14 Manuscript/MakeFiguresFiles/makeFigure3.R

@@ -17,141 +17,25 @@ library(scales)
 library(bigPint)
 library(matrixStats)
 
-getPCP <- function(nC){
-  
-  set.seed(1)
-  colList = scales::hue_pal()(nC+1)
-  colList[2] = "#E9AA0D"
-  colList[3] = "#EA502F"
-  k = cutree(hc, k=nC)
-  
-  yMin = min(sigDatas[,1:nColumns])
-  yMax = max(sigDatas[,1:nColumns])
-  
-  ###########################
-  
-  sbsDF <- data.frame()
-  for (i in 1:nC){
-    x = as.data.frame(sigDatas[which(k==i),])
-    xNames = rownames(x)
-    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
-    sbsDF = rbind(sbsDF, data.frame(Cluster = paste("Cluster", i), FDR = xFDR))
-  }
-  
-  plot_clusters = lapply(1:nC, function(i){
-    j = rev(order(table(k)))[i]
-    x = as.data.frame(sigDatas[which(k==j),])
-    nGenes = nrow(x)
-    x$cluster = "color"
-    x$cluster2 = factor(x$cluster)
-    xNames = rownames(x)
-    xFDR = metrics[which(metrics$ID %in% xNames),]$padj
-    scatMatMetrics = list()
-    scatMatMetrics[[currPair]] = metrics[which(metrics$ID %in% xNames),]
-    scatMatMetrics[[currPair]]$FDR = 10e-10
-    scatMatMetrics[[currPair]]$ID = as.factor(as.character(scatMatMetrics[[currPair]]$ID))
-    
-    plotDatas = datas[, c(ncol(datas), 1:ncol(datas)-1)]
-    plotData = data[,c(9,1:6)]
-    plotData[,c(2:7)] <- log(plotData[,c(2:7)] +1)
-    
-    fileName = paste(getwd(), "/", outDir, "/", currPair, "_SM_", nC, "_", i, ".jpg", sep="")
-    ret <- plotDEG(data = plotDatas, dataMetrics = scatMatMetrics, option="scatterPoints", threshVar = "padj", threshVal = 0.05, degPointColor = colList[i+1], fileName=fileName)
-    
-    x$ID = xNames
-    saveRDS(xNames, file=paste0("Sig_", nC,"_", i,".Rds"))
-    
-    pcpDat <- melt(x[,c(1:(nColumns+1))], id.vars="ID")
[…]

### 2018-11-15 Manuscript/Text/manuscript.tex
Additional file 20

@@
 
+  \subsection*{Additional file 20 --- Tables listing DEGs for contrasts.}
+    \textbf{Table 1:} IDs of 1,914 DEGs in our diet main effect.
+    \textbf{Table 2:} IDs of 43 DEGs in our virus main effect.
+    \textbf{Table 3:} IDs of 178 DEGs in our NR versus VR contrast.
+    \textbf{Table 4:} IDs of 376 DEGs in our VC versus VR contrast.
+    \textbf{Table 5:} IDs of 774 DEGs in our NC versus VR contrast.
+    \textbf{Table 6:} IDs of 955 DEGs in our VC versus NR contrast. 
+    \textbf{Table 7:} IDs of 941 DEGs in our NC versus NR contrast. 
+    \textbf{Table 8:} IDs of 125 resistance candidate genes.
+    \textbf{Table 9:} IDs of 122 tolerance candidate genes. (XLS).
+
 \end{backmatter}

### 2018-11-15 ResistanceTolerance/KruskalWelch/functionRT.R

@@ -1,76 +1,0 @@
-functionRT <- function(data, type, colNum, PVal){
-  rsq <- function (x, y) cor(x, y) ^ 2
-  
-  Variables <- read_csv("~/HoneyBeePaper/Variables.csv")
-  myVar <- as.data.frame(Variables[,colNum])[,1]
-  
-  readFile <- c("../../N_V/DESeq2/ClusterStandard/Sig_4_", "../../ResistanceTolerance/Clustering_Tolerance/Sig_4_", "../../ResistanceTolerance/Clustering_Resistance/Sig_4_")
-  strVar <- c("virus", "tolerance", "resistance")
-  
-  getR2 <- function(rf, sv){
-    dfR2 <- data.frame(matrix(ncol = 2, nrow = 0))
-    colnames(dfR2) <- c("Group", "R2")
-    allR2 <- c()
-    for (i in 1:4){
-      temp <- readRDS(paste0(rf, i, ".Rds")) # different
-      temp2 <- data[which(rownames(data) %in% temp),]
-      temp3 = c()
-      allR2 <- c(allR2, temp)
-      if (nrow(temp2)>0){
-        for (j in 1:nrow(temp2)){
-          temp3[j] <- rsq(as.numeric(temp2[j,]), myVar)
-        } 
-      }
-      dfR2 <- rbind(dfR2, data.frame(Group = rep(paste0("Cluster",i), length(temp3)), R2 = temp3))
-    }
-    dfR2Box <- data.frame(matrix(ncol = 2, nrow = 0))
-    colnames(dfR2Box) <- c("Group", "R2")
-    dataMin <- data[-which(rownames(data) %in% allR2),]
-    temp3 = c()
-    if (nrow(dataMin)>0){
-      for (i in 1:nrow(dataMin)){
-        temp3[i] <- rsq(as.numeric(dataMin[i,]), myVar)
-      } 
-    }
-    temp3 <- temp3[-which(is.na(temp3))]
-    dfR2Box <- rbind(dfR2Box, data.frame(Group = rep(paste0("data"), length(temp3)), R2 = temp3))
-    list(dfR2, dfR2Box)
-  }
-  
-  virusR2 <- getR2(readFile[1], strVar[1])
-  plotVirus <- rbind(virusR2[[1]], virusR2[[2]])
-  toleranceR2 <- getR2(readFile[2], strVar[2])
[…]

### 2018-11-15 ResistanceTolerance/KruskalWelch/getPVal.R

@@ -1,19 +1,0 @@
-library(readr)
-library(ggplot2)
-library(EDASeq)
-library(dplyr)
-library(genefilter)
-library(nlme)
-library(onewaytests)
-source("functionRT.R")
-
-data <- as.data.frame(readRDS("../../NC_NR_VC_VR/data/data.Rds"))
-
-PVal <- data.frame()
-PVal <- functionRT(data=data, type="IAPV", colNum=7, PVal=PVal)
-dev.off()
-PVal <- functionRT(data=data, type="SBV", colNum=6, PVal=PVal)
-dev.off()
-PVal <- functionRT(data=data, type="Mortality", colNum=5, PVal=PVal)
-
-saveRDS(PVal, "PVal.Rds")

### 2019-01-04 Manuscript/Text/manuscript.tex
Draft submitted

@@
 Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2).
-Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online (https://lrutter.github.io/bigPint).
+Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online \cite{bigpint}.
 To understand how to interpret the visualizations used in this section, please read (https://lrutter.github.io/bigPint/articles/plotIntro.html).
@@
 
+\section*{Declarations}
+
 \begin{backmatter}
@@
   The data discussed in this publication have been deposited in NCBI's Gene Expression Omnibus \cite{geo} and are accessible through GEO Series accession number GSE121885 (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE121885).
-The scripts to reproduce analyses and figures in this publication are available online (https://github.com/lrutter/HoneyBeePaper).
+Scripts to reproduce visualizations in this paper are available online (https://github.com/lrutter/HoneyBeePaper).
+Information about bigPint visualizations, including how to intrepret them are available online (https://lrutter.github.io/bigPint).
+Lists of BeeBase IDs for contrast DEGs from this study are found in Additional File 20.
 
@@
 \section*{Author's contributions}
-  LR performed the bioinformatic and statistical analyses, produced the figures and tables, and drafted the manuscript.
-BB conceptualized the study and critically revised the manuscript.
-AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq.
-JCT contributed to experimental design and laboratory experiments.
+  LR performed the bioinformatic and statistical analyses and produced the figures and tables.
+LR and ALT drafted the manuscript.
+ALT, AGD, JCT, BCB contributed to experimental design.
+AGD and JCT carried out laboratory experiments.
+AGD processed samples for virus and RNA-seq.
 DC advised on statistical analyses and visualization.
-    %All authors read and approved the final manuscript.
+All authors revised and approved the manuscript.
 
@@
 \begin{figure}[h!]
-\caption{\csentence{Parallel coordinate plots of the 1,019 DEGs after hierarchical clustering of size four between the virus-infected and control groups of the Galbraith data \cite{galbraith}.}
+\caption{\csentence{Parallel coordinate plots of the 1,019 virus-related DEGs of the Galbraith data \cite{galbraith}.}
 Parallel coordinate plots of the 1,019 DEGs after hierarchical clustering of size four between the virus-infected and control groups of the Galbraith study.
@@
 \begin{figure}[h!]
-\caption{\csentence{Parallel coordinate plots of the 43 DEGs after hierarchical clustering of size four between the virus-infected and control groups of our study.}
+\caption{\csentence{Parallel coordinate plots of the 43 virus-related DEGs of our data.}
 Parallel coordinate plots of the 43 DEGs after hierarchical clustering of size four between the virus-infected and control groups of our study.
@@
 \begin{figure}[h!]
-\caption{\csentence{Gene ontology analysis results for the 122 DEGs related to our ``tolerance'' hypothesis and for the 125 DEGs related to our ``resistance'' hypothesis.}
+\caption{\csentence{Gene ontology results for the 122 ``tolerance'' and 125 ``resistance'' DEG candidates in our data.}
 GO analysis results for the 122 DEGs related to our ``tolerance'' hypothesis (A) and for the 125 DEGs related to our ``resistance'' hypothesis (B).

### 2019-02-18 Manuscript/Text/manuscript.tex
Most reviewer revisions

@@
 In particular, honey bees are confronted with a less diverse selection of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}.
-Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}.
-Understanding how low diversity diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.
-
-Viral infection was a comparatively minor problem in honey bees until the last century when the ectoparasitic varroa mite (\textit{Varroa destructor}) spread worldwide \cite{miteSpread}.
-This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}.
+Reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}.
+Understanding how low diversity diets (i.e.
+monofloral diets) affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.
+Indeed, differing qualities of monofloral diets have been shown to affect nurse bee physiology and tolerance to parasites \cite{DiPasquale}.
+
+Viral infection was considered a comparatively minor problem in honey bees until the last century when the ectoparasitic varroa mite (\textit{Varroa destructor}) spread worldwide \cite{miteProblem, miteProblem2, miteProblem3}.
+This mite feeds on honey bee hemolymph and/or fat body tissue \cite{hemolymph, fatbody}, and is believed to decrease lipid and glycogen reserves and reduce protein synthesis in bees \cite{miteNutrition}.
+Additionally, it transmits multiple viruses and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}.
 More than 20 honey bee viruses have been identified \cite{numVirus}.
-One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV), a positive-sense RNA virus of the family Dicistroviridae \cite{fam}.
+One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV), a positive-sense RNA virus of the family \textit{Dicistroviridae} \cite{fam}.
 IAPV infection causes shivering wings, decreased locomotion, muscle spasms, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}.
-IAPV has demonstrated higher infectious capacities than other honey bee viruses under certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.
+IAPV has demonstrated higher infectious capacity \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.
 
@@
 Following up on these findings, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality.
-For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on vigor (tolerance) \cite{resTol1}.
+For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function that reduces the pathogen load of the host (resistance, \cite{Medzhitov}) or if it is due to indirect effects of good nutrition on the ability of the host to reduce pathogen impacts without affecting pathogen load (tolerance, \cite{Medzhitov}).
 Transcriptomics is one means to better understand the mechanistic underpinnings of dietary and viral effects on honey bee health.
-Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection.
+Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection, which can help us generate candidate gene lists to further investigate the relative roles of tolerance and resistance.
 This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a ``buffer'' against other stressors \cite{AdamTothReview}.
 
-There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have been several such studies in model organisms.
+There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith, Brutscher, Chen, Ryabov, Doublet} in honey bees, but there have been several such studies in model organisms.
 Model insect studies can inform studies of honey bee transcriptomic responses, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms.
 Previous Drosophila studies that examined various diet effects have found gene expression changes related to immunity, metabolism, cell cycle activity, DNA binding, transcription, and insulin signaling \cite{sugarFat, sugarProtein, jHormone, alaux2}.
-While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.
+While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2, Chen}.
+Previous transcriptomic studies have identified genes serving links between metabolism and antiviral defense in honey bees \cite{metabolismVirus, metabolismVirus1}; see \cite{metabolismVirusReview} for an overview.
 Numerous studies on the transcriptomic effects of virus infection in model insect organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}.
@@
 Rockrose pollen is generally considered less nutritious than chestnut pollen because it contains smaller amounts of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}.
+For specific quantitative differences between these two pollen groups, please see \cite{DiPasquale}.
 We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section).
@@
 Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally-mated honey bees may also reflect more realistic environmental and genetic conditions.
-Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data.
 To achieve this comparison, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
@@
 It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees.
-We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.
+We suggest that in-depth data visualization approaches (including scatterplot matrices, parallel coordinate plots, and litre plots from the bigPint software package) can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.
 
@@
 We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus.
-Bees fed rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).
+Bees fed rockrose pollen had significantly elevated mortality with virus infection compared to non-inoculated controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).
 
@@
 The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant.
-We compared IAPV titers  based on pairwise comparisons: Bees fed rockrose pollen had significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03).
-However, bees fed chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02).
-Overall, we interpreted these findings to mean that high-quality chestnut pollen could partially ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).
+We compared IAPV titers  based on pairwise comparisons: Bees fed rockrose pollen had significantly elevated IAPV titers with virus infection compared to non-inoculated controls (Benjamini Hochberg, p $<$ 7.56e-03).
+However, bees fed chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to non-inoculated controls (Benjamini Hochberg, p = 6.29e-02).
+While many of the non-inoculated treatment groups showed some RT-qPCR amplification (non-inoculated average Ct=33.92; inoculated average Ct=24.9), and thus have virus titers calculable on a standard curve, these Ct levels are similar to those deemed uninfected in previous studies \cite{galbraith}.
+Overall, we interpreted these findings to mean that high-quality chestnut pollen could partially reduce high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).
 
@@
 We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1).
+There were only four genes that were DEGs in both our diet main effect and our virus main effect (GB48747, GB47214, GB42908, and GB42507).
 In the diet factor, more DEGs were upregulated in the more-nutritious chestnut group (\textit{n} = 1,033) than in the less-nutritious rockrose group (\textit{n} = 881).
@@
 While these reported DEG counts are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table 1, Additional file 1 and Additional file 18).
-
+For our DEG analysis, we used R software version 3.3.3 \cite{R}.
+
+We performed GO analysis to statistically assign our DEGs to predefined bins based on their functional characteristics, allowing us to better understand the biological processes of our DEGs.
 GO analysis of the chestnut-upregulated DEGs revealed the following over-represented biological functions: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1).
@@
 We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset.
-Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level.
-Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2).
+Note that the Galbraith dataset contained three individual bees per treatment group as a single pooled sample, while our dataset contained 16 individual bees per treatment group in 8 RNA-seq samples.
+Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and non-inoculated honey bees better than our dataset (Additional file 2).
 Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online \cite{bigpint}.
-To understand how to interpret the visualizations used in this section, please read (https://lrutter.github.io/bigPint/articles/plotIntro.html).
+For more information about the visualizations used here, please refer to  (https://lindsayrutter.github.io/bigPint/articles/plotIntro.html).
 
 We used parallel coordinate lines superimposed onto side-by-side boxplots to visualize the DEGs associated with virus infection in the two studies.
-The background side-by-side boxplot represents the distribution of \textit{all} genes in the data, and each parallel coordinate line represents one DEG.
+The background side-by-side boxplot represents the distribution of \textit{all} genes in the data (all 15,314 genes in our count table), and each parallel coordinate line represents one DEG.
 In a parallel coordinate line, connections between samples with positive correlations should be flat, while connections between samples with negative correlations should be crossed.
@@
 
-The 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments (Figure 3).
+The 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments.
+The few inconsistent replicates we observed (such as V.1 from Cluster 1 and V.2 from Cluster 4) were small enough that consistent differences between the treatment groups remained apparent across the samples (Figure 3).
 In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4).
@@
 As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14).
+Specifically, we arbitrarily subsetted the samples so each one was represented once in each of these four files (i.e.
+Additional File 11 shows samples 1-3; Additional File 12 shows samples 4-6; Additional File 13 shows samples 7-9; and Additional File 14 shows samples 10-12).
 We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: They deviated more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.
@@
 The majority of these genes were upregulated in response to high quality diet, and these genes were over-represented for functions such as nutrient signaling metabolism (insulin resistance), immune response (Notch signaling and JaK-STAT pathways), and carbon metabolism (Supplementary table 2, Additional file 1).
-These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into innate immune responses.
+These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into immune responses.
 
@@
 We also found genes related to transcriptional regulation and muscle contraction.
-The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14).
+The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14) and baseline viral titers observed in our control bees (Figure 2).
 
@@
 These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.
+Thus, our study uses transcriptome data to generate lists of candidate genes that can be the focus of future investigations to better experimentally test putative roles of tolerance and resistance genes in this system.
 
@@
 In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on worker honey bees that were fed sugar and artificial pollen diets, whereas our study focused on worker honey bees that were fed bee-collected monofloral diets.
-Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group.
+Also, Galbraith’s bees were sampled at 24 hours while ours were sampled at 36 hours.
+Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and observations to determine behaviorally symptomatic bees whereas we used whole bodies and categorized only into inoculated vs.
+non-inoculated groups.
 There are also differences in the hours post inoculation and possible differences in the inoculation amount between the studies.
@@
 Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}.
+In brief, our virus inoculum was prepared by injection of infectious virus particles (derived from infected adults) into white-eyed honey bee pupae; these pupae were then homogenized and virus particles enriched and resuspended.
+This inoculum was then characterized for presence of acute bee paralysis virus, black queen cell virus, DWV, IAPV, Kashmir bee virus, and SBV.
+Experimental infection tests of adult bees and honey bee cell cultures \cite{carrillo} showed that only IAPV is amplified in adult bees.
+To infect caged bees for these experiments, newly emerged bees from 15 healthy colonies at the Iowa State University research apiary were homogeneously mixed, then counted into clear acrylic cages in groups of 35 bees per cage.
+Cages were then presented with open feeders containing 30\% sucrose solution (control) or 30\% sucrose solution containing a 1:1000 dilution of viral inoculum (treatment).
+Dietary treatments were then added (described below).
+To quantify virus titers, two live bees were randomly sampled at 36 hpi from each of 9-10 randomly selected cages.
+Virus levels were then measured via RT-qPCR and quantified against a standard curve, identically to methods described in \cite{carrillo, adamInt}.
+
 A linear mixed effects model was used to relate the mortality rates and IAPV titers to the main and interaction effects of the diet and virus factors.
-The model was fitted to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''.
+The model was fit to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''.
 A random (intercept) effect for experimental setup was included in the model.
@@
 However, such an experimental design does not reflect real-world conditions for honey bees as they rarely face a total lack of pollen \cite{DiPasquale}.
+Moreover, younger larvae tend to be fed pollen diets, whereas older larvae tend to be fed nectar diets.
+By focusing on pollen diets, our study design reflects natural diet conditions for larvae of a specific age category \cite{ageMatters}.
 Second, in studies that compared honey bee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet \cite{DiPasquale}.
@@
 From these, six live honey bees were randomly selected from each cage 36 hours post inoculation and placed into tubes \cite{carrillo}.
+In summary, 8 samples (representing two bees each) were sequenced per experimental condition (i.e., 32 samples sequenced).
 Tubes were kept on dry ice and then transferred into a -80C freezer until processing.
@@
 
-We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs).
+We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs: NR versus NC, NR versus VR, NR versus VC, NC versus VR, NC versus VC, and VR versus VC).
 We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs).
@@
 However, in cases of large DEG lists, these visualization tools had overplotting problems (where multiple objects are drawn on top of one another, making it impossible to detect individual values).
-To remedy this problem, we first standardized each DEG to have a mean of zero and standard deviation of unity \cite{Chandrasekhar, deSouto}.
+To remedy this problem, we first standardized each DEG to have a mean of zero and standard deviation of unity for its read counts across its samples \cite{Chandrasekhar, deSouto}.
 Then, we performed hierarchical clustering on the standardized DEGs using Ward's linkage.
@@
 To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (tolerance), we created contrasts of interest (Table \ref{tbl:contrasts}).
-In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the chestnut group within the virus infected bees but not upregulated in the chestnut group within the non-infected bees.
+In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the chestnut group within the virus infected bees but not upregulated in the chestnut group within the non-inoculated bees.
 Our interpretation of these genes is that they represent those that are only activated in infected bees that are fed a high quality diet.
-We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the chestnut group for both the virus infected bees and non-infected bees.
+We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the chestnut group for both the virus infected bees and non-inoculated bees.
 Our interpretation of these genes is that they represent those that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not.
@@
 We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates.
-Hence, we wanted to explore whether our DEG read counts correlated with pathogen response metrics, including IAPV titers, sacbrood bee virus (SBV) titers, and mortality rates.
+Hence, we wanted to explore whether our DEG read counts correlated with pathogen response metrics, including IAPV titers, sacbrood virus (SBV; also present in our inoculum \cite{carrillo, adamInt}), and mortality rates.
+We explored correlation with SBV because our inoculum \cite{carrillo} does contain SBV, and bees from both inoculated and non-inoculated groups do exhibit detectable SBV titers.
 For this process, we considered virus main effect DEGs (Figure 4), ``tolerance candidate'' DEGs (Additional file 15), and ``resistance candidate'' DEGs (Additional file 16).
@@
 Clusters 1, 2, and 4 seem to represent DEGs that were overexpressed in the virus inoculated group, and Cluster 3 seems to represent DEGs that were overexpressed in the non-inoculated control group.
-In general, the DEGs appeared as expected, but there is rather noticeable deviation of the first replicate from the virus-treated sample (``V.1'') from the other virus-treated replicates in Cluster 1.}
+In general, the DEGs appeared as expected, but there is rather noticeable deviation of the first replicate from the virus-treated sample (``V.1'') from the other virus-treated replicates in Cluster 1.
+We also note a deviation of the second replicate from the virus-treated samples (``V.2'') from the other virus-treated replicates in Cluster 4.}
 \end{figure}
@@
 Parallel coordinate plots of the 43 DEGs after hierarchical clustering of size four between the virus-infected and control groups of our study.
-``N'' represents non-infected control group, and ``V'' represents treatment of virus.
+``N'' represents non-inoculated control group, and ``V'' represents treatment of virus.
 The vertical red line indicates the distinction between treatment groups.
@@
 \hline
-BeeBase ID & Gene Name & Known functions & Us & Galbraith \\ 
-\hline
-GB41545 & \makecell{MD-2-related \\ lipid-recognition \\ protein-like} & \makecell{Implicated in lipid recognition, \\ particularly in the recognition of \\ pathogen related products} & N & - \\
-\hline
-GB50955 & \makecell{Protein \\ argonaute-2} & \makecell{Interacts with small interfering RNAs \\ to form RNA-induced silencing \\ complexes which target and cleave \\ transcripts that are mostly from \\ viruses and transposons} & V & V \\
-\hline
-GB48755 & \makecell{UBA-like \\ domain-containing \\ protein 2} & \makecell{Found in diverse proteins involved \\ in ubiquitin/proteasome \\ pathways} & V & V \\
-\hline
-GB47407 & Histone H4 & \makecell{Capable of affecting transcription, \\ DNA repair, and DNA replication \\ when post-transcriptionally modified} & V & V \\
-\hline
-GB42313 & \makecell{Leishmanolysin-like \\ peptidase} & \makecell{Encodes a protein involved in cell \\ migration and invasion; implicated in \\ mitotic progression in D. melanogaster} & V & V \\
-\hline
-GB50813 & \makecell{Rho guanine \\ nucleotide \\ exchange factor 11} & \makecell{Implicated in regulation of apoptopic \\ processes, cell growth, signal \\ transduction, and transcription} & V & V \\
-\hline
-GB54503 & \makecell{Thioredoxin \\ domain-containing \\ protein} & \makecell{Serves as a general protein \\ disulphide oxidoreductase} & N & - \\
-\hline
-GB53500 & \makecell{Transcriptional \\ regulator Myc-B} & \makecell{Regulator gene that codes for a \\ transcription factor} & V & V \\
-\hline
-GB51305 & Tropomyosin-like & \makecell{Related to protein involved in muscle \\ contraction} & N & N \\
-\hline
-GB50178 & \makecell{Cilia and \\ flagella-associated \\ protein 61-like} & \makecell{Induces components required for \\ wild-type motility and \\ stable assembly of motile cilia} & V & V \\
+\makecell{BeeBase ID, \\ NCBI Gene ID} & Gene Name & Known functions & Us & Galbraith \\ 
+\hline
+\makecell{GB41545, \\ 409187} & \makecell{MD-2-related \\ lipid-recognition \\ protein-like} & \makecell{Implicated in lipid recognition, \\ particularly in the recognition of \\ pathogen related products} & N & - \\
+\hline
+\makecell{GB50955, \\ 411577} & \makecell{Protein \\ argonaute-2} & \makecell{Interacts with small interfering RNAs \\ to form RNA-induced silencing \\ complexes which target and cleave \\ transcripts that are mostly from \\ viruses and transposons} & V & V \\
+\hline
+\makecell{GB48755, \\ 727455} & \makecell{UBA-like \\ domain-containing \\ protein 2} & \makecell{Found in diverse proteins involved \\ in ubiquitin/proteasome \\ pathways} & V & V \\
+\hline
+\makecell{GB47407, \\ 406132} & Histone H4 & \makecell{Capable of affecting transcription, \\ DNA repair, and DNA replication \\ when post-transcriptionally modified} & V & V \\
+\hline
+\makecell{GB42313, \\ 409923} & \makecell{Leishmanolysin-like \\ peptidase} & \makecell{Encodes a protein involved in cell \\ migration and invasion; implicated in \\ mitotic progression in D.
+melanogaster} & V & V \\
+\hline
+\makecell{GB50813, \\ 410127} & \makecell{Rho guanine \\ nucleotide \\ exchange factor 11} & \makecell{Implicated in regulation of apoptopic \\ processes, cell growth, signal \\ transduction, and transcription} & V & V \\
+\hline
+\makecell{GB54503, \\ 411255} & \makecell{Thioredoxin \\ domain-containing \\ protein} & \makecell{Serves as a general protein \\ disulphide oxidoreductase} & N & - \\
+\hline
+\makecell{GB53500, \\ 100576392} & \makecell{Transcriptional \\ regulator Myc-B} & \makecell{Regulator gene that codes for a \\ transcription factor} & V & V \\
+\hline
+\makecell{GB51305, \\ 551252} & Tropomyosin-like & \makecell{Related to protein involved in muscle \\ contraction} & N & N \\
+\hline
+\makecell{GB50178, \\ 726905} & \makecell{Cilia and \\ flagella-associated \\ protein 61-like} & \makecell{Induces components required for \\ wild-type motility and \\ stable assembly of motile cilia} & V & V \\
 \hline
@@
 \hline
-NC vs NR & 941 & \makecell{Genes that change expression \\ due to diet effect in \\ uninfected bees} & \makecell{Supplementary \\ tables 4 and 5, \\ Additional file 1} \\
+NC vs NR & 941 & \makecell{Genes that change expression \\ due to diet effect in \\ non-inoculated bees} & \makecell{Supplementary \\ tables 4 and 5, \\ Additional file 1} \\
 \hline
@@
     \textbf{Table 4:} GO analysis results for the 601 DEGs that were upregulated in the NC treatment from the NC versus NR treatment pair analysis.
-These DEGs represent genes that are upregulated when non-infected honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
+These DEGs represent genes that are upregulated when non-inoculated honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
     \textbf{Table 5:} GO analysis results for the 340 DEGs that were upregulated in the NR treatment from the NC versus NR treatment pair analysis.
-These DEGs represent genes that are upregulated when non-infected honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
+These DEGs represent genes that are upregulated when non-inoculated honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
     \textbf{Table 6:} GO analysis results for the 247 DEGs that were upregulated in the VC treatment from the VC versus VR treatment pair analysis.
@@
     Example litre plots of the nine DEGs with the lowest FDR values from the 43 virus-related DEGs of our dataset.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 Most of the magenta points (representing the 144 combinations of samples between treatment groups for a given DEG) do not reflect the expected pattern as clearly compared to what we saw in the litre plots of the Galbraith data.
@@
     Example litre plots of the nine DEGs with the lowest FDR values from the 365 DEGs in Cluster 1 (originally shown in Figure 3) of the Galbraith dataset.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 Most of the light orange points (representing the nine combinations of samples between treatment groups for a given DEG) deviate from the \textit{x=y} line in a tight bundle as expected (PNG).
@@
     Example litre plots of the nine DEGs with the lowest FDR values from the 327 DEGs in Cluster 2 (originally shown in Figure 3) of the Galbraith dataset.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 Most of the dark orange points (representing the nine combinations of samples between treatment groups for a given DEG) deviate from the \textit{x=y} line in a compact clump as expected.
@@
 The data has been standardized.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots.
@@
 The data has been standardized.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots (PNG).
@@
 The data has been standardized.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 We confirm that the DEGs mostly follow the expected structure, with their placement deviating from the \textit{x=y} line in the treatment scatterplots, but adhering to the \textit{x=y} line in the replicate scatterplots.
-We also see that the second replicate from the virus-treated sample (``V.2'') may be somewhat inconsistent in these DEGs, as its presence in the replicate scatterplots results in the DEGs unexpectedly deviating from the \textit{x=y} line and its presence in the treatment scatterplots results in the DEGs unexpectedly adhering to the \textit{x=y} line (PNG).
+We also see that the second replicate from the virus-treated sample (``V.2'') may be somewhat inconsistent in these DEGs, as its presence in the replicate scatterplots results in the DEGs unexpectedly deviating from the \textit{x=y} line and its presence in the treatment scatterplots results in the DEGs unexpectedly adhering to the \textit{x=y} line.
+This inconsistent sample was something we observed in Figure 3 (PNG).
 
@@
 The data has been standardized.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 We see that, compared to the scatterplot matrices from certain clusters of the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, sometimes unexpectedly deviating from the \textit{x=y} line in the replicate plots and sometimes unexpectedly adhering to the \textit{x=y} line in the treatment plots (PNG).
@@
 The data has been standardized.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 We see that, compared to the scatterplot matrices from certain clusters of the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, and most of them unexpectedly adhere to the \textit{x=y} line in the treatment plots (PNG).
@@
 The data has been standardized.
-``N'' represents non-infected control samples and ``V'' represents virus-treated samples.
+``N'' represents non-inoculated control samples and ``V'' represents virus-treated samples.
 We see that, compared to the scatterplot matrices from certain clusters of the Galbraith data, the 43 DEGs from this subset of six samples from our data do not paint as clear of a picture, and most of them unexpectedly deviate from the \textit{x=y} line in the virus-related replicate plots (PNG).
@@
     Parallel coordinate plots of the 122 DEGs after hierarchical clustering of size four between the ``tolerance'' candidate DEGs.
-Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality chestnut diet, and ``R'' represents low-quality rockrose diet.
+Here ``N'' represents non-inoculated control group, ``V'' represents treatment of virus, ``C'' represents high-quality chestnut diet, and ``R'' represents low-quality rockrose diet.
 The vertical red line indicates the distinction between treatment groups.
@@
     Parallel coordinate plots of the 125 DEGs after hierarchical clustering of size four between the ``resistance'' candidate DEGs.
-Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality chestnut diet, and ``R'' represents low-quality rockrose diet.
+Here ``N'' represents non-inoculated control group, ``V'' represents treatment of virus, ``C'' represents high-quality chestnut diet, and ``R'' represents low-quality rockrose diet.
 The vertical red line indicates the distinction between treatment groups.

### 2019-02-18 Manuscript/MakeFiguresFiles/makeFigure2.R

@@ -29,7 +29,7 @@ summary(glht(mortcomp2, linfct=mcp(Treatment="Tukey")))
 
 # IAPV (All)
 labelDF = data.frame(plot.labels=c("NR","VC","VR","NC"), labels = c("a","ab","b","a"), V1 = c(4.25, 5.35, 8.57, 3.85))
-plotIAPV = ggplot(d, aes(x=Treatment, y=logIAPV)) + geom_boxplot(fill="paleturquoise2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) + ylab("Log IAPV Titer") +theme_gray() + theme(text=element_text(size=20)) + ylim(2,9)
+plotIAPV = ggplot(d, aes(x=Treatment, y=logIAPV)) + geom_boxplot(fill="paleturquoise2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) + ylab("Log viral ge/100 ng RNA") +theme_gray() + theme(text=element_text(size=20)) + ylim(2,9)
 
 # IAPV (N vs V)
 labelDF = data.frame(plot.labels=c("N","V"), labels = c("a","b"), V1 = c(4.3, 8.7))

### 2019-03-04 Manuscript/Text/manuscript.tex
Final revisions

@@
 To our knowledge, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense \cite{interactingDV}.
-In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure).
+In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV inoculation, high quality diet without IAPV inoculation, low quality diet with IAPV inoculation, and high quality diet with IAPV inoculation).
 For our diet factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.).
@@
 For specific quantitative differences between these two pollen groups, please see \cite{DiPasquale}.
+Throughout this paper, we refer to our four treatment groups as ``NR'' (non-inoculation and low quality pollen), ``NC'' (non-inoculation and high quality pollen), ``VR'' (IAPV inoculation and low quality pollen), and ``VC'' (IAPV inoculation and high quality pollen).
 We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section).
@@
 It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees.
-We suggest that in-depth data visualization approaches (including scatterplot matrices, parallel coordinate plots, and litre plots from the bigPint software package) can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.
+We suggest that in-depth data visualization approaches (including scatterplot matrices, parallel coordinate plots, and litre plots from the bigPint software package \cite{bigpint}) can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.
+
+\section*{Methods}
+
+\subsection*{Mortality and virus titers}
+
+Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}.
+In brief, our virus inoculum was prepared by injection of infectious virus particles (derived from infected adults) into white-eyed honey bee pupae; these pupae were then homogenized and virus particles enriched and resuspended.
+This inoculum was then characterized for presence of acute bee paralysis virus, black queen cell virus, deformed wing virus (DWV), IAPV, Kashmir bee virus, and sacbrood bee virus (SBV).
+Experimental infection tests of adult bees and honey bee cell cultures \cite{carrillo} showed that only IAPV is amplified in adult bees.
+To infect caged bees for these experiments, newly emerged bees from 15 healthy colonies at the Iowa State University research apiary were homogeneously mixed, then counted into clear acrylic cages in groups of 35 bees per cage.
+Cages were then presented with open feeders containing 30\% sucrose solution (control) or 30\% sucrose solution containing a 1:1000 dilution of viral inoculum (treatment).
+Dietary treatments were then added (described below).
+To quantify virus titers, two live bees were randomly sampled at 36 hpi from each of 9-10 randomly selected cages.
+Virus levels were then measured via RT-qPCR and quantified against a standard curve, identically to methods described in \cite{carrillo, adamInt}.
+
+A linear mixed effects model was used to relate the mortality rates and IAPV titers to the main and interaction effects of the diet and virus factors.
+The model was fit to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''.
+A random (intercept) effect for experimental setup was included in the model.
+Post-hoc pairwise comparisons of the four (diet and virus combination) treatment groups were performed and Benjamini-Hochberg adjusted p-values were calculated to limit familywise Type I error rates \cite{bh}.
+
+\subsection*{Design of two-factor experiment}
+
+For our nutrition factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.).
+Rockrose pollen is generally considered less nutritious than chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}.
+For our virus factor, one level contained bees that were inoculated with IAPV and another level contained bees that were not inoculated with IAPV.
+This experimental design resulted in four treatment groups (low quality rockrose pollen without IAPV exposure, high quality chestnut pollen without IAPV exposure, low quality rockrose pollen with IAPV exposure, and high quality chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.
+
+There are several reasons why our design focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
+First, when assessing diet diversity, a sugar diet is often used as a control.
+However, such an experimental design does not reflect real-world conditions for honey bees as they rarely face a total lack of pollen \cite{DiPasquale}.
+Moreover, younger larvae tend to be fed pollen diets, whereas older larvae tend to be fed nectar diets.
+By focusing on pollen diets, our study design reflects natural diet conditions for larvae of a specific age category \cite{ageMatters}.
+Second, in studies that compared honey bee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet \cite{DiPasquale}.
+Third, as was previously mentioned, honey bees are now confronted with less diverse sources of pollen.
+As a result, there is a need to better understand how monofloral diets affect honey bee health.
+
+\subsection*{RNA extraction}
+
+Fifteen cages per treatment were originally produced for monitoring of mortality.
+From these, six live honey bees were randomly selected from each cage 36 hours post inoculation and placed into tubes \cite{carrillo}.
+In summary, 8 samples (representing two bees each) were sequenced per experimental condition (i.e., 32 samples sequenced).
+Tubes were kept on dry ice and then transferred into a -80C freezer until processing.
+From the fifteen possible cages, eight were randomly selected for RNA-sequencing.
+From these eight cages, two of the honey bees per cage were randomly selected from the original six live honey bees per cage.
+These two bees were combined to form a pooled sample representing the cage.
+Whole body RNA from each pool was extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment.
+Samples were suspended in water to 200-400 ng/$\mu$l.
+All samples were then tested on a Bioanalyzer at the Iowa State University DNA Facility to ensure quality (RIN $>$ 8).
+
+\subsection*{Gene expression}
+
+Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing 2500 in rapid run mode; Category: Single End 100 cycle sequencing).
+A standard Illumina mRNA library was prepared by the DNA facility.
+Reads were aligned to the BeeBase Version 3.2 genome \cite{hbGenome} from the Hymenoptera Genome Database \cite{hymenopteraDB} using the programs GMAP and GSNAP \cite{gsnap}.
+There were four lanes of sequencing with 24 samples per lane.
+Each sample was run twice.
+Approximately 75-90\% of reads were mapped to the honey bee genome.
+Each lane produced around 13 million single-end 100 basepair reads.
+
+We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs: NR versus NC, NR versus VR, NR versus VC, NC versus VR, NC versus VC, and VR versus VC).
+We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs).
+We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honey bees \cite{galbraith}.
+We tested each DEG analysis using recommended parameters with DESeq2 \cite{deseq2}, edgeR \cite{edger}, and LimmaVoom \cite{limma}.
+In all cases, we used a false discovery rate (FDR) threshold of 0.05 \cite{benjamini}.
+Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).
+The eulerr shiny application was used to construct Venn diagram overlap images \cite{euler}.
+In the end, we focused on the DEG results from DESeq2 \cite{deseq2} as this pipeline was also used in the Galbraith study \cite{galbraith}.
+We used the independent filtering process built into the DESeq2 software that mitigates multiple comparison corrections on genes with no power rather than defining one filtering threshold.
+
+\subsection*{Comparison to prior studies on transcriptomic response to viral infection}
+
+We compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith} who also addressed honey bee transcriptomic responses to virus infection.
+We applied the same downstream bioinformatics analyses between our count table and the count table provided in the Galbraith study.
+When we applied our bioinformatics pipeline to the Galbraith count table, we obtained different differential expression counts compared to the results published in the Galbraith study.
+However, there was substantial overlap and we considered this justification to use the differential expression list we obtained in order to keep the downstream bioinformatics analyses as similar as possible between the two datasets (Additional file 17).
+
+We used honey bees from naturally-mated colonies, whereas Galbraith et al.
+\cite{galbraith} used honey bees from single-drone colonies.
+In light of this, we should expect the Galbraith et al.
+dataset to contain lower genetic variation between its replicates and higher signal:to:noise ratios than our dataset.
+We use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
+
+\subsection*{Visualization}
+
+We used an array of visualization tools as part of our analysis.
+We used the PCA plot \cite{pca} from the DESeq2 package, a well-known and established tool.
+Along with that, we used lesser-known multivariate visualization tools from our R package called bigPint \cite{bigpint}.
+Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds}) to assess the variability between the replicates and the treatments in our data.
+We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines \cite{jds}.
+
+Furthermore, we used statistical graphics to better understand patterns in our DEGs.
+However, in cases of large DEG lists, these visualization tools had overplotting problems (where multiple objects are drawn on top of one another, making it impossible to detect individual values).
+To remedy this problem, we first standardized each DEG to have a mean of zero and standard deviation of unity for its read counts across its samples \cite{Chandrasekhar, deSouto}.
+Then, we performed hierarchical clustering on the standardized DEGs using Ward's linkage.
+This process divided large DEG lists into smaller clusters of similar patterns, which allowed us to more efficiently visualize the different types of patterns within large DEG lists (see Figures 3 and 4 for examples).
+
+\subsection*{Gene ontology}
+
+DEGs were uploaded as a background list to DAVID Bioinformatics Resources 6.7 \cite{davidBio, davidBio2}.
+The overrepresented gene ontology (GO) terms of DEGs were determined using the BEEBASE\_ID identifier option (honey bee gene model) in the DAVID software.
+To fine-tune the GO term list, only terms correlating to Biological Processes were considered.
+The refined GO term list was then imported into REVIGO \cite{revigo}, which uses semantic similarity measures to cluster long lists of GO terms.
+
+\subsection*{Probing tolerance versus resistance}
+
+To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (tolerance), we created contrasts of interest (Table \ref{tbl:contrasts}).
+In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the chestnut group within the virus inoculated bees but not upregulated in the chestnut group within the non-inoculated bees.
+Our interpretation of these genes is that they represent those that are only activated in inoculated bees that are fed a high quality diet.
+We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the chestnut group for both the virus inoculated bees and non-inoculated bees.
+Our interpretation of these genes is that they represent those that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not.
+We then determined how many genes fell into these two categories and analyzed their GO terminologies.
+
+\subsection*{Post hoc analysis}
+
+We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates.
+Hence, we wanted to explore whether our DEG read counts correlated with pathogen response metrics, including IAPV titers, SBV (also present in our inoculum \cite{carrillo, adamInt}), and mortality rates.
+We explored correlation with SBV because our inoculum \cite{carrillo} does contain SBV, and bees from both inoculated and non-inoculated groups do exhibit detectable SBV titers.
+For this process, we considered virus main effect DEGs (Figure 4), ``tolerance candidate'' DEGs (Additional file 15), and ``resistance candidate'' DEGs (Additional file 16).
+For each DEG in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the pathogen response metrics across its 24 samples.
+We then used the Kruskal–Wallis test to determine if the distribution of the R-squared values in any of the DEG clusters significantly differed from those in the non-DEG genes (the rest of the data).
+As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate DEGs, and virus-related DEGs) and pathogen response measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests.
 
@@
 GO analysis of the chestnut-upregulated DEGs revealed the following over-represented biological functions: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1).
+These encompassed almost all of the overrepresented biological functions in chestnut-upregulated DEGs conditioned on non-inoculation (i.e.
+upregulated in the ``NC'' group compared to the ``NR'' group; Supplementary table 4, Additional file 1) and inoculation (i.e.
+upregulated in the ``VC'' group compared to the ``VR'' group; Supplementary table 6, Additional file 1).
 GO analysis of the rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).
+We note that Fanconi anemia pathways was also the only GO term discovered in rockrose DEGs conditioned on viral inoculation (i.e.
+upregulated in the ``VR'' group compared to the ``VC'' group) (Supplementary table 7, Additional file 1).
+However, Fanconi anemia pathways were not found in rockrose DEGs conditioned on non-inoculation (i.e.
+upregulated in the ``NR'' group compared to the ``NC'' group) (Supplementary table 5, Additional file 1).
 
@@
 
+Supplementary tables 1-9 in Additional file 20 contain complete DEG lists for all comparisons performed in this study.
+
 \subsection*{Transcriptomic data visualization and comparison to a previous study}
@@
 Note that the Galbraith dataset contained three individual bees per treatment group as a single pooled sample, while our dataset contained 16 individual bees per treatment group in 8 RNA-seq samples.
-Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and non-inoculated honey bees better than our dataset (Additional file 2).
+Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the inoculated and non-inoculated honey bees better than our dataset (Additional file 2).
 Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online \cite{bigpint}.
@@
 
-We next used repLIcate TREatment (``litre'') plots, which we recently developed for our bigPint software package.
+We next used repLIcate TREatment (``litre'') plots, which we recently developed for our bigPint software package \cite{bigpint}.
 Litre plots allow users to visualize one DEG onto the Cartesian coordinates of one scatterplot matrix.
@@
 
-Finally, we used scatterplot matrices from the bigPint software to further assess the DEGs.
+Finally, we used scatterplot matrices from the bigPint software to further assess the DEGs \cite{bigpint}.
 A scatterplot matrix is another effective multivariate visualization tool that plots read count distributions across all genes and samples.
@@
 The transcriptomic response to virus infection in our experiment was fairly limited.
-We found only 43 differentially expressed transcripts, some with known immune functions such as a gene with similarity to MD-2 lipid recognition protein and argonaute-2, a protein that plays a central role in RNA silencing (Table \ref{tbl:virusGenes}).
-We also found genes related to transcriptional regulation and muscle contraction.
+We found only 43 differentially expressed transcripts, some with known immune functions such as an MD-2 lipid recognition protein that is particularly implicated in the recognition of pathogen products and argonaute-2, a protein that plays a central role in RNA silencing (Table \ref{tbl:virusGenes}).
+We also found genes related to transcriptional regulation, including Histone H4, Rho guanine nucleotide exchange factor 11, and transcriptional regular Myc-B, which is a regular gene that codes for a transcription factor.
+We additionally found Tropomyosin-like, a gene involved in muscle contraction.
 The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14) and baseline viral titers observed in our control bees (Figure 2).
@@
 The different factors between these two studies may be critical because particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed.
-This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs.
-deformed wing virus (DWV) \cite{Ryabov}.
+This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, SBV vs.
+DWV \cite{Ryabov}.
 Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.
@@
 
-\section*{Methods}
-
-\subsection*{Mortality and virus titers}
-
-Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}.
-In brief, our virus inoculum was prepared by injection of infectious virus particles (derived from infected adults) into white-eyed honey bee pupae; these pupae were then homogenized and virus particles enriched and resuspended.
-This inoculum was then characterized for presence of acute bee paralysis virus, black queen cell virus, DWV, IAPV, Kashmir bee virus, and SBV.
-Experimental infection tests of adult bees and honey bee cell cultures \cite{carrillo} showed that only IAPV is amplified in adult bees.
-To infect caged bees for these experiments, newly emerged bees from 15 healthy colonies at the Iowa State University research apiary were homogeneously mixed, then counted into clear acrylic cages in groups of 35 bees per cage.
-Cages were then presented with open feeders containing 30\% sucrose solution (control) or 30\% sucrose solution containing a 1:1000 dilution of viral inoculum (treatment).
-Dietary treatments were then added (described below).
-To quantify virus titers, two live bees were randomly sampled at 36 hpi from each of 9-10 randomly selected cages.
-Virus levels were then measured via RT-qPCR and quantified against a standard curve, identically to methods described in \cite{carrillo, adamInt}.
-
-A linear mixed effects model was used to relate the mortality rates and IAPV titers to the main and interaction effects of the diet and virus factors.
-The model was fit to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''.
-A random (intercept) effect for experimental setup was included in the model.
-Post-hoc pairwise comparisons of the four (diet and virus combination) treatment groups were performed and Benjamini-Hochberg adjusted p-values were calculated to limit familywise Type I error rates \cite{bh}.
-
-\subsection*{Design of two-factor experiment}
-
-For our nutrition factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.).
-Rockrose pollen is generally considered less nutritious than chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}.
-For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV.
-This experimental design resulted in four treatment groups (rockrose pollen without IAPV exposure, chestnut pollen without IAPV exposure, rockrose pollen with IAPV exposure, and chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.
-
-There are several reasons why our design focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets).
-First, when assessing diet diversity, a sugar diet is often used as a control.
-However, such an experimental design does not reflect real-world conditions for honey bees as they rarely face a total lack of pollen \cite{DiPasquale}.
-Moreover, younger larvae tend to be fed pollen diets, whereas older larvae tend to be fed nectar diets.
-By focusing on pollen diets, our study design reflects natural diet conditions for larvae of a specific age category \cite{ageMatters}.
-Second, in studies that compared honey bee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet \cite{DiPasquale}.
-Third, as was previously mentioned, honey bees are now confronted with less diverse sources of pollen.
-As a result, there is a need to better understand how monofloral diets affect honey bee health.
-
-\subsection*{RNA extraction}
-
-Fifteen cages per treatment were originally produced for monitoring of mortality.
-From these, six live honey bees were randomly selected from each cage 36 hours post inoculation and placed into tubes \cite{carrillo}.
-In summary, 8 samples (representing two bees each) were sequenced per experimental condition (i.e., 32 samples sequenced).
-Tubes were kept on dry ice and then transferred into a -80C freezer until processing.
-From the fifteen possible cages, eight were randomly selected for RNA-sequencing.
-From these eight cages, two of the honey bees per cage were randomly selected from the original six live honey bees per cage.
-These two bees were combined to form a pooled sample representing the cage.
-Whole body RNA from each pool was extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment.
-Samples were suspended in water to 200-400 ng/$\mu$l.
-All samples were then tested on a Bioanalyzer at the Iowa State University DNA Facility to ensure quality (RIN $>$ 8).
-
-\subsection*{Gene expression}
-
-Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing 2500 in rapid run mode; Category: Single End 100 cycle sequencing).
-A standard Illumina mRNA library was prepared by the DNA facility.
-Reads were aligned to the BeeBase Version 3.2 genome \cite{hbGenome} from the Hymenoptera Genome Database \cite{hymenopteraDB} using the programs GMAP and GSNAP \cite{gsnap}.
-There were four lanes of sequencing with 24 samples per lane.
-Each sample was run twice.
-Approximately 75-90\% of reads were mapped to the honey bee genome.
-Each lane produced around 13 million single-end 100 basepair reads.
-
-We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs: NR versus NC, NR versus VR, NR versus VC, NC versus VR, NC versus VC, and VR versus VC).
-We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs).
-We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honey bees \cite{galbraith}.
-We tested each DEG analysis using recommended parameters with DESeq2 \cite{deseq2}, edgeR \cite{edger}, and LimmaVoom \cite{limma}.
-In all cases, we used a false discovery rate (FDR) threshold of 0.05 \cite{benjamini}.
-Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).
-The eulerr shiny application was used to construct Venn diagram overlap images \cite{euler}.
-In the end, we focused on the DEG results from DESeq2 \cite{deseq2} as this pipeline was also used in the Galbraith study \cite{galbraith}.
-We used the independent filtering process built into the DESeq2 software that mitigates multiple comparison corrections on genes with no power rather than defining one filtering threshold.
-
-\subsection*{Comparison to prior studies on transcriptomic response to viral infection}
-
-We compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith} who also addressed honey bee transcriptomic responses to virus infection.
-We applied the same downstream bioinformatics analyses between our count table and the count table provided in the Galbraith study.
-When we applied our bioinformatics pipeline to the Galbraith count table, we obtained different differential expression counts compared to the results published in the Galbraith study.
-However, there was substantial overlap and we considered this justification to use the differential expression list we obtained in order to keep the downstream bioinformatics analyses as similar as possible between the two datasets (Additional file 17).
-
-We used honey bees from naturally-mated colonies, whereas Galbraith et al.
-\cite{galbraith} used honey bees from single-drone colonies.
-In light of this, we should expect the Galbraith et al.
-dataset to contain lower genetic variation between its replicates and higher signal:to:noise ratios than our dataset.
-We use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets.
-
-\subsection*{Visualization}
-
-We used an array of visualization tools as part of our analysis.
-We used the PCA plot \cite{pca} from the DESeq2 package, a well-known and established tool.
-Along with that, we used lesser-known multivariate visualization tools from our work-in-progress R package called bigPint.
-Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds}) to assess the variability between the replicates and the treatments in our data.
-We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines \cite{jds}.
-
-Furthermore, we used statistical graphics to better understand patterns in our DEGs.
-However, in cases of large DEG lists, these visualization tools had overplotting problems (where multiple objects are drawn on top of one another, making it impossible to detect individual values).
-To remedy this problem, we first standardized each DEG to have a mean of zero and standard deviation of unity for its read counts across its samples \cite{Chandrasekhar, deSouto}.
-Then, we performed hierarchical clustering on the standardized DEGs using Ward's linkage.
-This process divided large DEG lists into smaller clusters of similar patterns, which allowed us to more efficiently visualize the different types of patterns within large DEG lists (see Figures 3 and 4 for examples).
-
-\subsection*{Gene ontology}
-
-DEGs were uploaded as a background list to DAVID Bioinformatics Resources 6.7 \cite{davidBio, davidBio2}.
-The overrepresented gene ontology (GO) terms of DEGs were determined using the BEEBASE\_ID identifier option (honey bee gene model) in the DAVID software.
-To fine-tune the GO term list, only terms correlating to Biological Processes were considered.
-The refined GO term list was then imported into REVIGO \cite{revigo}, which uses semantic similarity measures to cluster long lists of GO terms.
-
-\subsection*{Probing tolerance versus resistance}
-
-To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (tolerance), we created contrasts of interest (Table \ref{tbl:contrasts}).
-In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the chestnut group within the virus infected bees but not upregulated in the chestnut group within the non-inoculated bees.
-Our interpretation of these genes is that they represent those that are only activated in infected bees that are fed a high quality diet.
-We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the chestnut group for both the virus infected bees and non-inoculated bees.
-Our interpretation of these genes is that they represent those that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not.
-We then determined how many genes fell into these two categories and analyzed their GO terminologies.
-
-\subsection*{Post hoc analysis}
-
-We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates.
-Hence, we wanted to explore whether our DEG read counts correlated with pathogen response metrics, including IAPV titers, sacbrood virus (SBV; also present in our inoculum \cite{carrillo, adamInt}), and mortality rates.
-We explored correlation with SBV because our inoculum \cite{carrillo} does contain SBV, and bees from both inoculated and non-inoculated groups do exhibit detectable SBV titers.
-For this process, we considered virus main effect DEGs (Figure 4), ``tolerance candidate'' DEGs (Additional file 15), and ``resistance candidate'' DEGs (Additional file 16).
-For each DEG in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the pathogen response metrics across its 24 samples.
-We then used the Kruskal–Wallis test to determine if the distribution of the R-squared values in any of the DEG clusters significantly differed from those in the non-DEG genes (the rest of the data).
-As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate DEGs, and virus-related DEGs) and pathogen response measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests.
-
 \end{doublespacing} % Lindsay added
@@
   The data discussed in this publication have been deposited in NCBI's Gene Expression Omnibus \cite{geo} and are accessible through GEO Series accession number GSE121885 (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE121885).
-Scripts to reproduce visualizations in this paper are available online (https://github.com/lrutter/HoneyBeePaper).
-Information about bigPint visualizations, including how to intrepret them are available online (https://lrutter.github.io/bigPint).
+Scripts to reproduce visualizations in this paper are available online (https://github.com/lindsayrutter/HoneyBeePaper).
+Information about bigPint visualizations, including how to intrepret them are available online (https://lindsayrutter.github.io/bigPint).
 Lists of BeeBase IDs for contrast DEGs from this study are found in Additional File 20.
@@
 \caption{\csentence{Parallel coordinate plots of the 43 virus-related DEGs of our data.}
-Parallel coordinate plots of the 43 DEGs after hierarchical clustering of size four between the virus-infected and control groups of our study.
+Parallel coordinate plots of the 43 DEGs after hierarchical clustering of size four between the virus-inoculated and control groups of our study.
 ``N'' represents non-inoculated control group, and ``V'' represents treatment of virus.
@@
 \hline
-VC vs VR & 376 & \makecell{Genes that change expression \\ due to diet effect in \\ infected bees} & \makecell{Supplementary \\ tables 6 and 7, \\ Additional file 1} \\
+VC vs VR & 376 & \makecell{Genes that change expression \\ due to diet effect in \\ inoculated bees} & \makecell{Supplementary \\ tables 6 and 7, \\ Additional file 1} \\
 \hline
@@
 \hline
-\makecell{VC upregulated in VC vs VR, but \\ NC not upregulated in NC vs NR} & 125 & \makecell{``Resistance'' genes that turn \\ on by good diet only in \\ infected bees} & Figure 5B \\
+\makecell{VC upregulated in VC vs VR, but \\ NC not upregulated in NC vs NR} & 125 & \makecell{``Resistance'' genes that turn \\ on by good diet only in \\ inoculated bees} & Figure 5B \\
 \hline
@@
     \textbf{Table 6:} GO analysis results for the 247 DEGs that were upregulated in the VC treatment from the VC versus VR treatment pair analysis.
-These DEGs represent genes that are upregulated when infected honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
+These DEGs represent genes that are upregulated when inoculated honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
     \textbf{Table 7:} GO analysis results for the 129 DEGs that were upregulated in the VR treatment from the VC versus VR treatment pair analysis.
-These DEGs represent genes that are upregulated when infected honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
+These DEGs represent genes that are upregulated when inoculated honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
     \textbf{Table 8:} Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor.

### 2019-03-04 PostHocDEGPatterns/functionRT.R

@@ -1,10 +1,11 @@
 functionRT <- function(data, type, colNum, PVal){
   rsq <- function (x, y) cor(x, y) ^ 2
   
-  Variables <- read_csv("~/HoneyBeePaper/Variables.csv")
+  #Variables <- read_csv("../Variables.csv")
+  Variables <- read_csv("../PathogenResponse/day3Mortality.csv")
   myVar <- as.data.frame(Variables[,colNum])[,1]
   
-  readFile <- c("../../N_V/DESeq2/ClusterStandard/Sig_4_", "../../ResistanceTolerance/Clustering_Tolerance/Sig_4_", "../../ResistanceTolerance/Clustering_Resistance/Sig_4_")
+  readFile <- c("../N_V/DESeq2/ClusterStandard/Sig_4_", "../ResistanceTolerance/Clustering_Tolerance/Sig_4_", "../ResistanceTolerance/Clustering_Resistance/Sig_4_")
   strVar <- c("virus", "tolerance", "resistance")
   
   getR2 <- function(rf, sv){

### 2019-03-04 PostHocDEGPatterns/getPVal.R

@@ -7,7 +7,7 @@ library(nlme)
 library(onewaytests)
 source("functionRT.R")
 
-data <- as.data.frame(readRDS("../../NC_NR_VC_VR/data/data.Rds"))
+data <- as.data.frame(readRDS("../NC_NR_VC_VR/data/data.Rds"))
 
 PVal <- data.frame()
 PVal <- functionRT(data=data, type="IAPV", colNum=7, PVal=PVal)

### 2019-06-09 Manuscript/Text/manuscript.tex
Accepted changes

@@
 For our virus factor, one level contained bees that were inoculated with IAPV and another level contained bees that were not inoculated with IAPV.
-This experimental design resulted in four treatment groups (low quality rockrose pollen without IAPV exposure, high quality chestnut pollen without IAPV exposure, low quality rockrose pollen with IAPV exposure, and high quality chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.
+This experimental design resulted in four treatment groups (NR: low quality rockrose pollen without IAPV exposure; NC: high quality chestnut pollen without IAPV exposure; VR: low quality rockrose pollen with IAPV exposure; VC: high quality chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.
 
@@
 
-We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1).
+In bees collected 36 hours post treatment, we observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1).
 There were only four genes that were DEGs in both our diet main effect and our virus main effect (GB48747, GB47214, GB42908, and GB42507).
@@
 The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1.
+Distribution of the R-squared values for DEG cluster read counts and pathogen response metrics is provided in Additional file 19.
 An overall trend emerges to suggest that DEGs may have significantly larger correlation with the pathogen response measurements compared to non-DEGs.
@@
 For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally-mated colonies) was limited by numerous extraneous variables between these studies.
-In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on worker honey bees that were fed sugar and artificial pollen diets, whereas our study focused on worker honey bees that were fed bee-collected monofloral diets.
+
+In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on worker honey bees that were fed sugar and artificial pollen diets, whereas we used whole bodies and categorized only into inoculated vs.
+non-inoculated groups; noise may have been introduced through different responses in asymptomatic bees.
 Also, Galbraith’s bees were sampled at 24 hours while ours were sampled at 36 hours.
@@
 
+\section*{Abbreviations}
+
+IAPV: Israeli acute paralysis virus;  DWV: deformed wing virus; SBV: sacbrood virus; NR: non-inoculation, low quality (\textit{Cistus} sp.) pollen treatment; NC: non-inoculation, high quality pollen (\textit{Castanea} sp.) pollen treatment; VR: IAPV-inoculated, low quality (\textit{Cistus} sp.) pollen treatment; VC: IAPV-inoculated, high quality pollen (\textit{Castanea} sp.) pollen treatment; REML: restricted maximum likelihood;
+DEG: differentially expressed gene; GO: gene ontology; PCD: principal component analysis.
+
 \section*{Declarations}
@@
 \section*{Funding}
-  This work was supported by the United States Department of Agriculture, Agriculture and Food Research Initiative (USDA-AFRI) 2011-04894.
+  This work was supported by the United States Department of Agriculture, Agriculture and Food Research Initiative (USDA-AFRI) 2011-04894, but it had no role in the design, analysis, or interpretation of the study.
 
@@
   LR performed the bioinformatic and statistical analyses and produced the figures and tables.
-LR and ALT drafted the manuscript.
-ALT, AGD, JCT, BCB contributed to experimental design.
+LR, AGD, and ALT drafted the manuscript.
+All authors contributed to the conception of the project, the experimental design, and the interpretation of the data.
 AGD and JCT carried out laboratory experiments.
@@
   \subsection*{Additional file 1 --- Supplementary tables.}
-    \textbf{Table 1:} Number of DEGs across three analysis pipelines for (A) the diet main effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study.
-For the diet effects, ``C'' represents chestnut diet and ``R'' represents rockrose diet.
-For the virus effects, ``N'' represents control non-inoculated and ``V'' represents virus-inoculated.
-    \textbf{Table 2:} Pathways related to the 1,033 DEGs that were upregulated in the chestnut treatment from the diet main effect.
-    \textbf{Table 3:} Pathways related to the 881 DEGs that were upregulated in the rockrose treatment from the diet main effect.
-    \textbf{Table 4:} GO analysis results for the 601 DEGs that were upregulated in the NC treatment from the NC versus NR treatment pair analysis.
-These DEGs represent genes that are upregulated when non-inoculated honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
-    \textbf{Table 5:} GO analysis results for the 340 DEGs that were upregulated in the NR treatment from the NC versus NR treatment pair analysis.
-These DEGs represent genes that are upregulated when non-inoculated honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
-    \textbf{Table 6:} GO analysis results for the 247 DEGs that were upregulated in the VC treatment from the VC versus VR treatment pair analysis.
-These DEGs represent genes that are upregulated when inoculated honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
-    \textbf{Table 7:} GO analysis results for the 129 DEGs that were upregulated in the VR treatment from the VC versus VR treatment pair analysis.
-These DEGs represent genes that are upregulated when inoculated honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
+  In all tables, ``C'' represents chestnut diet, ``R'' represents rockrose diet, ``N'' represents control non-inoculated, and ``V'' represents virus-inoculated.
+    \textbf{Table 1:} Number of DEGs across three analysis pipelines for the (A) diet main effect in our study, (B) virus main effect in our study, and (C) virus main effect in the Galbraith study.
+    \textbf{Table 2:} Pathways related to the 1,033 DEGs upregulated in the chestnut treatment from the diet main effect.
+    \textbf{Table 3:} Pathways related to the 881 DEGs upregulated in the rockrose treatment from the diet main effect.
+    \textbf{Table 4:} GO analysis results for the 601 DEGs upregulated in the NC treatment from the NC versus NR treatment pair analysis.
+These DEGs represent genes upregulated in non-inoculated honey bees given high quality chestnut pollen versus low quality rockrose pollen.
+    \textbf{Table 5:} GO analysis results for the 340 DEGs upregulated in the NR treatment from the NC versus NR treatment pair analysis.
+These DEGs represent genes upregulated in non-inoculated honey bees given low quality rockrose pollen versus high quality chestnut pollen.
+    \textbf{Table 6:} GO analysis results for the 247 DEGs upregulated in the VC treatment from the VC versus VR treatment pair analysis.
+These DEGs represent genes upregulated in inoculated honey bees given high quality chestnut pollen versus low quality rockrose pollen.
+    \textbf{Table 7:} GO analysis results for the 129 DEGs upregulated in the VR treatment from the VC versus VR treatment pair analysis.
+These DEGs represent genes upregulated in inoculated honey bees given low quality rockrose pollen versus high quality chestnut pollen.
     \textbf{Table 8:} Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor.
-``C'' represents chestnut diet, ``R'' represents rockrose diet, ``V'' represents virus-inoculated, and ``N'' represents control non-inoculated.
     \textbf{Table 9:} Kruskal-Wallis p-value and Bonferroni corrections for the 36 combinations of DEG lists, pathogen response metrics, and cluster number. (XLS).

### 2019-06-09 Manuscript/MakeFiguresFiles/makeFigure2.R

@@ -29,7 +29,7 @@ summary(glht(mortcomp2, linfct=mcp(Treatment="Tukey")))
 
 # IAPV (All)
 labelDF = data.frame(plot.labels=c("NR","VC","VR","NC"), labels = c("a","ab","b","a"), V1 = c(4.25, 5.35, 8.57, 3.85))
-plotIAPV = ggplot(d, aes(x=Treatment, y=logIAPV)) + geom_boxplot(fill="paleturquoise2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) + ylab("Log viral ge/100 ng RNA") +theme_gray() + theme(text=element_text(size=20)) + ylim(2,9)
+plotIAPV = ggplot(d, aes(x=Treatment, y=logIAPV)) + geom_boxplot(fill="paleturquoise2") + geom_text(data = labelDF, size = 8, aes(x = plot.labels, y = V1, label = labels)) + ylab("Log IAPV Titer") +theme_gray() + theme(text=element_text(size=20)) + ylim(2,9)
 
 # IAPV (N vs V)
 labelDF = data.frame(plot.labels=c("N","V"), labels = c("a","b"), V1 = c(4.3, 8.7))
